### 2014-10-22 tex/parametrized-learning.tex
Adding short tex description

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}

\title{Calibrated, Parametrized Learning}
\author{Kyle Cranmer and Daniel Whiteson}
%\date{}							% Activate to display a given date or no date

\begin{document}
\maketitle

\section{Introduction}
%\subsection{}

Notation:

\begin{itemize}
 \item $x$: a vector of observed quantities for each event (energy, momenta, angles of particles)
 \item $\theta$: parameters of a statistical model
\item $f_c(x| \theta)$:  probability density function (statistical model) for $x$ for category $c\in\{0,1\}$.
\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta)$: real-valued score from a parametrized machine learning classification algorithm (or any map $s: X\otimes \Theta \to\mathbb{R}$)
\item $f_c( s | \theta )$ The probability density function for $s$ implied by $f_c(x|\theta)$ and $s(x)$
\item $f_c( s(\cdot; \theta) | \theta )$ The probability density function for $s$ implied by $f_c(x|\theta)$ and $s(x;\theta)$

\end{itemize}

\subsection{Typical usage of machine learning in HEP}

In high-energy physics (HEP) we typically are searching for or measuring the properties of some 
class of events, generically referred to as \textit{signal}, in the presence of a separate class 
of \textit{background} events. For each event we measure some quantities $x$ that have corresponding distributions 
$f_0(x|\theta)$ for background and $f_1(x|\theta)$ for signal.  Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, c_i\}$ generated with some nominal values of the parameters $\theta_0$, where $c=0$ corresponds to background and $c=1$ corresponds to signal. The resulting classifier is denoted $s(x)$. Based on this classifier and large samples of synthetic data drawn from $f_c(x | theta)$ we construct the distribution  $f_c(s | \theta)$. An example of the distributions of the distribution of $s$ for the signal and background events with $\theta=\theta_0$ is shown in Figure~\ref{fig:tmva}.

\begin{figure}[htbp]
\begin{center}
 \includegraphics[height=2in]{example-TMVA-BDT.pdf}
 \includegraphics[height=2in]{example-TMVA-ROC.pdf}
\caption{Left: an example of the distributions $f_0(s|\theta)$ and $f_1(s|\theta)$ when the classifier $s$ is a boosted-decision tree (BDT). Right: the corresponding ROC curve (right) for this and other classifiers. Figures taken from TMVA manual.}
\label{fig:tmva}
\end{center}
\end{figure}

These steps lead to a subsequent statistical analysis where one observes in data $\{x_e\}$, where $e$ is an event index running from $1$ to $n$. For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution. In particular, one forms the statistical model
\begin{equation}\label{eq:typicalML}
p( \{ x_e \} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e) \,|\, \theta) \,\right] \; ,
\end{equation}
where $\mu=0$ is the null (background-only) hypothesis and $\mu>0$ is the alternate (signal-plus-background) hypothesis.\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known.} Typically, we are interested in inference on $\mu$ and $\theta$ are nuisance parameters; though, sometimes $\theta$ may include some components that we are also wish to infer (like the mass of a new particle that affects the distribution $x$ for the signal events).

\subsection{Comments on typical usage of machine learning in HEP}

Nuisance parameters are an after thought in the typical usage of machine learning in HEP. In fact, most machine learning discussions would only consider $f_0(x)$ and $f_1(x)$. However, as experimentalists we know that we must account for various forms of systematic uncertainty, parametrized by $\theta$. In practice, we take the classifier as fixed and then propagate uncertainty through the classifier as in Eq.~\ref{eq:typicalML}. Building the distribution $f(s(x)|\theta)$ for values of $\theta$ other than the nominal $\theta_0$ used to train the classifier can be thought of as a calibration necessary for classical statistical inference; however, this classifier is clearly not optimal for $\theta \ne \theta_0$.

\subsection{A more powerful  approach}

The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $s(x; \theta)$. Creating the family of classifiers is straight forward, one simply augments the training data with $x$ examples drawn from several values of $\theta$ and then includes the corresponding value of $\theta$ as an input to the classifier. Thus $\{x_e,c_e\} \to \{x_e,\theta_e, c_e\}$ leading to a parametrized learner $s(x)\to s(x;\theta)$. This leads to a complication: one does not know the value of $\theta$ to use when evaluating the parametrized classifier $s(x;\theta)$.  However, when performing statistical inference one will evaluate the likelihood of the data at a given value of $\theta$, so this is a natural choice.  Thus one arrives at the generalization of Eq.~\ref{eq:typicalML}
\begin{equation}\label{eq:parametrizedML}
p( \{ x_e \} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e;\theta) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e;\theta) \,|\, \theta) \,\right] \; .
\end{equation}

The construction in Eq.~\ref{eq:parametrizedML} presents a computational challenge as for each value of $\theta$ one has a new mapping $s: x\to\mathbb{R}$. The most naive realization of this would be to generate synthetic data $\{x_e\}$ according to $f_c(x|\theta)$, evaluate $s(x_e ; \theta)$, and then use a histogram, kernel density, or other non-parametric density estimation procedure to estimate $f_c( s | \theta)$. This approach is not realistic in situations where $f_c(x|\theta)$ is realized with expensive computer simulations; thus, some interpolation strategy or emulator over a fixed set of parameter points $\{\theta_i\}$ would typically be used.

\end{document}

### 2014-10-22 TMVAWrapper/TMVAWrapper.h

@@ -46,7 +46,6 @@ public:
 
 protected:
 
-  RooListProxy _actualVars;
 
   RooAbsReal *_rab[10];
   
@@ -59,12 +58,15 @@ protected:
 
 private:
 
+
   void findVariables(TXMLNode *node);
   void getListOfVars(char weights[100]);
   
   TMVA::Reader *reader ;
   mutable Float_t local_features[10];
   Float_t target;
+
+  RooListProxy _actualVars;
 
   ClassDef(TMVAWrapper,1) // Your description goes here...
 };

### 2014-11-06 tex/parametrized-learning.tex
more writing

@@
 The resulting classifier is denoted $s(x)$.
-Based on this classifier and large samples of synthetic data drawn from $f_c(x | theta)$ we construct the distribution  $f_c(s | \theta)$.
+Based on this classifier and large samples of synthetic data drawn from $f_c(x | \theta)$ we construct the distribution  $f_c(s | \theta)$.
 An example of the distributions of the distribution of $s$ for the signal and background events with $\theta=\theta_0$ is shown in Figure~\ref{fig:tmva}.
@@
 
+\section{Possible Extensions}
+The approach describe thus far has been agnostic as to the type of classifier used to produce $s(x)$ and $s(x; \theta)$.
+In practice, we have used standard packages for the training of these learners.
+In particular, the loss function for these learning algorithms is based on classification performance of individual $(x_e, c_e)$ feature $\to$ target examples.
+However, in overarching statistical procedure that uses the statistical model in Eq.~\ref{eq:typicalML} or \ref{eq:parametrizedML} the quantity we wish to optimize is a property of an entire dataset $\{x_e\}$ or ensembles of datasets that can be generated from the statistical models $f_c(x|\theta)$.
+
+For example, in typical hypothesis testing problems posed in the Neyman-Pearson context one would be interested in minimizing type II error under a fixed rate of typeI error rate.
+In the presence of systematic uncertainty and nuisance parameters, the loss function may become more complicated.
+The recent Higgs boson machine learning challenge hosted by Kaggle, the evaluation was the approximate median significance (AMS) that characterized the power of a hypothesis test between $\mu=0$ (the background only hypothesis) and $\mu=\mu_0$ (the signal+background hypothesis, where $\mu_0$ is a small number reflecting the small Higgs signal that would be found in addition to the much larger background process) in the presence of uncertainty on the background normalization.
+Interestingly, the top entries to the challenge did not improve by focusing on this alternative loss function 
+in training their classifier, but instead used it for the optimization of a threshold on the score (a working point on the ROC curve) that was related to the way the challenge was designed.
+This suggests that the standard
+classification loss functions lead to similar optimizations and that refinements for the custom loss functions
+are not easy to come by.
+
+More generally, there is an avenue of research associated to custom loss functions specifically designed 
+for these real world statistical problems. One direction might be to choose loss functions that make 
+some balance between the standard classification loss, take into account robustness with respect to nuisance parameters, and population-level quantities like type I and II error.
+
+\section*{Acknowldgements}
+We would like to thank Yann LeCun, Philip Stark for their feedback on the
+project early in its conception and Bal\'azs K\'egl for discussions about the Kaggle challenge and 
+non-standard loss functions.
+KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 
+KC would like to thank UC-Irvine for their hospitality while this research was carried out and the 
+Moore and Sloan foundations for their generous support of the data science environment at NYU.
+
 \end{document}

### 2014-11-06 ttbar_resonance.py
initial code from Daniel

@@ -1,0 +1,298 @@
+#!/usr/bin/env python
+# https://github.com/svenkreiss/PyROOTUtils/blob/master/PyROOTUtils/Graph.py  
+__author__ = "Kyle Cranmer <kyle.cranmer@nyu.edu"
+__version__ = "0.1"
+
+'''
+This is a research work in progress.
+
+Define model mu_s*Gaus(x|alpha,sigma)+mu_b*flat(x)
+Generate {x} for several {alpha}
+Calculate power (expected significance) for some alpha using profile likelihood approach
+1) Train NN for alpha=0. 
+	1a) make (histfactory/on the fly) model for NN with alpha variations
+		- calculate power
+	1b) make pdf on the fly for each value of alpha
+2) Train NN with {x,alpha}
+	a) make histfactory model for NN with alpha variations using same alpha as input to NN
+		- calculate power
+	b) make pdf on the fly for NN with alpha variations using same alpha as input to NN
+		- calculate power
+'''
+
+
+import ROOT
+import numpy as np
+
+
+from sklearn import svm, linear_model, gaussian_process
+from sklearn.neural_network import BernoulliRBM
+from sklearn.ensemble import AdaBoostClassifier
+from sklearn.tree import DecisionTreeClassifier
+from sklearn.externals import joblib
+
+
+import matplotlib.pyplot as plt
+
+import os.path
+
+
+def createdPdfForFixed():
+	'''
+	Read in learner saved in fixed.pkl
+	Evaluate outputs for several parameter points.
+	Generate histograms for each point.
+	(to do:	create parametrized pdf that interpolates across these pdfs)
+	'''
+	clf = joblib.load('fixed.pkl') 
+
+	trainAndTarget = np.loadtxt('traindata.dat')
+	traindata = trainAndTarget[:,0:2]
+	targetdata = trainAndTarget[:,2]
+
+	massPoints = np.unique(traindata[:,1])
+
+
+	fixedhists=[]
+	c1 = ROOT.TCanvas()
+	for j, name in enumerate(['sig','bkg']):
+		for i, mass in enumerate(massPoints):
+			#bkg part
+			#plt.hist(outputs[i*chunk+shift: \
+			#	(i+1)*chunk+shift], 30, alpha=0.3)
+			#sig part
+	
[…]

### 2014-11-08 ttbar_resonance.py
TMVAWrapper working, but histfuncs have problems

@@ -150,6 +150,15 @@ def createPdfForAdaptive_tree(tree):
 		morph.specialIntegratorConfig(ROOT.kTRUE).method1D().setLabel('RooBinIntegrator')
 		print morph
 
+	# make dataset, add to workspace
+	w.factory('mwwbb[500,7000]')
+	w.factory('mx[350,1600]')
+	w.factory('target[-1,2]')
+	w.defineSet('inputvars','mwwbb,mx')
+	w.defineSet('treevars','mwwbb,mx,target')
+	alldata = ROOT.RooDataSet('alldata','',tree, w.set('treevars'))
+	getattr(w,'import')(alldata)
+
 	w.Print()
 	w.writeToFile("workspace_adaptive_tree.root")
 
@@ -158,6 +167,24 @@ def createPdfForAdaptive():
         f = ROOT.TFile("ttbar_14tev_jes1_eval.root")
         nt = f.Get("nto")
         createPdfForAdaptive_tree(nt)
+
+def plotScore():
+	ROOT.gSystem.Load( 'TMVAWrapper/libTMVAWrapper' )
+	ROOT.gROOT.ProcessLine(".L RooBSplineBases.cxx+")
+	ROOT.gROOT.ProcessLine(".L RooBSpline.cxx+")
+	
+	f = ROOT.TFile('workspace_adaptive_tree.root','r')
+	w = f.Get('w')
+	inputVars = ROOT.RooArgList(w.set('inputvars'))
+	inputVars.Print()
+	nn = ROOT.TMVAWrapper('nn','nn',inputVars,"TMVARegression_ttbar_14tev_jes1.root_MLP.weights.xml")
+	frame = w.var('mwwbb').frame()
+	for x in np.linspace(400,1600,20):
+		w.var('mx').setVal(x)
+		nn.plotOn(frame)
+	c1 = ROOT.TCanvas("c2",'',400,400)
+	frame.Draw()
+	c1.SaveAs('tmva.pdf')
 
 
 def plotAdaptive():
@@ -194,8 +221,8 @@ def fitAdaptive():
 	f = ROOT.TFile('workspace_adaptive_tree.root','r')
 	w = f.Get('w')
 
-	w.factory('CompositeFunctionPdf::sigtemplate(sigmorphfunc)')
-	w.factory('CompositeFunctionPdf::bkgtemplate(bkgmorphfunc)')
[…]

### 2014-11-08 ttbar_resonance.py
fixed initial # naming problem

@@ -62,7 +62,7 @@ def createdPdfForFixed():
 			#	(i+1)*chunk+shift], 30, alpha=0.3)
 			#sig part
 	
-			hist = ROOT.TH1F('{0}hist{1}'.format(name,i),"hist",30,-0.1,1.2)
+			hist = ROOT.TH1F('h{0}hist{1}'.format(name,i),"hist",30,-0.1,1.2)
 			fixedhists.append(hist)
 			for val in outputs[i*chunk+j*shift: (i+1)*chunk+j*shift]:
 				hist.Fill(val)
@@ -102,7 +102,7 @@ def createPdfForAdaptive_tree(tree):
                         var_points[int(tree.target)].append(tree.mx)
                         ind = var_points[int(tree.target)].index(tree.mx)
 
-			hist = ROOT.TH1F('{0}hist{1}'.format(int(tree.target),ind),"hist",bins,low,high)
+			hist = ROOT.TH1F('h{0}hist{1}'.format(int(tree.target),ind),"hist",bins,low,high)
 			adaptivehists[int(tree.target)].append(hist)
 
 #                if (i%1000==0):
@@ -132,12 +132,12 @@ def createPdfForAdaptive_tree(tree):
                 for ind in range(0,len(var_points[target])):
                         print "Building RooWorld stuff for target",target," index ",ind
                         print "   mx = ", var_points[target][ind], " mean = ", adaptivehists[target][ind].GetMean(), " rms = ", adaptivehists[target][ind].GetRMS()
-                        datahist = ROOT.RooDataHist('{0}datahist{1}'.format(target,ind),"hist", 
+                        datahist = ROOT.RooDataHist('dh{0}datahist{1}'.format(target,ind),"hist", 
                                                     ROOT.RooArgList(s), adaptivehists[target][ind])
                 
 			order=1
 			s.setBins(bins)
-			histpdf = ROOT.RooHistPdf('{0}histpdf{1}'.format(target,ind),"hist", 
+			histpdf = ROOT.RooHistPdf('hp{0}histpdf{1}'.format(target,ind),"hist", 
 				ROOT.RooArgSet(s), datahist,order)
[…]

### 2014-11-08 ttbar_resonance.py
EDIT not working

@@ -233,11 +233,11 @@ def fitAdaptive():
 	c1 = ROOT.TCanvas('c1')
 	sframe = w.var('score').frame()
 	w.pdf('sigtemplate').plotOn(sframe)
-	w.pdf('m0morph').plotOn(sframe)
-	w.pdf('m1morph').plotOn(sframe)
-	w.pdf('sigtemplate').plotOn(sframe)
-	w.pdf('bkgtemplate').plotOn(sframe)
-	w.pdf('template').plotOn(sframe)
+	w.pdf('m0morph').plotOn(sframe,ROOT.RooFit.LineColor(ROOT.kGreen))
+	w.pdf('m1morph').plotOn(sframe,ROOT.RooFit.LineColor(ROOT.kRed))
+	w.pdf('sigtemplate').plotOn(sframe,ROOT.RooFit.LineColor(ROOT.kGreen))
+	w.pdf('bkgtemplate').plotOn(sframe,ROOT.RooFit.LineColor(ROOT.kRed))
+	w.pdf('template').plotOn(sframe,ROOT.RooFit.LineColor(ROOT.kBlack))
 	w.pdf('template').plotOn(sframe,ROOT.RooFit.Components('sigtemplate'),ROOT.RooFit.LineColor(ROOT.kRed))
 	w.pdf('template').plotOn(sframe,ROOT.RooFit.Components('bkgtemplate'),ROOT.RooFit.LineColor(ROOT.kGreen))
 	sframe.Draw()
@@ -255,13 +255,16 @@ def fitAdaptive():
 	nn = ROOT.TMVAWrapper('nn','nn',inputVars,"TMVARegression_ttbar_14tev_jes1.root_MLP.weights.xml")
 	#weightfile = "TMVARegression_alphavary.root_MLP.weights.xml"
 
+	print "about to import"
 	print "get val = ",	nn.getVal()
 	getattr(w,'import')(ROOT.RooArgSet(nn),ROOT.RooFit.RecycleConflictNodes()) 
 	w.Print()	
-	return
-
+	print "ok, almost done"
 	#create nll based on pdf(NN(x,mu) | mu)
-	w.factory('EDIT::pdf(template,score=nn)')
+	w.factory('EDIT::pdftemp(template,score=nn)')
+	return
+	w.factory('EDIT::pdf(pdftemp,mu=mx)')
+	w.factory('EDIT::pdf(template,score=nn,mu=mx)')
[…]

### 2014-11-08 ttbar_resonance.py
structure basically working for physics example

@@ -121,10 +121,10 @@ def createPdfForAdaptive_tree(tree):
 
         # build RooWorld stuff
 	w = ROOT.RooWorkspace('w')
-	w.factory('mu[{0},{1}]'.format( var_points[0][0],var_points[0][len(var_points[0])-1]))
+	w.factory('mx[{0},{1}]'.format( var_points[0][0],var_points[0][len(var_points[0])-1]))
 	w.factory('score[{0},{1}]'.format(low,high))
 	s = w.var('score')
-	mu = w.var('mu')
+	mu = w.var('mx')
 
         adpativedatahists=[[],[]]
         adpativehistpdfs=[[],[]]
@@ -156,8 +156,16 @@ def createPdfForAdaptive_tree(tree):
 	w.factory('target[-1,2]')
 	w.defineSet('inputvars','mwwbb,mx')
 	w.defineSet('treevars','mwwbb,mx,target')
+	w.defineSet('obsvars','mwwbb')
 	alldata = ROOT.RooDataSet('alldata','',tree, w.set('treevars'))
+	alldata.Print()
+	toydata = alldata.reduce(ROOT.RooFit.Cut('mx==1000.'))
+	toydata.Print()
+	obsdata = toydata.reduce(ROOT.RooFit.SelectVars(w.set('obsvars')))
+	obsdata.Print()
+	obsdata = ROOT.RooDataSet('obsdata','',tree, w.set('obsvars'))
 	getattr(w,'import')(alldata)
+	getattr(w,'import')(obsdata)
 
 	w.Print()
 	w.writeToFile("workspace_adaptive_tree.root")
@@ -203,7 +211,7 @@ def plotAdaptive():
 	c1.SetLogy();
 
 	for val in np.linspace(400,1500,100):
-		w.var('mu').setVal(val)
+		w.var('mx').setVal(val)
 		w.pdf('m1morph').plotOn(frame,ROOT.RooFit.LineColor(ROOT.kRed))
 		w.pdf('m0morph').plotOn(frame,ROOT.RooFit.LineColor(ROOT.kBlue))
 	frame.Draw()
@@ -227,7 +235,7 @@ def fitAdaptive():
 	w.factory('Uniform::baseline(score)')
 	w.factory('SUM::template(sigfrac[0,1]*sigtemplate,const[0.01]*baseline,bkgtemplate)')
 
-	mu = w.var('mu')
[…]

### 2014-11-12 tex/parametrized-learning.tex
progress on note after UCI seminar

@@
  \item $x$: a vector of observed quantities for each event (energy, momenta, angles of particles)
+ \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$
  \item $\theta$: parameters of a statistical model
@@
 \begin{equation}\label{eq:typicalML}
-p( \{ x_e \} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e) \,|\, \theta) \,\right] \; ,
+f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e) \,|\, \theta) \,\right] \; ,
 \end{equation}
@@
 \begin{equation}\label{eq:parametrizedML}
-p( \{ x_e \} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e;\theta) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e;\theta) \,|\, \theta) \,\right] \; .
+f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e;\theta) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e;\theta) \,|\, \theta) \,\right] \; .
 \end{equation}
@@
 
+\subsection{A different starting point}
+
+Instead of evolving from the traditional usage of ML in HEP, let's start from scratch with $f_c(x|\theta)$ and a well defined objective.
+
+\subsubsection{Simple Hypothesis Testing}
+
+Consider the situation of simple hypothesis testing between the background-only $\mu=0$ and signal+background hypothesis $\mu=\mu_0=\nu_s/(\nu_s+\nu_b)$ for $\theta=\theta_0$.
+Let us assume that not only can we use the $f_c(x|\theta_0)$ as a generative model, but that we can evaluate the probability density at any point $x$.
+The statistical model is given by
+\begin{equation}\label{eq:NP}
+f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( x_e \, |\,  \theta)  + (1-\mu)\, f_0( x_e \,|\, \theta) \,\right] \; .
+\end{equation}
+The objective is to find a  test statistic $T(D)$ (a map from the space of the data to $\mathbb{R}$)  
+and a threshold $k_\alpha$ that minimizes the rate of type-II error (i.e.
+$P(T(D) < k_\alpha | \mu=\mu_0, \theta_0) = \beta$ under the constraint of a fixed rate of type-I error ( i.e.
+$P(T(D) > k_\alpha | \mu=0, \theta_0) = \alpha$ ).
+The Neyman-Pearson lemma states that the optimal test statistic is the likelihood ratio, or equivalently, the log-likelihood ratio 
+\begin{equation}
+T =  \log \frac{f(D | \mu=\mu_0, \theta_0)}{f(D | \mu=0, \theta_0)}
+\end{equation}
+When the $x_e$ are i.i.d., then the log-likelihood ratio expands into a sum
+\begin{equation}
+T =   \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(x_e | \theta_0) }{ f_0(x_e | \theta_0) } \right]}_{q(x_e)} + c_2\;,
+\end{equation}
+where $c_1=\mu/(1-\mu)$ and $c_2=\log(1-\mu)$ are constants for the simple hypothesis test.
+
+Here we see that the optimal $T$ for the experiment is composed of a sum over events of a linear linear function of the per-event function $q(x)$.
+A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.
+
+The important part of the per-event function $q(x)$ is that it defines contours in the feature space $x$.
+These contours are also equivalent to the function $f_1(x|\theta_0)/f_0(x|\theta_0)$, the signal-to-background ratio.
+If we can find a machine learner $s(x)$ that is a monotonic function of $f_1(x|\theta_0)/f_0(x|\theta_0)$ it will share the same contours.
+Then the remaining challenge is to find the appropriate rescaling that gives back  linear function $q(x)$.
+
+Postulate:
+\[
+T' = \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(s(x_e) | \theta_0) }{ f_0(s(x_e) | \theta_0) } \right]}_{q(s_e)} \;,
+\]
+leads to an equivalent test, where $s_e \equiv s(x_e)$
+\[
+f_c(s) = \int dx \, \delta(s-s(x)) \, f_c(x)  \,  |\partial s / \partial x|^{-1}
+\]
+
+Note, in order to form the hypothesis test, one still needs to build the build the $f_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$.
+
+Proof:
+
+First show $q(x_e) = q(s(x_e))$.
+\[
+\frac {f_1(s(x) | \theta_0) }{ f_0(s(x) | \theta_0) } = \frac { \int dx \delta(s-s(x)) \, f_1(x_e | \theta_0) J }{ \int dx \delta(s-s(x)) \, f_0(x_e | \theta_0) J } = \frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) }
+\]
+
+Second, show density is the same
+
+\[
+f_c(q_x) = \int dx \delta(q_x-q_x(x)) f_c(x) / |\partial q_x / \partial x|
+\]
+
+\[
+f_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, f_c(x) \, / |\partial q_s / \partial x|
+\]
+
+\[
+f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
+\]
+
+This is the test statistic that one would naturally get from constructing the likelihood function for the 1-d distributions of the output of the machine learner
+\begin{equation}\label{eq:NP}
+f( \{s_e\} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s_e \, |\,  \theta)  + (1-\mu)\, f_0( s_e \,|\, \theta) \,\right] \; .
+\end{equation}
+Thus this motivates Eq.~\ref{eq:typicalML} in the more general case.
+
+Question:
+Where does this fit in:
+\[
+T'' = \sum_{e=1}^n \underbrace{ \log \left[ 1+c_1 s(x_e) \right] }_{q'(s_e)} \;,
+\]
+
+Usefulness:
+If we can readily compute $f_c(x|\theta_0)$, then there would be no need for machine learning algorithms.
+However,  these densities are often difficult to compute though one can readily use them as a generative model to produce samples.
+In that case, one might wish to use a machine learning algorithm to find $s(x) \approx f_1(x|\theta_0) / f_0(x|\theta_0)$.
+
+Conclusion, while it is possible to start directly with the Neyman-Pearson loss function of type-II error under the constraint of fixed size and try to find a per-experiment learner $T(D)$, it is more convenient to cast the problem in terms of learning a per-event function $s(x)$, and then calibrate it to be a linear function of $q(x)$.
+
 \section*{Acknowldgements}

### 2014-11-12 plotsForTtbar.py

@@ -1,0 +1,34 @@
+from ROOT import *
+
+def plotsForTtbar():
+	f = TFile.Open("ttbar_14tev_alljes_eval.root")
+	tree = f.Get('nto')
+	c1 = TCanvas("c1","",600,600)
+	c1.SetObjectStat(0)
+	mx = 1000
+	jes = 1
+	sameness=""
+	colors = [kBlue,kRed]
+	hists = []
+	for i, target in enumerate([0,1]):
+		cuts = 'target=={target} && jes=={jes} && mx=={mx}'.format(jes=jes, target=target,mx=mx)
+		print cuts
+		histname = 'h{target}{jes}{mx}'.format(jes=jes, target=target,mx=mx)
+		hist = TH1F(histname,'',20,0,3000)
+		hists.append(hist)
+		tree.Project(histname,'mwwbb',cuts)
+
+	for i,hist in enumerate(hists):
+		hist.SetLineColor(colors[i])
+		hist.SetLineWidth(2)
+		hist.SetObjectStat(0)
+
+		#hist.SetFillColor(colors[i])
+		hist.DrawNormalized(sameness)
+		sameness='same'
+
+	c1.SaveAs('dists_for_ttbar.pdf')
+
+
+if __name__ == '__main__':
+	plotsForTtbar()

### 2014-11-12 ttbar_resonance.py

@@ -195,6 +195,14 @@ def plotScore():
 	c1.SaveAs('tmva.pdf')
 
 
+	frame = w.var('mx').frame()
+	w.var('mwwbb').setVal(800)
+	for x in np.linspace(400,1600,20):
+		w.var('mwwbb').setVal(x)
+		nn.plotOn(frame)
+	frame.Draw()
+	c1.SaveAs('tmva_vs_mx.pdf')
+
 def plotAdaptive():
 	'''
 	make plots of the output of the parametrized model
@@ -346,7 +354,7 @@ if __name__ == '__main__':
 	'''
 	The main function that calls the individual steps of the procedure
 	'''
-	#plotScore()
-	createPdfForAdaptive()
+	plotScore()
+	#createPdfForAdaptive()
 	#plotAdaptive()
-	fitAdaptive()
+	#fitAdaptive()

### 2014-11-13 tex/parametrized-learning.tex
proof needed for N-P equivalence

@@
 
-Postulate:
+\textbf{Postulate:}
 \[
@@
 
-Proof:
+\textbf{Proof (in progress):}
 
-First show $q(x_e) = q(s(x_e))$.
-\[
-\frac {f_1(s(x) | \theta_0) }{ f_0(s(x) | \theta_0) } = \frac { \int dx \delta(s-s(x)) \, f_1(x_e | \theta_0) J }{ \int dx \delta(s-s(x)) \, f_0(x_e | \theta_0) J } = \frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) }
-\]
-
-Second, show density is the same
+We want to show density is the same
 
@@
 f_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, f_c(x) \, / |\partial q_s / \partial x|
+\]
+sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $f_1(x)/f_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
+\[
+\frac {f_1(c | \theta_0) }{ f_0(c | \theta_0) } = 
+\frac { \int dx \delta(c-s(x)) \, f_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  } = 
+\frac {c \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, f_0(x | \theta_0)  } = c =\frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
 \]
@@
 
+\newpage
+
+Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $f_1(x)$ and $f_0(x)$ define 
+\[ 
+s(x)=\frac{f_1(x)}{f_2(x)}
+\].
+Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
+I wish to show
+\[
+\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
+\]
+$\forall x \ni s(x)=c$.
+
+Sketch of Proof: 
+let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$.
+The  coordinate perpendicular to $\Omega$ can be written $y = x+\epsilon \hat{n}$.
+In these coordinates, the integral $\int dx \to \int d\Omega d\epsilon$, thus
+\[
+g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon|
+\]
+Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega$.
+Thus
+\[
+g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|
+\]
+and we arrive at the result
+\[
+\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;.
+\]
+
+\newpage
+
 \section*{Acknowldgements}

### 2014-11-18 tex/parametrized-learning.tex
MLE and CMLE

@@
 								% TeX will automatically convert eps --> pdf in pdflatex		
-\usepackage{amssymb}
+\usepackage{amssymb, amsmath}
 
-\title{Calibrated, Parametrized Learning}
+\DeclareMathOperator*{\argmin}{arg\,min}
+\DeclareMathOperator*{\argmax}{arg\,max}
+
+\title{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}
+%\title{Calibrated, Parametrized Learning}
 \author{Kyle Cranmer and Daniel Whiteson}
@@
 
-Notation:
+In many areas of science, likelihood ratio tests  are established tools for statistical inference. 
+Directly constructing the likelihood ratio for high-dimensional observations 
+is often not possible or is computationally impractical. Here we demonstrate how 
+discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
+a generative model for the data is available for calibration.  We use the following notation
 
 \begin{itemize}
- \item $x$: a vector of observed quantities for each event (energy, momenta, angles of particles)
- \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$
+ \item $x$: a vector of features
+ \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
  \item $\theta$: parameters of a statistical model
-\item $f_c(x| \theta)$:  probability density function (statistical model) for $x$ for category $c\in\{0,1\}$.
-\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
-\item $s(x;\theta)$: real-valued score from a parametrized machine learning classification algorithm (or any map $s: X\otimes \Theta \to\mathbb{R}$)
-\item $f_c( s | \theta )$ The probability density function for $s$ implied by $f_c(x|\theta)$ and $s(x)$
-\item $f_c( s(\cdot; \theta) | \theta )$ The probability density function for $s$ implied by $f_c(x|\theta)$ and $s(x;\theta)$
+\item $f(x| \theta)$:  probability density  (statistical model) for $x$ 
+%\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
+\item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
+%\item $f( s | \theta )$ The probability density function for $s$ implied by $f(x|\theta)$ and $s(x)$
+\item $g( s(x; \theta_0, \theta_1) | \theta )$: The probability density  for $s$ implied by $f(x|\theta_0, \theta_1)$ and $s(x;\theta_0, \theta_1)$
+\end{itemize}
+We will assume the $x_e$ are i.i.d., so that $f(D|\theta) = \prod_{e=1}^n f(x_e | \theta)$.
 
-\end{itemize}
+In the setting where one is interested in simple hypothesis testing between a null $\theta=\theta_0$ against an alternate $\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio 
+\begin{equation}
+T(D) = \prod_{e=1}^n \frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)}
+\end{equation}
+is the most powerful test statistic. In order to evaluate $T(D)$, one must be able to evaluate the probability density 
+$f(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that 
+can act as generative model  for $f(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case 
+high energy physics where the simulation of particle detectors can only be done in the `forward mode'.
+
+Our main result is that one can form an equivalent test based on 
+%\begin{equation}
+%T'(D) = \prod_{e=1}^n \frac{ f(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ f(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
+%\end{equation}
+\begin{equation}
+T'(D) = \prod_{e=1}^n \frac{ g(s_e | \theta_0)}{ g(s_e | \theta_1)}
+\end{equation}
+if 
+\begin{equation}
+s_e = s(x_e; \theta_0, \theta_1) = \frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)} \; 
+\end{equation}
+or some monotonic function of $s$. This will be proven below.
+This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
+used to learn $s(x; \theta_1, \theta_0)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
+from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$.
+
+While the original goal for hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, the machine learning problem is an event-by-event classification problem.
+Of course, this follows from the fact that we assume the $x_e$ to be i.i.d.
+
+\section{Dimensionality reduction}
+
+ The target hypothesis test is be based on 
+\begin{equation}
+\ln T =   \sum_{e=1}^n \underbrace{\log \left[ \frac {f(x_e | \theta_0) }{ f(x_e | \theta_1) } \right]}_{q(x_e)} \;.
+\end{equation}
+Here we see that the optimal $T$ for the experiment is composed of a sum over events of a linear linear function of the per-event function $q(x)$.
+A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.
+
+The important part of the per-event function $q(x)$ is that it defines iso-contours in the feature space $x$.
+As we will show, our goal is to learn a monotonic function of $f(x|\theta_0)/f(x|\theta_1)$ that shares the same iso-contours.
+Then the remaining challenge is to find the appropriate rescaling that gives back  linear function $q(x)$.
+Our claim is that the generative model $f(x|\theta)$ can be used to calibrate $g(s|\theta)$ and that
+\begin{equation}
+\ln T' = \sum_{e=1}^n \underbrace{\log \left[ \frac {g(s_e | \theta_0) }{ g(s_e | \theta_1) } \right]}_{q(s_e)} \;,
+\end{equation}
+leads to an equivalent test. In particular, we need to show the density
+\begin{equation}
+f(q_x|\theta) = \int dx \delta(q_x-q_x(x)) f(x|\theta)  / | \hat{n} \cdot \nabla q_x  |
+\end{equation}
+is the same as
+\begin{equation}
+f(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, f(x|\theta) \, / | \hat{n} \cdot \nabla q_s  | \; .
+\end{equation}
+It is sufficient to show that $q(x_e) = q(s(x_e))$ $ \forall x\in\Omega_c$.
+
+For notational simplicity, let $f_0(x) = f(x|\theta_0)$, $f_1(x) = f(x|\theta_1)$, and $s(x)=s(x; \theta_1, \theta_0)$.
+The distribution of $x$ totally determines the distribution of $s$ via the change of variables $x\to s$. 
+In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
+Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to $\Omega_c$ at the point $x$. The induced density $g_1(c)$ is given by 
+\begin{equation}
+g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x)  / | \hat{n} \cdot \nabla s  |
+\end{equation}
+and a similar equation for $g_0(c)$. 
+%\textbf{Do we need Jacobian for x $\to$ s independent of delta function part, I think that's double counting?}
+
+\textbf{\flushleft Theorem 1:}
+We have the following equalities
+\begin{equation}
+\frac{g_1(c)}{g_0(c)} = s(x) = \frac{f_1(x)}{f_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
+\end{equation}
+\textbf{Proof}
+We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega_c$.
+Thus
+\begin{equation}
+g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x) / | \hat{n} \cdot \nabla s  |= s(x) \int d\Omega_c f_0(x)  / | \hat{n} \cdot \nabla s  | \;,
+\end{equation}
+and the integrals cancel in the likelihood ratio
+\begin{equation}
+\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega_c f_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_c f_0(x) / | \hat{n} \cdot \nabla s  |} = s(x) = \frac{f_1(x)}{f_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
+\end{equation}
+
+\bigskip
+In the case of simple hypothesis testing, $\theta_0$ and $\theta_1$ are specified and there is a unique map $s(x) =  s(x_e; \theta_0, \theta_1)$.
+In that case, the equivalent likelihood ratio test can be performed by first transforming the data to $D_s = \{s_1, \dots, s_e\}$, constructing the likelihoods
+\begin{equation}\label{eq:NP}
+g( D_s \,|\,  \theta) = \prod_{e=1}^n \,  g( s_e \, |\,  \theta)   \; 
+\end{equation}
+for $\theta=\{\theta_0,\theta_1\}$, and constructing the likelihood ratio based on $g(D_s|\theta_0)/g(D_s|\theta_1)$.
+
+\section{Composite hypotheses and the generalized likelihood ratio}
+
+In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative $\theta \in \Theta_0^C$, the generalized likelihood ratio\footnote{Also known as the profile likelihood ratio.} test is commonly used
+\begin{equation}
+\lambda(x) =  \frac{ \sup_{\theta \in \Theta_0} f(D | \theta)}{ \sup_{\theta \in \Theta} f(D | \theta)} \; .
+\end{equation}
+This generalized likelihood ratio can be used both for hypothesis tests in the presence of nuisance parameters or to create confidence intervals with or without nuisance parameters.
+Often, the parameter vector is broken into two components $\theta=(\mu,\nu)$, where the $\mu$ components are considered parameters of interest while the $\nu$ components are considered nuisance parameters.
+In that case $\Theta_0$ corresponds to all values of $\nu$ with $\mu$ fixed.
+
+Denote the maximum likelihood estimator
+\begin{equation}
+\hat{\theta} = \argmax_\theta  f(D | \theta)
+\end{equation}
+and the conditional maximum likelihood estimator
+\begin{equation}
+\hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  f(D | \theta) \; .
+\end{equation}
+
+It is not obvious that if we are working with the distributions $g(s|\theta)$ (for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the same estimators. 
+Fortunately, there is a construction based on $g(s|\theta)$ that works.
+The maximum likelihood estimate is the same as the value that maximizes the ratio to $f(D|\theta_1)$ for some fixed value of $\theta_1$.
+This allows us to use Theorem~1 to find
+\begin{equation}
+\hat{\theta} = \argmax_\theta \frac{ f(D | \theta)}{ f(D | \theta_1)} = \argmax_\theta  \sum \ln \frac{f(x_e | \theta)}{f(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{g(s(x_e; \theta, \theta_1) | \theta)}{g(s(x_e; \theta, \theta_1) |\theta_1)} \; .
+\end{equation}
+It is important that we include the denominator $g(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that change as we vary $\theta$.
+
+\section{Learning the correct mapping}
 
@@
 \textbf{Postulate:}
-\[
+\begin{equation}
 T' = \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(s(x_e) | \theta_0) }{ f_0(s(x_e) | \theta_0) } \right]}_{q(s_e)} \;,
-\]
+\end{equation}
 leads to an equivalent test, where $s_e \equiv s(x_e)$
-\[
+\begin{equation}
 f_c(s) = \int dx \, \delta(s-s(x)) \, f_c(x)  \,  |\partial s / \partial x|^{-1}
-\]
+\end{equation}
 
@@
 
-\[
+\begin{equation}
 f_c(q_x) = \int dx \delta(q_x-q_x(x)) f_c(x) / |\partial q_x / \partial x|
-\]
+\end{equation}
 
-\[
+\begin{equation}
 f_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, f_c(x) \, / |\partial q_s / \partial x|
-\]
+\end{equation}
 sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $f_1(x)/f_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
-\[
+\begin{equation}
 \frac {f_1(c | \theta_0) }{ f_0(c | \theta_0) } = 
@@
 \frac {c \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, f_0(x | \theta_0)  } = c =\frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
-\]
+\end{equation}
 
-\[
-f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
-\]
+%\begin{equation}
+%f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
+%\end{equation}
 
@@
 Where does this fit in:
-\[
+\begin{equation}
 T'' = \sum_{e=1}^n \underbrace{ \log \left[ 1+c_1 s(x_e) \right] }_{q'(s_e)} \;,
-\]
+\end{equation}
 
@@
 Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $f_1(x)$ and $f_0(x)$ define 
-\[ 
+\begin{equation} 
 s(x)=\frac{f_1(x)}{f_2(x)}
-\].
+\end{equation}.
 Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
 I wish to show
-\[
+\begin{equation}
 \frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
-\]
+\end{equation}
 $\forall x \ni s(x)=c$.
@@
 In these coordinates, the integral $\int dx \to \int d\Omega d\epsilon$, thus
-\[
+\begin{equation}
 g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon|
-\]
+\end{equation}
 Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega$.
 Thus
-\[
+\begin{equation}
 g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|
-\]
+\end{equation}
 and we arrive at the result
-\[
+\begin{equation}
 \frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;.
-\]
+\end{equation}
 
-\newpage
+\subsection{misc notes}
+
+The distribution of $s$ is fully determined by the mapping $s(x_e; \theta_0, \theta_1)$ and the distribution $f(x)$, which we can write
+\begin{equation}
+g(s_e) = \int dx \, \delta(s_e-s(x)) \, f(x)  \; .
+\end{equation}
+\textbf{NEED Jacobian for x $\to$ s independent of delta function part, or double counting?}
+
+\textbf{Theorem 1:}
+Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to the surface at the point $x$.
+The component perpendicular to $\Omega_c$ can be written $x+\epsilon \hat{n}$.
+In these coordinates, we rewrite integral $\int dx \to \int d\Omega_c d\epsilon$, thus
+\begin{equation}
+g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x)  / | \hat{n} \cdot \nabla q_s  |
+\end{equation}
+Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega_c$.
+Thus
+\begin{equation}
+g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x)  / | \hat{n} \cdot \nabla q_s  |
+\end{equation}
+and we arrive at the result
+\begin{equation}
+\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;\hspace{3em} \forall x\in\Omega_c.
+\end{equation}
+
+Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
+I wish to show
+\begin{equation}
+\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
+\end{equation}
+$\forall x \ni s(x)=c$.
+
+Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $f_1(x)$ and $f_0(x)$ define 
+\begin{equation} 
+s(x)=\frac{f_1(x)}{f_0(x)}
+\end{equation}.
+Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
+I wish to show
+\begin{equation}
+\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
+\end{equation}
+$\forall x \ni s(x)=c$.
+
+Sketch of Proof: 
+let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$.
+The  coordinate perpendicular to $\Omega$ can be written $y = x+\epsilon \hat{n}$.
+In these coordinates, the integral $\int dx \to \int d\Omega d\epsilon$, thus
+\begin{equation}
+g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon|
+\end{equation}
+Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega$.
+Thus
+\begin{equation}
+g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|
+\end{equation}
+and we arrive at the result
+\begin{equation}
+\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;.
+\end{equation}
+
+Note, in order to form the hypothesis test, one still needs to build the build the $f_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$.
+
+\textbf{Proof (in progress):}
+
+We want to show density is the same
+
+\begin{equation}
+f(q_x|\theta) = \int dx \delta(q_x-q_x(x)) f(x|\theta)  / | \hat{n} \cdot \nabla q_s  |
+\end{equation}
+
+\begin{equation}
+f(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, f(x|\theta) \, / | \hat{n} \cdot \nabla q_s  |
+\end{equation}
+sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $f_1(x)/f_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
+\begin{equation}
+\frac {f_1(c | \theta_0) }{ f_0(c | \theta_0) } = 
+\frac { \int dx \delta(c-s(x)) \, f_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  } = 
+\frac {c \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, f_0(x | \theta_0)  } = c =\frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
+\end{equation}
+
+%\begin{equation}
+%f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
+%\end{equation}
 

### 2014-11-18 tex/parametrized-learning.tex
finishing draft 1 of re-write

@@
 %\end{equation}
-\begin{equation}
+\begin{equation}\label{eq:equivLRtest}
 T'(D) = \prod_{e=1}^n \frac{ g(s_e | \theta_0)}{ g(s_e | \theta_1)}
@@
 Denote the maximum likelihood estimator
-\begin{equation}
+\begin{equation}\label{eq:mle}
 \hat{\theta} = \argmax_\theta  f(D | \theta)
@@
 
-\section{Learning the correct mapping}
+\section{Learning the correct mapping and its distribution}
 
-\subsection{Typical usage of machine learning in HEP}
+Thus far we have shown that likelihood ratio tests based on $f(x|\theta_0)/f(x|\theta_1)$ with high dimensional features $x$ can be reproduced via hypothesis tests based on the univariate densities $g(s|\theta)$ for the very special dimensionality reduction map $s(x|\theta_0, \theta_1)$.
+The motivation for this is that often it is not possible to evaluate the density $f(x|\theta)$ at a given point $x$.
+This approach is not useful if it is not possible to approximate $s(x|\theta_0, \theta_1)$ and $g(s|\theta)$ without evaluating the density $f(x|\theta)$.
+In order for this approach to be useful, we need to be able to approximate both based on samples $\{(x,\theta)\}$ drawn from the generative model $f(x|\theta)$.
 
-In high-energy physics (HEP) we typically are searching for or measuring the properties of some 
+Denote the approximate dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{g}(\hat{s}|\theta)$.
+In general we will be interested in the machine learning problem that approximates these distributions based on samples $\{(x, \theta)\}$ drawn from the generative model $f(x|\theta)$.
+In particular,
+is there a loss function such that the function $\hat{s}(x)$ that minimizes the expected loss leads to a function
+that is one-to-one with $f(x|\theta_0)/f(x|\theta_1)$?
+
+\subsection{The standard discriminative classification setting} For fixed $\theta_0$ and $\theta_1$ we can generate 
+large samples from each model and train a classifier. To be concrete, let's use $f(x|\theta_0)$ to generate training 
+data $(x_i, \theta_i=\theta_0, y_i=0)$ and $f(x|\theta_1)$ to generate training data $(x_i, \theta_i=\theta_1, y_i=1)$. 
+If we use the squared-loss function, then the expected loss is
+\begin{equation}
+\mathbb{E}[L] = \int dx f(x|\theta_0)  (  \hat{s}(x) )^2  + \int dx f(x|\theta_1)  ( 1 - \hat{s}(x) )^2 \; .
+\end{equation}
+The function $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ minimizes this expected loss.
+
+Proof: Consider a variation about $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ given by $\hat{s}'(x) = \hat{s}(x) +h(x)$. The change in the expected loss is given by 
+\begin{equation}
+\Delta = \mathbb{E}[L_{\hat{s}'}]-\mathbb{E}[L_{\hat{s}}] = \int dx f(x|\theta_0)  (h^2(x) + 2 h(x)  \hat{s}(x) )  + f(x|\theta_1)   (h^2(x) - 2 h(x) + 2 h(x) \hat{s}(x)  \; .
+\end{equation}
+using $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ we obtain
+\begin{equation}
+\Delta =  \int dx (f(x|\theta_0)+f(x|\theta_1))  h^2(x) >0    \; .
+\end{equation}
+Thus any variation on $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ has a larger expected loss.
+
+The conclusion is that standard classification with a quadratic loss function and training data as described above will approximate a discriminative classifier needed to produce an equivalent likelihood ratio test.
+
+Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g. histograms or kernel density estimation) to approximate $\hat{g}(x|\theta)$.
+
+This treatment shows that in the asymptotic limit of samples from the generative model, that we can approximate arbitrarily well the original likelihood ratio test.
+With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{g}(x)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test.
+This will depend in general on the ultimate goal of the test and is left as a subject for future work.
+
+\section{Typical usage of machine learning in HEP}
+
+In high-energy physics (HEP) we are often searching for the properties of some 
 class of events, generically referred to as \textit{signal}, in the presence of a separate class 
 of \textit{background} events. For each event we measure some quantities $x$ that have corresponding distributions 
-$f_0(x|\theta)$ for background and $f_1(x|\theta)$ for signal.
-Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, c_i\}$ generated with some nominal values of the parameters $\theta_0$, where $c=0$ corresponds to background and $c=1$ corresponds to signal.
+$f_b(x|\nu)$ for background and $f_s(x|\nu)$ for signal.
+Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, c_i\}$ generated with some nominal values of the parameters $\nu_0$, where $c=0$ corresponds to background and $c=1$ corresponds to signal.
 The resulting classifier is denoted $s(x)$.
-Based on this classifier and large samples of synthetic data drawn from $f_c(x | \theta)$ we construct the distribution  $f_c(s | \theta)$.
-An example of the distributions of the distribution of $s$ for the signal and background events with $\theta=\theta_0$ is shown in Figure~\ref{fig:tmva}.
+Based on this classifier and large samples of synthetic data drawn from $f_c(x | \nu)$ we construct the distribution  $g_c(s | \nu)$.
+An example of the distributions of the distribution of $s$ for the signal and background events with $\nu=\nu_0$ is shown in Figure~\ref{fig:tmva}.
 
@@
  \includegraphics[height=2in]{example-TMVA-ROC.pdf}
-\caption{Left: an example of the distributions $f_0(s|\theta)$ and $f_1(s|\theta)$ when the classifier $s$ is a boosted-decision tree (BDT).
+\caption{Left: an example of the distributions $f_b(s|\nu)$ and $f_s(s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).
 Right: the corresponding ROC curve (right) for this and other classifiers.
@@
 
-These steps lead to a subsequent statistical analysis where one observes in data $\{x_e\}$, where $e$ is an event index running from $1$ to $n$.
+These steps lead to a subsequent statistical analysis where one observes in data $D=(x_1, \dots, x_n)$.
 For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution.
@@
 \begin{equation}\label{eq:typicalML}
-f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e) \,|\, \theta) \,\right] \; ,
+f( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu g_s( s(x_e) \, |\,  \nu)  + (1-\mu)\, g_b( s(x_e) \,|\, \nu) \,\right] \; ,
 \end{equation}
-where $\mu=0$ is the null (background-only) hypothesis and $\mu>0$ is the alternate (signal-plus-background) hypothesis.\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known.} Typically, we are interested in inference on $\mu$ and $\theta$ are nuisance parameters; though, sometimes $\theta$ may include some components that we are also wish to infer (like the mass of a new particle that affects the distribution $x$ for the signal events).
+where $\mu=0$ is the null (background-only) hypothesis and $\mu>0$ is the alternate (signal-plus-background) hypothesis.\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood.} Typically, we are interested in inference on $\mu$ and $\nu$ are nuisance parameters.
 
@@
 Nuisance parameters are an after thought in the typical usage of machine learning in HEP.
-In fact, most machine learning discussions would only consider $f_0(x)$ and $f_1(x)$.
-However, as experimentalists we know that we must account for various forms of systematic uncertainty, parametrized by $\theta$.
+In fact, most machine learning discussions would only consider $f_b(x)$ and $f_s(x)$.
+However, as experimentalists we know that we must account for various forms of systematic uncertainty, parametrized by nuisance parameters $\nu$.
 In practice, we take the classifier as fixed and then propagate uncertainty through the classifier as in Eq.~\ref{eq:typicalML}.
-Building the distribution $f(s(x)|\theta)$ for values of $\theta$ other than the nominal $\theta_0$ used to train the classifier can be thought of as a calibration necessary for classical statistical inference; however, this classifier is clearly not optimal for $\theta \ne \theta_0$.
+Building the distribution $g(s|\nu)$ for values of $\nu$ other than the nominal $\nu_0$ used to train the classifier can be thought of as a calibration necessary for classical statistical inference; however, this classifier is clearly not optimal for $\nu \ne \nu_0$.
 
@@
 
-The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $s(x; \theta)$.
-Creating the family of classifiers is straight forward, one simply augments the training data with $x$ examples drawn from several values of $\theta$ and then includes the corresponding value of $\theta$ as an input to the classifier.
-Thus $\{x_e,c_e\} \to \{x_e,\theta_e, c_e\}$ leading to a parametrized learner $s(x)\to s(x;\theta)$.
-This leads to a complication: one does not know the value of $\theta$ to use when evaluating the parametrized classifier $s(x;\theta)$.
-However, when performing statistical inference one will evaluate the likelihood of the data at a given value of $\theta$, so this is a natural choice.
-Thus one arrives at the generalization of Eq.~\ref{eq:typicalML}
-\begin{equation}\label{eq:parametrizedML}
-f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e;\theta) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e;\theta) \,|\, \theta) \,\right] \; .
-\end{equation}
+The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $s(x; \nu_0, \nu_1)$.
+Creating the family of classifiers is straight forward, one simply augments the training data with $x$ examples drawn from several values of $\nu$ and then includes the corresponding value of $\nu$ as an input to the classifier.
+Thus $\{x_e,c_e\} \to \{x_e,\nu_{0,e},\nu_{1,e}, c_e\}$ leading to a parametrized learner $s(x)\to s(x;\nu_0, \nu_1)$.
+This leads to a complication: one does not know the value of $\nu$ to use when evaluating the parametrized classifier $s(x;\nu_0, \nu_1)$, so one cannot simply pre-compute $s_e$ before performing the likelihood fit.
+However, when performing a likelihood ratio test, one can calculate the maximum likelihood estimate and conditional maximum likelihood estimate according to Eq.~\ref{eq:mle} and then form the equivalent likelihood ratio test according to Eq.~\ref{eq:equivLRtest}.
 
-The construction in Eq.~\ref{eq:parametrizedML} presents a computational challenge as for each value of $\theta$ one has a new mapping $s: x\to\mathbb{R}$.
-The most naive realization of this would be to generate synthetic data $\{x_e\}$ according to $f_c(x|\theta)$, evaluate $s(x_e ; \theta)$, and then use a histogram, kernel density, or other non-parametric density estimation procedure to estimate $f_c( s | \theta)$.
-This approach is not realistic in situations where $f_c(x|\theta)$ is realized with expensive computer simulations; thus, some interpolation strategy or emulator over a fixed set of parameter points $\{\theta_i\}$ would typically be used.
+%Equivalently, once one has determined $\hat{\theta}$ and $\hat{\hat{\theta}}$, then one can specify the parameters of the mapping $s(x;\hat{\hat{\theta}}, \hat{\theta})$ and then following Eq.~\ref{eq:NP} form the likelihood
+%\begin{equation}\label{eq:parametrizedML}
+%g( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu g_s( s(x_e;\nu_0, \nu_1) \, |\,  \nu)  + (1-\mu)\, g_b( s(x_e;\nu_0, \nu_1) \,|\, \nu) \,\right] \; ,
+%\end{equation}
+%which is a natural generalization of Eq.~\ref{eq:typicalML}.
 
-\section{Possible Extensions}
-The approach describe thus far has been agnostic as to the type of classifier used to produce $s(x)$ and $s(x; \theta)$.
+%The construction in Eq.~\ref{eq:parametrizedML} presents a computational challenge as for each value of $\nu$ one has a new mapping $s: x\to\mathbb{R}$.
+The most naive realization of this would be to generate synthetic data $\{x_e\}$ according to $f_c(x|\nu)$, evaluate $s(x_e ; \nu_0, \nu_1)$, and then use a histogram, kernel density, or other non-parametric density estimation procedure to estimate $g_c( s | \nu)$.
+This approach is not realistic in situations where $f_c(x|\nu)$ is realized with expensive computer simulations; thus, some interpolation strategy or emulator over a fixed set of parameter points $\{\nu_i\}$ would typically be used.
+
+\section{Conclusions}
+
+We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $f(x|\theta)$ can be used to approximate statistical inference likelihoodased on the  ratio $f(x|\theta_0)/f(x|\theta_1)$ when it is not possible to evaluate the densities $f(x|\theta)$ for an arbitrary $x$.
+This approach leverages the power of machine learning in a classical statistical setting.
+
+%\section{Possible Extensions}
+%The approach describe thus far has been agnostic as to the type of classifier used to produce $s(x)$ and $s(x; \nu)$.
 In practice, we have used standard packages for the training of these learners.
@@
 However, in overarching statistical procedure that uses the statistical model in Eq.~\ref{eq:typicalML} or \ref{eq:parametrizedML} the quantity we wish to optimize is a property of an entire dataset $\{x_e\}$ or ensembles of datasets that can be generated from the statistical models $f_c(x|\theta)$.
-
-For example, in typical hypothesis testing problems posed in the Neyman-Pearson context one would be interested in minimizing type II error under a fixed rate of typeI error rate.
+%
+%For example, in typical hypothesis testing problems posed in the Neyman-Pearson context one would be interested in minimizing type II error under a fixed rate of typeI error rate.
 In the presence of systematic uncertainty and nuisance parameters, the loss function may become more complicated.
 The recent Higgs boson machine learning challenge hosted by Kaggle, the evaluation was the approximate median significance (AMS) that characterized the power of a hypothesis test between $\mu=0$ (the background only hypothesis) and $\mu=\mu_0$ (the signal+background hypothesis, where $\mu_0$ is a small number reflecting the small Higgs signal that would be found in addition to the much larger background process) in the presence of uncertainty on the background normalization.
-Interestingly, the top entries to the challenge did not improve by focusing on this alternative loss function 
-in training their classifier, but instead used it for the optimization of a threshold on the score (a working point on the ROC curve) that was related to the way the challenge was designed.
+%Interestingly, the top entries to the challenge did not improve by focusing on this alternative loss function 
+%in training their classifier, but instead used it for the optimization of a threshold on the score (a working point on the ROC curve) that was related to the way the challenge was designed.
 This suggests that the standard
-classification loss functions lead to similar optimizations and that refinements for the custom loss functions
-are not easy to come by.
-
-More generally, there is an avenue of research associated to custom loss functions specifically designed 
-for these real world statistical problems. One direction might be to choose loss functions that make 
-some balance between the standard classification loss, take into account robustness with respect to nuisance parameters, and population-level quantities like type I and II error.
-
-\subsection{A different starting point}
-
-Instead of evolving from the traditional usage of ML in HEP, let's start from scratch with $f_c(x|\theta)$ and a well defined objective.
-
-\subsubsection{Simple Hypothesis Testing}
-
-Consider the situation of simple hypothesis testing between the background-only $\mu=0$ and signal+background hypothesis $\mu=\mu_0=\nu_s/(\nu_s+\nu_b)$ for $\theta=\theta_0$.
+%classification loss functions lead to similar optimizations and that refinements for the custom loss functions
+%are not easy to come by.
+%
+%More generally, there is an avenue of research associated to custom loss functions specifically designed 
+%for these real world statistical problems. One direction might be to choose loss functions that make 
+%some balance between the standard classification loss, take into account robustness with respect to nuisance parameters, and population-level quantities like type I and II error.
+%
+%\subsection{A different starting point}
+%
+%Instead of evolving from the traditional usage of ML in HEP, let's start from scratch with $f_c(x|\theta)$ and a well defined objective.  
+%
+%\subsubsection{Simple Hypothesis Testing}
+%
+%Consider the situation of simple hypothesis testing between the background-only $\mu=0$ and signal+background hypothesis $\mu=\mu_0=\nu_s/(\nu_s+\nu_b)$ for $\theta=\theta_0$.
 Let us assume that not only can we use the $f_c(x|\theta_0)$ as a generative model, but that we can evaluate the probability density at any point $x$.
 The statistical model is given by
-\begin{equation}\label{eq:NP}
-f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( x_e \, |\,  \theta)  + (1-\mu)\, f_0( x_e \,|\, \theta) \,\right] \; .
-\end{equation}
-The objective is to find a  test statistic $T(D)$ (a map from the space of the data to $\mathbb{R}$)  
-and a threshold $k_\alpha$ that minimizes the rate of type-II error (i.e.
+%\begin{equation}\label{eq:NP}
+%f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( x_e \, |\,  \theta)  + (1-\mu)\, f_b( x_e \,|\, \theta) \,\right] \; .
+%\end{equation}
+%The objective is to find a  test statistic $T(D)$ (a map from the space of the data to $\mathbb{R}$)  
+%and a threshold $k_\alpha$ that minimizes the rate of type-II error (i.e.
 $P(T(D) < k_\alpha | \mu=\mu_0, \theta_0) = \beta$ under the constraint of a fixed rate of type-I error ( i.e.
 $P(T(D) > k_\alpha | \mu=0, \theta_0) = \alpha$ ).
-The Neyman-Pearson lemma states that the optimal test statistic is the likelihood ratio, or equivalently, the log-likelihood ratio 
-\begin{equation}
-T =  \log \frac{f(D | \mu=\mu_0, \theta_0)}{f(D | \mu=0, \theta_0)}
-\end{equation}
-When the $x_e$ are i.i.d., then the log-likelihood ratio expands into a sum
-\begin{equation}
-T =   \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(x_e | \theta_0) }{ f_0(x_e | \theta_0) } \right]}_{q(x_e)} + c_2\;,
-\end{equation}
-where $c_1=\mu/(1-\mu)$ and $c_2=\log(1-\mu)$ are constants for the simple hypothesis test.
-
-Here we see that the optimal $T$ for the experiment is composed of a sum over events of a linear linear function of the per-event function $q(x)$.
+%The Neyman-Pearson lemma states that the optimal test statistic is the likelihood ratio, or equivalently, the log-likelihood ratio 
+%\begin{equation}
+%T =  \log \frac{f(D | \mu=\mu_0, \theta_0)}{f(D | \mu=0, \theta_0)}
+%\end{equation}
+%When the $x_e$ are i.i.d., then the log-likelihood ratio expands into a sum
+%\begin{equation}
+%T =   \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(x_e | \theta_0) }{ f_b(x_e | \theta_0) } \right]}_{q(x_e)} + c_2\;,
+%\end{equation}
+%where $c_1=\mu/(1-\mu)$ and $c_2=\log(1-\mu)$ are constants for the simple hypothesis test.
+%
+%Here we see that the optimal $T$ for the experiment is composed of a sum over events of a linear linear function of the per-event function $q(x)$.
 A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.
-
-The important part of the per-event function $q(x)$ is that it defines contours in the feature space $x$.
-These contours are also equivalent to the function $f_1(x|\theta_0)/f_0(x|\theta_0)$, the signal-to-background ratio.
-If we can find a machine learner $s(x)$ that is a monotonic function of $f_1(x|\theta_0)/f_0(x|\theta_0)$ it will share the same contours.
+%
+%The important part of the per-event function $q(x)$ is that it defines contours in the feature space $x$.
+These contours are also equivalent to the function $f_1(x|\theta_0)/f_b(x|\theta_0)$, the signal-to-background ratio.
+If we can find a machine learner $s(x)$ that is a monotonic function of $f_1(x|\theta_0)/f_b(x|\theta_0)$ it will share the same contours.
 Then the remaining challenge is to find the appropriate rescaling that gives back  linear function $q(x)$.
-
-\textbf{Postulate:}
-\begin{equation}
-T' = \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(s(x_e) | \theta_0) }{ f_0(s(x_e) | \theta_0) } \right]}_{q(s_e)} \;,
-\end{equation}
-leads to an equivalent test, where $s_e \equiv s(x_e)$
-\begin{equation}
-f_c(s) = \int dx \, \delta(s-s(x)) \, f_c(x)  \,  |\partial s / \partial x|^{-1}
-\end{equation}
-
-Note, in order to form the hypothesis test, one still needs to build the build the $f_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$.
-
-\textbf{Proof (in progress):}
-
-We want to show density is the same
-
-\begin{equation}
-f_c(q_x) = \int dx \delta(q_x-q_x(x)) f_c(x) / |\partial q_x / \partial x|
-\end{equation}
-
-\begin{equation}
-f_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, f_c(x) \, / |\partial q_s / \partial x|
-\end{equation}
-sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $f_1(x)/f_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
-\begin{equation}
-\frac {f_1(c | \theta_0) }{ f_0(c | \theta_0) } = 
-\frac { \int dx \delta(c-s(x)) \, f_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  } = 
-\frac {c \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, f_0(x | \theta_0)  } = c =\frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
-\end{equation}
-
+%
+%\textbf{Postulate:}
 %\begin{equation}
-%f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
+%T' = \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(s(x_e) | \theta_0) }{ f_b(s(x_e) | \theta_0) } \right]}_{q(s_e)} \;,
 %\end{equation}
-
-This is the test statistic that one would naturally get from constructing the likelihood function for the 1-d distributions of the output of the machine learner
-\begin{equation}\label{eq:NP}
-f( \{s_e\} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s_e \, |\,  \theta)  + (1-\mu)\, f_0( s_e \,|\, \theta) \,\right] \; .
-\end{equation}
-Thus this motivates Eq.~\ref{eq:typicalML} in the more general case.
-
-Question:
-Where does this fit in:
-\begin{equation}
-T'' = \sum_{e=1}^n \underbrace{ \log \left[ 1+c_1 s(x_e) \right] }_{q'(s_e)} \;,
-\end{equation}
-
-Usefulness:
-If we can readily compute $f_c(x|\theta_0)$, then there would be no need for machine learning algorithms.
+%leads to an equivalent test, where $s_e \equiv s(x_e)$
+%\begin{equation}
+%f_c(s) = \int dx \, \delta(s-s(x)) \, f_c(x)  \,  |\partial s / \partial x|^{-1}
+%\end{equation}
+%
+%
+%
+%Note, in order to form the hypothesis test, one still needs to build the build the $f_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$. 
+%
+%
+%
+%\textbf{Proof (in progress):}
+%
+%We want to show density is the same 
+%
+%\begin{equation}
+%f_c(q_x) = \int dx \delta(q_x-q_x(x)) f_c(x) / |\partial q_x / \partial x|
+%\end{equation}
+%
+%\begin{equation}
+%f_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, f_c(x) \, / |\partial q_s / \partial x|
+%\end{equation}
+%sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $f_1(x)/f_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
+%\begin{equation}
+%\frac {f_1(c | \theta_0) }{ f_0(c | \theta_0) } = 
+%\frac { \int dx \delta(c-s(x)) \, f_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  } = 
+%\frac {c \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, f_0(x | \theta_0)  } = c =\frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
+%\end{equation}
+%
+%
+%
+%%\begin{equation}
+%%f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
+%%\end{equation}
+%
+%
+%This is the test statistic that one would naturally get from constructing the likelihood function for the 1-d distributions of the output of the machine learner
+%\begin{equation}\label{eq:NP}
+%f( \{s_e\} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s_e \, |\,  \theta)  + (1-\mu)\, f_0( s_e \,|\, \theta) \,\right] \; .
+%\end{equation}
+%Thus this motivates Eq.~\ref{eq:typicalML} in the more general case. 
+%
+%
+%Question:
+%Where does this fit in:
+%\begin{equation}
+%T'' = \sum_{e=1}^n \underbrace{ \log \left[ 1+c_1 s(x_e) \right] }_{q'(s_e)} \;,
+%\end{equation}
+%
+%
+%
+%Usefulness:
+%If we can readily compute $f_c(x|\theta_0)$, then there would be no need for machine learning algorithms.
 However,  these densities are often difficult to compute though one can readily use them as a generative model to produce samples.
 In that case, one might wish to use a machine learning algorithm to find $s(x) \approx f_1(x|\theta_0) / f_0(x|\theta_0)$.
-
-Conclusion, while it is possible to start directly with the Neyman-Pearson loss function of type-II error under the constraint of fixed size and try to find a per-experiment learner $T(D)$, it is more convenient to cast the problem in terms of learning a per-event function $s(x)$, and then calibrate it to be a linear function of $q(x)$.
-
-\newpage
-
-Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $f_1(x)$ and $f_0(x)$ define 
-\begin{equation} 
-s(x)=\frac{f_1(x)}{f_2(x)}
-\end{equation}.
-Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
-I wish to show
-\begin{equation}
-\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
-\end{equation}
-$\forall x \ni s(x)=c$.
-
-Sketch of Proof: 
-let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$.
+%
+%Conclusion, while it is possible to start directly with the Neyman-Pearson loss function of type-II error under the constraint of fixed size and try to find a per-experiment learner $T(D)$, it is more convenient to cast the problem in terms of learning a per-event function $s(x)$, and then calibrate it to be a linear function of $q(x)$.
+%
+%\newpage
+%
+%Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $f_1(x)$ and $f_0(x)$ define 
+%\begin{equation} 
+%s(x)=\frac{f_1(x)}{f_2(x)}
+%\end{equation}.
+%Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
+%I wish to show
+%\begin{equation}
+%\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
+%\end{equation}
+%$\forall x \ni s(x)=c$.
+%
+%Sketch of Proof: 
+%let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$.
 The  coordinate perpendicular to $\Omega$ can be written $y = x+\epsilon \hat{n}$.
 In these coordinates, the integral $\int dx \to \int d\Omega d\epsilon$, thus
-\begin{equation}
-g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon|
-\end{equation}
-Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega$.
-Thus
-\begin{equation}
-g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|
-\end{equation}
-and we arrive at the result
-\begin{equation}
-\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;.
-\end{equation}
-
-\subsection{misc notes}
-
-The distribution of $s$ is fully determined by the mapping $s(x_e; \theta_0, \theta_1)$ and the distribution $f(x)$, which we can write
-\begin{equation}
-g(s_e) = \int dx \, \delta(s_e-s(x)) \, f(x)  \; .
-\end{equation}
-\textbf{NEED Jacobian for x $\to$ s independent of delta function part, or double counting?}
-
-\textbf{Theorem 1:}
-Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to the surface at the point $x$.
+%\begin{equation}
+%g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon|
+%\end{equation}
+%Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega$.
+%Thus
+%\begin{equation}
+%g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|
+%\end{equation}
+%and we arrive at the result
+%\begin{equation}
+%\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;.
+%\end{equation}
+%
+%
+%
+%
+%\subsection{misc notes}
+%
+%
+%
+%
+%The distribution of $s$ is fully determined by the mapping $s(x_e; \theta_0, \theta_1)$ and the distribution $f(x)$, which we can write
+%\begin{equation}
+%g(s_e) = \int dx \, \delta(s_e-s(x)) \, f(x)  \; .
+%\end{equation}
+%\textbf{NEED Jacobian for x $\to$ s independent of delta function part, or double counting?}
+%
+%
+%\textbf{Theorem 1:}
+%Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to the surface at the point $x$.
 The component perpendicular to $\Omega_c$ can be written $x+\epsilon \hat{n}$.
 In these coordinates, we rewrite integral $\int dx \to \int d\Omega_c d\epsilon$, thus
-\begin{equation}
-g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x)  / | \hat{n} \cdot \nabla q_s  |
-\end{equation}
-Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega_c$.
-Thus
-\begin{equation}
-g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x)  / | \hat{n} \cdot \nabla q_s  |
-\end{equation}
-and we arrive at the result
-\begin{equation}
-\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;\hspace{3em} \forall x\in\Omega_c.
-\end{equation}
-
-Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
-I wish to show
-\begin{equation}
-\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
-\end{equation}
-$\forall x \ni s(x)=c$.
-
-Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $f_1(x)$ and $f_0(x)$ define 
-\begin{equation} 
-s(x)=\frac{f_1(x)}{f_0(x)}
-\end{equation}.
-Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
-I wish to show
-\begin{equation}
-\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
-\end{equation}
-$\forall x \ni s(x)=c$.
-
-Sketch of Proof: 
-let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$.
+%\begin{equation}
+%g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x)  / | \hat{n} \cdot \nabla q_s  |
+%\end{equation}
+%Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega_c$.
+%Thus
+%\begin{equation}
+%g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x)  / | \hat{n} \cdot \nabla q_s  |
+%\end{equation}
+%and we arrive at the result
+%\begin{equation}
+%\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;\hspace{3em} \forall x\in\Omega_c.
+%\end{equation}
+%
+%
+%
+%
+%
+%Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
+%I wish to show
+%\begin{equation}
+%\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
+%\end{equation}
+%$\forall x \ni s(x)=c$.
+%
+%
+%Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $f_1(x)$ and $f_0(x)$ define 
+%\begin{equation} 
+%s(x)=\frac{f_1(x)}{f_0(x)}
+%\end{equation}.
+%Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
+%I wish to show
+%\begin{equation}
+%\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
+%\end{equation}
+%$\forall x \ni s(x)=c$.
+%
+%Sketch of Proof: 
+%let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$.
 The  coordinate perpendicular to $\Omega$ can be written $y = x+\epsilon \hat{n}$.
 In these coordinates, the integral $\int dx \to \int d\Omega d\epsilon$, thus
-\begin{equation}
-g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon|
-\end{equation}
-Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega$.
-Thus
-\begin{equation}
-g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|
-\end{equation}
-and we arrive at the result
-\begin{equation}
-\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;.
-\end{equation}
-
-Note, in order to form the hypothesis test, one still needs to build the build the $f_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$.
-
-\textbf{Proof (in progress):}
-
-We want to show density is the same
-
-\begin{equation}
-f(q_x|\theta) = \int dx \delta(q_x-q_x(x)) f(x|\theta)  / | \hat{n} \cdot \nabla q_s  |
-\end{equation}
-
-\begin{equation}
-f(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, f(x|\theta) \, / | \hat{n} \cdot \nabla q_s  |
-\end{equation}
-sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $f_1(x)/f_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
-\begin{equation}
-\frac {f_1(c | \theta_0) }{ f_0(c | \theta_0) } = 
-\frac { \int dx \delta(c-s(x)) \, f_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  } = 
-\frac {c \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, f_0(x | \theta_0)  } = c =\frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
-\end{equation}
-
 %\begin{equation}
-%f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
+%g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon|
 %\end{equation}
+%Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega$.
+%Thus
+%\begin{equation}
+%g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|
+%\end{equation}
+%and we arrive at the result
+%\begin{equation}
+%\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;.
+%\end{equation}
+%
+%
+%
+%
+%
+%
+%
+%Note, in order to form the hypothesis test, one still needs to build the build the $f_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$. 
+%
+%
+%
+%\textbf{Proof (in progress):}
+%
+%We want to show density is the same 
+%
+%\begin{equation}
+%f(q_x|\theta) = \int dx \delta(q_x-q_x(x)) f(x|\theta)  / | \hat{n} \cdot \nabla q_s  |
+%\end{equation}
+%
+%\begin{equation}
+%f(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, f(x|\theta) \, / | \hat{n} \cdot \nabla q_s  |
+%\end{equation}
+%sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $f_1(x)/f_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
+%\begin{equation}
+%\frac {f_1(c | \theta_0) }{ f_0(c | \theta_0) } = 
+%\frac { \int dx \delta(c-s(x)) \, f_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  } = 
+%\frac {c \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, f_0(x | \theta_0)  } = c =\frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
+%\end{equation}
+%
+%
+%
+%%\begin{equation}
+%%f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
+%%\end{equation}
+%
+%
 
 \section*{Acknowldgements}
-We would like to thank Yann LeCun, Philip Stark for their feedback on the
+We would like to thank Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
 project early in its conception and Bal\'azs K\'egl for discussions about the Kaggle challenge and 
-non-standard loss functions.
+non-standard loss functions. KC would like to thank in particular Alex Ihler for challenging discussions that 
+led to a reformulation of the initial idea and Shimon Whiteson for patient feedback in that process. 
+KC would also like to thank Yuri Shirman for reassuring discussions on elementary vector calculus.
 KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 

### 2014-11-18 tex/parametrized-learning.tex
typo, author, acknowledgements

@@
 %\title{Calibrated, Parametrized Learning}
-\author{Kyle Cranmer and Daniel Whiteson}
+\author{Kyle Cranmer}
 %\date{}							% Activate to display a given date or no date
@@
 %\item $f( s | \theta )$ The probability density function for $s$ implied by $f(x|\theta)$ and $s(x)$
-\item $g( s(x; \theta_0, \theta_1) | \theta )$: The probability density  for $s$ implied by $f(x|\theta_0, \theta_1)$ and $s(x;\theta_0, \theta_1)$
-\end{itemize}
+\item $g( s | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $f(x|\theta_0, \theta_1)$ \end{itemize}
 We will assume the $x_e$ are i.i.d., so that $f(D|\theta) = \prod_{e=1}^n f(x_e | \theta)$.
@@
 \end{equation}
-or some monotonic function of $s$. This will be proven below.
+or some monotonic function of that ratio. This will be proven below.
 This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
@@
 
-While the original goal for hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, the machine learning problem is an event-by-event classification problem.
+While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, the machine learning problem is an event-by-event classification problem.
 Of course, this follows from the fact that we assume the $x_e$ to be i.i.d.
@@
 
- The target hypothesis test is be based on 
+ The target hypothesis test is based on 
 \begin{equation}
@@
 \end{equation}
-Here we see that the optimal $T$ for the experiment is composed of a sum over events of a linear linear function of the per-event function $q(x)$.
+Here we see that the optimal $T$ for the experiment is composed of a sum over events of a  linear function of the per-event function $q(x)$.
 A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.
@@
 Fortunately, there is a construction based on $g(s|\theta)$ that works.
-The maximum likelihood estimate is the same as the value that maximizes the ratio to $f(D|\theta_1)$ for some fixed value of $\theta_1$.
+The maximum likelihood estimate is the same as the value that maximizes the ratio with respect to $f(D|\theta_1)$ for some fixed value of $\theta_1$.
 This allows us to use Theorem~1 to find
@@
 \section*{Acknowldgements}
-We would like to thank Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
-project early in its conception and Bal\'azs K\'egl for discussions about the Kaggle challenge and 
-non-standard loss functions. KC would like to thank in particular Alex Ihler for challenging discussions that 
-led to a reformulation of the initial idea and Shimon Whiteson for patient feedback in that process. 
-KC would also like to thank Yuri Shirman for reassuring discussions on elementary vector calculus.
+KC would like to thank in particular 
+Daniel Whiteson for discussions and encouragement throughout the project, 
+Alex Ihler for challenging discussions that 
+led to a reformulation of the initial idea, and Shimon Whiteson for patient feedback in that process. 
+KC would also like to thank Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
+project early in its conception, Bal\'azs K\'egl for discussions about the Kaggle challenge and 
+non-standard loss functions, and Yuri Shirman for reassuring discussions on elementary vector calculus.
 KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 
-KC would like to thank UC-Irvine for their hospitality while this research was carried out and the 
+KC is grateful to UC-Irvine for their hospitality while this research was carried out and the 
 Moore and Sloan foundations for their generous support of the data science environment at NYU.

### 2014-11-19 tex/parametrized-learning.tex
add algorithm, better HEP text.

@@
 \usepackage{amssymb, amsmath}
+\usepackage{algorithm}
+\usepackage{algorithmic}
 
@@
 This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
-used to learn $s(x; \theta_1, \theta_0)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
+used to learn $s(x; \theta_0, \theta_1)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
 from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$.
+In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the discriminative classifier explicitly parametrized in terms of $(\theta_0, \theta_1)$.
 
 While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, the machine learning problem is an event-by-event classification problem.
-Of course, this follows from the fact that we assume the $x_e$ to be i.i.d.
+This follows from the fact that we assume the $x_e$ to be i.i.d.
 
@@
 
-\section{Composite hypotheses and the generalized likelihood ratio}
+\section{Composite hypotheses and the generalized likelihood ratio}\label{S:GLR}
 
@@
 
-\section{Learning the correct mapping and its distribution}
+\section{Learning the correct mapping and its distribution}\label{S:classifier}
 
@@
 
+\subsection{Training a parametrized, discriminative classifier}
+
+Our goal is to train a family of discriminative classifiers parametrized by $\theta_0$ and $\theta_1$, the 
+parameters associated to the null and alternate hypotheses, respectively. While this could be done independently
+for all $\theta_0$ and $\theta_1$, it is desirable and convenient to have a smooth evolution of the classification score as a function of the parameters.
+Thus, we anticipate a single learning stage based on training data with input $(x, \theta_0, \theta_1)_i$ and target $y_i$.
+Somewhat unusually, the unknown values of the parameters are taken
+as input to the classifier, as latent variables who's values will be specified via the enveloping (generalized) likelihood ratio test.
+We denote the learned family of classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based roughly on the following algorithmic flow.
+\begin{algorithm}
+\caption{Training the parametrized classifier}\label{euclid}
+\begin{algorithmic}
+\STATE initialize trainingData = \{\}
+\FOR{ $\theta_0$ in $\Theta$ }
+	\FOR{ $\theta_1$ in $\Theta$ }
+		\STATE generate $x_i^0 \sim f(x|\theta_0)$
+		\STATE append $\{ (x_i^0, \theta_0, \theta_1, y=0) \}$ to trainingData
+		\STATE generate $x_i^1 \sim f(x|\theta_1)$
+		\STATE append $\{ (x_i^1, \theta_0, \theta_1, y=1) \}$ to trainingData
+	\ENDFOR
+\ENDFOR
+\STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_0)$
+\end{algorithmic}
+\end{algorithm}
+
 \section{Typical usage of machine learning in HEP}
@@
 of \textit{background} events. For each event we measure some quantities $x$ that have corresponding distributions 
-$f_b(x|\nu)$ for background and $f_s(x|\nu)$ for signal.
-Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, c_i\}$ generated with some nominal values of the parameters $\nu_0$, where $c=0$ corresponds to background and $c=1$ corresponds to signal.
-The resulting classifier is denoted $s(x)$.
-Based on this classifier and large samples of synthetic data drawn from $f_c(x | \nu)$ we construct the distribution  $g_c(s | \nu)$.
-An example of the distributions of the distribution of $s$ for the signal and background events with $\nu=\nu_0$ is shown in Figure~\ref{fig:tmva}.
+$f_b(x|\nu)$ for background and $f_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing uncertainties in the underlying physics prediction or response of the measurement device.
+In the simple setup, the total model is a mixture of the signal and background components, and $\mu$ is the mixture coefficient associate dot the signal component.
+The generative model in this case is
+\begin{equation}\label{eq:hepGen}
+f( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu f_s( x_e \, |\,  \nu)  + (1-\mu)\, f_b( x_e \,|\, \nu) \,\right] \; ,
+\end{equation}
+New particle searches correspond to the hypothesis test $\mu=0$, and are generally formulated with the
+generalized likelihood ratio profiling over $\nu$.
+
+Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to background and $y=1$ corresponds to signal.
+Following the result of Section~\ref{S:classifier}, the resulting classifier approximates the likelihood ratio of the mixture components $f_s(x|\nu_0)/(f_s(x|\nu_0)+f_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $f(x|\mu=0,\nu_0)/f(x|\mu,\nu_0)$ for all $\mu$.
+The resulting classifier is denoted $\hat s(x)$.
+Based on this classifier and large samples of synthetic data drawn from $f_s(x | \nu)$ and $f_b(x | \nu)$ we construct the distributions  $g_s(\hat s | \nu)$ and $g_b(\hat s | \nu)$.
+An example of the distributions of the distribution of $\hat s$ for the signal and background events with $\nu=\nu_0$ is shown in Figure~\ref{fig:tmva}.
 
@@
  \includegraphics[height=2in]{example-TMVA-ROC.pdf}
-\caption{Left: an example of the distributions $f_b(s|\nu)$ and $f_s(s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).
+\caption{Left: an example of the distributions $g_b(\hat s|\nu)$ and $g_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).
 Right: the corresponding ROC curve (right) for this and other classifiers.
-Figures taken from TMVA manual.}
+(Figures taken from TMVA manual.)}
 \label{fig:tmva}
@@
 \end{figure}
+\bigskip
 
@@
 \begin{equation}\label{eq:typicalML}
-f( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu g_s( s(x_e) \, |\,  \nu)  + (1-\mu)\, g_b( s(x_e) \,|\, \nu) \,\right] \; ,
+g( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu g_s( \hat s(x_e) \, |\,  \nu)  + (1-\mu)\, g_b( \hat s(x_e) \,|\, \nu) \,\right] \; ,
 \end{equation}
@@
 In practice, we take the classifier as fixed and then propagate uncertainty through the classifier as in Eq.~\ref{eq:typicalML}.
-Building the distribution $g(s|\nu)$ for values of $\nu$ other than the nominal $\nu_0$ used to train the classifier can be thought of as a calibration necessary for classical statistical inference; however, this classifier is clearly not optimal for $\nu \ne \nu_0$.
+Building the distribution $g(\hat s|\nu)$ for values of $\nu$ other than the nominal $\nu_0$ used to train the classifier can be thought of as a calibration necessary for classical statistical inference; however, this classifier is clearly not optimal for $\nu \ne \nu_0$.
 
@@
 
-The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $s(x; \nu_0, \nu_1)$.
+The standard use of machine learning in HEP can be improved by training a parametrized, discriminative classifier corresponding to the generalized likelihood ratio test 
+\begin{equation}
+\lambda(\mu) = \frac{f(D|\mu, \hat{\hat{\nu}})}{f(D|\hat \mu, {\hat{\nu}})} \;,
+\end{equation}
+following the approach outlined in Section~\ref{S:GLR}.
+
+There is an interesting distinction between this approach and the standard use in which the classifier is trained for a fixed $\nu_0$.
+In the standard use one trains a classifier for signal vs.
+background, which is equivalent (in an ideal setting) to training a classifier for  null (background-only) vs.
+alternate (signal-plus-backgound) as
+\begin{equation}
+ \frac{f(x| 0, \nu_0)}{f(x|\hat \mu, \nu_0)} =  \frac{f_b(x| \nu_0)}{ \mu f_s( x_e \, |\,  \nu_0)  + (1-\mu)\, f_b( x_e \,|\, \nu_0)} = \left[ c_1 + c_2   \frac{f_s(x| \nu_0)}{ f_b( x_e \,|\, \nu_0)} \right ]^{-1} \; ,
+\end{equation}
+and $c_1$ and $c_2$ are constants.
+Specifically, the two approaches likelihood ratios are in one-to-one correspondence, so an ideal algorithm would lead to equivalent tests.
+In contrast, in the case of the generalized likelihood ratio test
+\begin{equation}
+ \frac{f(x| 0, \hat{\hat{ \nu}})}{f(x|\hat \mu, \hat\nu)} =  \frac{f_b(x| \hat{\hat{ \nu}})}{ \hat \mu f_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, f_b( x_e \,|\, \hat \nu)} \; ,
+\end{equation}
+the background components don't cancel and there is an additional term $f_b(x| \hat{\hat{ \nu}})/f_b(x| {\hat{ \nu}})$.
+In practice, with classifiers of finite capacity, there will be some tradeoff between taking into account this additional term and the more challenging learning problem when $\mu$ is very small.
+
+%
+%The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $\hat s(x; \nu_0, \nu_1)$.
 Creating the family of classifiers is straight forward, one simply augments the training data with $x$ examples drawn from several values of $\nu$ and then includes the corresponding value of $\nu$ as an input to the classifier.
-Thus $\{x_e,c_e\} \to \{x_e,\nu_{0,e},\nu_{1,e}, c_e\}$ leading to a parametrized learner $s(x)\to s(x;\nu_0, \nu_1)$.
-This leads to a complication: one does not know the value of $\nu$ to use when evaluating the parametrized classifier $s(x;\nu_0, \nu_1)$, so one cannot simply pre-compute $s_e$ before performing the likelihood fit.
+Thus $\{x_e,c_e\} \to \{x_e,\nu_{0,e},\nu_{1,e}, c_e\}$ leading to a parametrized learner $\hat s(x)\to \hat s(x;\nu_0, \nu_1)$.
+This leads to a complication: one does not know the value of $\nu$ to use when evaluating the parametrized classifier $\hat s(x;\nu_0, \nu_1)$, so one cannot simply pre-compute $\hat s_e$ before performing the likelihood fit.
 However, when performing a likelihood ratio test, one can calculate the maximum likelihood estimate and conditional maximum likelihood estimate according to Eq.~\ref{eq:mle} and then form the equivalent likelihood ratio test according to Eq.~\ref{eq:equivLRtest}.
@@
 \section*{Acknowldgements}
-KC would like to thank in particular 
+KC would like to thank  
 Daniel Whiteson for discussions and encouragement throughout the project, 

### 2014-11-19 tex/parametrized-learning.tex
embedding the classifier into the likelihood

@@
 As we will show, our goal is to learn a monotonic function of $f(x|\theta_0)/f(x|\theta_1)$ that shares the same iso-contours.
-Then the remaining challenge is to find the appropriate rescaling that gives back  linear function $q(x)$.
+Then the remaining challenge is to find the appropriate rescaling that gives back  linear function of $q(x)$.
 Our claim is that the generative model $f(x|\theta)$ can be used to calibrate $g(s|\theta)$ and that
@@
 Denote the approximate dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{g}(\hat{s}|\theta)$.
-In general we will be interested in the machine learning problem that approximates these distributions based on samples $\{(x, \theta)\}$ drawn from the generative model $f(x|\theta)$.
+In general we will be interested in the machine learning problem that approximates these distributions based on samples $\{x_i\}$ drawn from the generative model $f(x|\theta)$.
 In particular,
@@
 large samples from each model and train a classifier. To be concrete, let's use $f(x|\theta_0)$ to generate training 
-data $(x_i, \theta_i=\theta_0, y_i=0)$ and $f(x|\theta_1)$ to generate training data $(x_i, \theta_i=\theta_1, y_i=1)$. 
+data $(x_i,  y_i=0)$ and $f(x|\theta_1)$ to generate training data $(x_i , y_i=1)$. 
 If we use the squared-loss function, then the expected loss is
@@
 
-Proof: Consider a variation about $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ given by $\hat{s}'(x) = \hat{s}(x) +h(x)$. The change in the expected loss is given by 
+Proof (Sketch): Consider a variation about $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ given by $\hat{s}'(x) = \hat{s}(x) +h(x)$. The change in the expected loss is given by 
 \begin{equation}
@@
 
-This treatment shows that in the asymptotic limit of samples from the generative model, that we can approximate arbitrarily well the original likelihood ratio test.
+This treatment shows that in the asymptotic limit of large samples from the generative model, that we can approximate arbitrarily well the original likelihood ratio test.
 With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{g}(x)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test.
-This will depend in general on the ultimate goal of the test and is left as a subject for future work.
+This will depend in general on the ultimate goal of the test.
+We know that in the case of composite hypothesis tests that there is in general, not uniformly most powerful test, thus it is likely that a decision theoretic approach taking into account some weighting or utility over the space $\Theta$ is necessary.
+This is left as a subject for future work.
 
@@
 
-Our goal is to train a family of discriminative classifiers parametrized by $\theta_0$ and $\theta_1$, the 
+We are left with the practical question of how to train a family of discriminative classifiers parametrized by $\theta_0$ and $\theta_1$, the 
 parameters associated to the null and alternate hypotheses, respectively. While this could be done independently
@@
 Somewhat unusually, the unknown values of the parameters are taken
-as input to the classifier, as latent variables who's values will be specified via the enveloping (generalized) likelihood ratio test.
+as input to the classifier, as latent variables whose values will be specified via the enveloping (generalized) likelihood ratio test.
 We denote the learned family of classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based roughly on the following algorithmic flow.
@@
 \ENDFOR
-\STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_0)$
+\STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_1)$
 \end{algorithmic}
-\end{algorithm}
+\end{algorithm}%\vspace{-1.5em}
+
+While the function $f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ will minimize the expected squared loss based on 
+training data produced according to the algorithm above, it is not clear how training data from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real world classifier with finite capacity.
+This is left as an area for future work.
+
+\subsection{Embedding the classifier into the likelihood}
+
+In most settings that make use of likelihood ratio tests, the probability densities directly model the observed data via $f(x|\theta)$.
+In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$.
+In the parametrized setting this it is not possible to pre-compute $\hat s(x; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$, so we must computationally we must embed the classifier into the likelihood function to carry out the composition $g\circ s$.
+A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and discriminative classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}.
+
+In both cases, constructing the density $\hat g(\hat s|\theta)$ requires running the generative model at $\theta$, and in the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize $\theta$ needs access to the generative model $f(x|\theta)$.
+This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
+In practice, one may want to interpolate the distribution between discrete values of $\theta$ to produce a continuous parametrization for $\hat g(s | \theta)$.
+In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
 

### 2014-11-19 tex/orig-parametrized-learning.tex

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}

\title{Calibrated, Parametrized Learning}
\author{Kyle Cranmer and Daniel Whiteson}
%\date{}							% Activate to display a given date or no date

\begin{document}
\maketitle

\section{Introduction}
%\subsection{}

Notation:

\begin{itemize}
 \item $x$: a vector of observed quantities for each event (energy, momenta, angles of particles)
 \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$
 \item $\theta$: parameters of a statistical model
\item $f_c(x| \theta)$:  probability density function (statistical model) for $x$ for category $c\in\{0,1\}$.
\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta)$: real-valued score from a parametrized machine learning classification algorithm (or any map $s: X\otimes \Theta \to\mathbb{R}$)
\item $f_c( s | \theta )$ The probability density function for $s$ implied by $f_c(x|\theta)$ and $s(x)$
\item $f_c( s(\cdot; \theta) | \theta )$ The probability density function for $s$ implied by $f_c(x|\theta)$ and $s(x;\theta)$

\end{itemize}

\subsection{Typical usage of machine learning in HEP}

In high-energy physics (HEP) we typically are searching for or measuring the properties of some 
class of events, generically referred to as \textit{signal}, in the presence of a separate class 
of \textit{background} events. For each event we measure some quantities $x$ that have corresponding distributions 
$f_0(x|\theta)$ for background and $f_1(x|\theta)$ for signal.  Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, c_i\}$ generated with some nominal values of the parameters $\theta_0$, where $c=0$ corresponds to background and $c=1$ corresponds to signal. The resulting classifier is denoted $s(x)$. Based on this classifier and large samples of synthetic data drawn from $f_c(x | \theta)$ we construct the distribution  $f_c(s | \theta)$. An example of the distributions of the distribution of $s$ for the signal and background events with $\theta=\theta_0$ is shown in Figure~\ref{fig:tmva}.

\begin{figure}[htbp]
\begin{center}
 \includegraphics[height=2in]{example-TMVA-BDT.pdf}
 \includegraphics[height=2in]{example-TMVA-ROC.pdf}
\caption{Left: an example of the distributions $f_0(s|\theta)$ and $f_1(s|\theta)$ when the classifier $s$ is a boosted-decision tree (BDT). Right: the corresponding ROC curve (right) for this and other classifiers. Figures taken from TMVA manual.}
\label{fig:tmva}
\end{center}
\end{figure}

These steps lead to a subsequent statistical analysis where one observes in data $\{x_e\}$, where $e$ is an event index running from $1$ to $n$. For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution. In particular, one forms the statistical model
\begin{equation}\label{eq:typicalML}
f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e) \,|\, \theta) \,\right] \; ,
\end{equation}
where $\mu=0$ is the null (background-only) hypothesis and $\mu>0$ is the alternate (signal-plus-background) hypothesis.\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known.} Typically, we are interested in inference on $\mu$ and $\theta$ are nuisance parameters; though, sometimes $\theta$ may include some components that we are also wish to infer (like the mass of a new particle that affects the distribution $x$ for the signal events).

\subsection{Comments on typical usage of machine learning in HEP}

Nuisance parameters are an after thought in the typical usage of machine learning in HEP. In fact, most machine learning discussions would only consider $f_0(x)$ and $f_1(x)$. However, as experimentalists we know that we must account for various forms of systematic uncertainty, parametrized by $\theta$. In practice, we take the classifier as fixed and then propagate uncertainty through the classifier as in Eq.~\ref{eq:typicalML}. Building the distribution $f(s(x)|\theta)$ for values of $\theta$ other than the nominal $\theta_0$ used to train the classifier can be thought of as a calibration necessary for classical statistical inference; however, this classifier is clearly not optimal for $\theta \ne \theta_0$.

\subsection{A more powerful  approach}

The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $s(x; \theta)$. Creating the family of classifiers is straight forward, one simply augments the training data with $x$ examples drawn from several values of $\theta$ and then includes the corresponding value of $\theta$ as an input to the classifier. Thus $\{x_e,c_e\} \to \{x_e,\theta_e, c_e\}$ leading to a parametrized learner $s(x)\to s(x;\theta)$. This leads to a complication: one does not know the value of $\theta$ to use when evaluating the parametrized classifier $s(x;\theta)$.  However, when performing statistical inference one will evaluate the likelihood of the data at a given value of $\theta$, so this is a natural choice.  Thus one arrives at the generalization of Eq.~\ref{eq:typicalML}
\begin{equation}\label{eq:parametrizedML}
f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s(x_e;\theta) \, |\,  \theta)  + (1-\mu)\, f_0( s(x_e;\theta) \,|\, \theta) \,\right] \; .
\end{equation}

The construction in Eq.~\ref{eq:parametrizedML} presents a computational challenge as for each value of $\theta$ one has a new mapping $s: x\to\mathbb{R}$. The most naive realization of this would be to generate synthetic data $\{x_e\}$ according to $f_c(x|\theta)$, evaluate $s(x_e ; \theta)$, and then use a histogram, kernel density, or other non-parametric density estimation procedure to estimate $f_c( s | \theta)$. This approach is not realistic in situations where $f_c(x|\theta)$ is realized with expensive computer simulations; thus, some interpolation strategy or emulator over a fixed set of parameter points $\{\theta_i\}$ would typically be used.

\section{Possible Extensions}
The approach describe thus far has been agnostic as to the type of classifier used to produce $s(x)$ and $s(x; \theta)$. In practice, we have used standard packages for the training of these learners. In particular, the loss function for these learning algorithms is based on classification performance of individual $(x_e, c_e)$ feature $\to$ target examples. However, in overarching statistical procedure that uses the statistical model in Eq.~\ref{eq:typicalML} or \ref{eq:parametrizedML} the quantity we wish to optimize is a property of an entire dataset $\{x_e\}$ or ensembles of datasets that can be generated from the statistical models $f_c(x|\theta)$.

For example, in typical hypothesis testing problems posed in the Neyman-Pearson context one would be interested in minimizing type II error under a fixed rate of typeI error rate. In the presence of systematic uncertainty and nuisance parameters, the loss function may become more complicated. The recent Higgs boson machine learning challenge hosted by Kaggle, the evaluation was the approximate median significance (AMS) that characterized the power of a hypothesis test between $\mu=0$ (the background only hypothesis) and $\mu=\mu_0$ (the signal+background hypothesis, where $\mu_0$ is a small number reflecting the small Higgs signal that would be found in addition to the much larger background process) in the presence of uncertainty on the background normalization. 
Interestingly, the top entries to the challenge did not improve by focusing on this alternative loss function 
in training their classifier, but instead used it for the optimization of a threshold on the score (a working point on the ROC curve) that was related to the way the challenge was designed. This suggests that the standard 
classification loss functions lead to similar optimizations and that refinements for the custom loss functions
are not easy to come by.

More generally, there is an avenue of research associated to custom loss functions specifically designed 
for these real world statistical problems. One direction might be to choose loss functions that make 
some balance between the standard classification loss, take into account robustness with respect to nuisance parameters, and population-level quantities like type I and II error.

\subsection{A different starting point}

Instead of evolving from the traditional usage of ML in HEP, let's start from scratch with $f_c(x|\theta)$ and a well defined objective.

\subsubsection{Simple Hypothesis Testing}

Consider the situation of simple hypothesis testing between the background-only $\mu=0$ and signal+background hypothesis $\mu=\mu_0=\nu_s/(\nu_s+\nu_b)$ for $\theta=\theta_0$.  Let us assume that not only can we use the $f_c(x|\theta_0)$ as a generative model, but that we can evaluate the probability density at any point $x$.  The statistical model is given by
\begin{equation}\label{eq:NP}
f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( x_e \, |\,  \theta)  + (1-\mu)\, f_0( x_e \,|\, \theta) \,\right] \; .
\end{equation}
The objective is to find a  test statistic $T(D)$ (a map from the space of the data to $\mathbb{R}$)  
and a threshold $k_\alpha$ that minimizes the rate of type-II error (i.e. $P(T(D) < k_\alpha | \mu=\mu_0, \theta_0) = \beta$ under the constraint of a fixed rate of type-I error ( i.e.  $P(T(D) > k_\alpha | \mu=0, \theta_0) = \alpha$ ).
The Neyman-Pearson lemma states that the optimal test statistic is the likelihood ratio, or equivalently, the log-likelihood ratio 
\begin{equation}
T =  \log \frac{f(D | \mu=\mu_0, \theta_0)}{f(D | \mu=0, \theta_0)}
\end{equation}
When the $x_e$ are i.i.d., then the log-likelihood ratio expands into a sum
\begin{equation}
T =   \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(x_e | \theta_0) }{ f_0(x_e | \theta_0) } \right]}_{q(x_e)} + c_2\;,
\end{equation}
where $c_1=\mu/(1-\mu)$ and $c_2=\log(1-\mu)$ are constants for the simple hypothesis test.

Here we see that the optimal $T$ for the experiment is composed of a sum over events of a linear linear function of the per-event function $q(x)$. A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.

The important part of the per-event function $q(x)$ is that it defines contours in the feature space $x$. These contours are also equivalent to the function $f_1(x|\theta_0)/f_0(x|\theta_0)$, the signal-to-background ratio. If we can find a machine learner $s(x)$ that is a monotonic function of $f_1(x|\theta_0)/f_0(x|\theta_0)$ it will share the same contours. Then the remaining challenge is to find the appropriate rescaling that gives back  linear function $q(x)$.

\textbf{Postulate:}
\[
T' = \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(s(x_e) | \theta_0) }{ f_0(s(x_e) | \theta_0) } \right]}_{q(s_e)} \;,
\]
leads to an equivalent test, where $s_e \equiv s(x_e)$
\[
f_c(s) = \int dx \, \delta(s-s(x)) \, f_c(x)  \,  |\partial s / \partial x|^{-1}
\]

Note, in order to form the hypothesis test, one still needs to build the build the $f_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$.

\textbf{Proof (in progress):}

We want to show density is the same

\[
f_c(q_x) = \int dx \delta(q_x-q_x(x)) f_c(x) / |\partial q_x / \partial x|
\]

\[
f_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, f_c(x) \, / |\partial q_s / \partial x|
\]
sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $f_1(x)/f_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
\[
\frac {f_1(c | \theta_0) }{ f_0(c | \theta_0) } = 
\frac { \int dx \delta(c-s(x)) \, f_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  } = 
\frac {c \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, f_0(x | \theta_0)  } = c =\frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
\]

%\[
%f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
%\]

This is the test statistic that one would naturally get from constructing the likelihood function for the 1-d distributions of the output of the machine learner
\begin{equation}\label{eq:NP}
f( \{s_e\} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s_e \, |\,  \theta)  + (1-\mu)\, f_0( s_e \,|\, \theta) \,\right] \; .
\end{equation}
Thus this motivates Eq.~\ref{eq:typicalML} in the more general case.

Question:
Where does this fit in:
\[
T'' = \sum_{e=1}^n \underbrace{ \log \left[ 1+c_1 s(x_e) \right] }_{q'(s_e)} \;,
\]

Usefulness:
If we can readily compute $f_c(x|\theta_0)$, then there would be no need for machine learning algorithms. However,  these densities are often difficult to compute though one can readily use them as a generative model to produce samples. In that case, one might wish to use a machine learning algorithm to find $s(x) \approx f_1(x|\theta_0) / f_0(x|\theta_0)$.

Conclusion, while it is possible to start directly with the Neyman-Pearson loss function of type-II error under the constraint of fixed size and try to find a per-experiment learner $T(D)$, it is more convenient to cast the problem in terms of learning a per-event function $s(x)$, and then calibrate it to be a linear function of $q(x)$.

\newpage

Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $f_1(x)$ and $f_0(x)$ define 
\[ 
s(x)=\frac{f_1(x)}{f_2(x)}
\].
Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
I wish to show
\[
\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
\]
$\forall x \ni s(x)=c$.

Sketch of Proof: 
let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$. The  coordinate perpendicular to $\Omega$ can be written $y = x+\epsilon \hat{n}$. In these coordinates, the integral $\int dx \to \int d\Omega d\epsilon$, thus
\[
g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon|
\]
Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega$.
Thus
\[
g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|
\]
and we arrive at the result
\[
\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;.
\]

\newpage

Discriminative classifiers calibrated with generative models for likelihood ratio based tests

In many areas of science, likelihood ratio tests  are established tools for statistical inference. 
Directly constructing the likelihood ratio for high-dimensional observations 
is often not possible or is computationally impractical. Here we demonstrate how 
discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
a generative model for the data is available for calibration.

\subsection{Simple Hypothesis Tests and the Neyman Pearson Lemma}
\subsection{Composite Hypothesis  and generalized likelihood ratio tests}

\section*{Acknowldgements}
We would like to thank Yann LeCun, Philip Stark for their feedback on the
project early in its conception and Bal\'azs K\'egl for discussions about the Kaggle challenge and 
non-standard loss functions.
KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 
KC would like to thank UC-Irvine for their hospitality while this research was carried out and the 
Moore and Sloan foundations for their generous support of the data science environment at NYU.

\end{document}

### 2014-11-20 tex/parametrized-learning.tex
references

@@
 \usepackage{algorithmic}
+\usepackage{xfrac}
 
@@
 This follows from the fact that we assume the $x_e$ to be i.i.d.
+
+\section{Comments on classification and frequentist hypothesis tests}
+
+Significant literature exists around generative classifiers and discriminative classifiers.
+Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule.
+In contrast, discriminative classifiers model the posterior $p(y|x)$ directly.
+For classification tasks, one then thresholds on $p(y|x)$.
+In both cases this description in terms of a posterior requires a prior distribution for $p(y)$, which is either modeled explicitly or learned from the training data.
+This formulation of classification is somewhat confusing in the setting of the current work.
+
+First of all, the generative model $f(x|\theta)$ here is a generative statistical model, not a generative classifier.
+We think of the  $f(x|\theta)$ along the lines of a traditional scientific theory, able to make predictions about $x$ and being motivated by domain-specific considerations.
+For example, in the context of high energy particle physics $f(x|\theta)$ is based on quantum field theory and a detailed simulation of the particle detector and data processing algorithms that transform raw sensor data into the feature vector $x$.
+The point here is to make the distinction of $f(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.
+
+Secondly, we are not interested in thresholded classification on individual events $x_e$, but are interested in a thresholded hypothesis test on an entire data set based on the likelihood ratio test statistic $T(D)$.
+Additionally, we know that both discriminative and generative classifier scores are often poorly calibrated. 
+We wish to have well calibrated statements, but not about $p(y|x)$ but p-values associated to $P(T(D) > k |\theta)$.
+
+Lastly, in the setting of frequentist hypothesis we do not have a prior $\pi(\theta)$. 
+While we can use the generative models to produce training data $(x,y=0)$ generated 
+from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$, the relative mix $p(y)$ 
+is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then 
+\begin{equation}
+p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{f(x|\theta_1)}{f(x|\theta_0)+f(x|\theta_1))} \;,
+\end{equation}
+which is monotone with the desired likelihood ratio $f(x|\theta_1)/f(x|\theta_0)$ and free from the
+complications of the prior $\pi(\theta)$ that is undefined in the frequentist context.
+
+Since the prior $p(y)$ has no clear meaning for the test at hand and because the classifier score $p(x|y)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
 
@@
 
+Related work:
+"Classification with hybrid generative/discriminative models"
+
+Refs to include:
+\cite{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,TommiJaakkola}
+
 \section*{Acknowldgements}
@@
 
+\bibliographystyle{unsrt}
+\bibliography{learning.bib}
+%\begin{thebibliography}
+%\parametrized-learning}
+%\end{thebibliography
+
 \end{document}

### 2014-11-20 tex/parametrized-learning.tex
style file and bib

@@
 \documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
+
+\usepackage{jmlr2e}
+
+%\jmlrheading{1}{2014}{X-XX}{4/00}{10/00}{Kyle Cranmer}
+
+%\ShortHeadings{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}{Kyle Cranmer}
+
 \usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
@@
 \begin{document}
+
+\author{\name Kyle Cranmer \email kyle.cranmer@nyu.edu \\
+       \addr Center for Cosmology and Particle Physics\\
+       Center for Data Science\\
+       New York University \\
+       New York, NY 10003, USA
+}
+
+%\editor{Leslie Pack Kaelbling}
+\editor{}
+
+\maketitle
+
+\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
+words
+\end{abstract}
+%
+%\begin{keywords}
+%  Bayesian Networks, Mixture Models, Chow-Liu Trees
+%\end{keywords}
+
 \maketitle
@@
 
+\subsection{Related work}
+
+Refs to include:
+\cite{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,TommiJaakkola}
+
+\cite{JMLR:v14:tong13a}
+
 \section{Conclusions}
@@
 
-Related work:
-"Classification with hybrid generative/discriminative models"
-
-Refs to include:
-\cite{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,TommiJaakkola}
-
 \section*{Acknowldgements}
@@
 
-\bibliographystyle{unsrt}
+\bibliographystyle{natbib}
 \bibliography{learning.bib}

### 2014-11-20 tex/parametrized-learning.tex
line numbers

@@
 \usepackage{jmlr2e}
-
+\usepackage{lineno}
+\linenumbers
 %\jmlrheading{1}{2014}{X-XX}{4/00}{10/00}{Kyle Cranmer}
@@
 
-Since the prior $p(y)$ has no clear meaning for the test at hand and because the classifier score $p(x|y)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
+Since the prior $p(y)$ has no clear meaning for the test at hand and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
 

### 2015-03-15 tex/parametrized-learning.tex
tweaks and decomposing test between mixture models

@@
 %\item $f( s | \theta )$ The probability density function for $s$ implied by $f(x|\theta)$ and $s(x)$
-\item $g( s | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $f(x|\theta_0, \theta_1)$ \end{itemize}
+\item $g( s | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $f(x|\theta)$ \end{itemize}
 We will assume the $x_e$ are i.i.d., so that $f(D|\theta) = \prod_{e=1}^n f(x_e | \theta)$.
@@
 \begin{equation}
-s_e = s(x_e; \theta_0, \theta_1) = \frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)} \; 
+s_e = s(x_e; \theta_0, \theta_1) = M\left(\frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)} \right) \; 
 \end{equation}
-or some monotonic function of that ratio. This will be proven below.
+where $M$ is any monotonic function. This will be proven below.
 This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
@@
 Additionally, we know that both discriminative and generative classifier scores are often poorly calibrated. 
-We wish to have well calibrated statements, but not about $p(y|x)$ but p-values associated to $P(T(D) > k |\theta)$.
+We wish to have well calibrated statements regarding p-values defined by $P(T(D) > k |\theta)$, not well calibrated posterior probabilities $p(y|x)$.
 
-Lastly, in the setting of frequentist hypothesis we do not have a prior $\pi(\theta)$. 
+Lastly, in the setting of frequentist hypothesis tests, we do not have a prior $\pi(\theta)$. 
 While we can use the generative models to produce training data $(x,y=0)$ generated 
@@
 \end{equation}
-It is sufficient to show that $q(x_e) = q(s(x_e))$ $ \forall x\in\Omega_c$.
+It is sufficient to show that $q(x) = q(s(x))$ $ \forall x\in\Omega_c$.
 
 For notational simplicity, let $f_0(x) = f(x|\theta_0)$, $f_1(x) = f(x|\theta_1)$, and $s(x)=s(x; \theta_1, \theta_0)$.
-The distribution of $x$ totally determines the distribution of $s$ via the change of variables $x\to s$. 
+The distribution of $x$ totally determines the distribution of $s$. 
 In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
@@
 
-Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g. histograms or kernel density estimation) to approximate $\hat{g}(x|\theta)$.
+Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g. histograms or kernel density estimation) to approximate $\hat{g}(\hat{s}|\theta)$.
 
 This treatment shows that in the asymptotic limit of large samples from the generative model, that we can approximate arbitrarily well the original likelihood ratio test.
-With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{g}(x)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test.
+With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{g}(\hat{s}|\theta)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test.
 This will depend in general on the ultimate goal of the test.
-We know that in the case of composite hypothesis tests that there is in general, not uniformly most powerful test, thus it is likely that a decision theoretic approach taking into account some weighting or utility over the space $\Theta$ is necessary.
+We know that in the case of composite hypothesis tests that there is in general no uniformly most powerful test, thus it is likely that a decision theoretic approach taking into account some weighting or utility over the space $\Theta$ is necessary.
 This is left as a subject for future work.
@@
 In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$.
-In the parametrized setting this it is not possible to pre-compute $\hat s(x; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$, so we must computationally we must embed the classifier into the likelihood function to carry out the composition $g\circ s$.
+In the parametrized setting this it is not possible to pre-compute $\hat s(x; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$, so  we must embed the classifier into the likelihood function to carry out the composition $g\circ s$.
 A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and discriminative classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}.
@@
 of \textit{background} events. For each event we measure some quantities $x$ that have corresponding distributions 
-$f_b(x|\nu)$ for background and $f_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing uncertainties in the underlying physics prediction or response of the measurement device.
-In the simple setup, the total model is a mixture of the signal and background components, and $\mu$ is the mixture coefficient associate dot the signal component.
-The generative model in this case is
+$f_b(x|\nu)$ for background and $f_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing 
+uncertainties in the underlying physics prediction or response of the measurement device. In the simple setup, the 
+total model is a mixture of the signal and background components, and $\mu$ is the mixture coefficient associate 
+dot the signal component. The generative model in this case is
 \begin{equation}\label{eq:hepGen}
@@
 and $c_1$ and $c_2$ are constants.
-Specifically, the two approaches likelihood ratios are in one-to-one correspondence, so an ideal algorithm would lead to equivalent tests.
+Specifically, the two  likelihood ratios are in one-to-one correspondence, so an ideal algorithm would lead to equivalent tests.
 In contrast, in the case of the generalized likelihood ratio test
@@
 In practice, with classifiers of finite capacity, there will be some tradeoff between taking into account this additional term and the more challenging learning problem when $\mu$ is very small.
+
+\subsection{Decomposing tests between mixture models into their components}
+
+It is common that the generative model for the low-level features is a mixture model of several components
+\begin{equation}
+f(x|\theta)=\sum_c w_c(\theta) f_c(x| \theta) \;.
+\end{equation}
+In the case of particle physics, the distributions $f(x|\theta)$ is not a Gaussian Mixture Model, but mixture of  complicated distributions associated to relatively few types of particle interactions.
+Moreover, when searching for a new particle, the null hypothesis would correspond to some of the coefficients $w_c=0$ while the alternate ``signal-plus-background'' hypothesis would have $0<w_{c\in \rm signal} \ll w_{c\in \rm background}$.
+In some cases $w_{c\in \rm signal} / w_{c\in \rm background} <10^{-6}$, which means the alternate hypothesis is a small perturbation to the null hypothesis.
+This can be a challenge for typical classifiers because they should devote their capacity to the region where $f_{c\in \rm signal}(x) / f_{c\in \rm background}(x)$ is relatively large.
+Lastly, even when the distributions $f_c(x|\theta)$ are well known, it is often the case that the coefficients are uncertain or treated as completely unknown.
+These all present challenges to machine learning algorithms that aim to learn $s(x; \theta_0, \theta_1)$.
+
+%\begin{equation}
+%\frac{f(x|\theta_0)}{f(x|\theta_1)}= \frac{\sum_c w_c(\theta_0) f_c(x| \theta_0)}{\sum_c w_c(\theta_1) f_c(x| \theta_1)} 
+%\end{equation}
+%
+%\begin{equation}
+%\frac{f(x|\theta_0)}{f(x|\theta_1)}= \sum_c \frac{ w_c(\theta_0) f_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)} 
+%\end{equation}
+
+However, it is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
+\begin{eqnarray}
+\frac{f(x|\theta_0)}{f(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) f_c(x| \theta_0)}{\sum_c w_c(\theta_1) f_c(x| \theta_1)} 
+\\
+%&=& \sum_c \frac{ w_c(\theta_0) f_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)} \\
+%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)}{ w_c(\theta_0) f_c(x| \theta_0)}  \right]^{-1} \\
+&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ f_{c'}(x| \theta_1)}{  f_c(x| \theta_0)}  \right]^{-1} \\
+&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ g_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ g_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
+\end{eqnarray}
+The second line is a trivial, but useful decomposition into pair-wise classification between $f_{c'}(x|\theta_1)$ and $f_c(x|\theta_0)$.
+The third line uses the previous results to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
+In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $g_{c}(s_{c,c',\theta_0, \theta_1}| \theta_1)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).
 

### 2015-03-26 tex/parametrized-learning.tex
removed pedantic parts, added description of other work

@@
 
+%\title{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}
 \title{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}
@@
 \begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-words
+We demonstrate how discriminative classifiers can be used to approximate 
+(generalized) likelihood ratio tests over high-dimensional data when a generative model 
+for the data is available for training and calibration.  
 \end{abstract}
@@
  \item $\theta$: parameters of a statistical model
-\item $f(x| \theta)$:  probability density  (statistical model) for $x$ 
+\item $f(x| \theta)$:  probability density  (statistical model) for $x$ given $\theta$
 %\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
@@
 We will assume the $x_e$ are i.i.d., so that $f(D|\theta) = \prod_{e=1}^n f(x_e | \theta)$.
+
+\newpage
 
@@
 if 
-\begin{equation}
-s_e = s(x_e; \theta_0, \theta_1) = M\left(\frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)} \right) \; 
+\begin{equation}\label{eq:montonic}
+s_e = s(x_e; \theta_0, \theta_1) = m\left(\, f(x_e|\theta_0) / f(x_e|\theta_1) \,\right) \; 
+%s_e = s(x_e; \theta_0, \theta_1) = m\left(\frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)} \right) \; 
 \end{equation}
-where $M$ is any monotonic function. This will be proven below.
+where $m$ is any strictly increasing or decreasing function. This result will be proven below.
 This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
@@
 
-While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, the machine learning problem is an event-by-event classification problem.
+While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, we are able to reformulate it such that the machine learning problem is an event-by-event classification problem.
 This follows from the fact that we assume the $x_e$ to be i.i.d.
@@
 
-Significant literature exists around generative classifiers and discriminative classifiers.
+Significant literature exists around generative and discriminative classifiers~\citep{AndrewY.Ng}.
 Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule.
@@
 In both cases this description in terms of a posterior requires a prior distribution for $p(y)$, which is either modeled explicitly or learned from the training data.
-This formulation of classification is somewhat confusing in the setting of the current work.
+This familiar formulation of classification may lead to some confusion in the setting of the current work.
 
-First of all, the generative model $f(x|\theta)$ here is a generative statistical model, not a generative classifier.
+The first possible source of confusion we wish to avoid is that  $f(x|\theta)$  is a generative \textit{statistical model} for the features $x$, not a generative classifier.
 We think of the  $f(x|\theta)$ along the lines of a traditional scientific theory, able to make predictions about $x$ and being motivated by domain-specific considerations.
 For example, in the context of high energy particle physics $f(x|\theta)$ is based on quantum field theory and a detailed simulation of the particle detector and data processing algorithms that transform raw sensor data into the feature vector $x$.
-The point here is to make the distinction of $f(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.
+Moreover, we are not attempting to learn the generative model $f(x|\theta)$, we are taking it as given and trying to learn the corresponding likelihood ratio test.
+%The point here is to make the distinction of $f(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.
 
-Secondly, we are not interested in thresholded classification on individual events $x_e$, but are interested in a thresholded hypothesis test on an entire data set based on the likelihood ratio test statistic $T(D)$.
+The second possible source of confusion is that the likelihood ratio $T(D)$ is aimed at tests based on the entire dataset; we are not interested in thresholded classification on individual events $x_e$.
 Additionally, we know that both discriminative and generative classifier scores are often poorly calibrated. 
-We wish to have well calibrated statements regarding p-values defined by $P(T(D) > k |\theta)$, not well calibrated posterior probabilities $p(y|x)$.
+For instance, often we wish to have well calibrated p-values defined by $P(T(D) > k |\theta)$, not well calibrated posterior probabilities $p(y|x)$.
 
@@
 \end{equation}
-which is monotone with the desired likelihood ratio $f(x|\theta_1)/f(x|\theta_0)$ and free from the
-complications of the prior $\pi(\theta)$ that is undefined in the frequentist context.
+which is monotonic with the desired likelihood ratio $f(x|\theta_1)/f(x|\theta_0)$.
+Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
 
-Since the prior $p(y)$ has no clear meaning for the test at hand and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
+% and we need not
+%interpret the balance of the training data in terms of a prior probability for the two hypotheses.
+%free from the
+%complications of the prior $\pi(\theta)$ that is undefined in the frequentist context.
 
-\section{Dimensionality reduction}
+\section{Dimensionality reduction and calibration}
 
@@
 The important part of the per-event function $q(x)$ is that it defines iso-contours in the feature space $x$.
-As we will show, our goal is to learn a monotonic function of $f(x|\theta_0)/f(x|\theta_1)$ that shares the same iso-contours.
+As we will show, our goal is to learn a monotonic function of $f(x|\theta_0)/f(x|\theta_1)$, which will share the same iso-contours.
 Then the remaining challenge is to find the appropriate rescaling that gives back  linear function of $q(x)$.
-Our claim is that the generative model $f(x|\theta)$ can be used to calibrate $g(s|\theta)$ and that
+Our claim is that the generative model $f(x|\theta)$ can be used to calibrate the density $g(s|\theta)$ and that
 \begin{equation}
@@
 \end{equation}
-leads to an equivalent test. In particular, we need to show the density
+leads to an equivalent test.
+
+For notational simplicity, let $f_0(x) = f(x|\theta_0)$, $f_1(x) = f(x|\theta_1)$, and $s(x)=s(x; \theta_1, \theta_0)$.
+The distribution of $x$ totally determines the distribution of $s$. 
+In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
+Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to $\Omega_c$ at the point $x$.
+
+We need to show the density
 \begin{equation}
@@
 \end{equation}
-It is sufficient to show that $q(x) = q(s(x))$ $ \forall x\in\Omega_c$.
-
-For notational simplicity, let $f_0(x) = f(x|\theta_0)$, $f_1(x) = f(x|\theta_1)$, and $s(x)=s(x; \theta_1, \theta_0)$.
-The distribution of $x$ totally determines the distribution of $s$. 
-In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
-Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to $\Omega_c$ at the point $x$. The induced density $g_1(c)$ is given by 
+It is sufficient to show that $q_x(x) = q_s(s(x))$ $ \forall x\in\Omega_c$. The function $q_s(s)$ is based on the induced densities $g_0(s)$ and $g_1(s)$.  The induced density $g_1(c)$ is given by 
 \begin{equation}
@@
 \textbf{\flushleft Theorem 1:}
-We have the following equalities
+We have the following equality
 \begin{equation}
-\frac{g_1(c)}{g_0(c)} = s(x) = \frac{f_1(x)}{f_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
+\frac{g_1(c)}{g_0(c)} =  \frac{f_1(x)}{f_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
 \end{equation}
 \textbf{Proof}
-We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega_c$.
+For $x\in \Omega_c$, we can factor out of the integral the constant $f_1(x)/f_0(x)$.
 Thus
 \begin{equation}
-g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x) / | \hat{n} \cdot \nabla s  |= s(x) \int d\Omega_c f_0(x)  / | \hat{n} \cdot \nabla s  | \;,
+g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x) / | \hat{n} \cdot \nabla s  |= \frac{f_1(x)}{f_0(x)} \int d\Omega_c f_0(x)  / | \hat{n} \cdot \nabla s  | \;,
 \end{equation}
@@
 \begin{equation}
-\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega_c f_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_c f_0(x) / | \hat{n} \cdot \nabla s  |} = s(x) = \frac{f_1(x)}{f_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
+\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)} \frac{\int d\Omega_c f_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_c f_0(x) / | \hat{n} \cdot \nabla s  |} = \frac{f_1(x)}{f_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
 \end{equation}
 
-\bigskip
-In the case of simple hypothesis testing, $\theta_0$ and $\theta_1$ are specified and there is a unique map $s(x) =  s(x_e; \theta_0, \theta_1)$.
+One can think of the ratio $g_1(s)/g_0(s)$ as a way of calibrating the the discriminative classifier and correcting for the monotonic transformation $m$ of the desired likelihood ratio as in Eq.~\ref{eq:montonic}.
+
+%\bigskip
+%In the case of simple hypothesis testing, $\theta_0$ and $\theta_1$ are specified and there is a unique map $s(x) =  s(x_e; \theta_0, \theta_1)$.
 In that case, the equivalent likelihood ratio test can be performed by first transforming the data to $D_s = \{s_1, \dots, s_e\}$, constructing the likelihoods
-\begin{equation}\label{eq:NP}
-g( D_s \,|\,  \theta) = \prod_{e=1}^n \,  g( s_e \, |\,  \theta)   \; 
-\end{equation}
-for $\theta=\{\theta_0,\theta_1\}$, and constructing the likelihood ratio based on $g(D_s|\theta_0)/g(D_s|\theta_1)$.
+%\begin{equation}\label{eq:NP}
+%g( D_s \,|\,  \theta) = \prod_{e=1}^n \,  g( s_e \, |\,  \theta)   \; 
+%\end{equation}
+%for $\theta=\{\theta_0,\theta_1\}$, and constructing the likelihood ratio based on $g(D_s|\theta_0)/g(D_s|\theta_1)$.
 
@@
 Fortunately, there is a construction based on $g(s|\theta)$ that works.
-The maximum likelihood estimate is the same as the value that maximizes the ratio with respect to $f(D|\theta_1)$ for some fixed value of $\theta_1$.
-This allows us to use Theorem~1 to find
+The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as the value that maximizes the likelihood ratio with respect to $f(D|\theta_1)$ for some fixed value of $\theta_1$.
+This allows us to use Theorem~1 to reformulate the maximum likelihood estimate
 \begin{equation}
@@
 \end{equation}
-It is important that we include the denominator $g(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that change as we vary $\theta$.
+It is important that we include the denominator $g(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.
 
@@
 In general we will be interested in the machine learning problem that approximates these distributions based on samples $\{x_i\}$ drawn from the generative model $f(x|\theta)$.
-In particular,
-is there a loss function such that the function $\hat{s}(x)$ that minimizes the expected loss leads to a function
-that is one-to-one with $f(x|\theta_0)/f(x|\theta_1)$?
+The first step in this direction is to confirm that a discriminative classifier obtained from common training procedures will yield a function that is one-to-one with $f(x|\theta_0)/f(x|\theta_1)$.
+%In particular, 
+%is there a loss function such that the function $\hat{s}(x)$ that minimizes the expected loss leads to a function
+%that is one-to-one with $f(x|\theta_0)/f(x|\theta_1)$?
 
@@
 data $(x_i,  y_i=0)$ and $f(x|\theta_1)$ to generate training data $(x_i , y_i=1)$.
-If we use the squared-loss function, then the expected loss is
-\begin{equation}
-\mathbb{E}[L] = \int dx f(x|\theta_0)  (  \hat{s}(x) )^2  + \int dx f(x|\theta_1)  ( 1 - \hat{s}(x) )^2 \; .
-\end{equation}
-The function $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ minimizes this expected loss.
+With balanced training data  -- $\mbox{p(y=1)=p(y=0)=\sfrac{1}{2}}$ -- a quadratic loss function will lead to classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$, which is  monotonic with the desired likelihood ratio $f(x|\theta_0)/f(x|\theta_1)$.
+Thus, standard classification approaches will lead to discriminative classifiers needed to produce an equivalent likelihood ratio test.
+Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g.
+histograms or kernel density estimation) to approximate $\hat{g}(\hat{s}|\theta)$.
 
-Proof (Sketch): Consider a variation about $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ given by $\hat{s}'(x) = \hat{s}(x) +h(x)$. The change in the expected loss is given by 
-\begin{equation}
-\Delta = \mathbb{E}[L_{\hat{s}'}]-\mathbb{E}[L_{\hat{s}}] = \int dx f(x|\theta_0)  (h^2(x) + 2 h(x)  \hat{s}(x) )  + f(x|\theta_1)   (h^2(x) - 2 h(x) + 2 h(x) \hat{s}(x)  \; .
-\end{equation}
-using $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ we obtain
-\begin{equation}
-\Delta =  \int dx (f(x|\theta_0)+f(x|\theta_1))  h^2(x) >0    \; .
-\end{equation}
-Thus any variation on $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ has a larger expected loss.
+%
+%If we use the squared-loss function, then the expected loss is
+%\begin{equation}
+%\mathbb{E}[L] = \int dx f(x|\theta_0)  (  \hat{s}(x) )^2  + \int dx f(x|\theta_1)  ( 1 - \hat{s}(x) )^2 \; .
+%\end{equation}
+%The function $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ minimizes this expected loss.
+%
+%Proof (Sketch): Consider a variation about $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ given by $\hat{s}'(x) = \hat{s}(x) +h(x)$. The change in the expected loss is given by 
+%\begin{equation}
+%\Delta = \mathbb{E}[L_{\hat{s}'}]-\mathbb{E}[L_{\hat{s}}] = \int dx f(x|\theta_0)  (h^2(x) + 2 h(x)  \hat{s}(x) )  + f(x|\theta_1)   (h^2(x) - 2 h(x) + 2 h(x) \hat{s}(x)  \; .
+%\end{equation}
+%using $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ we obtain
+%\begin{equation}
+%\Delta =  \int dx (f(x|\theta_0)+f(x|\theta_1))  h^2(x) >0    \; .
+%\end{equation}
+%Thus any variation on $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ has a larger expected loss.
 
-The conclusion is that standard classification with a quadratic loss function and training data as described above will approximate a discriminative classifier needed to produce an equivalent likelihood ratio test.
+%The conclusion is that standard classification with a quadratic loss function and training data as described above will approximate a discriminative classifier needed to produce an equivalent likelihood ratio test.
 
-Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g. histograms or kernel density estimation) to approximate $\hat{g}(\hat{s}|\theta)$.
+%Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g.
+histograms or kernel density estimation) to approximate $\hat{g}(\hat{s}|\theta)$.
 
-This treatment shows that in the asymptotic limit of large samples from the generative model, that we can approximate arbitrarily well the original likelihood ratio test.
+Thus, in the  limit of large samples from the generative model,  we can approximate arbitrarily well the original likelihood ratio test.
 With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{g}(\hat{s}|\theta)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test.
@@
 We denote the learned family of classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based roughly on the following algorithmic flow.
-\begin{algorithm}
-\caption{Training the parametrized classifier}\label{euclid}
+\begin{algorithm}[ht]
+\caption{Training of the parametrized classifier.}\label{alg:training}
 \begin{algorithmic}
@@
 While the function $f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ will minimize the expected squared loss based on 
-training data produced according to the algorithm above, it is not clear how training data from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real world classifier with finite capacity.
+training data produced according to Algorithm~\ref{alg:training}, it is not clear how training data from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real world classifier with finite capacity.
 This is left as an area for future work.
@@
 
-In most settings that make use of likelihood ratio tests, the probability densities directly model the observed data via $f(x|\theta)$.
-In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$.
-In the parametrized setting this it is not possible to pre-compute $\hat s(x; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$, so  we must embed the classifier into the likelihood function to carry out the composition $g\circ s$.
-A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and discriminative classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}.
+In most settings that make use of likelihood ratio tests, the likelihood is based directly on some approximation of density for the observed data via $\hat{f}(x|\theta)$.
+Approximating the density $\hat{f}(x|\theta)$ is difficult for high-dimensional data, which motivates the use of the dimensionality reduction map $\hat{s}(x)$ and likelihood ratio tests based on the density $\hat{g}(\hat{s}|\theta)$.
+In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$.
+In the parametrized setting this it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$, so  we must embed the classifier into the likelihood function to carry out the composition $\hat{g}\circ \hat{s}$.
+A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and discriminative classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
-In both cases, constructing the density $\hat g(\hat s|\theta)$ requires running the generative model at $\theta$, and in the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize $\theta$ needs access to the generative model $f(x|\theta)$.
+In both cases, constructing the density $\hat g(\hat s|\theta)$ requires running the generative model at $\theta$.
+In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $f(x|\theta)$.
 This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
-In practice, one may want to interpolate the distribution between discrete values of $\theta$ to produce a continuous parametrization for $\hat g(s | \theta)$.
+In practice, one may want to interpolate the distribution between discrete values of $\theta$ to produce a continuous parametrization for $\hat g(\hat s | \theta)$.
 In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
@@
 
-In high-energy physics (HEP) we are often searching for the properties of some 
+In high-energy physics (HEP) we are often searching for some 
 class of events, generically referred to as \textit{signal}, in the presence of a separate class 
-of \textit{background} events. For each event we measure some quantities $x$ that have corresponding distributions 
+of \textit{background} events.
+Generalized likelihood ratio tests are used widely in HEP ~\citep{Cowan:2010js}, most notably in the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
+For each event we measure some quantities $x$ that have corresponding distributions
 $f_b(x|\nu)$ for background and $f_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing 
@@
 
-Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to background and $y=1$ corresponds to signal.
-Following the result of Section~\ref{S:classifier}, the resulting classifier approximates the likelihood ratio of the mixture components $f_s(x|\nu_0)/(f_s(x|\nu_0)+f_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $f(x|\mu=0,\nu_0)/f(x|\mu,\nu_0)$ for all $\mu$.
+Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $f_b(x|\nu_0)$  and $y=1$ corresponds to signal density $f_s(x|\nu_0)$ (not the signal-plus-background).
+The resulting classifier approximates the regression function $f_s(x|\nu_0)/(f_s(x|\nu_0)+f_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $f(x|\mu=0,\nu_0)/f(x|\mu,\nu_0)$ for all $\mu$.
 The resulting classifier is denoted $\hat s(x)$.
@@
 \end{figure}
-\bigskip
+%\bigskip
 
-These steps lead to a subsequent statistical analysis where one observes in data $D=(x_1, \dots, x_n)$.
+These steps lead to a subsequent statistical analysis where one observes in data \\
+${D=(x_1, \dots, x_n)}$.
 For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution.
@@
 The third line uses the previous results to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
-In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $g_{c}(s_{c,c',\theta_0, \theta_1}| \theta_1)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).
+In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $g_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).
 
@@
 
-Refs to include:
-\cite{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,TommiJaakkola}
+%Refs to include:
+%\cite{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,}
 
-\cite{JMLR:v14:tong13a}
+In \cite{ClaytonScott,JMLR:v14:tong13a}, the authors consider the machine learning problem associated to Neyman-Pearson hypothesis testing.
+As in this work, they consider the situation where one does not have access to the underlying distributions, but only has i.i.d.
+samples from each hypothesis.
+This work generalizes that goal from the Neyman-Pearson setting to generalized likelihood ratio tests and emphasizes the connection with classification.
+Perhaps a  formal treatment similar to the Neyman-Pearson case can be brought to bear in this more general setting.
+In \citep{TommiJaakkola}, the authors explore a way of leveraging generative models to derive kernel functions for use in discriminative methods.
+This interesting work is distinct from the point made here in which the generative model is being used for the purpose of providing training data and calibration.
+In \citep{BiancaZadrozny}, the authors emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this.
+In contrast to that work, we are not interested in calibrated probability estimates for $p(y|x)$ for individual events, but instead we use the calibration to correct for non-linear transformations of the target likelihood ratio and, perhaps, to provide calibrated p-values based on those likelihood ratio tests.
+
+Mention also \citep{McCallum}
 
@@
 KC would also like to thank Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
-project early in its conception, Bal\'azs K\'egl for discussions about the Kaggle challenge and 
-non-standard loss functions, and Yuri Shirman for reassuring discussions on elementary vector calculus.
+project early in its conception, Bal\'azs K\'egl for discussions about the Kaggle challenge and feedback to the draft, 
+and Yuri Shirman for reassuring cross checks of the Theorem.
 KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 

### 2015-03-26 tex/parametrized-learning.tex
reworked other work section

@@
 We demonstrate how discriminative classifiers can be used to approximate 
-(generalized) likelihood ratio tests over high-dimensional data when a generative model 
+generalized likelihood ratio tests over high-dimensional data when a generative model 
 for the data is available for training and calibration.  
@@
 discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
-a generative model for the data is available for calibration.  We use the following notation
+a generative model for the data is available for training and calibration.  We use the following notation
 
@@
 Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
+Similar points are made in~\citep{ClaytonScott}.
 
@@
 The second line is a trivial, but useful decomposition into pair-wise classification between $f_{c'}(x|\theta_1)$ and $f_c(x|\theta_0)$.
-The third line uses the previous results to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
+The third line uses Theorem 1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
 In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $g_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).
@@
 
-In \cite{ClaytonScott,JMLR:v14:tong13a}, the authors consider the machine learning problem associated to Neyman-Pearson hypothesis testing.
+\cite{ClaytonScott} and \cite{JMLR:v14:tong13a} consider the machine learning problem associated to Neyman-Pearson hypothesis testing.
 As in this work, they consider the situation where one does not have access to the underlying distributions, but only has i.i.d.
@@
 Perhaps a  formal treatment similar to the Neyman-Pearson case can be brought to bear in this more general setting.
-In \citep{TommiJaakkola}, the authors explore a way of leveraging generative models to derive kernel functions for use in discriminative methods.
+\cite{TommiJaakkola} explore a way of leveraging generative models to derive kernel functions for use in discriminative methods.
 This interesting work is distinct from the point made here in which the generative model is being used for the purpose of providing training data and calibration.
-In \citep{BiancaZadrozny}, the authors emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this.
+\cite{McCallum}
+ consider a hybrid generative/discriminative classifier; however, the goal of that work is not to leverage a generative model for the data, but to use both approaches to learn different subsets of the parameters in a single hybrid classifier.
+\cite{BiancaZadrozny} emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this.
 In contrast to that work, we are not interested in calibrated probability estimates for $p(y|x)$ for individual events, but instead we use the calibration to correct for non-linear transformations of the target likelihood ratio and, perhaps, to provide calibrated p-values based on those likelihood ratio tests.
-
-Mention also \citep{McCallum}
+\cite{Ihler2004} take on a different problem (tests of statistical independence) by using machine learning algorithms to find  scalar maps from the high-dimensional feature space that achieve the desired statistical goal when the fundamental high-dimensional test is intractable.
 

### 2015-04-09 tex/parametrized-learning.tex
notation change, about to restructure text

@@
  \item $\theta$: parameters of a statistical model
-\item $f(x| \theta)$:  probability density  (statistical model) for $x$ given $\theta$
+\item $p(x| \theta)$:  probability density  (statistical model) for $x$ given $\theta$
 %\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
 \item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
-%\item $f( s | \theta )$ The probability density function for $s$ implied by $f(x|\theta)$ and $s(x)$
-\item $g( s | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $f(x|\theta)$ \end{itemize}
-We will assume the $x_e$ are i.i.d., so that $f(D|\theta) = \prod_{e=1}^n f(x_e | \theta)$.
+%\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
+\item $p( s | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$ \end{itemize}
+We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.
 
@@
 \begin{equation}
-T(D) = \prod_{e=1}^n \frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)}
+T(D) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
 \end{equation}
 is the most powerful test statistic. In order to evaluate $T(D)$, one must be able to evaluate the probability density 
-$f(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that 
-can act as generative model  for $f(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case 
+$p(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that 
+can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case 
 high energy physics where the simulation of particle detectors can only be done in the `forward mode'.
@@
 %\begin{equation}
-%T'(D) = \prod_{e=1}^n \frac{ f(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ f(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
+%T'(D) = \prod_{e=1}^n \frac{ p(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ p(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
 %\end{equation}
 \begin{equation}\label{eq:equivLRtest}
-T'(D) = \prod_{e=1}^n \frac{ g(s_e | \theta_0)}{ g(s_e | \theta_1)}
+T'(D) = \prod_{e=1}^n \frac{ p(s_e | \theta_0)}{ p(s_e | \theta_1)}
 \end{equation}
@@
 \begin{equation}\label{eq:montonic}
-s_e = s(x_e; \theta_0, \theta_1) = m\left(\, f(x_e|\theta_0) / f(x_e|\theta_1) \,\right) \; 
-%s_e = s(x_e; \theta_0, \theta_1) = m\left(\frac{ f(x_e|\theta_0)}{ f(x_e|\theta_1)} \right) \; 
+s_e = s(x_e; \theta_0, \theta_1) = m\left(\, p(x_e|\theta_0) / p(x_e|\theta_1) \,\right) \; 
+%s_e = s(x_e; \theta_0, \theta_1) = m\left(\frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)} \right) \; 
 \end{equation}
@@
 used to learn $s(x; \theta_0, \theta_1)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
-from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$.
+from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$.
 In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the discriminative classifier explicitly parametrized in terms of $(\theta_0, \theta_1)$.
@@
 
-The first possible source of confusion we wish to avoid is that  $f(x|\theta)$  is a generative \textit{statistical model} for the features $x$, not a generative classifier.
-We think of the  $f(x|\theta)$ along the lines of a traditional scientific theory, able to make predictions about $x$ and being motivated by domain-specific considerations.
-For example, in the context of high energy particle physics $f(x|\theta)$ is based on quantum field theory and a detailed simulation of the particle detector and data processing algorithms that transform raw sensor data into the feature vector $x$.
-Moreover, we are not attempting to learn the generative model $f(x|\theta)$, we are taking it as given and trying to learn the corresponding likelihood ratio test.
-%The point here is to make the distinction of $f(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.
+The first possible source of confusion we wish to avoid is that  $p(x|\theta)$  is a generative \textit{statistical model} for the features $x$, not a generative classifier.
+We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory, able to make predictions about $x$ and being motivated by domain-specific considerations.
+For example, in the context of high energy particle physics $p(x|\theta)$ is based on quantum field theory and a detailed simulation of the particle detector and data processing algorithms that transform raw sensor data into the feature vector $x$.
+Moreover, we are not attempting to learn the generative model $p(x|\theta)$, we are taking it as given and trying to learn the corresponding likelihood ratio test.
+%The point here is to make the distinction of $p(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.
 
@@
 While we can use the generative models to produce training data $(x,y=0)$ generated 
-from $f(x|\theta_0)$ and $(x,y=1)$ generated from $f(x|\theta_1)$, the relative mix $p(y)$ 
+from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$, the relative mix $p(y)$ 
 is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then 
 \begin{equation}
-p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{f(x|\theta_1)}{f(x|\theta_0)+f(x|\theta_1))} \;,
+p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{p(x|\theta_1)}{p(x|\theta_0)+p(x|\theta_1))} \;,
 \end{equation}
-which is monotonic with the desired likelihood ratio $f(x|\theta_1)/f(x|\theta_0)$.
+which is monotonic with the desired likelihood ratio $p(x|\theta_1)/p(x|\theta_0)$.
 Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
@@
 \begin{equation}
-\ln T =   \sum_{e=1}^n \underbrace{\log \left[ \frac {f(x_e | \theta_0) }{ f(x_e | \theta_1) } \right]}_{q(x_e)} \;.
+\ln T =   \sum_{e=1}^n \underbrace{\log \left[ \frac {p(x_e | \theta_0) }{ p(x_e | \theta_1) } \right]}_{q(x_e)} \;.
 \end{equation}
@@
 The important part of the per-event function $q(x)$ is that it defines iso-contours in the feature space $x$.
-As we will show, our goal is to learn a monotonic function of $f(x|\theta_0)/f(x|\theta_1)$, which will share the same iso-contours.
+As we will show, our goal is to learn a monotonic function of $p(x|\theta_0)/p(x|\theta_1)$, which will share the same iso-contours.
 Then the remaining challenge is to find the appropriate rescaling that gives back  linear function of $q(x)$.
-Our claim is that the generative model $f(x|\theta)$ can be used to calibrate the density $g(s|\theta)$ and that
+Our claim is that the generative model $p(x|\theta)$ can be used to calibrate the density $p(s|\theta)$ and that
 \begin{equation}
-\ln T' = \sum_{e=1}^n \underbrace{\log \left[ \frac {g(s_e | \theta_0) }{ g(s_e | \theta_1) } \right]}_{q(s_e)} \;,
+\ln T' = \sum_{e=1}^n \underbrace{\log \left[ \frac {p(s_e | \theta_0) }{ p(s_e | \theta_1) } \right]}_{q(s_e)} \;,
 \end{equation}
@@
 
-For notational simplicity, let $f_0(x) = f(x|\theta_0)$, $f_1(x) = f(x|\theta_1)$, and $s(x)=s(x; \theta_1, \theta_0)$.
+For notational simplicity, let $p_0(x) = p(x|\theta_0)$, $p_1(x) = p(x|\theta_1)$, and $s(x)=s(x; \theta_1, \theta_0)$.
 The distribution of $x$ totally determines the distribution of $s$. 
@@
 \begin{equation}
-f(q_x|\theta) = \int dx \delta(q_x-q_x(x)) f(x|\theta)  / | \hat{n} \cdot \nabla q_x  | 
+p(q_x|\theta) = \int dx \delta(q_x-q_x(x)) p(x|\theta)  / | \hat{n} \cdot \nabla q_x  | 
 \end{equation}
@@
 \begin{equation}
-f(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, f(x|\theta) \, / | \hat{n} \cdot \nabla q_s  | \; .
+p(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, p(x|\theta) \, / | \hat{n} \cdot \nabla q_s  | \; .
 \end{equation}
-It is sufficient to show that $q_x(x) = q_s(s(x))$ $ \forall x\in\Omega_c$. The function $q_s(s)$ is based on the induced densities $g_0(s)$ and $g_1(s)$.  The induced density $g_1(c)$ is given by 
+It is sufficient to show that $q_x(x) = q_s(s(x))$ $ \forall x\in\Omega_c$. The function $q_s(s)$ is based on the induced densities $p_0(s)$ and $p_1(s)$.  The induced density $p_1(c)$ is given by 
 \begin{equation}
-g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x)  / | \hat{n} \cdot \nabla s  |
+p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega_c p_1(x)  / | \hat{n} \cdot \nabla s  |
 \end{equation}
-and a similar equation for $g_0(c)$. 
+and a similar equation for $p_0(c)$. 
 %\textbf{Do we need Jacobian for x $\to$ s independent of delta function part, I think that's double counting?}
@@
 \begin{equation}
-\frac{g_1(c)}{g_0(c)} =  \frac{f_1(x)}{f_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
+\frac{p_1(c)}{p_0(c)} =  \frac{p_1(x)}{p_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
 \end{equation}
 \textbf{Proof}
-For $x\in \Omega_c$, we can factor out of the integral the constant $f_1(x)/f_0(x)$.
+For $x\in \Omega_c$, we can factor out of the integral the constant $p_1(x)/p_0(x)$.
 Thus
 \begin{equation}
-g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x) / | \hat{n} \cdot \nabla s  |= \frac{f_1(x)}{f_0(x)} \int d\Omega_c f_0(x)  / | \hat{n} \cdot \nabla s  | \;,
+p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega_c p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_c p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
 \end{equation}
@@
 \begin{equation}
-\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)} \frac{\int d\Omega_c f_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_c f_0(x) / | \hat{n} \cdot \nabla s  |} = \frac{f_1(x)}{f_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
+\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)} \frac{\int d\Omega_c p_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_c p_0(x) / | \hat{n} \cdot \nabla s  |} = \frac{p_1(x)}{p_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
 \end{equation}
 
-One can think of the ratio $g_1(s)/g_0(s)$ as a way of calibrating the the discriminative classifier and correcting for the monotonic transformation $m$ of the desired likelihood ratio as in Eq.~\ref{eq:montonic}.
+One can think of the ratio $p_1(s)/p_0(s)$ as a way of calibrating the the discriminative classifier and correcting for the monotonic transformation $m$ of the desired likelihood ratio as in Eq.~\ref{eq:montonic}.
 
@@
 %\begin{equation}\label{eq:NP}
-%g( D_s \,|\,  \theta) = \prod_{e=1}^n \,  g( s_e \, |\,  \theta)   \; 
+%p( D_s \,|\,  \theta) = \prod_{e=1}^n \,  p( s_e \, |\,  \theta)   \; 
 %\end{equation}
-%for $\theta=\{\theta_0,\theta_1\}$, and constructing the likelihood ratio based on $g(D_s|\theta_0)/g(D_s|\theta_1)$.
+%for $\theta=\{\theta_0,\theta_1\}$, and constructing the likelihood ratio based on $p(D_s|\theta_0)/p(D_s|\theta_1)$.
 
-\section{Composite hypotheses and the generalized likelihood ratio}\label{S:GLR}
+\section{Composite hypotheses and the generalized likelihood ratio}\label{?}l{S:GLR}
 
@@
 \begin{equation}
-\lambda(x) =  \frac{ \sup_{\theta \in \Theta_0} f(D | \theta)}{ \sup_{\theta \in \Theta} f(D | \theta)} \; .
+\lambda(x) =  \frac{ \sup_{\theta \in \Theta_0} p(D | \theta)}{ \sup_{\theta \in \Theta} p(D | \theta)} \; .
 \end{equation}
@@
 \begin{equation}\label{eq:mle}
-\hat{\theta} = \argmax_\theta  f(D | \theta)
+\hat{\theta} = \argmax_\theta  p(D | \theta)
 \end{equation}
@@
 \begin{equation}
-\hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  f(D | \theta) \; .
+\hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p(D | \theta) \; .
 \end{equation}
 
-It is not obvious that if we are working with the distributions $g(s|\theta)$ (for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the same estimators. 
-Fortunately, there is a construction based on $g(s|\theta)$ that works.
-The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as the value that maximizes the likelihood ratio with respect to $f(D|\theta_1)$ for some fixed value of $\theta_1$.
+It is not obvious that if we are working with the distributions $p(s|\theta)$ (for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the same estimators. 
+Fortunately, there is a construction based on $p(s|\theta)$ that works.
+The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as the value that maximizes the likelihood ratio with respect to $p(D|\theta_1)$ for some fixed value of $\theta_1$.
 This allows us to use Theorem~1 to reformulate the maximum likelihood estimate
 \begin{equation}
-\hat{\theta} = \argmax_\theta \frac{ f(D | \theta)}{ f(D | \theta_1)} = \argmax_\theta  \sum \ln \frac{f(x_e | \theta)}{f(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{g(s(x_e; \theta, \theta_1) | \theta)}{g(s(x_e; \theta, \theta_1) |\theta_1)} \; .
+\hat{\theta} = \argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} = \argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
 \end{equation}
-It is important that we include the denominator $g(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.
+It is important that we include the denominator $p(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.
 
@@
 
-Thus far we have shown that likelihood ratio tests based on $f(x|\theta_0)/f(x|\theta_1)$ with high dimensional features $x$ can be reproduced via hypothesis tests based on the univariate densities $g(s|\theta)$ for the very special dimensionality reduction map $s(x|\theta_0, \theta_1)$.
-The motivation for this is that often it is not possible to evaluate the density $f(x|\theta)$ at a given point $x$.
-This approach is not useful if it is not possible to approximate $s(x|\theta_0, \theta_1)$ and $g(s|\theta)$ without evaluating the density $f(x|\theta)$.
-In order for this approach to be useful, we need to be able to approximate both based on samples $\{(x,\theta)\}$ drawn from the generative model $f(x|\theta)$.
+Thus far we have shown that likelihood ratio tests based on $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via hypothesis tests based on the univariate densities $p(s|\theta)$ for the very special dimensionality reduction map $s(x|\theta_0, \theta_1)$.
+The motivation for this is that often it is not possible to evaluate the density $p(x|\theta)$ at a given point $x$.
+This approach is not useful if it is not possible to approximate $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ without evaluating the density $p(x|\theta)$.
+In order for this approach to be useful, we need to be able to approximate both based on samples $\{(x,\theta)\}$ drawn from the generative model $p(x|\theta)$.
 
 Denote the approximate dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{g}(\hat{s}|\theta)$.
-In general we will be interested in the machine learning problem that approximates these distributions based on samples $\{x_i\}$ drawn from the generative model $f(x|\theta)$.
-The first step in this direction is to confirm that a discriminative classifier obtained from common training procedures will yield a function that is one-to-one with $f(x|\theta_0)/f(x|\theta_1)$.
+In general we will be interested in the machine learning problem that approximates these distributions based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
+The first step in this direction is to confirm that a discriminative classifier obtained from common training procedures will yield a function that is one-to-one with $p(x|\theta_0)/p(x|\theta_1)$.
 %In particular, 
 %is there a loss function such that the function $\hat{s}(x)$ that minimizes the expected loss leads to a function
-%that is one-to-one with $f(x|\theta_0)/f(x|\theta_1)$?
+%that is one-to-one with $p(x|\theta_0)/p(x|\theta_1)$?
 
 \subsection{The standard discriminative classification setting} For fixed $\theta_0$ and $\theta_1$ we can generate 
-large samples from each model and train a classifier. To be concrete, let's use $f(x|\theta_0)$ to generate training 
-data $(x_i,  y_i=0)$ and $f(x|\theta_1)$ to generate training data $(x_i , y_i=1)$.
-With balanced training data  -- $\mbox{p(y=1)=p(y=0)=\sfrac{1}{2}}$ -- a quadratic loss function will lead to classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$, which is  monotonic with the desired likelihood ratio $f(x|\theta_0)/f(x|\theta_1)$.
+large samples from each model and train a classifier. To be concrete, let's use $p(x|\theta_0)$ to generate training 
+data $(x_i,  y_i=0)$ and $p(x|\theta_1)$ to generate training data $(x_i , y_i=1)$.
+With balanced training data  -- $\mbox{p(y=1)=p(y=0)=\sfrac{1}{2}}$ -- a quadratic loss function will lead to classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the desired likelihood ratio $p(x|\theta_0)/p(x|\theta_1)$.
 Thus, standard classification approaches will lead to discriminative classifiers needed to produce an equivalent likelihood ratio test.
@@
 %\begin{equation}
-%\mathbb{E}[L] = \int dx f(x|\theta_0)  (  \hat{s}(x) )^2  + \int dx f(x|\theta_1)  ( 1 - \hat{s}(x) )^2 \; .
+%\mathbb{E}[L] = \int dx p(x|\theta_0)  (  \hat{s}(x) )^2  + \int dx p(x|\theta_1)  ( 1 - \hat{s}(x) )^2 \; .
 %\end{equation}
-%The function $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ minimizes this expected loss.
+%The function $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ minimizes this expected loss.
 %
-%Proof (Sketch): Consider a variation about $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ given by $\hat{s}'(x) = \hat{s}(x) +h(x)$. The change in the expected loss is given by 
+%Proof (Sketch): Consider a variation about $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ given by $\hat{s}'(x) = \hat{s}(x) +h(x)$. The change in the expected loss is given by 
 %\begin{equation}
-%\Delta = \mathbb{E}[L_{\hat{s}'}]-\mathbb{E}[L_{\hat{s}}] = \int dx f(x|\theta_0)  (h^2(x) + 2 h(x)  \hat{s}(x) )  + f(x|\theta_1)   (h^2(x) - 2 h(x) + 2 h(x) \hat{s}(x)  \; .
+%\Delta = \mathbb{E}[L_{\hat{s}'}]-\mathbb{E}[L_{\hat{s}}] = \int dx p(x|\theta_0)  (h^2(x) + 2 h(x)  \hat{s}(x) )  + p(x|\theta_1)   (h^2(x) - 2 h(x) + 2 h(x) \hat{s}(x)  \; .
 %\end{equation}
-%using $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ we obtain
+%using $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ we obtain
 %\begin{equation}
-%\Delta =  \int dx (f(x|\theta_0)+f(x|\theta_1))  h^2(x) >0    \; .
+%\Delta =  \int dx (p(x|\theta_0)+p(x|\theta_1))  h^2(x) >0    \; .
 %\end{equation}
-%Thus any variation on $\hat{s}(x) = f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ has a larger expected loss.
+%Thus any variation on $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ has a larger expected loss.
 
@@
 	\FOR{ $\theta_1$ in $\Theta$ }
-		\STATE generate $x_i^0 \sim f(x|\theta_0)$
+		\STATE generate $x_i^0 \sim p(x|\theta_0)$
 		\STATE append $\{ (x_i^0, \theta_0, \theta_1, y=0) \}$ to trainingData
-		\STATE generate $x_i^1 \sim f(x|\theta_1)$
+		\STATE generate $x_i^1 \sim p(x|\theta_1)$
 		\STATE append $\{ (x_i^1, \theta_0, \theta_1, y=1) \}$ to trainingData
@@
 
-While the function $f(x|\theta_1)/(f(x|\theta_0)+f(x|\theta_1))$ will minimize the expected squared loss based on 
+While the function $p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ will minimize the expected squared loss based on 
 training data produced according to Algorithm~\ref{alg:training}, it is not clear how training data from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real world classifier with finite capacity.
@@
 
-In both cases, constructing the density $\hat g(\hat s|\theta)$ requires running the generative model at $\theta$.
-In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $f(x|\theta)$.
+In both cases, constructing the density $\hat p(\hat s|\theta)$ requires running the generative model at $\theta$.
+In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $p(x|\theta)$.
 This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
-In practice, one may want to interpolate the distribution between discrete values of $\theta$ to produce a continuous parametrization for $\hat g(\hat s | \theta)$.
+In practice, one may want to interpolate the distribution between discrete values of $\theta$ to produce a continuous parametrization for $\hat p(\hat s | \theta)$.
 In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
@@
 For each event we measure some quantities $x$ that have corresponding distributions
-$f_b(x|\nu)$ for background and $f_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing 
+$p_b(x|\nu)$ for background and $p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing 
 uncertainties in the underlying physics prediction or response of the measurement device. In the simple setup, the 
@@
 \begin{equation}\label{eq:hepGen}
-f( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu f_s( x_e \, |\,  \nu)  + (1-\mu)\, f_b( x_e \,|\, \nu) \,\right] \; ,
+p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( x_e \, |\,  \nu)  + (1-\mu)\, p_b( x_e \,|\, \nu) \,\right] \; ,
 \end{equation}
@@
 
-Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $f_b(x|\nu_0)$  and $y=1$ corresponds to signal density $f_s(x|\nu_0)$ (not the signal-plus-background).
-The resulting classifier approximates the regression function $f_s(x|\nu_0)/(f_s(x|\nu_0)+f_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $f(x|\mu=0,\nu_0)/f(x|\mu,\nu_0)$ for all $\mu$.
+Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density $p_s(x|\nu_0)$ (not the signal-plus-background).
+The resulting classifier approximates the regression function $p_s(x|\nu_0)/(p_s(x|\nu_0)+p_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$.
 The resulting classifier is denoted $\hat s(x)$.
-Based on this classifier and large samples of synthetic data drawn from $f_s(x | \nu)$ and $f_b(x | \nu)$ we construct the distributions  $g_s(\hat s | \nu)$ and $g_b(\hat s | \nu)$.
+Based on this classifier and large samples of synthetic data drawn from $p_s(x | \nu)$ and $p_b(x | \nu)$ we construct the distributions  $p_s(\hat s | \nu)$ and $p_b(\hat s | \nu)$.
 An example of the distributions of the distribution of $\hat s$ for the signal and background events with $\nu=\nu_0$ is shown in Figure~\ref{fig:tmva}.
@@
  \includegraphics[height=2in]{example-TMVA-ROC.pdf}
-\caption{Left: an example of the distributions $g_b(\hat s|\nu)$ and $g_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).
+\caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).
 Right: the corresponding ROC curve (right) for this and other classifiers.
@@
 \begin{equation}\label{eq:typicalML}
-g( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu g_s( \hat s(x_e) \, |\,  \nu)  + (1-\mu)\, g_b( \hat s(x_e) \,|\, \nu) \,\right] \; ,
+p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( \hat s(x_e) \, |\,  \nu)  + (1-\mu)\, p_b( \hat s(x_e) \,|\, \nu) \,\right] \; ,
 \end{equation}
@@
 Nuisance parameters are an after thought in the typical usage of machine learning in HEP.
-In fact, most machine learning discussions would only consider $f_b(x)$ and $f_s(x)$.
+In fact, most machine learning discussions would only consider $p_b(x)$ and $p_s(x)$.
 However, as experimentalists we know that we must account for various forms of systematic uncertainty, parametrized by nuisance parameters $\nu$.
 In practice, we take the classifier as fixed and then propagate uncertainty through the classifier as in Eq.~\ref{eq:typicalML}.
-Building the distribution $g(\hat s|\nu)$ for values of $\nu$ other than the nominal $\nu_0$ used to train the classifier can be thought of as a calibration necessary for classical statistical inference; however, this classifier is clearly not optimal for $\nu \ne \nu_0$.
+Building the distribution $p(\hat s|\nu)$ for values of $\nu$ other than the nominal $\nu_0$ used to train the classifier can be thought of as a calibration necessary for classical statistical inference; however, this classifier is clearly not optimal for $\nu \ne \nu_0$.
 
@@
 \begin{equation}
-\lambda(\mu) = \frac{f(D|\mu, \hat{\hat{\nu}})}{f(D|\hat \mu, {\hat{\nu}})} \;,
+\lambda(\mu) = \frac{p(D|\mu, \hat{\hat{\nu}})}{p(D|\hat \mu, {\hat{\nu}})} \;,
 \end{equation}
@@
 \begin{equation}
- \frac{f(x| 0, \nu_0)}{f(x|\hat \mu, \nu_0)} =  \frac{f_b(x| \nu_0)}{ \mu f_s( x_e \, |\,  \nu_0)  + (1-\mu)\, f_b( x_e \,|\, \nu_0)} = \left[ c_1 + c_2   \frac{f_s(x| \nu_0)}{ f_b( x_e \,|\, \nu_0)} \right ]^{-1} \; ,
+ \frac{p(x| 0, \nu_0)}{p(x|\hat \mu, \nu_0)} =  \frac{p_b(x| \nu_0)}{ \mu p_s( x_e \, |\,  \nu_0)  + (1-\mu)\, p_b( x_e \,|\, \nu_0)} = \left[ c_1 + c_2   \frac{p_s(x| \nu_0)}{ p_b( x_e \,|\, \nu_0)} \right ]^{-1} \; ,
 \end{equation}
@@
 \begin{equation}
- \frac{f(x| 0, \hat{\hat{ \nu}})}{f(x|\hat \mu, \hat\nu)} =  \frac{f_b(x| \hat{\hat{ \nu}})}{ \hat \mu f_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, f_b( x_e \,|\, \hat \nu)} \; ,
+ \frac{p(x| 0, \hat{\hat{ \nu}})}{p(x|\hat \mu, \hat\nu)} =  \frac{p_b(x| \hat{\hat{ \nu}})}{ \hat \mu p_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, p_b( x_e \,|\, \hat \nu)} \; ,
 \end{equation}
-the background components don't cancel and there is an additional term $f_b(x| \hat{\hat{ \nu}})/f_b(x| {\hat{ \nu}})$.
+the background components don't cancel and there is an additional term $p_b(x| \hat{\hat{ \nu}})/p_b(x| {\hat{ \nu}})$.
 In practice, with classifiers of finite capacity, there will be some tradeoff between taking into account this additional term and the more challenging learning problem when $\mu$ is very small.
@@
 \begin{equation}
-f(x|\theta)=\sum_c w_c(\theta) f_c(x| \theta) \;.
+p(x|\theta)=\sum_c w_c(\theta) p_c(x| \theta) \;.
 \end{equation}
-In the case of particle physics, the distributions $f(x|\theta)$ is not a Gaussian Mixture Model, but mixture of  complicated distributions associated to relatively few types of particle interactions.
+In the case of particle physics, the distributions $p(x|\theta)$ is not a Gaussian Mixture Model, but mixture of  complicated distributions associated to relatively few types of particle interactions.
 Moreover, when searching for a new particle, the null hypothesis would correspond to some of the coefficients $w_c=0$ while the alternate ``signal-plus-background'' hypothesis would have $0<w_{c\in \rm signal} \ll w_{c\in \rm background}$.
 In some cases $w_{c\in \rm signal} / w_{c\in \rm background} <10^{-6}$, which means the alternate hypothesis is a small perturbation to the null hypothesis.
-This can be a challenge for typical classifiers because they should devote their capacity to the region where $f_{c\in \rm signal}(x) / f_{c\in \rm background}(x)$ is relatively large.
-Lastly, even when the distributions $f_c(x|\theta)$ are well known, it is often the case that the coefficients are uncertain or treated as completely unknown.
+This can be a challenge for typical classifiers because they should devote their capacity to the region where $p_{c\in \rm signal}(x) / p_{c\in \rm background}(x)$ is relatively large.
+Lastly, even when the distributions $p_c(x|\theta)$ are well known, it is often the case that the coefficients are uncertain or treated as completely unknown.
 These all present challenges to machine learning algorithms that aim to learn $s(x; \theta_0, \theta_1)$.
@@
 %\begin{equation}
-%\frac{f(x|\theta_0)}{f(x|\theta_1)}= \frac{\sum_c w_c(\theta_0) f_c(x| \theta_0)}{\sum_c w_c(\theta_1) f_c(x| \theta_1)} 
+%\frac{p(x|\theta_0)}{p(x|\theta_1)}= \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_c w_c(\theta_1) p_c(x| \theta_1)} 
 %\end{equation}
@@
 %\begin{equation}
-%\frac{f(x|\theta_0)}{f(x|\theta_1)}= \sum_c \frac{ w_c(\theta_0) f_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)} 
+%\frac{p(x|\theta_0)}{p(x|\theta_1)}= \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} 
 %\end{equation}
@@
 \begin{eqnarray}
-\frac{f(x|\theta_0)}{f(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) f_c(x| \theta_0)}{\sum_c w_c(\theta_1) f_c(x| \theta_1)} 
+\frac{p(x|\theta_0)}{p(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_c w_c(\theta_1) p_c(x| \theta_1)} 
 \\
-%&=& \sum_c \frac{ w_c(\theta_0) f_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)} \\
-%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) f_{c'}(x| \theta_1)}{ w_c(\theta_0) f_c(x| \theta_0)}  \right]^{-1} \\
-&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ f_{c'}(x| \theta_1)}{  f_c(x| \theta_0)}  \right]^{-1} \\
-&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ g_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ g_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
+%&=& \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \\
+%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)}{ w_c(\theta_0) p_c(x| \theta_0)}  \right]^{-1} \\
+&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \\
+&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
 \end{eqnarray}
-The second line is a trivial, but useful decomposition into pair-wise classification between $f_{c'}(x|\theta_1)$ and $f_c(x|\theta_0)$.
+The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.
 The third line uses Theorem 1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
-In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $g_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).
+In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).
 
@@
 %\begin{equation}\label{eq:parametrizedML}
-%g( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu g_s( s(x_e;\nu_0, \nu_1) \, |\,  \nu)  + (1-\mu)\, g_b( s(x_e;\nu_0, \nu_1) \,|\, \nu) \,\right] \; ,
+%p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( s(x_e;\nu_0, \nu_1) \, |\,  \nu)  + (1-\mu)\, p_b( s(x_e;\nu_0, \nu_1) \,|\, \nu) \,\right] \; ,
 %\end{equation}
@@
 %The construction in Eq.~\ref{eq:parametrizedML} presents a computational challenge as for each value of $\nu$ one has a new mapping $s: x\to\mathbb{R}$.
-The most naive realization of this would be to generate synthetic data $\{x_e\}$ according to $f_c(x|\nu)$, evaluate $s(x_e ; \nu_0, \nu_1)$, and then use a histogram, kernel density, or other non-parametric density estimation procedure to estimate $g_c( s | \nu)$.
-This approach is not realistic in situations where $f_c(x|\nu)$ is realized with expensive computer simulations; thus, some interpolation strategy or emulator over a fixed set of parameter points $\{\nu_i\}$ would typically be used.
+The most naive realization of this would be to generate synthetic data $\{x_e\}$ according to $p_c(x|\nu)$, evaluate $s(x_e ; \nu_0, \nu_1)$, and then use a histogram, kernel density, or other non-parametric density estimation procedure to estimate $p_c( s | \nu)$.
+This approach is not realistic in situations where $p_c(x|\nu)$ is realized with expensive computer simulations; thus, some interpolation strategy or emulator over a fixed set of parameter points $\{\nu_i\}$ would typically be used.
 
@@
 
-We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $f(x|\theta)$ can be used to approximate statistical inference likelihoodased on the  ratio $f(x|\theta_0)/f(x|\theta_1)$ when it is not possible to evaluate the densities $f(x|\theta)$ for an arbitrary $x$.
+We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate statistical inference likelihoodased on the  ratio $p(x|\theta_0)/p(x|\theta_1)$ when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$.
 This approach leverages the power of machine learning in a classical statistical setting.
@@
 In particular, the loss function for these learning algorithms is based on classification performance of individual $(x_e, c_e)$ feature $\to$ target examples.
-However, in overarching statistical procedure that uses the statistical model in Eq.~\ref{eq:typicalML} or \ref{eq:parametrizedML} the quantity we wish to optimize is a property of an entire dataset $\{x_e\}$ or ensembles of datasets that can be generated from the statistical models $f_c(x|\theta)$.
+However, in overarching statistical procedure that uses the statistical model in Eq.~\ref{eq:typicalML} or \ref{eq:parametrizedML} the quantity we wish to optimize is a property of an entire dataset $\{x_e\}$ or ensembles of datasets that can be generated from the statistical models $p_c(x|\theta)$.
 %
@@
 %
-%Instead of evolving from the traditional usage of ML in HEP, let's start from scratch with $f_c(x|\theta)$ and a well defined objective.  
+%Instead of evolving from the traditional usage of ML in HEP, let's start from scratch with $p_c(x|\theta)$ and a well defined objective.  
 %
@@
 %Consider the situation of simple hypothesis testing between the background-only $\mu=0$ and signal+background hypothesis $\mu=\mu_0=\nu_s/(\nu_s+\nu_b)$ for $\theta=\theta_0$.
-Let us assume that not only can we use the $f_c(x|\theta_0)$ as a generative model, but that we can evaluate the probability density at any point $x$.
+Let us assume that not only can we use the $p_c(x|\theta_0)$ as a generative model, but that we can evaluate the probability density at any point $x$.
 The statistical model is given by
 %\begin{equation}\label{eq:NP}
-%f( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( x_e \, |\,  \theta)  + (1-\mu)\, f_b( x_e \,|\, \theta) \,\right] \; .
+%p( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu p_1( x_e \, |\,  \theta)  + (1-\mu)\, p_b( x_e \,|\, \theta) \,\right] \; .
 %\end{equation}
@@
 %\begin{equation}
-%T =  \log \frac{f(D | \mu=\mu_0, \theta_0)}{f(D | \mu=0, \theta_0)}
+%T =  \log \frac{p(D | \mu=\mu_0, \theta_0)}{p(D | \mu=0, \theta_0)}
 %\end{equation}
@@
 %\begin{equation}
-%T =   \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(x_e | \theta_0) }{ f_b(x_e | \theta_0) } \right]}_{q(x_e)} + c_2\;,
+%T =   \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {p_1(x_e | \theta_0) }{ p_b(x_e | \theta_0) } \right]}_{q(x_e)} + c_2\;,
 %\end{equation}
-%where $c_1=\mu/(1-\mu)$ and $c_2=\log(1-\mu)$ are constants for the simple hypothesis test.
+%where $c_1=\mu/(1-\mu)$ and $c_2=\lop(1-\mu)$ are constants for the simple hypothesis test.
 %
@@
 %The important part of the per-event function $q(x)$ is that it defines contours in the feature space $x$.
-These contours are also equivalent to the function $f_1(x|\theta_0)/f_b(x|\theta_0)$, the signal-to-background ratio.
-If we can find a machine learner $s(x)$ that is a monotonic function of $f_1(x|\theta_0)/f_b(x|\theta_0)$ it will share the same contours.
+These contours are also equivalent to the function $p_1(x|\theta_0)/p_b(x|\theta_0)$, the signal-to-background ratio.
+If we can find a machine learner $s(x)$ that is a monotonic function of $p_1(x|\theta_0)/p_b(x|\theta_0)$ it will share the same contours.
 Then the remaining challenge is to find the appropriate rescaling that gives back  linear function $q(x)$.
@@
 %\begin{equation}
-%T' = \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {f_1(s(x_e) | \theta_0) }{ f_b(s(x_e) | \theta_0) } \right]}_{q(s_e)} \;,
+%T' = \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {p_1(s(x_e) | \theta_0) }{ p_b(s(x_e) | \theta_0) } \right]}_{q(s_e)} \;,
 %\end{equation}
@@
 %\begin{equation}
-%f_c(s) = \int dx \, \delta(s-s(x)) \, f_c(x)  \,  |\partial s / \partial x|^{-1}
+%p_c(s) = \int dx \, \delta(s-s(x)) \, p_c(x)  \,  |\partial s / \partial x|^{-1}
 %\end{equation}
@@
 %
-%Note, in order to form the hypothesis test, one still needs to build the build the $f_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$. 
+%Note, in order to form the hypothesis test, one still needs to build the build the $p_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$. 
 %
@@
 %\begin{equation}
-%f_c(q_x) = \int dx \delta(q_x-q_x(x)) f_c(x) / |\partial q_x / \partial x|
+%p_c(q_x) = \int dx \delta(q_x-q_x(x)) p_c(x) / |\partial q_x / \partial x|
 %\end{equation}
@@
 %\begin{equation}
-%f_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, f_c(x) \, / |\partial q_s / \partial x|
+%p_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, p_c(x) \, / |\partial q_s / \partial x|
 %\end{equation}
-%sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $f_1(x)/f_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
+%sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $p_1(x)/p_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
 %\begin{equation}
-%\frac {f_1(c | \theta_0) }{ f_0(c | \theta_0) } = 
-%\frac { \int dx \delta(c-s(x)) \, f_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  } = 
-%\frac {c \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, f_0(x | \theta_0)  } = c =\frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
+%\frac {p_1(c | \theta_0) }{ p_0(c | \theta_0) } = 
+%\frac { \int dx \delta(c-s(x)) \, p_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  } = 
+%\frac {c \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, p_0(x | \theta_0)  } = c =\frac {p_1(x | \theta_0) }{ p_0(x | \theta_0) } 
 %\end{equation}
@@
 %%\begin{equation}
-%%f_c(q_s) = \int ds \delta(q_s-q_s(s)) \,f_c(s) \, / |\partial q_s / \partial s|
+%%p_c(q_s) = \int ds \delta(q_s-q_s(s)) \,p_c(s) \, / |\partial q_s / \partial s|
 %%\end{equation}
@@
 %\begin{equation}\label{eq:NP}
-%f( \{s_e\} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu f_1( s_e \, |\,  \theta)  + (1-\mu)\, f_0( s_e \,|\, \theta) \,\right] \; .
+%p( \{s_e\} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu p_1( s_e \, |\,  \theta)  + (1-\mu)\, p_0( s_e \,|\, \theta) \,\right] \; .
 %\end{equation}
@@
 %Usefulness:
-%If we can readily compute $f_c(x|\theta_0)$, then there would be no need for machine learning algorithms.
+%If we can readily compute $p_c(x|\theta_0)$, then there would be no need for machine learning algorithms.
 However,  these densities are often difficult to compute though one can readily use them as a generative model to produce samples.
-In that case, one might wish to use a machine learning algorithm to find $s(x) \approx f_1(x|\theta_0) / f_0(x|\theta_0)$.
+In that case, one might wish to use a machine learning algorithm to find $s(x) \approx p_1(x|\theta_0) / p_0(x|\theta_0)$.
 %
@@
 %
-%Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $f_1(x)$ and $f_0(x)$ define 
+%Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $p_1(x)$ and $p_0(x)$ define 
 %\begin{equation} 
-%s(x)=\frac{f_1(x)}{f_2(x)}
+%s(x)=\frac{p_1(x)}{p_2(x)}
 %\end{equation}.
-%Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
+%Define $p_1(c) = \int dx \delta(c - s(x) ) \, p_1(x)$ and $p_0(c) = \int dx \delta(c - s(x) ) \, p_0(x)$.
 %I wish to show
 %\begin{equation}
-%\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
+%\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)}
 %\end{equation}
@@
 %\begin{equation}
-%g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon|
+%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon|
 %\end{equation}
-%Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega$.
+%Similarly for $p_0(c)$. We can factor out of the integral $s(x)=p_1(x)/p_0(x)$ since it is constant over $\Omega$.
 %Thus
 %\begin{equation}
-%g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|
+%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|
 %\end{equation}
@@
 %\begin{equation}
-%\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;.
+%\frac{p_1(c)}{p_0(c)} = \frac{s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{p_1(x)}{p_0(x)} \;.
 %\end{equation}
@@
 %
-%The distribution of $s$ is fully determined by the mapping $s(x_e; \theta_0, \theta_1)$ and the distribution $f(x)$, which we can write
+%The distribution of $s$ is fully determined by the mapping $s(x_e; \theta_0, \theta_1)$ and the distribution $p(x)$, which we can write
 %\begin{equation}
-%g(s_e) = \int dx \, \delta(s_e-s(x)) \, f(x)  \; .
+%p(s_e) = \int dx \, \delta(s_e-s(x)) \, p(x)  \; .
 %\end{equation}
@@
 %\begin{equation}
-%g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega_c f_1(x)  / | \hat{n} \cdot \nabla q_s  |
+%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega_c p_1(x)  / | \hat{n} \cdot \nabla q_s  |
 %\end{equation}
-%Similarly for $g_0(c)$. We can factor out of the integral $s(x)=f_1(x)/f_0(x)$ since it is constant over $\Omega_c$.
+%Similarly for $p_0(c)$. We can factor out of the integral $s(x)=p_1(x)/p_0(x)$ since it is constant over $\Omega_c$.
 %Thus
 %\begin{equation}
-%g_1(c) = \int dx \delta(c-s(x)) f_1(x) = \int d\Omega f_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega f_0(x)  / | \hat{n} \cdot \nabla q_s  |
+%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega p_0(x)  / | \hat{n} \cdot \nabla q_s  |
 %\end{equation}
@@
 %\begin{equation}
-%\frac{g_1(c)}{g_0(c)} = \frac{s(x) \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega f_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{f_1(x)}{f_0(x)} \;\hspace{3em} \forall x\in\Omega_c.
+%\frac{p_1(c)}{p_0(c)} = \frac{s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{p_1(x)}{p_0(x)} \;\hspace{3em} \forall x\in\Omega_c.
 %\end{equation}
@@
 %
-%Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
+%Define $p_1(c) = \int dx \delta(c - s(x) ) \, p_1(x)$ and $p_0(c) = \int dx \delta(c - s(x) ) \, p_0(x)$.
 %I wish to show
 %\begin{equation}
-%\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
+%\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)}
 %\end{equation}
@@
 %
-%Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $f_1(x)$ and $f_0(x)$ define 
+%Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $p_1(x)$ and $p_0(x)$ define 
 %\begin{equation} 
-%s(x)=\frac{f_1(x)}{f_0(x)}
+%s(x)=\frac{p_1(x)}{p_0(x)}
 %\end{equation}.
-%Define $g_1(c) = \int dx \delta(c - s(x) ) \, f_1(x)$ and $g_0(c) = \int dx \delta(c - s(x) ) \, f_0(x)$.
+%Define $p_1(c) = \int dx \delta(c - s(x) ) \, p_1(x)$ and $p_0(c) = \int dx \delta(c - s(x) ) \, p_0(x)$.
 %I wish to show
 %\begin{equation}
-%\frac{g_1(c)}{g_0(c)} = \frac{f_1(x)}{f_0(x)}
+%\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)}
 %\end{equation}
@@
 %\begin{equation}
-%f(q_x|\theta) = \int dx \delta(q_x-q_x(x)) f(x|\theta)  / | \hat{n} \cdot \nabla q_s  |
+%p(q_x|\theta) = \int dx \delta(q_x-q_x(x)) p(x|\theta)  / | \hat{n} \cdot \nabla q_s  |
 %\end{equation}
@@
 %\begin{equation}
-%f(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, f(x|\theta) \, / | \hat{n} \cdot \nabla q_s  |
+%p(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, p(x|\theta) \, / | \hat{n} \cdot \nabla q_s  |
 %\end{equation}
-%sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $f_1(x)/f_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
+%sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $p_1(x)/p_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
 %\begin{equation}
-%\frac {f_1(c | \theta_0) }{ f_0(c | \theta_0) } = 
-%\frac { \int dx \delta(c-s(x)) \, f_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  } = 
-%\frac {c \int dx \delta(c-s(x)) \, f_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, f_0(x | \theta_0)  } = c =\frac {f_1(x | \theta_0) }{ f_0(x | \theta_0) } 
+%\frac {p_1(c | \theta_0) }{ p_0(c | \theta_0) } = 
+%\frac { \int dx \delta(c-s(x)) \, p_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  } = 
+%\frac {c \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, p_0(x | \theta_0)  } = c =\frac {p_1(x | \theta_0) }{ p_0(x | \theta_0) } 
 %\end{equation}

### 2015-04-09 tex/parametrized-learning.tex
first pass at new introduction

@@
 discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
-a generative model for the data is available for training and calibration.  We use the following notation
+a generative model for the data is available for training and calibration.
 
+As a concrete example, consider searches for new particles at the Large Hadron Collider. 
+The ATLAS and CMS experiments have published several hundred papers where the 
+final statistical result was formulated as a hypothesis test or confidence interval 
+using a generalized likelihood ratio test~\citep{Cowan:2010js}. This includes
+the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and 
+subsequent measurement of its properties.  
+The bulk of the likelihood ratio tests at the LHC are based on the distribution of a single event-level feature
+that discriminates between a process of interest (labeled \textit{signal}) and various other processes 
+(labeled \textit{background}).
+
+To improve the statistical power of these tests, hundreds of these searches have utilized discriminative
+classifiers trained on labeled data.
+Typically using libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citep{Hocker:2007ht}, though recently there has been progress in deep networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history.
+The use of classifiers is effective, but an intermediate step not obviously connected to the target likelihood ratio test.
+
+The labeled data used to train the classifiers are not obtained from the experimental apparatus. 
+Instead they are generated from a 
+detailed simulation, which includes a quantum mechanical description of the underlying particle 
+interactions, an exquisitely detailed simulation of the particle detectors with hundreds of millions of 
+sensors, and the data processing pipeline that reduces the data for a single event to tens or hundreds of high-level features used as input to the classifier. 
+It takes several minutes for this generative model to simulate a single collision or \textit{event} with Monte Carlo
+techniques. While particle physics offers an extreme example, it is common in science that generative 
+models are available while it not feasible to evaluate the probability density function that is implicitly being 
+sampled by the generative model.
+
+This paper describes how one can utilize generative statistical models for the data in conjunction with 
+discriminative classifiers to construct generalized likelihood ratio tests that are equivalent to 
+the one that would have been obtained if it were possible to evaluate the probability 
+density implicitly being sampled by the generative model.
+
+\subsection{Notation and Assumptions}
+
+We use the following ntoootation:
 \begin{itemize}
@@
 %\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
-\item $p( s | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$ \end{itemize}
+\item $p( s_{\theta_0, \theta_1} | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$ \end{itemize}
 We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.
 
-\newpage
+%\newpage
+\subsection{Prelude}
 
@@
 can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case 
-high energy physics where the simulation of particle detectors can only be done in the `forward mode'.
+high energy physics where the simulation of particle detectors can only be done in the `forward mode'. This same setting has been considered by \cite{ClaytonScott} and \cite{JMLR:v14:tong13a}.
 
-Our main result is that one can form an equivalent test based on 
+The main result of this paper is that one can form an equivalent test based on 
 %\begin{equation}
@@
 
-\section{Comments on classification and frequentist hypothesis tests}
+\subsection{Comments on classification and frequentist hypothesis tests}
 
@@
 
-\section{Composite hypotheses and the generalized likelihood ratio}\label{?}l{S:GLR}
+\section{Composite hypotheses and the generalized likelihood ratio}\label{S:GLR}
 

### 2015-05-12 tex/parametrized-learning.tex
Updated intro, whiteson(s) ref.

--- 2015-04-09 tex/parametrized-learning.tex
@@
 \title{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}
+\title{Approximating generalized likelihood ratio tests with calibrated discriminative classifiers}
 %\title{Calibrated, Parametrized Learning}
@@
 
-In many areas of science, likelihood ratio tests  are established tools for statistical inference. 
+In many areas of science, (generalized) likelihood ratio tests  are established tools for statistical inference. 
 Directly constructing the likelihood ratio for high-dimensional observations 
@@
 discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
-a generative model for the data is available for training and calibration.  We use the following notation
+a generative model for the data is available for training and calibration.
 
+As a concrete example, consider searches for new particles at the Large Hadron Collider. 
+The ATLAS and CMS experiments have published several hundred papers where the 
+final statistical result was formulated as a hypothesis test or confidence interval 
+using a generalized likelihood ratio test~\citep{Cowan:2010js}. This includes
+the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and 
+subsequent measurement of its properties.  
+The bulk of the likelihood ratio tests at the LHC are based on the distribution of a single event-level feature
+that discriminates between a hypothesized process of interest (labeled \textit{signal}) and various other processes 
+(labeled \textit{background}).
+
+To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers.
+Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citep{Hocker:2007ht} are commonly used.
+Recently there has been progress in using deep networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history.
+%The use of classifiers is effective, but an intermediate step not obviously connected to the target likelihood ratio test.
+
+%Collisions are usually selected with supervised learning methods 
+As noted in ~\citep{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
+A key point is that evaluating the loss for classification is a per-event operation, while evaluating the loss for the mass measurement is a per-experiment operation involving on a dataset with many events.
+They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process.
+%This paper shows that (to the extent a generalized likelihood ratio test in the full feature space is close to  optimal) one can achieve optimal results with a  specific per-event procedure by reformulating the target likelihood ratio in terms of a per-event classification followed by an specific calibration process.
+
+Generalized likelihood ratio tests enjoy a number of optimal properties in the asymptotic limit and are widely used even away from that limit.
+The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss.
+This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process.
+Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.
+
+%This paper describes how one can utilize generative statistical models for the data in conjunction with discriminative classifiers to construct generalized likelihood ratio tests that are equivalent to
+%the one that would have been obtained if it were possible to evaluate the probability 
+%density implicitly being sampled by the generative model.
+%
+%This paper describes how one can utilize generative statistical models for the data in conjunction with discriminative classifiers to construct generalized likelihood ratio tests that are equivalent to
+%the one that would have been obtained if it were possible to evaluate the probability 
+%density implicitly being sampled by the generative model.
+
+%The labeled data used to train the classifiers are not obtained from the experimental apparatus. 
+%Instead they are generated from a 
+%detailed simulation, which includes a quantum mechanical description of the underlying particle 
+%interactions, an exquisitely detailed simulation of the particle detectors with hundreds of millions of 
+%sensors, and the data processing pipeline that reduces the data for a single event to tens or hundreds of high-level features used as input to the classifier. 
+%It takes several minutes for this generative model to simulate a single collision or \textit{event} with Monte Carlo
+%techniques. While particle physics offers an extreme example, it is common in science that generative 
+%models are available while it not feasible to evaluate the probability density function that is implicitly being 
+%sampled by the generative model.
+
+\subsection{Notation and Assumptions}
+
+We use the following notation:
 \begin{itemize}
- \item $x$: a vector of features
+ \item $x$: a vector of features for an event
  \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
@@
 where $m$ is any strictly increasing or decreasing function. This result will be proven below.
-This allows us to recast the original likelihood ratio test into an alternate form in which a discriminative classifier is 
-used to learn $s(x; \theta_0, \theta_1)$. The discriminative classifier can be trained with data $(x,y=0)$ generated 
+This allows us to recast the original likelihood ratio test into an alternate form in which supervised learning is used to train the discriminative classifier $s(x; \theta_0, \theta_1)$.
+The discriminative classifier can be trained with data $(x,y=0)$ generated
 from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$.
-In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the discriminative classifier explicitly parametrized in terms of $(\theta_0, \theta_1)$.
+In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the classifier  parametrized in terms of $(\theta_0, \theta_1)$.
 
-While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, we are able to reformulate it such that the machine learning problem is an event-by-event classification problem.
+While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, we are able to reformulate it such that the machine learning problem is a per-event classification problem.
 This follows from the fact that we assume the $x_e$ to be i.i.d.
 
-\section{Comments on classification and frequentist hypothesis tests}
+\subsection{Comments on classification and frequentist hypothesis tests}
 
@@
 
-The first possible source of confusion we wish to avoid is that  $p(x|\theta)$  is a generative \textit{statistical model} for the features $x$, not a generative classifier.
+The first possible source of confusion we wish to avoid is that here $p(x|\theta)$  is a generative \textit{statistical model} for the features $x$, not a generative classifier.
 We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory, able to make predictions about $x$ and being motivated by domain-specific considerations.
-For example, in the context of high energy particle physics $p(x|\theta)$ is based on quantum field theory and a detailed simulation of the particle detector and data processing algorithms that transform raw sensor data into the feature vector $x$.
-Moreover, we are not attempting to learn the generative model $p(x|\theta)$, we are taking it as given and trying to learn the corresponding likelihood ratio test.
+For example, in the context of HEP $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$.
+Moreover, we are not attempting to learn the generative model $p(x|\theta)$, we are taking it as given and trying to learn the corresponding (generalized) likelihood ratio test.
 %The point here is to make the distinction of $p(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.
 
-The second possible source of confusion is that the likelihood ratio $T(D)$ is aimed at tests based on the entire dataset; we are not interested in thresholded classification on individual events $x_e$.
-Additionally, we know that both discriminative and generative classifier scores are often poorly calibrated. 
-For instance, often we wish to have well calibrated p-values defined by $P(T(D) > k |\theta)$, not well calibrated posterior probabilities $p(y|x)$.
+The second possible source of confusion is that 
+we are not directly interested in calibrating the classification score in terms of a per-event posterior probability $p(y|x)$. 
+Instead, we are interested in the approximation of the per-experiment likelihood ratio, which might be used to calculate p-values defined by $P(T(D) > k |\theta)$.
+%
+%%the likelihood ratio $T(D)$ is aimed at tests based on the entire dataset; 
+%we are not interested in thresholded classification on individual events $x_e$.  
+%Knowing that classifier scores are often poorly calibrated is an immediate concern.
+
+%We wish to have well calibrated p-values defined by $P(T(D) > k |\theta)$, not well calibrated posterior probabilities $p(y|x)$.
 
@@
 
-We need to show the density
+We need to show that for all $x$, the density
 \begin{equation}
@@
 \end{equation}
-is the same as
+is equal to the density
 \begin{equation}
@@
 In general we will be interested in the machine learning problem that approximates these distributions based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
-The first step in this direction is to confirm that a discriminative classifier obtained from common training procedures will yield a function that is one-to-one with $p(x|\theta_0)/p(x|\theta_1)$.
+The first step in this direction is to confirm that a discriminative classifier obtained from common supervised learning algorithms will yield a function that is one-to-one with $p(x|\theta_0)/p(x|\theta_1)$.
 %In particular, 
@@
 
-\subsection{The standard discriminative classification setting} For fixed $\theta_0$ and $\theta_1$ we can generate 
-large samples from each model and train a classifier. To be concrete, let's use $p(x|\theta_0)$ to generate training 
+\subsection{The standard discriminative classification setting} 
+For fixed $\theta_0$ and $\theta_1$ we can generate 
+large samples from each model and train a classifier. To be concrete, let us use $p(x|\theta_0)$ to generate training 
 data $(x_i,  y_i=0)$ and $p(x|\theta_1)$ to generate training data $(x_i , y_i=1)$.
-With balanced training data  -- $\mbox{p(y=1)=p(y=0)=\sfrac{1}{2}}$ -- a quadratic loss function will lead to classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the desired likelihood ratio $p(x|\theta_0)/p(x|\theta_1)$.
-Thus, standard classification approaches will lead to discriminative classifiers needed to produce an equivalent likelihood ratio test.
-Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g.
+With balanced training data   ($p(y=1)=p(y=0)=\sfrac{1}{2}$) a quadratic loss function will lead to classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the desired likelihood ratio $p(x|\theta_0)/p(x|\theta_1)$.
+Thus, standard classification approaches with various surrogate loss functions lead to discriminative classifiers that approximate the desired likelihood ratio tests.
+Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
 histograms or kernel density estimation) to approximate $\hat{g}(\hat{s}|\theta)$.
@@
 
-Thus, in the  limit of large samples from the generative model,  we can approximate arbitrarily well the original likelihood ratio test.
+In the  limit of large samples from the generative model,  we can approximate the original likelihood ratio test arbitrarily well.
 With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{g}(\hat{s}|\theta)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test.

### 2015-05-12 tex/parametrized-learning.tex
more cleanup

@@
 
-Significant literature exists around generative and discriminative classifiers~\citep{AndrewY.Ng}.
+Vast literature exists around generative and discriminative classifiers~\citep{AndrewY.Ng}.
 Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule.
@@
 \end{equation}
-Here we see that the optimal $T$ for the experiment is composed of a sum over events of a  linear function of the per-event function $q(x)$.
-A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.
+Here we see that the test statistic $T$ for the experiment is composed of a sum over events of the per-event function $q(x)$.
+A sum over a monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.
 
@@
 
-Thus far we have shown that likelihood ratio tests based on $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via hypothesis tests based on the univariate densities $p(s|\theta)$ for the very special dimensionality reduction map $s(x|\theta_0, \theta_1)$.
-The motivation for this is that often it is not possible to evaluate the density $p(x|\theta)$ at a given point $x$.
-This approach is not useful if it is not possible to approximate $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ without evaluating the density $p(x|\theta)$.
-In order for this approach to be useful, we need to be able to approximate both based on samples $\{(x,\theta)\}$ drawn from the generative model $p(x|\theta)$.
+Thus far we have shown that likelihood ratio tests based on $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via hypothesis tests based on the univariate densities $p(s|\theta)$ for the particular dimensionality reduction map $s(x|\theta_0, \theta_1)$.
+% Often it is not possible to evaluate the density $p(x|\theta)$ at a given point $x$.  
+%This approach will not be useful if it is not possible to approximate $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ without evaluating the density $p(x|\theta)$. 
+%In order for this approach to be useful, we need to be able to approximate both based on samples $\{(x,\theta)\}$ drawn from the generative model $p(x|\theta)$.  
+In order for this approach to be useful in a setting where it is not possible to evaluate the density $p(x|\theta)$, we need to be able to approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
 
@@
 data $(x_i,  y_i=0)$ and $p(x|\theta_1)$ to generate training data $(x_i , y_i=1)$.
-With balanced training data   ($p(y=1)=p(y=0)=\sfrac{1}{2}$) a quadratic loss function will lead to classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the desired likelihood ratio $p(x|\theta_0)/p(x|\theta_1)$.
-Thus, standard classification approaches with various surrogate loss functions lead to discriminative classifiers that approximate the desired likelihood ratio tests.
+With balanced training data   ($p(y=1)=p(y=0)=\sfrac{1}{2}$) a quadratic loss function will lead to classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the desired per-event likelihood ratio $q(x)$.
+Thus, standard supervised learning algorithms with various surrogate loss functions lead to discriminative classifiers that approximate a monotonic function of per-event likelihood ratio $q(x)$.
 Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
@@
 Somewhat unusually, the unknown values of the parameters are taken
-as input to the classifier, as latent variables whose values will be specified via the enveloping (generalized) likelihood ratio test.
+as input to the classifier, and are latent variables whose values will be specified via the enveloping (generalized) likelihood ratio test.
 We denote the learned family of classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based roughly on the following algorithmic flow.

### 2015-05-13 tex/parametrized-learning.tex
added measurement section.

@@
 and the conditional maximum likelihood estimator
-\begin{equation}
+\begin{equation}\label{eq:cmle}
 \hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p(D | \theta) \; .
@@
 In most settings that make use of likelihood ratio tests, the likelihood is based directly on some approximation of density for the observed data via $\hat{f}(x|\theta)$.
-Approximating the density $\hat{f}(x|\theta)$ is difficult for high-dimensional data, which motivates the use of the dimensionality reduction map $\hat{s}(x)$ and likelihood ratio tests based on the density $\hat{g}(\hat{s}|\theta)$.
+In general, approximating the density $\hat{f}(x|\theta)$ is difficult for high-dimensional data, which motivates the use of the dimensionality reduction map $\hat{s}(x)$ and likelihood ratio tests based on the univariate density $\hat{g}(\hat{s}|\theta)$.
 In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$.
-In the parametrized setting this it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$, so  we must embed the classifier into the likelihood function to carry out the composition $\hat{g}\circ \hat{s}$.
-A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and discriminative classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
+In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{g}\circ \hat{s}$.
+A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
@@
 This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
-In practice, one may want to interpolate the distribution between discrete values of $\theta$ to produce a continuous parametrization for $\hat p(\hat s | \theta)$.
-In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
+In our HEP examples we have chosen to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$.
+%In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
+One can easily imagine a number of approaches to embedding the classifier and estimating the density $\hat p(\hat s|\theta)$ and the relative merits of those approaches will depend critically on the dimensionality of $\theta$ and the computational cost of the generative model.
+We leave a more general strategy for this overarching optimization problem as an area of future work.
 
-\section{Typical usage of machine learning in HEP}
+\section{Applications in HEP}
 
-In high-energy physics (HEP) we are often searching for some 
+In HEP we are often searching for some 
 class of events, generically referred to as \textit{signal}, in the presence of a separate class 
-of \textit{background} events.
-Generalized likelihood ratio tests are used widely in HEP ~\citep{Cowan:2010js}, most notably in the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
-For each event we measure some quantities $x$ that have corresponding distributions
+of \textit{background} events.  For each event we measure some quantities $x$ that have corresponding distributions 
 $p_b(x|\nu)$ for background and $p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing 
-uncertainties in the underlying physics prediction or response of the measurement device. In the simple setup, the 
-total model is a mixture of the signal and background components, and $\mu$ is the mixture coefficient associate 
-dot the signal component. The generative model in this case is
+uncertainties in the underlying physics prediction or response of the measurement device. The 
+total model is a mixture of the signal and background, and $\mu$ is the mixture coefficient associate 
+to the signal component. 
 \begin{equation}\label{eq:hepGen}
@@
 \end{equation}
-New particle searches correspond to the hypothesis test $\mu=0$, and are generally formulated with the
-generalized likelihood ratio profiling over $\nu$.
+New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
+generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\citep{Cowan:2010js}.
+This approach was used for the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
 
-Often machine learning classification algorithms are trained on large samples of synthetic data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density $p_s(x|\nu_0)$ (not the signal-plus-background).
+%Generalized likelihood ratio tests are used widely in HEP ~\citep{Cowan:2010js}, most notably in the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
+
+\subsection{Typical usage pattern}
+
+%Often machine learning classification algorithms are trained on 
+In this setting, large samples of synthetic data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density $p_s(x|\nu_0)$.
+Importantly, the $y=1$ label corresponds to the signal only, and not to the alternate signal-plus-background hypothesis.
 The resulting classifier approximates the regression function $p_s(x|\nu_0)/(p_s(x|\nu_0)+p_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$.
-The resulting classifier is denoted $\hat s(x)$.
-Based on this classifier and large samples of synthetic data drawn from $p_s(x | \nu)$ and $p_b(x | \nu)$ we construct the distributions  $p_s(\hat s | \nu)$ and $p_b(\hat s | \nu)$.
+Associating the $y=1$ label to the signal component has advantages because it helps the classifier focus its capacity on the relevant regions in the feature space, particularly when the signal is a very small perturbation to the background.
+%At the LHC one can have $\mu \sim 10^{-6}$. 
+%Note that the classifier is not optimal for $\nu\ne\nu_0$.
+
+ Once the classifier is trained, large samples of synthetic data drawn from $p_s(x | \nu)$ and $p_b(x | \nu)$, then we estimate the distributions  $\hat{p}_s(\hat s | \nu)$ and $\hat{p}_b(\hat s | \nu)$ continuously parametrized in $\nu$.
 An example of the distributions of the distribution of $\hat s$ for the signal and background events with $\nu=\nu_0$ is shown in Figure~\ref{fig:tmva}.
@@
 
-These steps lead to a subsequent statistical analysis where one observes in data \\
+These steps feed into a subsequent statistical test based on the observed data \\
 ${D=(x_1, \dots, x_n)}$.
 For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution.
-In particular, one forms the statistical model
+In particular, one forms the statistical model\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood.}
 \begin{equation}\label{eq:typicalML}
-p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( \hat s(x_e) \, |\,  \nu)  + (1-\mu)\, p_b( \hat s(x_e) \,|\, \nu) \,\right] \; ,
+p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu \hat{p}_s( \hat s(x_e) \, |\,  \nu)  + (1-\mu)\, \hat{p}_b( \hat s(x_e) \,|\, \nu) \,\right] \; .
 \end{equation}
-where $\mu=0$ is the null (background-only) hypothesis and $\mu>0$ is the alternate (signal-plus-background) hypothesis.\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood.} Typically, we are interested in inference on $\mu$ and $\nu$ are nuisance parameters.
+%where $\mu=0$ is the null (background-only) hypothesis and $\mu>0$ is the alternate (signal-plus-background) hypothesis.\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood.}
+%Typically, we are interested in inference on $\mu$ and $\nu$ are nuisance parameters.
 
@@
 Nuisance parameters are an after thought in the typical usage of machine learning in HEP.
-In fact, most machine learning discussions would only consider $p_b(x)$ and $p_s(x)$.
+In fact, most discussions would related to the training and optimizing the classifier only consider $p_b(x)$ and $p_s(x)$ with $\nu=\nu_0$ being implicit.
 However, as experimentalists we know that we must account for various forms of systematic uncertainty, parametrized by nuisance parameters $\nu$.
@@
 background, which is equivalent (in an ideal setting) to training a classifier for  null (background-only) vs.
-alternate (signal-plus-backgound) as
-\begin{equation}
- \frac{p(x| 0, \nu_0)}{p(x|\hat \mu, \nu_0)} =  \frac{p_b(x| \nu_0)}{ \mu p_s( x_e \, |\,  \nu_0)  + (1-\mu)\, p_b( x_e \,|\, \nu_0)} = \left[ c_1 + c_2   \frac{p_s(x| \nu_0)}{ p_b( x_e \,|\, \nu_0)} \right ]^{-1} \; ,
-\end{equation}
-and $c_1$ and $c_2$ are constants.
-Specifically, the two  likelihood ratios are in one-to-one correspondence, so an ideal algorithm would lead to equivalent tests.
+alternate (signal-plus-backgound) since the resulting regression functions are one-to-one with each other.
+% as 
+%\begin{equation}
+% \frac{p(x| 0, \nu_0)}{p(x|\hat \mu, \nu_0)} =  \frac{p_b(x| \nu_0)}{ \mu p_s( x_e \, |\,  \nu_0)  + (1-\mu)\, p_b( x_e \,|\, \nu_0)} = \left[ c_1 + c_2   \frac{p_s(x| \nu_0)}{ p_b( x_e \,|\, \nu_0)} \right ]^{-1} \; ,
+%\end{equation}
+%and $c_1$ and $c_2$ are constants. Specifically, the two  likelihood ratios are in one-to-one correspondence, so an ideal algorithm would lead to equivalent tests. 
 In contrast, in the case of the generalized likelihood ratio test 
-\begin{equation}
+\begin{equation}\label{eq:hep_improved}
  \frac{p(x| 0, \hat{\hat{ \nu}})}{p(x|\hat \mu, \hat\nu)} =  \frac{p_b(x| \hat{\hat{ \nu}})}{ \hat \mu p_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, p_b( x_e \,|\, \hat \nu)} \; ,
@@
 
-It is common that the generative model for the low-level features is a mixture model of several components
+In this section we generalize the capacity focusing technique of training classifiers to discriminate between components of a mixture model.
+First, we generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
 \begin{equation}
@@
 \end{equation}
-In the case of particle physics, the distributions $p(x|\theta)$ is not a Gaussian Mixture Model, but mixture of  complicated distributions associated to relatively few types of particle interactions.
+%In the case of particle physics, the distributions $p(x|\theta)$ is not a Gaussian Mixture Model, but mixture of  complicated distributions associated to relatively few types of particle interactions.
 Moreover, when searching for a new particle, the null hypothesis would correspond to some of the coefficients $w_c=0$ while the alternate ``signal-plus-background'' hypothesis would have $0<w_{c\in \rm signal} \ll w_{c\in \rm background}$.
@@
 These all present challenges to machine learning algorithms that aim to learn $s(x; \theta_0, \theta_1)$.
-
+%
 %\begin{equation}
@@
 %\end{equation}
-
-However, it is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
-\begin{eqnarray}
-\frac{p(x|\theta_0)}{p(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_c w_c(\theta_1) p_c(x| \theta_1)} 
+%
+It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
+\begin{eqnarray}\label{eq:decompose}
+\frac{p(x|\theta_0)}{p(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} 
 \\
@@
 In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).
+Equation~\ref{eq:decompose} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
 
@@
 This approach is not realistic in situations where $p_c(x|\nu)$ is realized with expensive computer simulations; thus, some interpolation strategy or emulator over a fixed set of parameter points $\{\nu_i\}$ would typically be used.
+
+\subsection{Measuring particle properties}
+
+While the original motivation for this work was to improve the treatment of systematic uncertainties in new particle searches by parametrizing the classifier in terms of the nuisance parameters $\nu$, the same approach can be used to for parameters of interest.
+In the case of new particle searches the parameter of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
+In the case of measuring particle properties the distribution of the features also depend on parameters of interest, such as a particle's mass and quantum numbers, so that one has $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$ includes both parameters of interest and nuisance parameters.
+
+This is a significant step forward in the usage of machine learning in HEP, because classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
+The work of  ~\citep{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
+This approach also offers the advantage that it explicitly relates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
 

### 2015-05-13 tex/parametrized-learning.tex
more

@@
 In the case of new particle searches the parameter of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
-In the case of measuring particle properties the distribution of the features also depend on parameters of interest, such as a particle's mass and quantum numbers, so that one has $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$ includes both parameters of interest and nuisance parameters.
+When measuring particle properties the distribution of the features also depend on parameters of interest, such as a particle's mass and quantum numbers, so that one has $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$ includes both parameters of interest and nuisance parameters.
 
@@
 This approach also offers the advantage that it explicitly relates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
+
+Another approach that is similar is the so-called matrix element method in which one tries to directly calculate the target likelihood ratio by performing a computationally intensive integral associated to the detector response.
+The detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector.
+Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time per event to calculate the likelihood ratio $q(x)$.
+The work here can be seen as aiming at the same conceptual target, but utilizing machine learning to overcome the complexity of the detector simulation.
+It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage.
+In practice, the matrix element method has only been used for searches and measurement of a single physical parameter.
+
+Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson and (if we are so lucky) measurement of the mass of supersymmetric particles in cascade decays.
+Both of these examples involve datasets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
+%The state of the art approaches for supersymmetric particle masses typically involve high-level features engineered based on physics intuition
+%Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
+The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects.
 
@@
 
-We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate statistical inference likelihoodased on the  ratio $p(x|\theta_0)/p(x|\theta_1)$ when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$.
+We have shown how a generalized likelihood ratio test on a dataset of multiple i.i.d.
+events with several features can be reformulated in terms of the distributions of discriminative classifier output.
+This allows one to turn the optimization of the procedure into optimization of a per-event classification problem and a calibration procedure.
+
+%We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate statistical inference likelihood based on the  ratio $p(x|\theta_0)/p(x|\theta_1)$ when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$.
+
 This approach leverages the power of machine learning in a classical statistical setting.

### 2015-05-14 tex/parametrized-learning.tex
new refs

@@
 In the case of new particle searches the parameter of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
-When measuring particle properties the distribution of the features also depend on parameters of interest, such as a particle's mass and quantum numbers, so that one has $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$ includes both parameters of interest and nuisance parameters.
+When measuring particle properties the distribution of the features also depend on parameters of interest, such as a particle's mass and quantum numbers.
+This is easily accommodated by extending $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$ includes both parameters of interest and nuisance parameters.
 
-This is a significant step forward in the usage of machine learning in HEP, because classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
+This formalism represents is a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
 The work of  ~\citep{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
-This approach also offers the advantage that it explicitly relates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
+This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
 
-Another approach that is similar is the so-called matrix element method in which one tries to directly calculate the target likelihood ratio by performing a computationally intensive integral associated to the detector response.
-The detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector.
+Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citep{Volobouev:2011vb}.
+In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method.
 Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time per event to calculate the likelihood ratio $q(x)$.
@@
 It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage.
-In practice, the matrix element method has only been used for searches and measurement of a single physical parameter.
+In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citep{Aaltonen:2010yz}).
 
-Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson and (if we are so lucky) measurement of the mass of supersymmetric particles in cascade decays.
+Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citep{Allanach:2000kt}.
 Both of these examples involve datasets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
@@
 %Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
-The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects.
+The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citep{Bolognesi:2012mm}.
 
@@
 
-We have shown how a generalized likelihood ratio test on a dataset of multiple i.i.d.
-events with several features can be reformulated in terms of the distributions of discriminative classifier output.
-This allows one to turn the optimization of the procedure into optimization of a per-event classification problem and a calibration procedure.
+We have outlined a technique to reformulate generalized likelihood ratio test over a high-dimensional dataset with multiple events in terms of a univariate density of a classifier score. 
+We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$
+This allows one to turn the optimization of the procedure for the per-experiment test into optimization of a per-event classification problem and a calibration procedure.
+This approach leverages the power of machine learning in a classical statistical setting.
 
 %We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate statistical inference likelihood based on the  ratio $p(x|\theta_0)/p(x|\theta_1)$ when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$.
-
-This approach leverages the power of machine learning in a classical statistical setting.
 

### 2015-05-14 tex/parametrized-learning.tex
new conclusion

@@
 subsequent measurement of its properties.
+
 The bulk of the likelihood ratio tests at the LHC are based on the distribution of a single event-level feature
@@
 (labeled \textit{background}).
-
 To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers.
 Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citep{Hocker:2007ht} are commonly used.
-Recently there has been progress in using deep networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history.
+Recently, there has been progress in using deep networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history.
 %The use of classifiers is effective, but an intermediate step not obviously connected to the target likelihood ratio test.
@@
 %Collisions are usually selected with supervised learning methods 
+While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citep{Dempster1965}, that is no longer true in the context of parameter estimation or compound hypothesis tests with nuisance parameters.
 As noted in ~\citep{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
-A key point is that evaluating the loss for classification is a per-event operation, while evaluating the loss for the mass measurement is a per-experiment operation involving on a dataset with many events.
+A fundamental distinction is that evaluating the loss for classification is a per-event operation, while evaluating the loss for a mass measurement (and other estimation problems) is a per-experiment operation involving a dataset with many events.
 They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process.
@@
 we are not directly interested in calibrating the classification score in terms of a per-event posterior probability $p(y|x)$. 
-Instead, we are interested in the approximation of the per-experiment likelihood ratio, which might be used to calculate p-values defined by $P(T(D) > k |\theta)$.
+Instead, we are interested in the approximation of the per-experiment likelihood ratio, which can be used to calculate p-values defined by $P(T(D) > k |\theta)$.
 %
@@
 Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
-Similar points are made in~\citep{ClaytonScott}.
+Similar points have been made by~\citep{ClaytonScott}.
 
@@
 
- The target hypothesis test is based on 
+We are interested in reformulating the target likelihood ratio  
 \begin{equation}
@@
 As we will show, our goal is to learn a monotonic function of $p(x|\theta_0)/p(x|\theta_1)$, which will share the same iso-contours.
-Then the remaining challenge is to find the appropriate rescaling that gives back  linear function of $q(x)$.
+Then the remaining challenge is to find the appropriate monotonic function that gives back a linear function of $q(x)$.
 Our claim is that the generative model $p(x|\theta)$ can be used to calibrate the density $p(s|\theta)$ and that
@@
 
-\subsection{Training a parametrized, discriminative classifier}
+\subsection{Training a parametrized classifier}
 
@@
 We have outlined a technique to reformulate generalized likelihood ratio test over a high-dimensional dataset with multiple events in terms of a univariate density of a classifier score. 
-We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$
-This allows one to turn the optimization of the procedure for the per-experiment test into optimization of a per-event classification problem and a calibration procedure.
+We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$.
+Furthermore, this allows one to turn the optimization of the procedure for the per-experiment test into optimization of a per-event classification problem and a calibration procedure.
 This approach leverages the power of machine learning in a classical statistical setting.
+In particular, it extends the common usage of machine learning for simple hypothesis testing to include the parameter estimation, confidence intervals, and composite hypothesis testing.
 

### 2015-05-29 tex/nips2015-draft.tex
text changes, try out NIPS draft

--- 2015-05-14 tex/parametrized-learning.tex
@@
-\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
+\documentclass{article} % For LaTeX2e
+\usepackage{nips15submit_e,times}
+\usepackage{hyperref}
+\usepackage{url}
+%\documentstyle[nips14submit_09,times,art10]{article} % For LaTeX 2.09
 
-\usepackage{jmlr2e}
-\usepackage{lineno}
-\linenumbers
-%\jmlrheading{1}{2014}{X-XX}{4/00}{10/00}{Kyle Cranmer}
-
-%\ShortHeadings{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}{Kyle Cranmer}
-
-\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
-%\geometry{landscape}                		% Activate for for rotated page geometry
-%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
-								% TeX will automatically convert eps --> pdf in pdflatex		
 \usepackage{amssymb, amsmath}
@@
 \usepackage{xfrac}
-
 \DeclareMathOperator*{\argmin}{arg\,min}
@@
 
-%\title{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}
-\title{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}
+\renewcommand\refname{}
+
 \title{Approximating generalized likelihood ratio tests with calibrated discriminative classifiers}
-%\title{Calibrated, Parametrized Learning}
-\author{Kyle Cranmer}
-%\date{}							% Activate to display a given date or no date
+
+\author{
+Kyle S.~Cranmer\thanks{ Use footnote for providing further information
+about author (webpage, alternative address)---\emph{not} for acknowledging
+funding agencies.} \\
+Center for Cosmology and Particle Physics\\
+       Center for Data Science\\
+       New York University \\
+       New York, NY 10003, USA
+\texttt{kyle.cranmer@nyu.edu} \\
+%\And
+%Coauthor \\
+%Affiliation \\
+%Address \\
+%\texttt{email} \\
+%\AND
+%Coauthor \\
+%Affiliation \\
+%Address \\
+%\texttt{email} \\
+%\And
+%Coauthor \\
+%Affiliation \\
+%Address \\
+%\texttt{email} \\
+%\And
+%Coauthor \\
+%Affiliation \\
+%Address \\
+%\texttt{email} \\
+%(if needed)\\
+}
+
+% The \author macro works with any number of authors. There are two commands
+% used to separate the names and addresses of multiple authors: \And and \AND.
+%
+% Using \And between authors leaves it to \LaTeX{} to determine where to break
+% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
+% puts 3 of 4 authors names on the first line, and the last on the second
+% line, try using \AND instead of \And before the third author name.
+
+\newcommand{\fix}{\marginpar{FIX}}
+\newcommand{\new}{\marginpar{NEW}}
+
+%\nipsfinalcopy % Uncomment for camera-ready version
 
@@
 
-\author{\name Kyle Cranmer \email kyle.cranmer@nyu.edu \\
-       \addr Center for Cosmology and Particle Physics\\
-       Center for Data Science\\
-       New York University \\
-       New York, NY 10003, USA
-}
-
-%\editor{Leslie Pack Kaelbling}
-\editor{}
-
 \maketitle
 
-\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
+\begin{abstract}
 We demonstrate how discriminative classifiers can be used to approximate 
@@
 \end{abstract}
-%
-%\begin{keywords}
-%  Bayesian Networks, Mixture Models, Chow-Liu Trees
-%\end{keywords}
-
-\maketitle
 
@@
 
-In many areas of science, (generalized) likelihood ratio tests  are established tools for statistical inference. 
-Directly constructing the likelihood ratio for high-dimensional observations 
-is often not possible or is computationally impractical. Here we demonstrate how 
-discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
+In many areas of science, (generalized) likelihood ratio tests  are established tools for statistical inference~\cite{Wilks1938}. 
+It is increasingly common that a generative model is used to describe complex processes that tie 
+parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
+Directly evaluating the likelihood ratio in these cases is often not possible or is computationally impractical.
+Here we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when
 a generative model for the data is available for training and calibration.
@@
 final statistical result was formulated as a hypothesis test or confidence interval 
-using a generalized likelihood ratio test~\citep{Cowan:2010js}. This includes
-the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and 
+using a generalized likelihood ratio test~\cite{Cowan:2010js}. This includes
+the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa} and 
 subsequent measurement of its properties.
@@
 To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers.
-Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citep{Hocker:2007ht} are commonly used.
-Recently, there has been progress in using deep networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history.
+Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\cite{Hocker:2007ht} are commonly used.
+Recently, there has been progress in using deep networks~\cite{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\cite{HepML}, the largest Kaggle challenge in history.
 %The use of classifiers is effective, but an intermediate step not obviously connected to the target likelihood ratio test.
@@
 %Collisions are usually selected with supervised learning methods 
-While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citep{Dempster1965}, that is no longer true in the context of parameter estimation or compound hypothesis tests with nuisance parameters.
-As noted in ~\citep{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
+While classification accuracy can lead to optimal approaches for simple hypothesis tests~\cite{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters.
+As noted in ~\cite{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
 A fundamental distinction is that evaluating the loss for classification is a per-event operation, while evaluating the loss for a mass measurement (and other estimation problems) is a per-experiment operation involving a dataset with many events.
@@
 Generalized likelihood ratio tests enjoy a number of optimal properties in the asymptotic limit and are widely used even away from that limit.
+These asymptotic properties, the difficulty in specifying a prior over the parameters needed to evaluate a Bayes risk, and the vast literature associated to likelihood ratio tests has led many users of such tests to not be explicit about the loss function they wish to minimize for their experiment.
+Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself.
 The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss.
@@
 
-Vast literature exists around generative and discriminative classifiers~\citep{AndrewY.Ng}.
+Vast literature exists around generative and discriminative classifiers~\cite{AndrewY.Ng}.
 Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule.
@@
 Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
-Similar points have been made by~\citep{ClaytonScott}.
+Similar points have been made by~\cite{ClaytonScott}.
 
@@
 In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{g}\circ \hat{s}$.
-A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
+A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\cite{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
@@
 New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
-generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\citep{Cowan:2010js}.
-This approach was used for the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
+generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\cite{Cowan:2010js}.
+This approach was used for the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa}.
 
-%Generalized likelihood ratio tests are used widely in HEP ~\citep{Cowan:2010js}, most notably in the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
+%Generalized likelihood ratio tests are used widely in HEP ~\cite{Cowan:2010js}, most notably in the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa}.
 
@@
 
-\begin{figure}[htbp]
-\begin{center}
- \includegraphics[height=2in]{example-TMVA-BDT.pdf}
- \includegraphics[height=2in]{example-TMVA-ROC.pdf}
-\caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).
+%\begin{figure}[htbp]
+%\begin{center}
+% \includegraphics[height=2in]{example-TMVA-BDT.pdf}
+% \includegraphics[height=2in]{example-TMVA-ROC.pdf}
+%\caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).
 Right: the corresponding ROC curve (right) for this and other classifiers.
 (Figures taken from TMVA manual.)}
-\label{fig:tmva}
-\end{center}
-\end{figure}
-%\bigskip
+%\label{fig:tmva}
+%\end{center}
+%\end{figure}
+%%\bigskip
 
@@
 
-\subsection{Decomposing tests between mixture models into their components}
-
-In this section we generalize the capacity focusing technique of training classifiers to discriminate between components of a mixture model.
+%\subsection{Decomposing tests between mixture models into their components}
+%
+%In this section we generalize the capacity focusing technique of training classifiers to discriminate between components of a mixture model.
 First, we generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
-\begin{equation}
-p(x|\theta)=\sum_c w_c(\theta) p_c(x| \theta) \;.
-\end{equation}
-%In the case of particle physics, the distributions $p(x|\theta)$ is not a Gaussian Mixture Model, but mixture of  complicated distributions associated to relatively few types of particle interactions.
+%\begin{equation}
+%p(x|\theta)=\sum_c w_c(\theta) p_c(x| \theta) \;.
+%\end{equation}
+%%In the case of particle physics, the distributions $p(x|\theta)$ is not a Gaussian Mixture Model, but mixture of  complicated distributions associated to relatively few types of particle interactions.
 Moreover, when searching for a new particle, the null hypothesis would correspond to some of the coefficients $w_c=0$ while the alternate ``signal-plus-background'' hypothesis would have $0<w_{c\in \rm signal} \ll w_{c\in \rm background}$.
@@
 These all present challenges to machine learning algorithms that aim to learn $s(x; \theta_0, \theta_1)$.
-%
-%\begin{equation}
-%\frac{p(x|\theta_0)}{p(x|\theta_1)}= \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_c w_c(\theta_1) p_c(x| \theta_1)} 
-%\end{equation}
-%
-%\begin{equation}
-%\frac{p(x|\theta_0)}{p(x|\theta_1)}= \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} 
-%\end{equation}
-%
-It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
-\begin{eqnarray}\label{eq:decompose}
-\frac{p(x|\theta_0)}{p(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} 
-\\
-%&=& \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \\
-%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)}{ w_c(\theta_0) p_c(x| \theta_0)}  \right]^{-1} \\
-&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \\
-&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
-\end{eqnarray}
-The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.
+%%
+%%\begin{equation}
+%%\frac{p(x|\theta_0)}{p(x|\theta_1)}= \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_c w_c(\theta_1) p_c(x| \theta_1)} 
+%%\end{equation}
+%%
+%%\begin{equation}
+%%\frac{p(x|\theta_0)}{p(x|\theta_1)}= \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} 
+%%\end{equation}
+%%
+%It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
+%\begin{eqnarray}\label{eq:decompose}
+%\frac{p(x|\theta_0)}{p(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} 
+%\\
+%%&=& \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \\
+%%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)}{ w_c(\theta_0) p_c(x| \theta_0)}  \right]^{-1} \\
+%&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \\
+%&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
+%\end{eqnarray}
+%The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.
 The third line uses Theorem 1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
@@
 Equation~\ref{eq:decompose} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
-
 %
-%The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $\hat s(x; \nu_0, \nu_1)$.
+%
+%%
+%%The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $\hat s(x; \nu_0, \nu_1)$.
 Creating the family of classifiers is straight forward, one simply augments the training data with $x$ examples drawn from several values of $\nu$ and then includes the corresponding value of $\nu$ as an input to the classifier.
@@
 However, when performing a likelihood ratio test, one can calculate the maximum likelihood estimate and conditional maximum likelihood estimate according to Eq.~\ref{eq:mle} and then form the equivalent likelihood ratio test according to Eq.~\ref{eq:equivLRtest}.
-
-%Equivalently, once one has determined $\hat{\theta}$ and $\hat{\hat{\theta}}$, then one can specify the parameters of the mapping $s(x;\hat{\hat{\theta}}, \hat{\theta})$ and then following Eq.~\ref{eq:NP} form the likelihood
-%\begin{equation}\label{eq:parametrizedML}
-%p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( s(x_e;\nu_0, \nu_1) \, |\,  \nu)  + (1-\mu)\, p_b( s(x_e;\nu_0, \nu_1) \,|\, \nu) \,\right] \; ,
-%\end{equation}
-%which is a natural generalization of Eq.~\ref{eq:typicalML}.
-
-%The construction in Eq.~\ref{eq:parametrizedML} presents a computational challenge as for each value of $\nu$ one has a new mapping $s: x\to\mathbb{R}$.
+%
+%
+%%Equivalently, once one has determined $\hat{\theta}$ and $\hat{\hat{\theta}}$, then one can specify the parameters of the mapping $s(x;\hat{\hat{\theta}}, \hat{\theta})$ and then following Eq.~\ref{eq:NP} form the likelihood
+%%\begin{equation}\label{eq:parametrizedML}
+%%p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( s(x_e;\nu_0, \nu_1) \, |\,  \nu)  + (1-\mu)\, p_b( s(x_e;\nu_0, \nu_1) \,|\, \nu) \,\right] \; ,
+%%\end{equation}
+%%which is a natural generalization of Eq.~\ref{eq:typicalML}.
+%
+%%The construction in Eq.~\ref{eq:parametrizedML} presents a computational challenge as for each value of $\nu$ one has a new mapping $s: x\to\mathbb{R}$.
 The most naive realization of this would be to generate synthetic data $\{x_e\}$ according to $p_c(x|\nu)$, evaluate $s(x_e ; \nu_0, \nu_1)$, and then use a histogram, kernel density, or other non-parametric density estimation procedure to estimate $p_c( s | \nu)$.
@@
 This formalism represents is a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
-The work of  ~\citep{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
+The work of  ~\cite{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
 This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
 
-Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citep{Volobouev:2011vb}.
+Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\cite{Volobouev:2011vb}.
 In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method.
@@
 It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage.
-In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citep{Aaltonen:2010yz}).
+In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\cite{Aaltonen:2010yz}).
 
-Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citep{Allanach:2000kt}.
+Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\cite{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\cite{Allanach:2000kt}.
 Both of these examples involve datasets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
@@
 %Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
-The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citep{Bolognesi:2012mm}.
+The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\cite{Bolognesi:2012mm}.
 
@@
 
-%We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate statistical inference likelihood based on the  ratio $p(x|\theta_0)/p(x|\theta_1)$ when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$.
+%\section*{Acknowldgements}
+%KC would like to thank  
+%Daniel Whiteson for discussions and encouragement throughout the project, 
+%Alex Ihler for challenging discussions that 
+%led to a reformulation of the initial idea, and Shimon Whiteson for patient feedback in that process. 
+%KC would also like to thank Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
+%project early in its conception, Bal\'azs K\'egl for discussions about the Kaggle challenge and feedback to the draft, 
+%and Yuri Shirman for reassuring cross checks of the Theorem.
+%KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 
+%KC is grateful to UC-Irvine for their hospitality while this research was carried out and the 
+%Moore and Sloan foundations for their generous support of the data science environment at NYU.
 
-%\section{Possible Extensions}
-%The approach describe thus far has been agnostic as to the type of classifier used to produce $s(x)$ and $s(x; \nu)$.
-In practice, we have used standard packages for the training of these learners.
-In particular, the loss function for these learning algorithms is based on classification performance of individual $(x_e, c_e)$ feature $\to$ target examples.
-However, in overarching statistical procedure that uses the statistical model in Eq.~\ref{eq:typicalML} or \ref{eq:parametrizedML} the quantity we wish to optimize is a property of an entire dataset $\{x_e\}$ or ensembles of datasets that can be generated from the statistical models $p_c(x|\theta)$.
-%
-%For example, in typical hypothesis testing problems posed in the Neyman-Pearson context one would be interested in minimizing type II error under a fixed rate of typeI error rate.
-In the presence of systematic uncertainty and nuisance parameters, the loss function may become more complicated.
-The recent Higgs boson machine learning challenge hosted by Kaggle, the evaluation was the approximate median significance (AMS) that characterized the power of a hypothesis test between $\mu=0$ (the background only hypothesis) and $\mu=\mu_0$ (the signal+background hypothesis, where $\mu_0$ is a small number reflecting the small Higgs signal that would be found in addition to the much larger background process) in the presence of uncertainty on the background normalization.
-%Interestingly, the top entries to the challenge did not improve by focusing on this alternative loss function 
-%in training their classifier, but instead used it for the optimization of a threshold on the score (a working point on the ROC curve) that was related to the way the challenge was designed.
-This suggests that the standard
-%classification loss functions lead to similar optimizations and that refinements for the custom loss functions
-%are not easy to come by.
-%
-%More generally, there is an avenue of research associated to custom loss functions specifically designed 
-%for these real world statistical problems. One direction might be to choose loss functions that make 
-%some balance between the standard classification loss, take into account robustness with respect to nuisance parameters, and population-level quantities like type I and II error.
-%
-%\subsection{A different starting point}
-%
-%Instead of evolving from the traditional usage of ML in HEP, let's start from scratch with $p_c(x|\theta)$ and a well defined objective.  
-%
-%\subsubsection{Simple Hypothesis Testing}
-%
-%Consider the situation of simple hypothesis testing between the background-only $\mu=0$ and signal+background hypothesis $\mu=\mu_0=\nu_s/(\nu_s+\nu_b)$ for $\theta=\theta_0$.
-Let us assume that not only can we use the $p_c(x|\theta_0)$ as a generative model, but that we can evaluate the probability density at any point $x$.
-The statistical model is given by
-%\begin{equation}\label{eq:NP}
-%p( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu p_1( x_e \, |\,  \theta)  + (1-\mu)\, p_b( x_e \,|\, \theta) \,\right] \; .
-%\end{equation}
-%The objective is to find a  test statistic $T(D)$ (a map from the space of the data to $\mathbb{R}$)  
-%and a threshold $k_\alpha$ that minimizes the rate of type-II error (i.e.
-$P(T(D) < k_\alpha | \mu=\mu_0, \theta_0) = \beta$ under the constraint of a fixed rate of type-I error ( i.e.
-$P(T(D) > k_\alpha | \mu=0, \theta_0) = \alpha$ ).
-%The Neyman-Pearson lemma states that the optimal test statistic is the likelihood ratio, or equivalently, the log-likelihood ratio 
-%\begin{equation}
-%T =  \log \frac{p(D | \mu=\mu_0, \theta_0)}{p(D | \mu=0, \theta_0)}
-%\end{equation}
-%When the $x_e$ are i.i.d., then the log-likelihood ratio expands into a sum
-%\begin{equation}
-%T =   \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {p_1(x_e | \theta_0) }{ p_b(x_e | \theta_0) } \right]}_{q(x_e)} + c_2\;,
-%\end{equation}
-%where $c_1=\mu/(1-\mu)$ and $c_2=\lop(1-\mu)$ are constants for the simple hypothesis test.
-%
-%Here we see that the optimal $T$ for the experiment is composed of a sum over events of a linear linear function of the per-event function $q(x)$.
-A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.
-%
-%The important part of the per-event function $q(x)$ is that it defines contours in the feature space $x$.
-These contours are also equivalent to the function $p_1(x|\theta_0)/p_b(x|\theta_0)$, the signal-to-background ratio.
-If we can find a machine learner $s(x)$ that is a monotonic function of $p_1(x|\theta_0)/p_b(x|\theta_0)$ it will share the same contours.
-Then the remaining challenge is to find the appropriate rescaling that gives back  linear function $q(x)$.
-%
-%\textbf{Postulate:}
-%\begin{equation}
-%T' = \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {p_1(s(x_e) | \theta_0) }{ p_b(s(x_e) | \theta_0) } \right]}_{q(s_e)} \;,
-%\end{equation}
-%leads to an equivalent test, where $s_e \equiv s(x_e)$
-%\begin{equation}
-%p_c(s) = \int dx \, \delta(s-s(x)) \, p_c(x)  \,  |\partial s / \partial x|^{-1}
-%\end{equation}
-%
-%
-%
-%Note, in order to form the hypothesis test, one still needs to build the build the $p_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$. 
-%
-%
-%
-%\textbf{Proof (in progress):}
-%
-%We want to show density is the same 
-%
-%\begin{equation}
-%p_c(q_x) = \int dx \delta(q_x-q_x(x)) p_c(x) / |\partial q_x / \partial x|
-%\end{equation}
-%
-%\begin{equation}
-%p_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, p_c(x) \, / |\partial q_s / \partial x|
-%\end{equation}
-%sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $p_1(x)/p_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
-%\begin{equation}
-%\frac {p_1(c | \theta_0) }{ p_0(c | \theta_0) } = 
-%\frac { \int dx \delta(c-s(x)) \, p_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  } = 
-%\frac {c \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, p_0(x | \theta_0)  } = c =\frac {p_1(x | \theta_0) }{ p_0(x | \theta_0) } 
-%\end{equation}
-%
-%
-%
-%%\begin{equation}
-%%p_c(q_s) = \int ds \delta(q_s-q_s(s)) \,p_c(s) \, / |\partial q_s / \partial s|
-%%\end{equation}
-%
-%
-%This is the test statistic that one would naturally get from constructing the likelihood function for the 1-d distributions of the output of the machine learner
-%\begin{equation}\label{eq:NP}
-%p( \{s_e\} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu p_1( s_e \, |\,  \theta)  + (1-\mu)\, p_0( s_e \,|\, \theta) \,\right] \; .
-%\end{equation}
-%Thus this motivates Eq.~\ref{eq:typicalML} in the more general case. 
-%
-%
-%Question:
-%Where does this fit in:
-%\begin{equation}
-%T'' = \sum_{e=1}^n \underbrace{ \log \left[ 1+c_1 s(x_e) \right] }_{q'(s_e)} \;,
-%\end{equation}
-%
-%
-%
-%Usefulness:
-%If we can readily compute $p_c(x|\theta_0)$, then there would be no need for machine learning algorithms.
-However,  these densities are often difficult to compute though one can readily use them as a generative model to produce samples.
-In that case, one might wish to use a machine learning algorithm to find $s(x) \approx p_1(x|\theta_0) / p_0(x|\theta_0)$.
-%
-%Conclusion, while it is possible to start directly with the Neyman-Pearson loss function of type-II error under the constraint of fixed size and try to find a per-experiment learner $T(D)$, it is more convenient to cast the problem in terms of learning a per-event function $s(x)$, and then calibrate it to be a linear function of $q(x)$.
-%
-%\newpage
-%
-%Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $p_1(x)$ and $p_0(x)$ define 
-%\begin{equation} 
-%s(x)=\frac{p_1(x)}{p_2(x)}
-%\end{equation}.
-%Define $p_1(c) = \int dx \delta(c - s(x) ) \, p_1(x)$ and $p_0(c) = \int dx \delta(c - s(x) ) \, p_0(x)$.
-%I wish to show
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)}
-%\end{equation}
-%$\forall x \ni s(x)=c$.
-%
-%Sketch of Proof: 
-%let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$.
-The  coordinate perpendicular to $\Omega$ can be written $y = x+\epsilon \hat{n}$.
-In these coordinates, the integral $\int dx \to \int d\Omega d\epsilon$, thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon|
-%\end{equation}
-%Similarly for $p_0(c)$. We can factor out of the integral $s(x)=p_1(x)/p_0(x)$ since it is constant over $\Omega$.
-%Thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|
-%\end{equation}
-%and we arrive at the result
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{p_1(x)}{p_0(x)} \;.
-%\end{equation}
-%
-%
-%
-%
-%\subsection{misc notes}
-%
-%
-%
-%
-%The distribution of $s$ is fully determined by the mapping $s(x_e; \theta_0, \theta_1)$ and the distribution $p(x)$, which we can write
-%\begin{equation}
-%p(s_e) = \int dx \, \delta(s_e-s(x)) \, p(x)  \; .
-%\end{equation}
-%\textbf{NEED Jacobian for x $\to$ s independent of delta function part, or double counting?}
-%
-%
-%\textbf{Theorem 1:}
-%Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to the surface at the point $x$.
-The component perpendicular to $\Omega_c$ can be written $x+\epsilon \hat{n}$.
-In these coordinates, we rewrite integral $\int dx \to \int d\Omega_c d\epsilon$, thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega_c p_1(x)  / | \hat{n} \cdot \nabla q_s  |
-%\end{equation}
-%Similarly for $p_0(c)$. We can factor out of the integral $s(x)=p_1(x)/p_0(x)$ since it is constant over $\Omega_c$.
-%Thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega p_0(x)  / | \hat{n} \cdot \nabla q_s  |
-%\end{equation}
-%and we arrive at the result
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{p_1(x)}{p_0(x)} \;\hspace{3em} \forall x\in\Omega_c.
-%\end{equation}
-%
-%
-%
-%
-%
-%Define $p_1(c) = \int dx \delta(c - s(x) ) \, p_1(x)$ and $p_0(c) = \int dx \delta(c - s(x) ) \, p_0(x)$.
-%I wish to show
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)}
-%\end{equation}
-%$\forall x \ni s(x)=c$.
-%
-%
-%Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $p_1(x)$ and $p_0(x)$ define 
-%\begin{equation} 
-%s(x)=\frac{p_1(x)}{p_0(x)}
-%\end{equation}.
-%Define $p_1(c) = \int dx \delta(c - s(x) ) \, p_1(x)$ and $p_0(c) = \int dx \delta(c - s(x) ) \, p_0(x)$.
-%I wish to show
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)}
-%\end{equation}
-%$\forall x \ni s(x)=c$.
-%
-%Sketch of Proof: 
-%let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$.
-The  coordinate perpendicular to $\Omega$ can be written $y = x+\epsilon \hat{n}$.
-In these coordinates, the integral $\int dx \to \int d\Omega d\epsilon$, thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon|
-%\end{equation}
-%Similarly for $p_0(c)$. We can factor out of the integral $s(x)=p_1(x)/p_0(x)$ since it is constant over $\Omega$.
-%Thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|
-%\end{equation}
-%and we arrive at the result
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{p_1(x)}{p_0(x)} \;.
-%\end{equation}
-%
-%
-%
-%
-%
-%
-%
-%Note, in order to form the hypothesis test, one still needs to build the build the $p_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$. 
-%
-%
-%
-%\textbf{Proof (in progress):}
-%
-%We want to show density is the same 
-%
-%\begin{equation}
-%p(q_x|\theta) = \int dx \delta(q_x-q_x(x)) p(x|\theta)  / | \hat{n} \cdot \nabla q_s  |
-%\end{equation}
-%
-%\begin{equation}
-%p(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, p(x|\theta) \, / | \hat{n} \cdot \nabla q_s  |
-%\end{equation}
-%sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $p_1(x)/p_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
-%\begin{equation}
-%\frac {p_1(c | \theta_0) }{ p_0(c | \theta_0) } = 
-%\frac { \int dx \delta(c-s(x)) \, p_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  } = 
-%\frac {c \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, p_0(x | \theta_0)  } = c =\frac {p_1(x | \theta_0) }{ p_0(x | \theta_0) } 
-%\end{equation}
-%
-%
-%
-%%\begin{equation}
-%%p_c(q_s) = \int ds \delta(q_s-q_s(s)) \,p_c(s) \, / |\partial q_s / \partial s|
-%%\end{equation}
-%
-%
+\newpage
 
-\section*{Acknowldgements}
-KC would like to thank  
-Daniel Whiteson for discussions and encouragement throughout the project, 
-Alex Ihler for challenging discussions that 
-led to a reformulation of the initial idea, and Shimon Whiteson for patient feedback in that process. 
-KC would also like to thank Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
-project early in its conception, Bal\'azs K\'egl for discussions about the Kaggle challenge and feedback to the draft, 
-and Yuri Shirman for reassuring cross checks of the Theorem.
-KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 
-KC is grateful to UC-Irvine for their hospitality while this research was carried out and the 
-Moore and Sloan foundations for their generous support of the data science environment at NYU.
+\subsubsection*{References}
 
-\bibliographystyle{natbib}
+\vspace*{-2em}
+
+\small{
+\bibliographystyle{unsrt}
 \bibliography{learning.bib}
-%\begin{thebibliography}
-%\parametrized-learning}
-%\end{thebibliography
+
+}
 

### 2015-06-02 tex/parametrized-learning-aoas.tex
AoAS draft

--- 2015-05-14 tex/parametrized-learning.tex
@@
-\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
+\documentclass[aoas,preprint]{imsart}
 
-\usepackage{jmlr2e}
-\usepackage{lineno}
-\linenumbers
-%\jmlrheading{1}{2014}{X-XX}{4/00}{10/00}{Kyle Cranmer}
+\RequirePackage[OT1]{fontenc}
+\RequirePackage{amsthm,amsmath}
+\RequirePackage{natbib}
+%\RequirePackage[nameyear]{natbib}
+\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
 
-%\ShortHeadings{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}{Kyle Cranmer}
-
-\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
-%\geometry{landscape}                		% Activate for for rotated page geometry
-%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
-								% TeX will automatically convert eps --> pdf in pdflatex		
 \usepackage{amssymb, amsmath}
@@
 
-%\title{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}
-\title{Likelihood ratio tests constructed with discriminative classifiers and calibrated with generative models}
-\title{Approximating generalized likelihood ratio tests with calibrated discriminative classifiers}
-%\title{Calibrated, Parametrized Learning}
-\author{Kyle Cranmer}
-%\date{}							% Activate to display a given date or no date
+% settings
+%\pubyear{2005}
+%\volume{0}
+%\issue{0}
+%\firstpage{1}
+%\lastpage{8}
+\arxiv{arXiv:0000.0000}
+
+\startlocaldefs
+\numberwithin{equation}{section}
+\theoremstyle{plain}
+\newtheorem{thm}{Theorem}[section]
+\endlocaldefs
 
@@
 
-\author{\name Kyle Cranmer \email kyle.cranmer@nyu.edu \\
-       \addr Center for Cosmology and Particle Physics\\
-       Center for Data Science\\
+\begin{frontmatter}
+%\title{Approximating Generalized Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
+\title{Approximating  Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
+%\title{Approximating Likelihood Ratio Tests with Calibrated Classifiers}
+%\title{A Sample Document\thanksref{T1}}
+\runtitle{Approximating the Likelihood Ratio with Classifiers}
+%\thankstext{T1}{Footnote to the title with the ``thankstext'' command.}
+
+\begin{aug}
+\author{\fnms{Kyle} \snm{ Cranmer}\thanksref{t1}\ead[label=e1]{kyle.cranmer@nyu.edu}}
+%,
+%\author{\fnms{Second} \snm{Author}\thanksref{t3,m1,m2}\ead[label=e2]{second@somewhere.com}}
+%\and
+%\author{\fnms{Third} \snm{Author}\thanksref{t1,m2}
+%\ead[label=e3]{third@somewhere.com}
+%\ead[label=u1,url]{http://www.foo.com}}
+
+\thankstext{t1}{US National Science Foundation grants PHY-0854724 and PHY-0955626.}
+%\thankstext{t2}{First supporter of the project}
+%\thankstext{t3}{Second supporter of the project}
+\runauthor{K. Cranmer}
+
+\affiliation{Center for Cosmology and Particle Physics\thanksmark{m1} 
+       Center for Data Science\thanksmark{m2}
+       New York University}
+
+\address{Center for Cosmology and Particle Physics\\
        New York University \\
        New York, NY 10003, USA
-}
-
-%\editor{Leslie Pack Kaelbling}
-\editor{}
-
-\maketitle
+\phantom{E-mail:\ }\printead*{e1}}
+\end{aug}
 
 \begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-We demonstrate how discriminative classifiers can be used to approximate 
-generalized likelihood ratio tests over high-dimensional data when a generative model 
-for the data is available for training and calibration.  
+In many areas of science, generalized likelihood ratio tests  are established tools for statistical inference. 
+It is increasingly common that a generative model is used to describe complex processes that tie 
+parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
+Directly evaluating the likelihood ratio in these cases is often not possible or is computationally impractical.
+Here we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when
+a generative model for the data is available for training and calibration.
+
+%We demonstrate how discriminative classifiers can be used to approximate 
+%generalized likelihood ratio tests over high-dimensional data when a generative model 
+%for the data is available for training and calibration.  
 \end{abstract}
-%
-%\begin{keywords}
-%  Bayesian Networks, Mixture Models, Chow-Liu Trees
-%\end{keywords}
 
-\maketitle
+\begin{keyword}[class=MSC]
+\kwd[Primary ]{62P35}
+\kwd{62F99}
+\kwd[; secondary ]{62H30}
+\end{keyword}
+
+\begin{keyword}
+\kwd{likelihood ratio}
+\kwd{classification}
+\end{keyword}
+
+\end{frontmatter}
 
@@
 
-\bibliographystyle{natbib}
+%\bibliographystyle{natbib}
+\bibliographystyle{imsart-nameyear}
 \bibliography{learning.bib}

### 2015-06-02 tex/parametrized-learning-aoas.tex
adding AoAS draft

@@
 
-\affiliation{Center for Cosmology and Particle Physics\thanksmark{m1} 
-       Center for Data Science\thanksmark{m2}
+\affiliation{Center for Cosmology \and Particle Physics\thanksmark{m1} \\
+       \and Center for Data Science\thanksmark{m2}
        New York University}
@@
 \kwd{classification}
+\kwd{particle physics}
 \end{keyword}
@@
 \begin{center}
- \includegraphics[height=2in]{example-TMVA-BDT.pdf}
- \includegraphics[height=2in]{example-TMVA-ROC.pdf}
+ \includegraphics[height=1.7in]{example-TMVA-BDT.pdf}
+ \includegraphics[height=1.7in]{example-TMVA-ROC.pdf}
 \caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).

### 2015-06-02 tex/parametrized-learning.tex
adding AoAS draft

@@
 \begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-We demonstrate how discriminative classifiers can be used to approximate 
-generalized likelihood ratio tests over high-dimensional data when a generative model 
-for the data is available for training and calibration.  
+In many areas of science, generalized likelihood ratio tests  are established tools for statistical inference. 
+It is increasingly common that a generative model is used to describe complex processes that tie 
+parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
+Directly evaluating the likelihood ratio in these cases is often not possible or is computationally impractical.
+Here we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when
+a generative model for the data is available for training and calibration.
+
+%We demonstrate how discriminative classifiers can be used to approximate 
+%generalized likelihood ratio tests over high-dimensional data when a generative model 
+%for the data is available for training and calibration.  
 \end{abstract}
@@
 
-In many areas of science, (generalized) likelihood ratio tests  are established tools for statistical inference. 
-Directly constructing the likelihood ratio for high-dimensional observations 
-is often not possible or is computationally impractical. Here we demonstrate how 
-discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
+In many areas of science, (generalized) likelihood ratio tests  are established tools for statistical inference~\citep{Wilks1938}. 
+It is increasingly common that a generative model is used to describe complex processes that tie 
+parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
+Directly evaluating the likelihood ratio in these cases is often not possible or is computationally impractical.
+Here we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when
 a generative model for the data is available for training and calibration.
@@
 %Collisions are usually selected with supervised learning methods 
-While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citep{Dempster1965}, that is no longer true in the context of parameter estimation or compound hypothesis tests with nuisance parameters.
+While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citep{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters.
 As noted in ~\citep{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''

### 2015-06-03 tex/parametrized-learning-aoas.tex
updating related work for Neal

@@
 \DeclareMathOperator*{\argmax}{arg\,max}
+\newcommand{\citek}[1]{[\cite{#1}]}
 
@@
 %\title{Approximating Generalized Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
-\title{Approximating  Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
+%\title{Approximating  Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
+\title{Approximating  Likelihood Ratios with Calibrated Discriminative Classifiers}
 %\title{Approximating Likelihood Ratio Tests with Calibrated Classifiers}
 %\title{A Sample Document\thanksref{T1}}
-\runtitle{Approximating the Likelihood Ratio with Classifiers}
+\runtitle{Approximating Likelihood Ratios with Classifiers}
 %\thankstext{T1}{Footnote to the title with the ``thankstext'' command.}
@@
 \begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-In many areas of science, generalized likelihood ratio tests  are established tools for statistical inference. 
-It is increasingly common that a generative model is used to describe complex processes that tie 
-parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
-Directly evaluating the likelihood ratio in these cases is often not possible or is computationally impractical.
-Here we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when
+In particle physics likelihood ratio tests are established tools for statistical inference. 
+These tests are complicated by the fact that computer simulators are used as a generative model for 
+the data, but they do not provide a way to evaluate the likelihood function for a given
+observation. 
+We demonstrate how discriminative classifiers can be used to approximate the likelihood function when 
 a generative model for the data is available for training and calibration.  
-
-%We demonstrate how discriminative classifiers can be used to approximate 
-%generalized likelihood ratio tests over high-dimensional data when a generative model 
-%for the data is available for training and calibration.  
 \end{abstract}
@@
 
-In many areas of science, (generalized) likelihood ratio tests  are established tools for statistical inference~\cite{Wilks1938}. 
-It is increasingly common that a generative model is used to describe complex processes that tie 
+The likelihood function is the central object that summarizes the information from an experiment needed for inference of model parameters.
+The likelihood function is key to Bayesian inference, and many areas of science that report the results of classical hypothesis tests or confidence intervals use the (generalized or profile) likelihood ratio as a test statistic.
+%  are established tools for statistical inference~\citek{Wilks1938}. 
+It is increasingly common that a simulator (or generative model) is used to describe complex processes that tie 
 parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
@@
 final statistical result was formulated as a hypothesis test or confidence interval 
-using a generalized likelihood ratio test~\cite{Cowan:2010js}. This includes
-the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa} and 
+using a generalized  likelihood ratio test~\citek{Cowan:2010js}. This includes
+the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa} and 
 subsequent measurement of its properties.
@@
 To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers.
-Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\cite{Hocker:2007ht} are commonly used.
-Recently, there has been progress in using deep networks~\cite{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\cite{HepML}, the largest Kaggle challenge in history.
+Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citek{Hocker:2007ht} are commonly used.
+Recently, there has been progress in using deep networks~\citek{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citek{HepML}, the largest Kaggle challenge in history.
 %The use of classifiers is effective, but an intermediate step not obviously connected to the target likelihood ratio test.
@@
 %Collisions are usually selected with supervised learning methods 
-While classification accuracy can lead to optimal approaches for simple hypothesis tests~\cite{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters.
-As noted in ~\cite{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
-A fundamental distinction is that evaluating the loss for classification is a per-event operation, while evaluating the loss for a mass measurement (and other estimation problems) is a per-experiment operation involving a dataset with many events.
+While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citek{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters.
+As noted in ~\citek{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
+%In what follows, it will be important to distinguish between the per-event loss used for classification, and the per-experiment loss 
+A fundamental distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g.
+the variance of an estimator for the mass parameter) is a per-experiment operation involving a data set with many events.
 They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process.
@@
 
-Generalized likelihood ratio tests enjoy a number of optimal properties in the asymptotic limit and are widely used even away from that limit.
+In the context of machine learning or decision theory one starts with a well defined loss function.
+In order to define a Bayes risk one must also specify a prior over the parameters.
+%In practice, it can be difficult, particularly for large collaborations, to agree on such high-level topics.
+In practice, large particle physics collaborations have preferred to avoid specifying the 
+high-level loss function and prior, and instead accepted the properties of generalized likelihood ratio tests.
+Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself.
+
+The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss. 
+This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process.
+Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.
+In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.
+
+%\bigskip
+%
+%Generalized likelihood ratio tests enjoy a number of optimal properties in the asymptotic limit and are widely used even away from that limit.
 These asymptotic properties, the difficulty in specifying a prior over the parameters needed to evaluate a Bayes risk, and the vast literature associated to likelihood ratio tests has led many users of such tests to not be explicit about the loss function they wish to minimize for their experiment.
@@
 The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss.
-This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process.
+
+%Similarly,~\cite{Neal:2007zz} suggested using a classifier as a dimensionality reduction map to aid in the estimation of the likelihood function.
+We discuss the key differences with that work in Section~\ref{S:Related}.
+
+%This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process.
 Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.
@@
  \item $x$: a vector of features for an event
- \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
+ \item $D$: a data set of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
  \item $\theta$: parameters of a statistical model
@@
 %\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
+\item $y$: a class label used for training a classifier.
 \item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
 %\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
-\item $p( s_{\theta_0, \theta_1} | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$ \end{itemize}
+\item $p( s_{\theta_0, \theta_1} | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$ 
+\end{itemize}
 We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.
@@
 can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case 
-high energy physics where the simulation of particle detectors can only be done in the `forward mode'. This same setting has been considered by \cite{ClaytonScott} and \cite{JMLR:v14:tong13a}.
+high energy physics where the simulation of particle detectors can only be done in the `forward mode'.
+This same setting has been considered by \cite{ClaytonScott}, \cite{JMLR:v14:tong13a}, and \cite{Neal:2007zz}.
 
@@
 
-While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, we are able to reformulate it such that the machine learning problem is a per-event classification problem.
+While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire data set $D$, we are able to reformulate it such that the machine learning problem is a per-event classification problem.
 This follows from the fact that we assume the $x_e$ to be i.i.d.
@@
 
-Vast literature exists around generative and discriminative classifiers~\cite{AndrewY.Ng}.
+Vast literature exists around generative and discriminative classifiers~\citek{AndrewY.Ng}.
 Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule.
@@
 The first possible source of confusion we wish to avoid is that here $p(x|\theta)$  is a generative \textit{statistical model} for the features $x$, not a generative classifier.
-We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory, able to make predictions about $x$ and being motivated by domain-specific considerations.
-For example, in the context of HEP $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$.
+We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory or simulator, able to make predictions about $x$ and being motivated by domain-specific considerations.
+For example, in the context of HEP $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.
 Moreover, we are not attempting to learn the generative model $p(x|\theta)$, we are taking it as given and trying to learn the corresponding (generalized) likelihood ratio test.
@@
 %
-%%the likelihood ratio $T(D)$ is aimed at tests based on the entire dataset; 
+%%the likelihood ratio $T(D)$ is aimed at tests based on the entire data set; 
 %we are not interested in thresholded classification on individual events $x_e$.  
@@
 Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
-Similar points have been made by~\cite{ClaytonScott}.
+Similar points have been made by~\cite{ClaytonScott} and \cite{Neal:2007zz}.
 
@@
 This allows us to use Theorem~1 to reformulate the maximum likelihood estimate
-\begin{equation}
-\hat{\theta} = \argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} = \argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
+\begin{equation}\label{eq:mle_withs}
+\hat{\theta} = %\argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} = 
+\argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
 \end{equation}
@@
 \begin{algorithmic}
-\STATE initialize trainingData = \{\}
+\STATE initialize trainingData to \{\}
 \FOR{ $\theta_0$ in $\Theta$ }
@@
 In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{g}\circ \hat{s}$.
-A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\cite{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
+A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
@@
 New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
-generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\cite{Cowan:2010js}.
-This approach was used for the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa}.
+generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\citek{Cowan:2010js}.
+This approach was used for the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa}.
 
-%Generalized likelihood ratio tests are used widely in HEP ~\cite{Cowan:2010js}, most notably in the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa}.
+%Generalized likelihood ratio tests are used widely in HEP ~\citek{Cowan:2010js}, most notably in the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa}.
 
@@
 The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.
-The third line uses Theorem 1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
+The third line uses Theorem~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
 In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).
@@
 This formalism represents is a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
-The work of  ~\cite{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
+The work of  ~\citek{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
 This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
 
-Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\cite{Volobouev:2011vb}.
+Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citek{Volobouev:2011vb}.
 In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method.
@@
 It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage.
-In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\cite{Aaltonen:2010yz}).
+In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citek{Aaltonen:2010yz}).
 
-Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\cite{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\cite{Allanach:2000kt}.
-Both of these examples involve datasets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
+Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citek{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citek{Allanach:2000kt}.
+Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
 %The state of the art approaches for supersymmetric particle masses typically involve high-level features engineered based on physics intuition
 %Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
-The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\cite{Bolognesi:2012mm}.
+The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citek{Bolognesi:2012mm}.
 
-\subsection{Related work}
+\subsection{Related work}\label{S:Related}
 
 %Refs to include:
-%\cite{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,}
+%\citek{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,}
 
@@
 
+\cite{Neal:2007zz} also considered the problem of approximating the likelihood function when only a generative model is available.
+That work sketches a scheme in which one uses a classifier with both $x$ and $\theta$ as an input to serve as a dimensionality reduction map.
+The key distinction comes in the handling of $\theta$.
+Neal says  ``we cannot use the classifier on real data, since we don't know the correct value for [$\theta]$'' and goes on to outline an approach where one uses regression on a per-event basis to estimate $\hat{\theta}(x)$ and perform the composition $s(x; \hat{\theta}(x))$ (much like profiling).\footnote{Neal considers a lower-dimensional `bottlenecks' $\theta^*=g(\theta)$, which are not essential to the discussion here.}  This can lead to a significant loss of information since (at least in most particle physics examples) a single event carries little information about the true value of $\theta$, though the full data set $D$ may be informative -- for instance, a single observation would not be sufficient to estimate the variance of a distribution, though repeated observations would.
+The work of Neal correctly identifies this as an approximate procedure even in the case of a ideal classifier.
+In contrast, this does not eliminate the dependence of the classifier on $\theta$.%
+\footnote{As a technical point, in Neal's work, the focus is on approximating the likelihood function (up to a multiplicative constant), which is equivalent to evaluating the ratio  with respect to a fixed $\theta_1$ as in Eq.~\ref{eq:mle_withs}.
+In Neal's case, the dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is constant; however, in this approach the map ratio is explicitly parametrized in terms of $\theta$, so the ratio is important for canceling the corresponding Jacobian factors.} Instead, we embed a parametrized classifier into the likelihood and postpone the evaluation of the classifier to the point of evaluation of the likelihood when $\theta$ is known.
+This avoid the loss of information that occurs from the regression step $\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact result in the case of an ideal classifier.
+
 \section{Conclusions}
 
-We have outlined a technique to reformulate generalized likelihood ratio test over a high-dimensional dataset with multiple events in terms of a univariate density of a classifier score. 
+We have outlined a technique to reformulate generalized likelihood ratio test over a high-dimensional data set with multiple events in terms of a univariate density of a classifier score. 
 We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$.
@@
 In particular, the loss function for these learning algorithms is based on classification performance of individual $(x_e, c_e)$ feature $\to$ target examples.
-However, in overarching statistical procedure that uses the statistical model in Eq.~\ref{eq:typicalML} or \ref{eq:parametrizedML} the quantity we wish to optimize is a property of an entire dataset $\{x_e\}$ or ensembles of datasets that can be generated from the statistical models $p_c(x|\theta)$.
+However, in overarching statistical procedure that uses the statistical model in Eq.~\ref{eq:typicalML} or \ref{eq:parametrizedML} the quantity we wish to optimize is a property of an entire data set $\{x_e\}$ or ensembles of data sets that can be generated from the statistical models $p_c(x|\theta)$.
 %
@@
 KC would like to thank  
-Daniel Whiteson for discussions and encouragement throughout the project, 
+Daniel Whiteson for discussions and encouragement throughout the project and
 Alex Ihler for challenging discussions that 
-led to a reformulation of the initial idea, and Shimon Whiteson for patient feedback in that process. 
-KC would also like to thank Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
+led to a reformulation of the initial idea.
+KC would also like to thank Shimon Whiteson, Babak Shahbaba for advice in the presentation, Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
 project early in its conception, Bal\'azs K\'egl for discussions about the Kaggle challenge and feedback to the draft, 

### 2015-06-03 tex/parametrized-learning-aoas.tex
text about embedding s in p

@@
 In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$.
-In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{g}\circ \hat{s}$.
+In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{g}\circ \hat{s}$.
 A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.

### 2015-06-03 tex/parametrized-learning-aoas.tex
left over f, g -> p

--- 2015-06-03 tex/parametrized-learning-aoas.tex
@@
 
-Denote the approximate dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{g}(\hat{s}|\theta)$.
+Denote the approximate dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{p}(\hat{s}|\theta)$.
 In general we will be interested in the machine learning problem that approximates these distributions based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
@@
 Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
-histograms or kernel density estimation) to approximate $\hat{g}(\hat{s}|\theta)$.
+histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$.
 
@@
 %Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g.
-histograms or kernel density estimation) to approximate $\hat{g}(\hat{s}|\theta)$.
+histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$.
 
 In the  limit of large samples from the generative model,  we can approximate the original likelihood ratio test arbitrarily well.
-With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{g}(\hat{s}|\theta)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test.
+With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{p}(\hat{s}|\theta)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test.
 This will depend in general on the ultimate goal of the test.
@@
 
-In most settings that make use of likelihood ratio tests, the likelihood is based directly on some approximation of density for the observed data via $\hat{f}(x|\theta)$.
-In general, approximating the density $\hat{f}(x|\theta)$ is difficult for high-dimensional data, which motivates the use of the dimensionality reduction map $\hat{s}(x)$ and likelihood ratio tests based on the univariate density $\hat{g}(\hat{s}|\theta)$.
+In most settings that make use of likelihood ratio tests, the likelihood is based directly on some approximation of density for the observed data via $\hat{p}(x|\theta)$.
+In general, approximating the density $\hat{p}(x|\theta)$ is difficult for high-dimensional data, which motivates the use of the dimensionality reduction map $\hat{s}(x)$ and likelihood ratio tests based on the univariate density $\hat{p}(\hat{s}|\theta)$.
 In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$.
-In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{g}\circ \hat{s}$.
+In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{p}\circ \hat{s}$.
 A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.

### 2015-06-03 tex/parametrized-learning-aoas.tex
typo

@@
 
-These steps feed into a subsequent statistical test based on the observed data \\
+These steps feed into a subsequent statistical test based on the observed data 
 ${D=(x_1, \dots, x_n)}$.
 For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution.
-In particular, one forms the statistical model\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood.}
+In particular, one forms the statistical model\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood or marked Poisson model.}
 \begin{equation}\label{eq:typicalML}
@@
 The work of Neal correctly identifies this as an approximate procedure even in the case of a ideal classifier.
-In contrast, this does not eliminate the dependence of the classifier on $\theta$.%
+In contrast, the approach described here does not eliminate the dependence of the classifier on $\theta$.%
 \footnote{As a technical point, in Neal's work, the focus is on approximating the likelihood function (up to a multiplicative constant), which is equivalent to evaluating the ratio  with respect to a fixed $\theta_1$ as in Eq.~\ref{eq:mle_withs}.

### 2015-06-03 tex/parametrized-learning-aoas.tex
clarify text

@@
 \footnote{As a technical point, in Neal's work, the focus is on approximating the likelihood function (up to a multiplicative constant), which is equivalent to evaluating the ratio  with respect to a fixed $\theta_1$ as in Eq.~\ref{eq:mle_withs}.
-In Neal's case, the dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is constant; however, in this approach the map ratio is explicitly parametrized in terms of $\theta$, so the ratio is important for canceling the corresponding Jacobian factors.} Instead, we embed a parametrized classifier into the likelihood and postpone the evaluation of the classifier to the point of evaluation of the likelihood when $\theta$ is known.
-This avoid the loss of information that occurs from the regression step $\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact result in the case of an ideal classifier.
+In Neal's case, the dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is constant; however, in this approach the map ratio is explicitly parametrized in terms of $\theta$, so the ratio is important for canceling the corresponding Jacobian factors.} Instead, we embed a parametrized classifier into the likelihood and postpone the evaluation of the classifier to the point of evaluation of the likelihood when $\theta$ is explicitly being tested.
+This avoids the loss of information that occurs from the regression step $\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact result in the case of an ideal classifier.
 

### 2015-06-05 tex/parametrized-learning-aoas.tex
re-writing intro for ABC

@@
 a generative model for the data is available for training and calibration.
+This offers an approach to parametric inference when simulators are used that is complementary to approximate Bayesian computation.
 \end{abstract}
@@
 parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
-Directly evaluating the likelihood ratio in these cases is often not possible or is computationally impractical.
-Here we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when
-a generative model for the data is available for training and calibration.
+Directly evaluating the likelihood ratio in these cases is often impossible or is computationally impractical. 
+Approximate Bayesian Computation (ABC) is one approach to parameter inference in this simulation-based or likelihood-free setting~\citek{Rubin1984,Tavare1997,Marin2011}.
+Here we consider an alternative approach that can also be used in a classical setting where a prior over the parameters is not available.
+In particular, we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when a generative model for the data is available for training and calibration.
 
 As a concrete example, consider searches for new particles at the Large Hadron Collider. 
-The ATLAS and CMS experiments have published several hundred papers where the 
-final statistical result was formulated as a hypothesis test or confidence interval 
+The simulator that is sampling from $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.
+
+The ATLAS and CMS experiments have published  hundreds papers where the 
+final  result was formulated as a hypothesis test or confidence interval 
 using a generalized  likelihood ratio test~\citek{Cowan:2010js}. This includes
@@
 that discriminates between a hypothesized process of interest (labeled \textit{signal}) and various other processes 
-(labeled \textit{background}).
-To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers.
+(labeled \textit{background}). Typically,  pseudo-data from the simulator are used to approximate the density
+at various parameter points, and an interpolation algorithm is used to approximate the parametrized model~\citek{Cranmer:2012sba}.
+
+To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers that take advantage of a high dimensional feature vector $x$.
 Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citek{Hocker:2007ht} are commonly used.
@@
 
-In the context of machine learning or decision theory one starts with a well defined loss function.
-In order to define a Bayes risk one must also specify a prior over the parameters.
-%In practice, it can be difficult, particularly for large collaborations, to agree on such high-level topics.
-In practice, large particle physics collaborations have preferred to avoid specifying the 
-high-level loss function and prior, and instead accepted the properties of generalized likelihood ratio tests.
-Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself.
+The initial goal of this work was to extend the typical usage of discriminative classifiers in HEP to be robust to 
+nuisance parameters in the simulators.
+The scope of the result expanded once it became clear that this offers a way to approximate the likelihood function $p(x|\theta)$ in what is typically considered the likelihood-free setting.
+This approach is complementary to ABC as it does not require a prior over the parameters and can also be used in the classical (frequentist) setting.
+In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.
 
-The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss. 
-This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process.
+%In the context of machine learning or decision theory one starts with a well defined loss function.
+%In order to define a Bayes risk one must also specify a prior over the parameters.
+%%In practice, it can be difficult, particularly for large collaborations, to agree on such high-level topics.
+%In practice, large particle physics collaborations have preferred to avoid specifying the 
+%high-level loss function and prior, and instead accepted the properties of generalized likelihood ratio tests.
+%Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself.
+%
+%The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss. 
+%This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process.
 Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.
-In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.
+%In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.
 
@@
  \item $\theta$: parameters of a statistical model
-\item $p(x| \theta)$:  probability density  (statistical model) for $x$ given $\theta$
+\item $p(x| \theta)$:  probability density  (simulation-based model) for $x$ given $\theta$
 %\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
@@
 
-The main result of this paper is that one can form an equivalent test based on 
+The main result of this paper is to generalize the observation that one can form an equivalent test based on 
 %\begin{equation}
@@
 
-While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire data set $D$, we are able to reformulate it such that the machine learning problem is a per-event classification problem.
+Here we see that the original goal for simple hypothesis testing (i.e.
+to make a decision to accept or reject the null hypothesis based on the entire data set $D$) has been reformulated into a per-event classification problem.
 This follows from the fact that we assume the $x_e$ to be i.i.d.
@@
 
-The first possible source of confusion we wish to avoid is that here $p(x|\theta)$  is a generative \textit{statistical model} for the features $x$, not a generative classifier.
+The first possible source of confusion we wish to avoid is that here $p(x|\theta)$  is a  \textit{generative statistical model} for the features $x$, not a generative classifier.
 We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory or simulator, able to make predictions about $x$ and being motivated by domain-specific considerations.
-For example, in the context of HEP $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.
-Moreover, we are not attempting to learn the generative model $p(x|\theta)$, we are taking it as given and trying to learn the corresponding (generalized) likelihood ratio test.
+%For example, in the context of HEP $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.
+%Moreover, we are not attempting to learn the generative model $p(x|\theta)$, we are taking it as given and trying to learn the corresponding (generalized) likelihood ratio test.
 %The point here is to make the distinction of $p(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.
@@
 we are not directly interested in calibrating the classification score in terms of a per-event posterior probability $p(y|x)$. 
-Instead, we are interested in the approximation of the per-experiment likelihood ratio, which can be used to calculate p-values defined by $P(T(D) > k |\theta)$.
+Instead, we are interested in the approximation of the per-experiment likelihood function or likelihood ratio, which might be used for several purposes, including the calculation of p-values.
+% defined by $P(T(D) > k |\theta)$.
 %
@@
 
-Lastly, in the setting of frequentist hypothesis tests, we do not have a prior $\pi(\theta)$. 
+Lastly, in the setting of frequentist hypothesis tests and confidence intervals, we do not have a prior $\pi(\theta)$. 
 While we can use the generative models to produce training data $(x,y=0)$ generated 
@@
 \end{equation}
+This is equivalent to approximating the likelihood function for $\theta_0$  when $\theta_1$ is held fixed.
 Here we see that the test statistic $T$ for the experiment is composed of a sum over events of the per-event function $q(x)$.
-A sum over a monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.
+A sum over a monotonic, but non-linear function of $q(x)$ would not lead to an equivalent statistic.
 
@@
 \end{equation}
-leads to an equivalent test.
+leads to an equivalent statistic.
 

### 2015-06-05 tex/parametrized-learning-aoas.tex
Polishing

@@
 These tests are complicated by the fact that computer simulators are used as a generative model for 
-the data, but they do not provide a way to evaluate the likelihood function for a given
-observation. 
+the data, but they do not provide a way to evaluate the likelihood function. 
 We demonstrate how discriminative classifiers can be used to approximate the likelihood function when 
@@
 %In what follows, it will be important to distinguish between the per-event loss used for classification, and the per-experiment loss 
-A fundamental distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g.
+The key distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g.
 the variance of an estimator for the mass parameter) is a per-experiment operation involving a data set with many events.
@@
 
-The initial goal of this work was to extend the typical usage of discriminative classifiers in HEP to be robust to 
+The initial motivation for this work was to extend the typical usage of discriminative classifiers in HEP to be robust to 
 nuisance parameters in the simulators.
@@
 Our claim is that the generative model $p(x|\theta)$ can be used to calibrate the density $p(s|\theta)$ and that
-\begin{equation}
+\begin{equation}\label{eq:envoloping}
 \ln T' = \sum_{e=1}^n \underbrace{\log \left[ \frac {p(s_e | \theta_0) }{ p(s_e | \theta_1) } \right]}_{q(s_e)} \;,
@@
 \begin{equation}
-\lambda(x) =  \frac{ \sup_{\theta \in \Theta_0} p(D | \theta)}{ \sup_{\theta \in \Theta} p(D | \theta)} \; .
+\lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p(D | \theta)}{ \sup_{\theta \in \Theta} p(D | \theta)} \; .
 \end{equation}
@@
 %In order for this approach to be useful, we need to be able to approximate both based on samples $\{(x,\theta)\}$ drawn from the generative model $p(x|\theta)$.  
-In order for this approach to be useful in a setting where it is not possible to evaluate the density $p(x|\theta)$, we need to be able to approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
+In order for this approach to be useful in the likelihood-free setting where it is not possible to evaluate the density $p(x|\theta)$, we need to be able to approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
 
@@
 Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
-histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$.
+histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter points.
+In order to construct an approximate likelihood function continuous in the parameters $\theta$ HEP experiments have typically used some form of interpolation of the univariate score density over the parameters $\theta$.
+%The relative merits of various interpolation algorithms and the tradeoff between more samples from the generative model at a particular set of parameter points and fewer samples at more parameter points is left as a subject of future work.
 
@@
 
-In the  limit of large samples from the generative model,  we can approximate the original likelihood ratio test arbitrarily well.
+%%% KC, commenting this out late in the game
+%In the  limit of large samples from the generative model,  we can approximate the original likelihood ratio test arbitrarily well.
 With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{p}(\hat{s}|\theta)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test.
@@
 Somewhat unusually, the unknown values of the parameters are taken
-as input to the classifier, and are latent variables whose values will be specified via the enveloping (generalized) likelihood ratio test.
+as input to the classifier, and are latent variables whose values will be specified via the enveloping (generalized) likelihood ratio of Eq.~\ref{eq:envoloping}.
 We denote the learned family of classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based roughly on the following algorithmic flow.
@@
 In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{p}\circ \hat{s}$.
+The critical observation is that  if we postpone the evaluation of the classifier to the stage of evaluating the enveloping likelihood ratio, then we can identify the parameters $\theta$ that are being compared in the likelihood ratio with the parameters used to parametrize $s(x;\theta_0, \theta_1)$.
 A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
-In both cases, constructing the density $\hat p(\hat s|\theta)$ requires running the generative model at $\theta$.
+In both the fixed and parametrized cases, constructing the density $\hat p(\hat s|\theta)$ requires running the generative model at $\theta$.
 In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $p(x|\theta)$.
 This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
-In our HEP examples we have chosen to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$.
+In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$.
 %In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
@@
 %Often machine learning classification algorithms are trained on 
-In this setting, large samples of synthetic data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density $p_s(x|\nu_0)$.
+In this setting, large samples of pseudo-data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density $p_s(x|\nu_0)$.
 Importantly, the $y=1$ label corresponds to the signal only, and not to the alternate signal-plus-background hypothesis.
@@
 
- Once the classifier is trained, large samples of synthetic data drawn from $p_s(x | \nu)$ and $p_b(x | \nu)$, then we estimate the distributions  $\hat{p}_s(\hat s | \nu)$ and $\hat{p}_b(\hat s | \nu)$ continuously parametrized in $\nu$.
-An example of the distributions of the distribution of $\hat s$ for the signal and background events with $\nu=\nu_0$ is shown in Figure~\ref{fig:tmva}.
+ Once the classifier is trained, large samples of pseudo-data are drawn from $p_s(x | \nu)$ and $p_b(x | \nu)$ and we estimate the distributions  $\hat{p}_s(\hat s | \nu)$ and $\hat{p}_b(\hat s | \nu)$ continuously parametrized in $\nu$.
+An example of the distributions of $\hat s$ for the signal and background events with $\nu=\nu_0$ is shown in Figure~\ref{fig:tmva}.
 
@@
 %&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)}{ w_c(\theta_0) p_c(x| \theta_0)}  \right]^{-1} \\
-&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \\
+&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \\ \label{eq:decomposedResult}
 &=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} 
@@
 The third line uses Theorem~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
-In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before performing the generalized likelihood ratio test).
-Equation~\ref{eq:decompose} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
+In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio).
+Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
 
@@
 
-While the original motivation for this work was to improve the treatment of systematic uncertainties in new particle searches by parametrizing the classifier in terms of the nuisance parameters $\nu$, the same approach can be used to for parameters of interest.
+While the original motivation for this work was to improve the treatment of systematic uncertainties in new particle searches by parametrizing the classifier in terms of the nuisance parameters $\nu$, the same approach can be used to for parameter inference.
 In the case of new particle searches the parameter of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
-When measuring particle properties the distribution of the features also depend on parameters of interest, such as a particle's mass and quantum numbers.
+When measuring particle properties the distribution of the features also depend on parameters such as a particle's mass and quantum numbers.
 This is easily accommodated by extending $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$ includes both parameters of interest and nuisance parameters.
@@
 
+\bigskip
+
 \subsection{Related work}\label{S:Related}
@@
 This interesting work is distinct from the point made here in which the generative model is being used for the purpose of providing training data and calibration.
-\cite{McCallum}
- consider a hybrid generative/discriminative classifier; however, the goal of that work is not to leverage a generative model for the data, but to use both approaches to learn different subsets of the parameters in a single hybrid classifier.
+%\cite{McCallum} consider a hybrid generative/discriminative classifier; however, the goal of that work is not to leverage a generative model for the data, but to use both approaches to learn different subsets of the parameters in a single hybrid classifier.
 \cite{BiancaZadrozny} emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this.
@@
 KC would like to thank  
-Daniel Whiteson for discussions and encouragement throughout the project and
+Daniel Whiteson for encouragement throughout the project and
 Alex Ihler for challenging discussions that 
 led to a reformulation of the initial idea.
-KC would also like to thank Shimon Whiteson, Babak Shahbaba for advice in the presentation, Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
+KC would also like to thank Shimon Whiteson, Babak Shahbaba for advice in the presentation, Radford Neal for discussion of his earlier work, Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
 project early in its conception, Bal\'azs K\'egl for discussions about the Kaggle challenge and feedback to the draft, 
@@
 Moore and Sloan foundations for their generous support of the data science environment at NYU.
+
+\newpage
 

### 2015-06-05 tex/parametrized-learning-aoas.tex
new conclusions

@@
 We have outlined a technique to reformulate generalized likelihood ratio test over a high-dimensional data set with multiple events in terms of a univariate density of a classifier score. 
-We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$.
-Furthermore, this allows one to turn the optimization of the procedure for the per-experiment test into optimization of a per-event classification problem and a calibration procedure.
+We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a simulator  $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not possible to directly evaluate the likelihood $p(x|\theta)$.
 This approach leverages the power of machine learning in a classical statistical setting.
-In particular, it extends the common usage of machine learning for simple hypothesis testing to include the parameter estimation, confidence intervals, and composite hypothesis testing.
+It offers an alternative to approximate Bayesian computation for parameter inference in the likelihood-free setting that can also be used in the frequentist formalism without specifying a prior over the parameters.
+
+%Furthermore, this allows one to turn the optimization of the procedure for the per-experiment test into optimization of a per-event classification problem and a calibration procedure.
+
+%In particular, it extends the common usage of machine learning for simple hypothesis testing to include the parameter estimation, confidence intervals, and composite hypothesis testing.
 

### 2015-06-05 tex/parametrized-learning-aoas.tex
reorganizing

@@
 \usepackage{amssymb, amsmath}
+\usepackage{oubraces}
+\usepackage{mathtools}
 \usepackage{algorithm}
@@
 \newcommand{\citek}[1]{[\cite{#1}]}
+\newcommand{\sstar}{s^*}
+\newcommand{\sfunc}{s}
 
@@
 This approach is complementary to ABC as it does not require a prior over the parameters and can also be used in the classical (frequentist) setting.
+A strength of this approach is that it separates the quality of the approximation of the target likelihood from the quality of the calibration.
 In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.
@@
 \begin{equation}
-T(D) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
+T(D; \theta_0, \theta_1) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
 \end{equation}
@@
 \begin{equation}\label{eq:equivLRtest}
-T'(D) = \prod_{e=1}^n \frac{ p(s_e | \theta_0)}{ p(s_e | \theta_1)}
+T'(D; \theta_0, \theta_1) = \prod_{e=1}^n \frac{ p(s_e | \theta_0)}{ p(s_e | \theta_1)}
 \end{equation}
@@
 \begin{equation}
-\ln T =   \sum_{e=1}^n \underbrace{\log \left[ \frac {p(x_e | \theta_0) }{ p(x_e | \theta_1) } \right]}_{q(x_e)} \;.
+\ln T(D; \theta_0, \theta_1) =   \sum_{e=1}^n \underbrace{\log \left[ \frac {p(x_e | \theta_0) }{ p(x_e | \theta_1) } \right]}_{q(x_e)} \;.
 \end{equation}
-This is equivalent to approximating the likelihood function for $\theta_0$  when $\theta_1$ is held fixed.
 Here we see that the test statistic $T$ for the experiment is composed of a sum over events of the per-event function $q(x)$.
@@
 \begin{equation}\label{eq:envoloping}
-\ln T' = \sum_{e=1}^n \underbrace{\log \left[ \frac {p(s_e | \theta_0) }{ p(s_e | \theta_1) } \right]}_{q(s_e)} \;,
+\ln T'(D; \theta_0, \theta_1) = \sum_{e=1}^n \underbrace{\log \left[ \frac {p(s_e | \theta_0) }{ p(s_e | \theta_1) } \right]}_{q(s_e)} \;,
 \end{equation}
@@
 
-For notational simplicity, let $p_0(x) = p(x|\theta_0)$, $p_1(x) = p(x|\theta_1)$, and $s(x)=s(x; \theta_1, \theta_0)$.
+For notational simplicity, let $p_0(x) = p(x|\theta_0)$, $p_1(x) = p(x|\theta_1)$, and $\sfunc(x)=s(x; \theta_0, \theta_1)$.
 The distribution of $x$ totally determines the distribution of $s$. 
 In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
-Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to $\Omega_c$ at the point $x$.
+Let $\Omega_{\sstar}$ be the level set $\{x \mid s(x; \theta_0, \theta_1) = \sstar \}$ and \mbox{$\hat{n}=\nabla s(x) / |\nabla s(x)|$} be the orthonormal vector to $\Omega_{\sstar}$ at the point $x$.
 
@@
 \end{equation}
-It is sufficient to show that $q_x(x) = q_s(s(x))$ $ \forall x\in\Omega_c$. The function $q_s(s)$ is based on the induced densities $p_0(s)$ and $p_1(s)$.  The induced density $p_1(c)$ is given by 
+It is sufficient to show that $q_x(x) = q_s(s(x))$.
+%$ \forall x\in\Omega_{\sstar}$. 
+The function $q_s(s)$ is based on the induced densities $p_0(s)$ and $p_1(s)$.  The induced density $p_1(s)$ is given by 
 \begin{equation}
-p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega_c p_1(x)  / | \hat{n} \cdot \nabla s  |
+p_1(\sstar) = \int dx \delta(\sstar-s(x)) p_1(x) = \int d\Omega_{\sstar} p_1(x)  / | \hat{n} \cdot \nabla s  |
 \end{equation}
-and a similar equation for $p_0(c)$. 
+and a similar equation for $p_0(s)$. 
 %\textbf{Do we need Jacobian for x $\to$ s independent of delta function part, I think that's double counting?}
@@
 \begin{equation}
-\frac{p_1(c)}{p_0(c)} =  \frac{p_1(x)}{p_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
+\frac{p_1(s(x))}{p_0(s(x))} =  \frac{p_1(x)}{p_0(x)}  \; . %\;\hspace{3em} \forall x\in\Omega_{\sstar}\; .
 \end{equation}
 \textbf{Proof}
-For $x\in \Omega_c$, we can factor out of the integral the constant $p_1(x)/p_0(x)$.
+For $x\in \Omega_{\sstar}$, we can factor out of the integral the constant $p_1(x)/p_0(x)$.
 Thus
 \begin{equation}
-p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega_c p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_c p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
+p_1(\sstar) =  \int d\Omega_{\sstar} p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_{\sstar} p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
+%p_1(\sstar) = \int dx \delta(\sstar-s(x)) p_1(x) = \int d\Omega_{\sstar} p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_{\sstar} p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
 \end{equation}
@@
 \begin{equation}
-\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)} \frac{\int d\Omega_c p_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_c p_0(x) / | \hat{n} \cdot \nabla s  |} = \frac{p_1(x)}{p_0(x)}  \;\hspace{3em} \forall x\in\Omega_c.
+\frac{p_1(\sstar)}{p_0(\sstar)} = \frac{p_1(x)}{p_0(x)} \frac{\int d\Omega_{\sstar} p_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_{\sstar} p_0(x) / | \hat{n} \cdot \nabla s  |} = \frac{p_1(x)}{p_0(x)}  \;\hspace{3em} \forall x\in\Omega_{\sstar}.
 \end{equation}
@@
 %for $\theta=\{\theta_0,\theta_1\}$, and constructing the likelihood ratio based on $p(D_s|\theta_0)/p(D_s|\theta_1)$.
+
+\section{Embedding the classifier into the likelihood}
+
+Thus far we have shown that the target likelihood ratio $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via the univariate densities $p(s|\theta_0)/p(s|\theta_1)$ if the classifier $s(x|\theta_0, \theta_1)$ is a strictly increasing function of $p(x|\theta_0)/p(x|\theta_1)$.
+We now generalize from the ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the case where $\theta$ are continuous model parameters.
+ We postpone the practicalities of training the classifier and estimating the density to Section~\ref{S:classifier} and continue in the likelihood-free setting with idealized classifiers and their densities.
+
+In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$ for the observed data and never refer back to the original features $x_e$.
+In the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0, \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
+
+The critical observation is that  if we postpone the evaluation of the classifier to the stage of evaluating the enveloping likelihood ratio, then we can identify the value of the parameters  that are being compared in the likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
+\begin{equation}\label{eq:embedding}
+T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
+{p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
+\end{equation}
+This is equivalent to approximating the likelihood function for $\theta_0$  when $\theta_1$ is held fixed.
+
+%\begin{equation}
+%\frac{p(x|\theta_0)}{p(x|\theta_1)} = \frac{ p (s(x; \,\lefteqn{\overbracket[0.9pt]{\phantom{  \theta_0, \theta_1) | \,\, }}^{
+%\text{identify}}} 
+%\theta_0, \theta_1) | \theta_0)}
+%{p (s(x; \theta_0, \,\lefteqn{\underbracket[0.9pt]{\phantom{  \theta_1) | \,\, }}_{
+%\text{identify}}} 
+% \theta_1) | \theta_1) } \; .
+%\end{equation}
+
+%A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
+%
+%In both the fixed and parametrized cases, constructing the density $\hat p(\hat s|\theta)$ requires running the generative model at $\theta$.
+In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $p(x|\theta)$.
+This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
+In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$.
+%%In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
+%One can easily imagine a number of approaches to embedding the classifier and estimating the density $\hat p(\hat s|\theta)$ and the relative merits of those approaches will depend critically on the dimensionality of $\theta$ and the computational cost of the generative model.
+We leave a more general strategy for this overarching optimization problem as an area of future work.
 
@@
 
-Thus far we have shown that likelihood ratio tests based on $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via hypothesis tests based on the univariate densities $p(s|\theta)$ for the particular dimensionality reduction map $s(x|\theta_0, \theta_1)$.
+%Thus far we have shown that likelihood ratio tests based on $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via hypothesis tests based on the univariate densities $p(s|\theta)$ for the particular dimensionality reduction map $s(x|\theta_0, \theta_1)$.
 % Often it is not possible to evaluate the density $p(x|\theta)$ at a given point $x$.  
@@
 
-\subsection{The standard discriminative classification setting} 
+\subsection{The fixed discriminative classification setting} 
 For fixed $\theta_0$ and $\theta_1$ we can generate 
@@
 data $(x_i,  y_i=0)$ and $p(x|\theta_1)$ to generate training data $(x_i , y_i=1)$.
-With balanced training data   ($p(y=1)=p(y=0)=\sfrac{1}{2}$) a quadratic loss function will lead to classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the desired per-event likelihood ratio $q(x)$.
+With balanced training data   \mbox{($p(y=1)=p(y=0)=\sfrac{1}{2}$)} a quadratic loss function will lead to classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the desired per-event likelihood ratio $q(x)$.
 Thus, standard supervised learning algorithms with various surrogate loss functions lead to discriminative classifiers that approximate a monotonic function of per-event likelihood ratio $q(x)$.
+
 Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
 histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter points.
-In order to construct an approximate likelihood function continuous in the parameters $\theta$ HEP experiments have typically used some form of interpolation of the univariate score density over the parameters $\theta$.
+In order to construct an approximate likelihood function continuous in the parameters $\theta$, HEP experiments have typically used some form of interpolation of the univariate score density over the parameters $\theta$~\citek{Cranmer:2012sba}.
+
 %The relative merits of various interpolation algorithms and the tradeoff between more samples from the generative model at a particular set of parameter points and fewer samples at more parameter points is left as a subject of future work.
@@
 
-\subsection{Embedding the classifier into the likelihood}
+\subsection{Practicalities of embedding the classifier into the likelihood}
 
-In most settings that make use of likelihood ratio tests, the likelihood is based directly on some approximation of density for the observed data via $\hat{p}(x|\theta)$.
+%A  straightforward approach to parameter inference in the likelihood-free setting is to approximate the density $\hat{p}(x|\theta)$.
 In general, approximating the density $\hat{p}(x|\theta)$ is difficult for high-dimensional data, which motivates the use of the dimensionality reduction map $\hat{s}(x)$ and likelihood ratio tests based on the univariate density $\hat{p}(\hat{s}|\theta)$.
-In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$.
-In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{p}\circ \hat{s}$.
+%In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$.
+In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$; however, we can  embed the classifier as in Eq.~\ref{eq:embedding}.
+% into the likelihood function to carry out the composition $\hat{p}\circ \hat{s}$.
 The critical observation is that  if we postpone the evaluation of the classifier to the stage of evaluating the enveloping likelihood ratio, then we can identify the parameters $\theta$ that are being compared in the likelihood ratio with the parameters used to parametrize $s(x;\theta_0, \theta_1)$.
-A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
@@
 In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$.
+
+In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$.
+In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$; however, we can  embed the classifier as in Eq.~\ref{eq:embedding}.
+
+ A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
+
 %In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
@@
 The third line uses Theorem~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
-In the situation where the only free parameters of the mixture model are the coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio).
+In the situation where the only free parameters of the  model are the mixture coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio).
 Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
@@
 
-\subsection{Related work}\label{S:Related}
+\section{Related work}\label{S:Related}
 
@@
 Neal says  ``we cannot use the classifier on real data, since we don't know the correct value for [$\theta]$'' and goes on to outline an approach where one uses regression on a per-event basis to estimate $\hat{\theta}(x)$ and perform the composition $s(x; \hat{\theta}(x))$ (much like profiling).\footnote{Neal considers a lower-dimensional `bottlenecks' $\theta^*=g(\theta)$, which are not essential to the discussion here.}  This can lead to a significant loss of information since (at least in most particle physics examples) a single event carries little information about the true value of $\theta$, though the full data set $D$ may be informative -- for instance, a single observation would not be sufficient to estimate the variance of a distribution, though repeated observations would.
-The work of Neal correctly identifies this as an approximate procedure even in the case of a ideal classifier.
+The work of Neal correctly identifies this as an approximation of the target likelihood even in the case of a ideal classifier.
 In contrast, the approach described here does not eliminate the dependence of the classifier on $\theta$.%
@@
 This avoids the loss of information that occurs from the regression step $\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact result in the case of an ideal classifier.
+In both cases, the quality of the classifier is factorized from the calibration of its density.
 
@@
 We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a simulator  $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not possible to directly evaluate the likelihood $p(x|\theta)$.
+%A technique for decomposing this ratio when the generative model is a mixture of components was presented with the aim to help focus capacity of the classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small mixture coefficient.
 This approach leverages the power of machine learning in a classical statistical setting.
 It offers an alternative to approximate Bayesian computation for parameter inference in the likelihood-free setting that can also be used in the frequentist formalism without specifying a prior over the parameters.
+A strength of this approach is that it separates the quality of the approximation of the target likelihood from the quality of the calibration.
+The former is related to the ability of supervised learning approaches to  classification, which will continue to improve.
+The calibration procedure for a particular parameter point is fairly straight forward since it involves estimating a univariate density using a generative model of the data.
+The difficulty of the calibration stage is performing this calibration continuously in $\theta$.
+Different strategies to this calibration are anticipated depending on the dimensionality of $\theta$, the complexity of the resulting likelihood function, and the difficulties associated to running the simulator.
 

### 2015-06-05 tex/parametrized-learning-aoas.tex
Updated section 5

@@
 Our claim is that the generative model $p(x|\theta)$ can be used to calibrate the density $p(s|\theta)$ and that
-\begin{equation}\label{eq:envoloping}
+\begin{equation}\label{eq:enveloping}
 \ln T'(D; \theta_0, \theta_1) = \sum_{e=1}^n \underbrace{\log \left[ \frac {p(s_e | \theta_0) }{ p(s_e | \theta_1) } \right]}_{q(s_e)} \;,
@@
 %In order for this approach to be useful, we need to be able to approximate both based on samples $\{(x,\theta)\}$ drawn from the generative model $p(x|\theta)$.  
-In order for this approach to be useful in the likelihood-free setting where it is not possible to evaluate the density $p(x|\theta)$, we need to be able to approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
-
-Denote the approximate dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{p}(\hat{s}|\theta)$.
-In general we will be interested in the machine learning problem that approximates these distributions based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
-The first step in this direction is to confirm that a discriminative classifier obtained from common supervised learning algorithms will yield a function that is one-to-one with $p(x|\theta_0)/p(x|\theta_1)$.
+In order for this approach to be useful in the likelihood-free setting, where it is not possible to evaluate the density $p(x|\theta)$, we need to be able to approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
+Denote the  dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{p}(\hat{s}|\theta)$.
+
+%In general we will be interested in two machine learning problems associated to  that approximates these distributions based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
+
+One strength of this approach is that it factorizes the approximation of the per-event likelihood ratio ($\hat{s}(x; \theta_0, \theta_1) \approx s(x; \theta_0, \theta_1)$) from the calibration procedure ($\hat p(\hat s| \theta) \approx p(\hat{s}|\theta)$).
+Thus, even if the classifier does a poor job of reproducing the level sets of the per-event likelihood ratio, the density of $\hat{s}$ can still be well calibrated.
+In that case, one might loose power, but the resulting inference will still be valid.
+This point was made by \cite{Neal:2007zz} and is well appreciated by the particle physics community that typically takes a conservative attitude towards the use of machine learning classifiers precisely due to concerns about the calibration $p$-values in the face of nuisance parameters associated to the simulator.
+
+%The first step in this direction is to confirm that a discriminative classifier obtained from common supervised learning algorithms will yield a function that is one-to-one with $p(x|\theta_0)/p(x|\theta_1)$.
 %In particular, 
@@
 
-Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
+%Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
 histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter points.
@@
 
-We are left with the practical question of how to train a family of discriminative classifiers parametrized by $\theta_0$ and $\theta_1$, the 
+In order to provide parameter inference in the likelihood-free setting, we must train a family of  classifiers parametrized by $\theta_0$ and $\theta_1$, the 
 parameters associated to the null and alternate hypotheses, respectively. While this could be done independently
@@
 Somewhat unusually, the unknown values of the parameters are taken
-as input to the classifier, and are latent variables whose values will be specified via the enveloping (generalized) likelihood ratio of Eq.~\ref{eq:envoloping}.
+as input to the classifier; their values will be specified via the enveloping (generalized) likelihood ratio of Eq.~\ref{eq:enveloping}.
 We denote the learned family of classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based roughly on the following algorithmic flow.
@@
 
+\bigskip
+
 While the function $p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ will minimize the expected squared loss based on 
@@
 
-In both the fixed and parametrized cases, constructing the density $\hat p(\hat s|\theta)$ requires running the generative model at $\theta$.
+Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
+histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter points.
+For an single parameter point, this is a tractable univariate density estimation problem. The challenge comes from the need to estimate this density for all values of $\theta$. 
+A straight forward approach would be to run the generative model on demand for any particular value of $\theta$. 
 In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $p(x|\theta)$.
 This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
-In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$.
-
+In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$~\citek{Cranmer:2012sba}.
+
+There is a mild technical challenge in embedding the classifier into the likelihood.
 In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$.
-In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$; however, we can  embed the classifier as in Eq.~\ref{eq:embedding}.
-
+In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$.
+Thus one must implement the embedding of the classifier as in Eq.~\ref{eq:embedding}.
 A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
-%In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
 One can easily imagine a number of approaches to embedding the classifier and estimating the density $\hat p(\hat s|\theta)$ and the relative merits of those approaches will depend critically on the dimensionality of $\theta$ and the computational cost of the generative model.

### 2015-06-05 tex/parametrized-learning-aoas.tex
edits of HEP section

@@
 The resulting classifier approximates the regression function $p_s(x|\nu_0)/(p_s(x|\nu_0)+p_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$.
-Associating the $y=1$ label to the signal component has advantages because it helps the classifier focus its capacity on the relevant regions in the feature space, particularly when the signal is a very small perturbation to the background.
+Associating the $y=1$ label to the signal component has advantages because it helps the classifier focus its capacity on the relevant regions in the feature space, particularly when the signal is a very small perturbation to the background (ie.
+$\mu \ll 1$).
 %At the LHC one can have $\mu \sim 10^{-6}$. 
@@
 
-The standard use of machine learning in HEP can be improved by training a parametrized, discriminative classifier corresponding to the generalized likelihood ratio test 
+The standard use of machine learning in HEP can be improved by training a parametrized classifier corresponding to the generalized likelihood ratio test 
 \begin{equation}
@@
 
-While the original motivation for this work was to improve the treatment of systematic uncertainties in new particle searches by parametrizing the classifier in terms of the nuisance parameters $\nu$, the same approach can be used to for parameter inference.
+While the original motivation for this work was to improve the treatment of systematic uncertainties in new particle searches by parametrizing the classifier in terms of the nuisance parameters $\nu$, the same approach can be used for parameter inference.
 In the case of new particle searches the parameter of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
@@
 
-This formalism represents is a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
-The work of  ~\citek{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
+This formalism represents a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
+The work of  ~\citek{Whiteson:2006ws} is similar as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
 This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
@@
 In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method.
-Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time per event to calculate the likelihood ratio $q(x)$.
+Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time to calculate the likelihood ratio $q(x)$ for a single event.
 The work here can be seen as aiming at the same conceptual target, but utilizing machine learning to overcome the complexity of the detector simulation.
@@
 %Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
-The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citek{Bolognesi:2012mm}.
+The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citek{Gao:2010qx,Bolognesi:2012mm}.
 
@@
 This avoids the loss of information that occurs from the regression step $\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact result in the case of an ideal classifier.
-In both cases, the quality of the classifier is factorized from the calibration of its density.
+In both cases, the quality of the classifier is factorized from the calibration of its density, which allows for valid inference even if there is a loss of power due to a non ideal classifier.
 
@@
 We have outlined a technique to reformulate generalized likelihood ratio test over a high-dimensional data set with multiple events in terms of a univariate density of a classifier score. 
-We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a simulator  $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not possible to directly evaluate the likelihood $p(x|\theta)$.
+We have shown that a parametrized family of discriminative classifiers $\hat s(x; \theta_0, \theta_1)$ trained and calibrated with a simulator  $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not possible to directly evaluate the likelihood $p(x|\theta)$.
 %A technique for decomposing this ratio when the generative model is a mixture of components was presented with the aim to help focus capacity of the classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small mixture coefficient.

### 2015-06-05 tex/parametrized-learning-aoas.tex
removing comments

@@
 The likelihood function is key to Bayesian inference, and many areas of science that report the results of classical hypothesis tests or confidence intervals use the (generalized or profile) likelihood ratio as a test statistic.
-%  are established tools for statistical inference~\citek{Wilks1938}. 
 It is increasingly common that a simulator (or generative model) is used to describe complex processes that tie 
@@
 Recently, there has been progress in using deep networks~\citek{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citek{HepML}, the largest Kaggle challenge in history.
-%The use of classifiers is effective, but an intermediate step not obviously connected to the target likelihood ratio test.
 
-%Collisions are usually selected with supervised learning methods 
 While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citek{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters.
 As noted in ~\citek{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
-%In what follows, it will be important to distinguish between the per-event loss used for classification, and the per-experiment loss 
 The key distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g.
@@
 They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process.
-%This paper shows that (to the extent a generalized likelihood ratio test in the full feature space is close to  optimal) one can achieve optimal results with a  specific per-event procedure by reformulating the target likelihood ratio in terms of a per-event classification followed by an specific calibration process.
 
@@
 In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.
-
-%In the context of machine learning or decision theory one starts with a well defined loss function.
-%In order to define a Bayes risk one must also specify a prior over the parameters.
-%%In practice, it can be difficult, particularly for large collaborations, to agree on such high-level topics.
-%In practice, large particle physics collaborations have preferred to avoid specifying the 
-%high-level loss function and prior, and instead accepted the properties of generalized likelihood ratio tests.
-%Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself.
-%
-%The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss. 
-%This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process.
-Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.
-%In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.
-
-%\bigskip
-%
-%Generalized likelihood ratio tests enjoy a number of optimal properties in the asymptotic limit and are widely used even away from that limit.
-These asymptotic properties, the difficulty in specifying a prior over the parameters needed to evaluate a Bayes risk, and the vast literature associated to likelihood ratio tests has led many users of such tests to not be explicit about the loss function they wish to minimize for their experiment.
-Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself.
-The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss.
-
-%Similarly,~\cite{Neal:2007zz} suggested using a classifier as a dimensionality reduction map to aid in the estimation of the likelihood function.
-We discuss the key differences with that work in Section~\ref{S:Related}.
-
-%This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process.
-Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.
-
-%This paper describes how one can utilize generative statistical models for the data in conjunction with discriminative classifiers to construct generalized likelihood ratio tests that are equivalent to
-%the one that would have been obtained if it were possible to evaluate the probability 
-%density implicitly being sampled by the generative model.
-%
-%This paper describes how one can utilize generative statistical models for the data in conjunction with discriminative classifiers to construct generalized likelihood ratio tests that are equivalent to
-%the one that would have been obtained if it were possible to evaluate the probability 
-%density implicitly being sampled by the generative model.
-
-%The labeled data used to train the classifiers are not obtained from the experimental apparatus. 
-%Instead they are generated from a 
-%detailed simulation, which includes a quantum mechanical description of the underlying particle 
-%interactions, an exquisitely detailed simulation of the particle detectors with hundreds of millions of 
-%sensors, and the data processing pipeline that reduces the data for a single event to tens or hundreds of high-level features used as input to the classifier. 
-%It takes several minutes for this generative model to simulate a single collision or \textit{event} with Monte Carlo
-%techniques. While particle physics offers an extreme example, it is common in science that generative 
-%models are available while it not feasible to evaluate the probability density function that is implicitly being 
-%sampled by the generative model.
 
@@
 We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory or simulator, able to make predictions about $x$ and being motivated by domain-specific considerations.
-%For example, in the context of HEP $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.
-%Moreover, we are not attempting to learn the generative model $p(x|\theta)$, we are taking it as given and trying to learn the corresponding (generalized) likelihood ratio test.
-%The point here is to make the distinction of $p(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.
 
@@
 Instead, we are interested in the approximation of the per-experiment likelihood function or likelihood ratio, which might be used for several purposes, including the calculation of p-values.
-% defined by $P(T(D) > k |\theta)$.
-%
-%%the likelihood ratio $T(D)$ is aimed at tests based on the entire data set; 
-%we are not interested in thresholded classification on individual events $x_e$.  
-%Knowing that classifier scores are often poorly calibrated is an immediate concern.
-
-%We wish to have well calibrated p-values defined by $P(T(D) > k |\theta)$, not well calibrated posterior probabilities $p(y|x)$.
 
@@
 Similar points have been made by~\cite{ClaytonScott} and \cite{Neal:2007zz}.
-
-% and we need not
-%interpret the balance of the training data in terms of a prior probability for the two hypotheses.
-%free from the
-%complications of the prior $\pi(\theta)$ that is undefined in the frequentist context.
 
@@
 
-%\bigskip
-%In the case of simple hypothesis testing, $\theta_0$ and $\theta_1$ are specified and there is a unique map $s(x) =  s(x_e; \theta_0, \theta_1)$.
-In that case, the equivalent likelihood ratio test can be performed by first transforming the data to $D_s = \{s_1, \dots, s_e\}$, constructing the likelihoods
-%\begin{equation}\label{eq:NP}
-%p( D_s \,|\,  \theta) = \prod_{e=1}^n \,  p( s_e \, |\,  \theta)   \; 
-%\end{equation}
-%for $\theta=\{\theta_0,\theta_1\}$, and constructing the likelihood ratio based on $p(D_s|\theta_0)/p(D_s|\theta_1)$.
-
 \section{Embedding the classifier into the likelihood}
@@
 This is equivalent to approximating the likelihood function for $\theta_0$  when $\theta_1$ is held fixed.
-
-%\begin{equation}
-%\frac{p(x|\theta_0)}{p(x|\theta_1)} = \frac{ p (s(x; \,\lefteqn{\overbracket[0.9pt]{\phantom{  \theta_0, \theta_1) | \,\, }}^{
-%\text{identify}}} 
-%\theta_0, \theta_1) | \theta_0)}
-%{p (s(x; \theta_0, \,\lefteqn{\underbracket[0.9pt]{\phantom{  \theta_1) | \,\, }}_{
-%\text{identify}}} 
-% \theta_1) | \theta_1) } \; .
-%\end{equation}
-
-%A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
-%
-%In both the fixed and parametrized cases, constructing the density $\hat p(\hat s|\theta)$ requires running the generative model at $\theta$.
-In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $p(x|\theta)$.
-This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
-In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$.
-%%In such cases, the properties of the interpolation algorithm should be part of the considerations of the over-arching optimization problem.
-%One can easily imagine a number of approaches to embedding the classifier and estimating the density $\hat p(\hat s|\theta)$ and the relative merits of those approaches will depend critically on the dimensionality of $\theta$ and the computational cost of the generative model.
-We leave a more general strategy for this overarching optimization problem as an area of future work.
 
@@
 
-%Thus far we have shown that likelihood ratio tests based on $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via hypothesis tests based on the univariate densities $p(s|\theta)$ for the particular dimensionality reduction map $s(x|\theta_0, \theta_1)$.
-% Often it is not possible to evaluate the density $p(x|\theta)$ at a given point $x$.  
-%This approach will not be useful if it is not possible to approximate $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ without evaluating the density $p(x|\theta)$. 
-%In order for this approach to be useful, we need to be able to approximate both based on samples $\{(x,\theta)\}$ drawn from the generative model $p(x|\theta)$.  
 In order for this approach to be useful in the likelihood-free setting, where it is not possible to evaluate the density $p(x|\theta)$, we need to be able to approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
 Denote the  dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{p}(\hat{s}|\theta)$.
-
-%In general we will be interested in two machine learning problems associated to  that approximates these distributions based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
 
@@
 This point was made by \cite{Neal:2007zz} and is well appreciated by the particle physics community that typically takes a conservative attitude towards the use of machine learning classifiers precisely due to concerns about the calibration $p$-values in the face of nuisance parameters associated to the simulator.
-
-%The first step in this direction is to confirm that a discriminative classifier obtained from common supervised learning algorithms will yield a function that is one-to-one with $p(x|\theta_0)/p(x|\theta_1)$.
-%In particular, 
-%is there a loss function such that the function $\hat{s}(x)$ that minimizes the expected loss leads to a function
-%that is one-to-one with $p(x|\theta_0)/p(x|\theta_1)$?
 
@@
 Thus, standard supervised learning algorithms with various surrogate loss functions lead to discriminative classifiers that approximate a monotonic function of per-event likelihood ratio $q(x)$.
-
-%Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
-histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter points.
-In order to construct an approximate likelihood function continuous in the parameters $\theta$, HEP experiments have typically used some form of interpolation of the univariate score density over the parameters $\theta$~\citek{Cranmer:2012sba}.
-
-%The relative merits of various interpolation algorithms and the tradeoff between more samples from the generative model at a particular set of parameter points and fewer samples at more parameter points is left as a subject of future work.
-
-%
-%If we use the squared-loss function, then the expected loss is
-%\begin{equation}
-%\mathbb{E}[L] = \int dx p(x|\theta_0)  (  \hat{s}(x) )^2  + \int dx p(x|\theta_1)  ( 1 - \hat{s}(x) )^2 \; .
-%\end{equation}
-%The function $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ minimizes this expected loss.
-%
-%Proof (Sketch): Consider a variation about $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ given by $\hat{s}'(x) = \hat{s}(x) +h(x)$. The change in the expected loss is given by 
-%\begin{equation}
-%\Delta = \mathbb{E}[L_{\hat{s}'}]-\mathbb{E}[L_{\hat{s}}] = \int dx p(x|\theta_0)  (h^2(x) + 2 h(x)  \hat{s}(x) )  + p(x|\theta_1)   (h^2(x) - 2 h(x) + 2 h(x) \hat{s}(x)  \; .
-%\end{equation}
-%using $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ we obtain
-%\begin{equation}
-%\Delta =  \int dx (p(x|\theta_0)+p(x|\theta_1))  h^2(x) >0    \; .
-%\end{equation}
-%Thus any variation on $\hat{s}(x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ has a larger expected loss.
-
-%The conclusion is that standard classification with a quadratic loss function and training data as described above will approximate a discriminative classifier needed to produce an equivalent likelihood ratio test.
-
-%Once the classifier is trained, we can use the generative model and any univariate density estimation technique (e.g.
-histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$.
-
-%%% KC, commenting this out late in the game
-%In the  limit of large samples from the generative model,  we can approximate the original likelihood ratio test arbitrarily well.
-With finite training data for $\hat{s}(x)$ and samples to approximate $\hat{p}(\hat{s}|\theta)$ it will be necessary to be more specific about the what loss function we are interested in for approximating the likelihood ratio test.
-This will depend in general on the ultimate goal of the test.
-We know that in the case of composite hypothesis tests that there is in general no uniformly most powerful test, thus it is likely that a decision theoretic approach taking into account some weighting or utility over the space $\Theta$ is necessary.
-This is left as a subject for future work.
 
@@
 
-%A  straightforward approach to parameter inference in the likelihood-free setting is to approximate the density $\hat{p}(x|\theta)$.
-In general, approximating the density $\hat{p}(x|\theta)$ is difficult for high-dimensional data, which motivates the use of the dimensionality reduction map $\hat{s}(x)$ and likelihood ratio tests based on the univariate density $\hat{p}(\hat{s}|\theta)$.
-%In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$.
-In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$; however, we can  embed the classifier as in Eq.~\ref{eq:embedding}.
-% into the likelihood function to carry out the composition $\hat{p}\circ \hat{s}$.
-The critical observation is that  if we postpone the evaluation of the classifier to the stage of evaluating the enveloping likelihood ratio, then we can identify the parameters $\theta$ that are being compared in the likelihood ratio with the parameters used to parametrize $s(x;\theta_0, \theta_1)$.
-
 Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
@@
 
-%Generalized likelihood ratio tests are used widely in HEP ~\citek{Cowan:2010js}, most notably in the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa}.
-
 \subsection{Typical usage pattern}
 
-%Often machine learning classification algorithms are trained on 
 In this setting, large samples of pseudo-data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density $p_s(x|\nu_0)$.
@@
 The resulting classifier approximates the regression function $p_s(x|\nu_0)/(p_s(x|\nu_0)+p_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$.
-Associating the $y=1$ label to the signal component has advantages because it helps the classifier focus its capacity on the relevant regions in the feature space, particularly when the signal is a very small perturbation to the background (ie.
+Associating the $y=1$ label to the signal component has advantages because it helps the classifier focus its capacity on the relevant regions in the feature space, particularly when the signal is a very small perturbation to the background (i.e.
 $\mu \ll 1$).
-%At the LHC one can have $\mu \sim 10^{-6}$. 
-%Note that the classifier is not optimal for $\nu\ne\nu_0$.
 
@@
 \end{equation}
-%where $\mu=0$ is the null (background-only) hypothesis and $\mu>0$ is the alternate (signal-plus-background) hypothesis.\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood.}
-%Typically, we are interested in inference on $\mu$ and $\nu$ are nuisance parameters.
 
@@
 alternate (signal-plus-backgound) since the resulting regression functions are one-to-one with each other.
-% as 
-%\begin{equation}
-% \frac{p(x| 0, \nu_0)}{p(x|\hat \mu, \nu_0)} =  \frac{p_b(x| \nu_0)}{ \mu p_s( x_e \, |\,  \nu_0)  + (1-\mu)\, p_b( x_e \,|\, \nu_0)} = \left[ c_1 + c_2   \frac{p_s(x| \nu_0)}{ p_b( x_e \,|\, \nu_0)} \right ]^{-1} \; ,
-%\end{equation}
-%and $c_1$ and $c_2$ are constants. Specifically, the two  likelihood ratios are in one-to-one correspondence, so an ideal algorithm would lead to equivalent tests. 
 In contrast, in the case of the generalized likelihood ratio test 
@@
 \end{equation}
-%In the case of particle physics, the distributions $p(x|\theta)$ is not a Gaussian Mixture Model, but mixture of  complicated distributions associated to relatively few types of particle interactions.
-Moreover, when searching for a new particle, the null hypothesis would correspond to some of the coefficients $w_c=0$ while the alternate ``signal-plus-background'' hypothesis would have $0<w_{c\in \rm signal} \ll w_{c\in \rm background}$.
-In some cases $w_{c\in \rm signal} / w_{c\in \rm background} <10^{-6}$, which means the alternate hypothesis is a small perturbation to the null hypothesis.
-This can be a challenge for typical classifiers because they should devote their capacity to the region where $p_{c\in \rm signal}(x) / p_{c\in \rm background}(x)$ is relatively large.
-Lastly, even when the distributions $p_c(x|\theta)$ are well known, it is often the case that the coefficients are uncertain or treated as completely unknown.
-These all present challenges to machine learning algorithms that aim to learn $s(x; \theta_0, \theta_1)$.
-%
-%\begin{equation}
-%\frac{p(x|\theta_0)}{p(x|\theta_1)}= \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_c w_c(\theta_1) p_c(x| \theta_1)} 
-%\end{equation}
-%
-%\begin{equation}
-%\frac{p(x|\theta_0)}{p(x|\theta_1)}= \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} 
-%\end{equation}
-%
 It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems. 
@@
 Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
-
-%
-%The idea here is to combine the calibration of the distributions of the classifier output and a more optimal family of classifiers $\hat s(x; \nu_0, \nu_1)$.
-Creating the family of classifiers is straight forward, one simply augments the training data with $x$ examples drawn from several values of $\nu$ and then includes the corresponding value of $\nu$ as an input to the classifier.
-Thus $\{x_e,c_e\} \to \{x_e,\nu_{0,e},\nu_{1,e}, c_e\}$ leading to a parametrized learner $\hat s(x)\to \hat s(x;\nu_0, \nu_1)$.
-This leads to a complication: one does not know the value of $\nu$ to use when evaluating the parametrized classifier $\hat s(x;\nu_0, \nu_1)$, so one cannot simply pre-compute $\hat s_e$ before performing the likelihood fit.
-However, when performing a likelihood ratio test, one can calculate the maximum likelihood estimate and conditional maximum likelihood estimate according to Eq.~\ref{eq:mle} and then form the equivalent likelihood ratio test according to Eq.~\ref{eq:equivLRtest}.
-
-%Equivalently, once one has determined $\hat{\theta}$ and $\hat{\hat{\theta}}$, then one can specify the parameters of the mapping $s(x;\hat{\hat{\theta}}, \hat{\theta})$ and then following Eq.~\ref{eq:NP} form the likelihood
-%\begin{equation}\label{eq:parametrizedML}
-%p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( s(x_e;\nu_0, \nu_1) \, |\,  \nu)  + (1-\mu)\, p_b( s(x_e;\nu_0, \nu_1) \,|\, \nu) \,\right] \; ,
-%\end{equation}
-%which is a natural generalization of Eq.~\ref{eq:typicalML}.
-
-%The construction in Eq.~\ref{eq:parametrizedML} presents a computational challenge as for each value of $\nu$ one has a new mapping $s: x\to\mathbb{R}$.
-The most naive realization of this would be to generate synthetic data $\{x_e\}$ according to $p_c(x|\nu)$, evaluate $s(x_e ; \nu_0, \nu_1)$, and then use a histogram, kernel density, or other non-parametric density estimation procedure to estimate $p_c( s | \nu)$.
-This approach is not realistic in situations where $p_c(x|\nu)$ is realized with expensive computer simulations; thus, some interpolation strategy or emulator over a fixed set of parameter points $\{\nu_i\}$ would typically be used.
 
@@
 Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
-%The state of the art approaches for supersymmetric particle masses typically involve high-level features engineered based on physics intuition
-%Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
 The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citek{Gao:2010qx,Bolognesi:2012mm}.
@@
 \section{Related work}\label{S:Related}
-
-%Refs to include:
-%\citek{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,}
 
@@
 
-%Furthermore, this allows one to turn the optimization of the procedure for the per-experiment test into optimization of a per-event classification problem and a calibration procedure.
-
-%In particular, it extends the common usage of machine learning for simple hypothesis testing to include the parameter estimation, confidence intervals, and composite hypothesis testing.
-
-%We have shown that a parametrized family of discriminative classifiers $s(x; \theta_0, \theta_1)$ trained and calibrated with a generative model $p(x|\theta)$ can be used to approximate statistical inference likelihood based on the  ratio $p(x|\theta_0)/p(x|\theta_1)$ when it is not possible to evaluate the densities $p(x|\theta)$ for an arbitrary $x$.
-
-%\section{Possible Extensions}
-%The approach describe thus far has been agnostic as to the type of classifier used to produce $s(x)$ and $s(x; \nu)$.
-In practice, we have used standard packages for the training of these learners.
-In particular, the loss function for these learning algorithms is based on classification performance of individual $(x_e, c_e)$ feature $\to$ target examples.
-However, in overarching statistical procedure that uses the statistical model in Eq.~\ref{eq:typicalML} or \ref{eq:parametrizedML} the quantity we wish to optimize is a property of an entire data set $\{x_e\}$ or ensembles of data sets that can be generated from the statistical models $p_c(x|\theta)$.
-%
-%For example, in typical hypothesis testing problems posed in the Neyman-Pearson context one would be interested in minimizing type II error under a fixed rate of typeI error rate.
-In the presence of systematic uncertainty and nuisance parameters, the loss function may become more complicated.
-The recent Higgs boson machine learning challenge hosted by Kaggle, the evaluation was the approximate median significance (AMS) that characterized the power of a hypothesis test between $\mu=0$ (the background only hypothesis) and $\mu=\mu_0$ (the signal+background hypothesis, where $\mu_0$ is a small number reflecting the small Higgs signal that would be found in addition to the much larger background process) in the presence of uncertainty on the background normalization.
-%Interestingly, the top entries to the challenge did not improve by focusing on this alternative loss function 
-%in training their classifier, but instead used it for the optimization of a threshold on the score (a working point on the ROC curve) that was related to the way the challenge was designed.
-This suggests that the standard
-%classification loss functions lead to similar optimizations and that refinements for the custom loss functions
-%are not easy to come by.
-%
-%More generally, there is an avenue of research associated to custom loss functions specifically designed 
-%for these real world statistical problems. One direction might be to choose loss functions that make 
-%some balance between the standard classification loss, take into account robustness with respect to nuisance parameters, and population-level quantities like type I and II error.
-%
-%\subsection{A different starting point}
-%
-%Instead of evolving from the traditional usage of ML in HEP, let's start from scratch with $p_c(x|\theta)$ and a well defined objective.  
-%
-%\subsubsection{Simple Hypothesis Testing}
-%
-%Consider the situation of simple hypothesis testing between the background-only $\mu=0$ and signal+background hypothesis $\mu=\mu_0=\nu_s/(\nu_s+\nu_b)$ for $\theta=\theta_0$.
-Let us assume that not only can we use the $p_c(x|\theta_0)$ as a generative model, but that we can evaluate the probability density at any point $x$.
-The statistical model is given by
-%\begin{equation}\label{eq:NP}
-%p( D \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu p_1( x_e \, |\,  \theta)  + (1-\mu)\, p_b( x_e \,|\, \theta) \,\right] \; .
-%\end{equation}
-%The objective is to find a  test statistic $T(D)$ (a map from the space of the data to $\mathbb{R}$)  
-%and a threshold $k_\alpha$ that minimizes the rate of type-II error (i.e.
-$P(T(D) < k_\alpha | \mu=\mu_0, \theta_0) = \beta$ under the constraint of a fixed rate of type-I error ( i.e.
-$P(T(D) > k_\alpha | \mu=0, \theta_0) = \alpha$ ).
-%The Neyman-Pearson lemma states that the optimal test statistic is the likelihood ratio, or equivalently, the log-likelihood ratio 
-%\begin{equation}
-%T =  \log \frac{p(D | \mu=\mu_0, \theta_0)}{p(D | \mu=0, \theta_0)}
-%\end{equation}
-%When the $x_e$ are i.i.d., then the log-likelihood ratio expands into a sum
-%\begin{equation}
-%T =   \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {p_1(x_e | \theta_0) }{ p_b(x_e | \theta_0) } \right]}_{q(x_e)} + c_2\;,
-%\end{equation}
-%where $c_1=\mu/(1-\mu)$ and $c_2=\lop(1-\mu)$ are constants for the simple hypothesis test.
-%
-%Here we see that the optimal $T$ for the experiment is composed of a sum over events of a linear linear function of the per-event function $q(x)$.
-A monotonic, but non-linear function of $q(x)$ would not lead to an equivalent hypothesis test.
-%
-%The important part of the per-event function $q(x)$ is that it defines contours in the feature space $x$.
-These contours are also equivalent to the function $p_1(x|\theta_0)/p_b(x|\theta_0)$, the signal-to-background ratio.
-If we can find a machine learner $s(x)$ that is a monotonic function of $p_1(x|\theta_0)/p_b(x|\theta_0)$ it will share the same contours.
-Then the remaining challenge is to find the appropriate rescaling that gives back  linear function $q(x)$.
-%
-%\textbf{Postulate:}
-%\begin{equation}
-%T' = \sum_{e=1}^n \underbrace{\log \left[ 1+c_1\frac {p_1(s(x_e) | \theta_0) }{ p_b(s(x_e) | \theta_0) } \right]}_{q(s_e)} \;,
-%\end{equation}
-%leads to an equivalent test, where $s_e \equiv s(x_e)$
-%\begin{equation}
-%p_c(s) = \int dx \, \delta(s-s(x)) \, p_c(x)  \,  |\partial s / \partial x|^{-1}
-%\end{equation}
-%
-%
-%
-%Note, in order to form the hypothesis test, one still needs to build the build the $p_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$. 
-%
-%
-%
-%\textbf{Proof (in progress):}
-%
-%We want to show density is the same 
-%
-%\begin{equation}
-%p_c(q_x) = \int dx \delta(q_x-q_x(x)) p_c(x) / |\partial q_x / \partial x|
-%\end{equation}
-%
-%\begin{equation}
-%p_c(q_s) = \int dx \delta(q_s-q_s(s(x))) \, p_c(x) \, / |\partial q_s / \partial x|
-%\end{equation}
-%sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $p_1(x)/p_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
-%\begin{equation}
-%\frac {p_1(c | \theta_0) }{ p_0(c | \theta_0) } = 
-%\frac { \int dx \delta(c-s(x)) \, p_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  } = 
-%\frac {c \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, p_0(x | \theta_0)  } = c =\frac {p_1(x | \theta_0) }{ p_0(x | \theta_0) } 
-%\end{equation}
-%
-%
-%
-%%\begin{equation}
-%%p_c(q_s) = \int ds \delta(q_s-q_s(s)) \,p_c(s) \, / |\partial q_s / \partial s|
-%%\end{equation}
-%
-%
-%This is the test statistic that one would naturally get from constructing the likelihood function for the 1-d distributions of the output of the machine learner
-%\begin{equation}\label{eq:NP}
-%p( \{s_e\} \,|\, \mu, \theta) = \prod_{e=1}^n \, \left[\, \mu p_1( s_e \, |\,  \theta)  + (1-\mu)\, p_0( s_e \,|\, \theta) \,\right] \; .
-%\end{equation}
-%Thus this motivates Eq.~\ref{eq:typicalML} in the more general case. 
-%
-%
-%Question:
-%Where does this fit in:
-%\begin{equation}
-%T'' = \sum_{e=1}^n \underbrace{ \log \left[ 1+c_1 s(x_e) \right] }_{q'(s_e)} \;,
-%\end{equation}
-%
-%
-%
-%Usefulness:
-%If we can readily compute $p_c(x|\theta_0)$, then there would be no need for machine learning algorithms.
-However,  these densities are often difficult to compute though one can readily use them as a generative model to produce samples.
-In that case, one might wish to use a machine learning algorithm to find $s(x) \approx p_1(x|\theta_0) / p_0(x|\theta_0)$.
-%
-%Conclusion, while it is possible to start directly with the Neyman-Pearson loss function of type-II error under the constraint of fixed size and try to find a per-experiment learner $T(D)$, it is more convenient to cast the problem in terms of learning a per-event function $s(x)$, and then calibrate it to be a linear function of $q(x)$.
-%
-%\newpage
-%
-%Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $p_1(x)$ and $p_0(x)$ define 
-%\begin{equation} 
-%s(x)=\frac{p_1(x)}{p_2(x)}
-%\end{equation}.
-%Define $p_1(c) = \int dx \delta(c - s(x) ) \, p_1(x)$ and $p_0(c) = \int dx \delta(c - s(x) ) \, p_0(x)$.
-%I wish to show
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)}
-%\end{equation}
-%$\forall x \ni s(x)=c$.
-%
-%Sketch of Proof: 
-%let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$.
-The  coordinate perpendicular to $\Omega$ can be written $y = x+\epsilon \hat{n}$.
-In these coordinates, the integral $\int dx \to \int d\Omega d\epsilon$, thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon|
-%\end{equation}
-%Similarly for $p_0(c)$. We can factor out of the integral $s(x)=p_1(x)/p_0(x)$ since it is constant over $\Omega$.
-%Thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|
-%\end{equation}
-%and we arrive at the result
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{p_1(x)}{p_0(x)} \;.
-%\end{equation}
-%
-%
-%
-%
-%\subsection{misc notes}
-%
-%
-%
-%
-%The distribution of $s$ is fully determined by the mapping $s(x_e; \theta_0, \theta_1)$ and the distribution $p(x)$, which we can write
-%\begin{equation}
-%p(s_e) = \int dx \, \delta(s_e-s(x)) \, p(x)  \; .
-%\end{equation}
-%\textbf{NEED Jacobian for x $\to$ s independent of delta function part, or double counting?}
-%
-%
-%\textbf{Theorem 1:}
-%Let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the orthonormal vector to the surface at the point $x$.
-The component perpendicular to $\Omega_c$ can be written $x+\epsilon \hat{n}$.
-In these coordinates, we rewrite integral $\int dx \to \int d\Omega_c d\epsilon$, thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega_c p_1(x)  / | \hat{n} \cdot \nabla q_s  |
-%\end{equation}
-%Similarly for $p_0(c)$. We can factor out of the integral $s(x)=p_1(x)/p_0(x)$ since it is constant over $\Omega_c$.
-%Thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega p_0(x)  / | \hat{n} \cdot \nabla q_s  |
-%\end{equation}
-%and we arrive at the result
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{p_1(x)}{p_0(x)} \;\hspace{3em} \forall x\in\Omega_c.
-%\end{equation}
-%
-%
-%
-%
-%
-%Define $p_1(c) = \int dx \delta(c - s(x) ) \, p_1(x)$ and $p_0(c) = \int dx \delta(c - s(x) ) \, p_0(x)$.
-%I wish to show
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)}
-%\end{equation}
-%$\forall x \ni s(x)=c$.
-%
-%
-%Given $x\in \mathbb{R}^n$ and two smooth, real-valued functions $p_1(x)$ and $p_0(x)$ define 
-%\begin{equation} 
-%s(x)=\frac{p_1(x)}{p_0(x)}
-%\end{equation}.
-%Define $p_1(c) = \int dx \delta(c - s(x) ) \, p_1(x)$ and $p_0(c) = \int dx \delta(c - s(x) ) \, p_0(x)$.
-%I wish to show
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{p_1(x)}{p_0(x)}
-%\end{equation}
-%$\forall x \ni s(x)=c$.
-%
-%Sketch of Proof: 
-%let $\Omega_{c}$ be the level set $\{x \mid s(x) = c \}$ and $\hat{n}=\nabla s(x) / |\nabla s(x)|$ be the perpendicular direction to the surface at the point $x$.
-The  coordinate perpendicular to $\Omega$ can be written $y = x+\epsilon \hat{n}$.
-In these coordinates, the integral $\int dx \to \int d\Omega d\epsilon$, thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon|
-%\end{equation}
-%Similarly for $p_0(c)$. We can factor out of the integral $s(x)=p_1(x)/p_0(x)$ since it is constant over $\Omega$.
-%Thus
-%\begin{equation}
-%p_1(c) = \int dx \delta(c-s(x)) p_1(x) = \int d\Omega p_1(x) / |\partial s(x)/\partial \epsilon| = s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|
-%\end{equation}
-%and we arrive at the result
-%\begin{equation}
-%\frac{p_1(c)}{p_0(c)} = \frac{s(x) \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|}{ \int d\Omega p_0(x) / |\partial s(x)/\partial \epsilon|} = \frac{p_1(x)}{p_0(x)} \;.
-%\end{equation}
-%
-%
-%
-%
-%
-%
-%
-%Note, in order to form the hypothesis test, one still needs to build the build the $p_0(T|\theta)$ to calibrate the threshold  $T>k_\alpha$ that gives rise to a test of size $\alpha$. 
-%
-%
-%
-%\textbf{Proof (in progress):}
-%
-%We want to show density is the same 
-%
-%\begin{equation}
-%p(q_x|\theta) = \int dx \delta(q_x-q_x(x)) p(x|\theta)  / | \hat{n} \cdot \nabla q_s  |
-%\end{equation}
-%
-%\begin{equation}
-%p(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, p(x|\theta) \, / | \hat{n} \cdot \nabla q_s  |
-%\end{equation}
-%sufficient to show $q(x_e) = q(s(x_e))$.  Here we use that $p_1(x)/p_0(x)=\textrm{const}$ for all points on the level-set $\Omega_c$ given by $ \delta(c-s(x))$. 
-%\begin{equation}
-%\frac {p_1(c | \theta_0) }{ p_0(c | \theta_0) } = 
-%\frac { \int dx \delta(c-s(x)) \, p_1(x_e | \theta_0) }{ \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  } = 
-%\frac {c \int dx \delta(c-s(x)) \, p_0(x | \theta_0)  }{ \int dx \delta(s-s(x)) \, p_0(x | \theta_0)  } = c =\frac {p_1(x | \theta_0) }{ p_0(x | \theta_0) } 
-%\end{equation}
-%
-%
-%
-%%\begin{equation}
-%%p_c(q_s) = \int ds \delta(q_s-q_s(s)) \,p_c(s) \, / |\partial q_s / \partial s|
-%%\end{equation}
-%
-%
-
 \section*{Acknowldgements}

### 2016-01-11 jasa/paper-jasa.tex
Bootstrap paper with JASA template

--- 2015-06-05 tex/parametrized-learning-aoas.tex
@@
-\documentclass[aoas,preprint]{imsart}
+\documentclass[12pt]{article}
+\usepackage{graphicx}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{url} % not crucial - just used below for the URL
 
-\RequirePackage[OT1]{fontenc}
-\RequirePackage{amsthm,amsmath}
-\RequirePackage{natbib}
-%\RequirePackage[nameyear]{natbib}
-\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{1}
 
-\usepackage{amssymb, amsmath}
-\usepackage{oubraces}
-\usepackage{mathtools}
-\usepackage{algorithm}
-\usepackage{algorithmic}
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{-.3in}%
+\addtolength{\topmargin}{-.8in}%
+
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{theorem}
 \usepackage{xfrac}
@@
 \DeclareMathOperator*{\argmax}{arg\,max}
-\newcommand{\citek}[1]{[\cite{#1}]}
 \newcommand{\sstar}{s^*}
@@
 
-% settings
-%\pubyear{2005}
-%\volume{0}
-%\issue{0}
-%\firstpage{1}
-%\lastpage{8}
-\arxiv{arXiv:0000.0000}
-
-\startlocaldefs
 \numberwithin{equation}{section}
@@
 \newtheorem{thm}{Theorem}[section]
-\endlocaldefs
 
@@
 
-\begin{frontmatter}
-%\title{Approximating Generalized Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
-%\title{Approximating  Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
-\title{Approximating  Likelihood Ratios with Calibrated Discriminative Classifiers}
-%\title{Approximating Likelihood Ratio Tests with Calibrated Classifiers}
-%\title{A Sample Document\thanksref{T1}}
-\runtitle{Approximating Likelihood Ratios with Classifiers}
-%\thankstext{T1}{Footnote to the title with the ``thankstext'' command.}
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
 
-\begin{aug}
-\author{\fnms{Kyle} \snm{ Cranmer}\thanksref{t1}\ead[label=e1]{kyle.cranmer@nyu.edu}}
-%,
-%\author{\fnms{Second} \snm{Author}\thanksref{t3,m1,m2}\ead[label=e2]{second@somewhere.com}}
-%\and
-%\author{\fnms{Third} \snm{Author}\thanksref{t1,m2}
-%\ead[label=e3]{third@somewhere.com}
-%\ead[label=u1,url]{http://www.foo.com}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\thankstext{t1}{US National Science Foundation grants PHY-0854724 and PHY-0955626.}
-%\thankstext{t2}{First supporter of the project}
-%\thankstext{t3}{Second supporter of the project}
-\runauthor{K. Cranmer}
+\if1\blind
+{
+  \title{\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers}
+  \author{Kyle Cranmer and Gilles Louppe\\
+          Center for Cosmology and Particle Physics, New York Univeristy}
+  \maketitle
+} \fi
 
-\affiliation{Center for Cosmology \and Particle Physics\thanksmark{m1} \\
-       \and Center for Data Science\thanksmark{m2}
-       New York University}
+\if0\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Title}
+\end{center}
+  \medskip
+} \fi
 
-\address{Center for Cosmology and Particle Physics\\
-       New York University \\
-       New York, NY 10003, USA
-\phantom{E-mail:\ }\printead*{e1}}
-\end{aug}
+\bigskip
+\begin{abstract}
 
-\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-In particle physics likelihood ratio tests are established tools for statistical inference. 
-These tests are complicated by the fact that computer simulators are used as a generative model for 
-the data, but they do not provide a way to evaluate the likelihood function. 
-We demonstrate how discriminative classifiers can be used to approximate the likelihood function when 
-a generative model for the data is available for training and calibration.
-This offers an approach to parametric inference when simulators are used that is complementary to approximate Bayesian computation.
+In particle physics likelihood ratio tests are established tools for statistical
+inference.  These tests are complicated by the fact that computer simulators are
+used as a generative model for the data, but they do not provide a way to
+evaluate the likelihood function. We demonstrate how discriminative classifiers
+can be used to approximate the likelihood function when a generative model for
+the data is available for training and calibration.  This offers an approach to
+parametric inference when simulators are used that is complementary to
+approximate Bayesian computation.
+
 \end{abstract}
 
-\begin{keyword}[class=MSC]
-\kwd[Primary ]{62P35}
-\kwd{62F99}
-\kwd[; secondary ]{62H30}
-\end{keyword}
+\noindent%
+{\it Keywords:}  likelihood ratio, classification, particle physics
+\vfill
 
-\begin{keyword}
-\kwd{likelihood ratio}
-\kwd{classification}
-\kwd{particle physics}
-\end{keyword}
-
-\end{frontmatter}
+\newpage
+\spacingset{1.45} % DON'T change the spacing!
 
 \section{Introduction}
-%\subsection{}
 
@@
 Directly evaluating the likelihood ratio in these cases is often impossible or is computationally impractical.
-Approximate Bayesian Computation (ABC) is one approach to parameter inference in this simulation-based or likelihood-free setting~\citek{Rubin1984,Tavare1997,Marin2011}.
+Approximate Bayesian Computation (ABC) is one approach to parameter inference in this simulation-based or likelihood-free setting~\cite{Rubin1984,Tavare1997,Marin2011}.
 Here we consider an alternative approach that can also be used in a classical setting where a prior over the parameters is not available.
@@
 As a concrete example, consider searches for new particles at the Large Hadron Collider.
-The simulator that is sampling from $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.
+The simulator that is sampling from $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\cite{Sjostrand:2006za,Agostinelli:2002hh}.
 
@@
 final  result was formulated as a hypothesis test or confidence interval
-using a generalized  likelihood ratio test~\citek{Cowan:2010js}. This includes
-the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa} and 
+using a generalized  likelihood ratio test~\cite{Cowan:2010js}. This includes
+the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa} and
 subsequent measurement of its properties.
@@
 (labeled \textit{background}). Typically,  pseudo-data from the simulator are used to approximate the density
-at various parameter points, and an interpolation algorithm is used to approximate the parametrized model~\citek{Cranmer:2012sba}.
+at various parameter points, and an interpolation algorithm is used to approximate the parametrized model~\cite{Cranmer:2012sba}.
 
 To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers that take advantage of a high dimensional feature vector $x$.
-Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citek{Hocker:2007ht} are commonly used.
-Recently, there has been progress in using deep networks~\citek{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citek{HepML}, the largest Kaggle challenge in history.
+Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\cite{Hocker:2007ht} are commonly used.
+Recently, there has been progress in using deep networks~\cite{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\cite{HepML}, the largest Kaggle challenge in history.
 
-While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citek{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters.
-As noted in ~\citek{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
+While classification accuracy can lead to optimal approaches for simple hypothesis tests~\cite{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters.
+As noted in ~\cite{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
 The key distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g.
@@
 
-Vast literature exists around generative and discriminative classifiers~\citek{AndrewY.Ng}.
+Vast literature exists around generative and discriminative classifiers~\cite{AndrewY.Ng}.
 Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule.
@@
 We denote the learned family of classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based roughly on the following algorithmic flow.
-\begin{algorithm}[ht]
-\caption{Training of the parametrized classifier.}\label{alg:training}
-\begin{algorithmic}
-\STATE initialize trainingData to \{\}
-\FOR{ $\theta_0$ in $\Theta$ }
-	\FOR{ $\theta_1$ in $\Theta$ }
-		\STATE generate $x_i^0 \sim p(x|\theta_0)$
-		\STATE append $\{ (x_i^0, \theta_0, \theta_1, y=0) \}$ to trainingData
-		\STATE generate $x_i^1 \sim p(x|\theta_1)$
-		\STATE append $\{ (x_i^1, \theta_0, \theta_1, y=1) \}$ to trainingData
-	\ENDFOR
-\ENDFOR
-\STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_1)$
-\end{algorithmic}
-\end{algorithm}%\vspace{-1.5em}
+% \begin{algorithm}[ht]
+% \caption{Training of the parametrized classifier.}\label{alg:training}
+% \begin{algorithmic}
+% \STATE initialize trainingData to \{\}
+% \FOR{ $\theta_0$ in $\Theta$ }
+% 	\FOR{ $\theta_1$ in $\Theta$ }
+% 		\STATE generate $x_i^0 \sim p(x|\theta_0)$
+% 		\STATE append $\{ (x_i^0, \theta_0, \theta_1, y=0) \}$ to trainingData
+% 		\STATE generate $x_i^1 \sim p(x|\theta_1)$
+% 		\STATE append $\{ (x_i^1, \theta_0, \theta_1, y=1) \}$ to trainingData
+% 	\ENDFOR
+% \ENDFOR
+% \STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_1)$
+% \end{algorithmic}
+% \end{algorithm}%\vspace{-1.5em}
 
@@
 This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
-In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$~\citek{Cranmer:2012sba}.
+In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$~\cite{Cranmer:2012sba}.
 
@@
 Thus one must implement the embedding of the classifier as in Eq.~\ref{eq:embedding}.
-A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
+A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\cite{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
@@
 New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
-generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\citek{Cowan:2010js}.
-This approach was used for the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa}.
+generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\cite{Cowan:2010js}.
+This approach was used for the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa}.
 
@@
 \begin{center}
- \includegraphics[height=1.7in]{example-TMVA-BDT.pdf}
- \includegraphics[height=1.7in]{example-TMVA-ROC.pdf}
+ \includegraphics[height=1.7in]{figures/example-TMVA-BDT.pdf}
+ \includegraphics[height=1.7in]{figures/example-TMVA-ROC.pdf}
 \caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).
@@
 This formalism represents a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
-The work of  ~\citek{Whiteson:2006ws} is similar as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
+The work of  ~\cite{Whiteson:2006ws} is similar as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
 This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
 
-Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citek{Volobouev:2011vb}.
+Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\cite{Volobouev:2011vb}.
 In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method.
@@
 It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage.
-In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citek{Aaltonen:2010yz}).
+In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\cite{Aaltonen:2010yz}).
 
-Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citek{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citek{Allanach:2000kt}.
+Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\cite{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\cite{Allanach:2000kt}.
 Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
-The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citek{Gao:2010qx,Bolognesi:2012mm}.
+The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\cite{Gao:2010qx,Bolognesi:2012mm}.
 
@@
 Perhaps a  formal treatment similar to the Neyman-Pearson case can be brought to bear in this more general setting.
+In a similarly titled work, \cite{Gutmann2014} advocate using the cross-validated classification accuracy as the similarity metric used in ABC.
+While the goal there is also parameter inference in the likelihood-free setting,  ``classifier ABC'' is very different than the approach presented here.
 \cite{TommiJaakkola} explore a way of leveraging generative models to derive kernel functions for use in discriminative methods.
@@
 
-\section*{Acknowldgements}
-KC would like to thank  
-Daniel Whiteson for encouragement throughout the project and
-Alex Ihler for challenging discussions that 
-led to a reformulation of the initial idea.
-KC would also like to thank Shimon Whiteson, Babak Shahbaba for advice in the presentation, Radford Neal for discussion of his earlier work, Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
-project early in its conception, Bal\'azs K\'egl for discussions about the Kaggle challenge and feedback to the draft, 
-and Yuri Shirman for reassuring cross checks of the Theorem. 
-KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 
-KC is grateful to UC-Irvine for their hospitality while this research was carried out and the 
-Moore and Sloan foundations for their generous support of the data science environment at NYU.
+\section{References}
 
-\newpage
-
-%\bibliographystyle{natbib}
-\bibliographystyle{imsart-nameyear}
+\bibliographystyle{apalike}
 \bibliography{learning.bib}
-%\begin{thebibliography}
-%\parametrized-learning}
-%\end{thebibliography
 

### 2016-01-12 jasa/paper-jasa.tex

@@
 \usepackage{xfrac}
+\usepackage{hyperref}
 
@@
 Directly evaluating the likelihood ratio in these cases is often impossible or is computationally impractical.
-Approximate Bayesian Computation (ABC) is one approach to parameter inference in this simulation-based or likelihood-free setting~\cite{Rubin1984,Tavare1997,Marin2011}.
+Approximate Bayesian Computation (ABC) is one approach to parameter inference in this simulation-based or likelihood-free setting~\citep{Rubin1984,Tavare1997,Marin2011}.
 Here we consider an alternative approach that can also be used in a classical setting where a prior over the parameters is not available.
@@
 As a concrete example, consider searches for new particles at the Large Hadron Collider.
-The simulator that is sampling from $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\cite{Sjostrand:2006za,Agostinelli:2002hh}.
+The simulator that is sampling from $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citep{Sjostrand:2006za,Agostinelli:2002hh}.
 
@@
 final  result was formulated as a hypothesis test or confidence interval
-using a generalized  likelihood ratio test~\cite{Cowan:2010js}. This includes
-the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa} and
+using a generalized  likelihood ratio test~\citep{Cowan:2010js}. This includes
+the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and
 subsequent measurement of its properties.
@@
 (labeled \textit{background}). Typically,  pseudo-data from the simulator are used to approximate the density
-at various parameter points, and an interpolation algorithm is used to approximate the parametrized model~\cite{Cranmer:2012sba}.
+at various parameter points, and an interpolation algorithm is used to approximate the parametrized model~\citep{Cranmer:2012sba}.
 
 To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers that take advantage of a high dimensional feature vector $x$.
-Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\cite{Hocker:2007ht} are commonly used.
-Recently, there has been progress in using deep networks~\cite{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\cite{HepML}, the largest Kaggle challenge in history.
+Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citep{Hocker:2007ht} are commonly used.
+Recently, there has been progress in using deep networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history.
 
-While classification accuracy can lead to optimal approaches for simple hypothesis tests~\cite{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters.
-As noted in ~\cite{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
+While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citep{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters.
+As noted in ~\citep{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
 The key distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g.
@@
 A strength of this approach is that it separates the quality of the approximation of the target likelihood from the quality of the calibration.
-In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.
+In Section~\ref{S:Related} we discuss the scheme sketched by \citep{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.
 
@@
 high energy physics where the simulation of particle detectors can only be done in the `forward mode'.
-This same setting has been considered by \cite{ClaytonScott}, \cite{JMLR:v14:tong13a}, and \cite{Neal:2007zz}.
+This same setting has been considered by \citep{ClaytonScott}, \citep{JMLR:v14:tong13a}, and \citep{Neal:2007zz}.
 
@@
 
-Vast literature exists around generative and discriminative classifiers~\cite{AndrewY.Ng}.
+Vast literature exists around generative and discriminative classifiers~\citep{AndrewY.Ng}.
 Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule.
@@
 Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
-Similar points have been made by~\cite{ClaytonScott} and \cite{Neal:2007zz}.
+Similar points have been made by~\citep{ClaytonScott} and \citep{Neal:2007zz}.
 
@@
 In that case, one might loose power, but the resulting inference will still be valid.
-This point was made by \cite{Neal:2007zz} and is well appreciated by the particle physics community that typically takes a conservative attitude towards the use of machine learning classifiers precisely due to concerns about the calibration $p$-values in the face of nuisance parameters associated to the simulator.
+This point was made by \citep{Neal:2007zz} and is well appreciated by the particle physics community that typically takes a conservative attitude towards the use of machine learning classifiers precisely due to concerns about the calibration $p$-values in the face of nuisance parameters associated to the simulator.
 
@@
 This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
-In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$~\cite{Cranmer:2012sba}.
+In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$~\citep{Cranmer:2012sba}.
 
@@
 Thus one must implement the embedding of the classifier as in Eq.~\ref{eq:embedding}.
-A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\cite{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
+A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
@@
 New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
-generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\cite{Cowan:2010js}.
-This approach was used for the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa}.
+generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\citep{Cowan:2010js}.
+This approach was used for the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
 
@@
 This formalism represents a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
-The work of  ~\cite{Whiteson:2006ws} is similar as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
+The work of  ~\citep{Whiteson:2006ws} is similar as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
 This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
 
-Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\cite{Volobouev:2011vb}.
+Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citep{Volobouev:2011vb}.
 In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method.
@@
 It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage.
-In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\cite{Aaltonen:2010yz}).
+In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citep{Aaltonen:2010yz}).
 
-Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\cite{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\cite{Allanach:2000kt}.
+Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citep{Allanach:2000kt}.
 Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
-The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\cite{Gao:2010qx,Bolognesi:2012mm}.
+The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citep{Gao:2010qx,Bolognesi:2012mm}.
 
@@
 
-\cite{ClaytonScott} and \cite{JMLR:v14:tong13a} consider the machine learning problem associated to Neyman-Pearson hypothesis testing.
+\citep{ClaytonScott} and \citep{JMLR:v14:tong13a} consider the machine learning problem associated to Neyman-Pearson hypothesis testing.
 As in this work, they consider the situation where one does not have access to the underlying distributions, but only has i.i.d.
@@
 Perhaps a  formal treatment similar to the Neyman-Pearson case can be brought to bear in this more general setting.
-In a similarly titled work, \cite{Gutmann2014} advocate using the cross-validated classification accuracy as the similarity metric used in ABC.
+In a similarly titled work, \citep{Gutmann2014} advocate using the cross-validated classification accuracy as the similarity metric used in ABC.
 While the goal there is also parameter inference in the likelihood-free setting,  ``classifier ABC'' is very different than the approach presented here.
-\cite{TommiJaakkola} explore a way of leveraging generative models to derive kernel functions for use in discriminative methods.
+\citep{TommiJaakkola} explore a way of leveraging generative models to derive kernel functions for use in discriminative methods.
 This interesting work is distinct from the point made here in which the generative model is being used for the purpose of providing training data and calibration.
-%\cite{McCallum} consider a hybrid generative/discriminative classifier; however, the goal of that work is not to leverage a generative model for the data, but to use both approaches to learn different subsets of the parameters in a single hybrid classifier.
-\cite{BiancaZadrozny} emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this.
+%\citep{McCallum} consider a hybrid generative/discriminative classifier; however, the goal of that work is not to leverage a generative model for the data, but to use both approaches to learn different subsets of the parameters in a single hybrid classifier.
+\citep{BiancaZadrozny} emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this.
 In contrast to that work, we are not interested in calibrated probability estimates for $p(y|x)$ for individual events, but instead we use the calibration to correct for non-linear transformations of the target likelihood ratio and, perhaps, to provide calibrated p-values based on those likelihood ratio tests.
-\cite{Ihler2004} take on a different problem (tests of statistical independence) by using machine learning algorithms to find  scalar maps from the high-dimensional feature space that achieve the desired statistical goal when the fundamental high-dimensional test is intractable.
+\citep{Ihler2004} take on a different problem (tests of statistical independence) by using machine learning algorithms to find  scalar maps from the high-dimensional feature space that achieve the desired statistical goal when the fundamental high-dimensional test is intractable.
 
-\cite{Neal:2007zz} also considered the problem of approximating the likelihood function when only a generative model is available.
+\citep{Neal:2007zz} also considered the problem of approximating the likelihood function when only a generative model is available.
 That work sketches a scheme in which one uses a classifier with both $x$ and $\theta$ as an input to serve as a dimensionality reduction map.
@@
 
-\section{References}
-
 \bibliographystyle{apalike}

### 2016-01-18 jasa/paper-jasa.tex
abstract

--- 2015-06-05 tex/parametrized-learning-aoas.tex
@@
-\documentclass[aoas,preprint]{imsart}
+\documentclass[12pt]{article}
+\usepackage{graphicx}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{url} % not crucial - just used below for the URL
 
-\RequirePackage[OT1]{fontenc}
-\RequirePackage{amsthm,amsmath}
-\RequirePackage{natbib}
-%\RequirePackage[nameyear]{natbib}
-\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{1}
 
-\usepackage{amssymb, amsmath}
-\usepackage{oubraces}
-\usepackage{mathtools}
-\usepackage{algorithm}
-\usepackage{algorithmic}
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-.5in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{-.3in}%
+\addtolength{\topmargin}{-.8in}%
+
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{theorem}
 \usepackage{xfrac}
+\usepackage{hyperref}
 
@@
 \newtheorem{thm}{Theorem}[section]
-\endlocaldefs
+
+\newcommand{\glnote}[1]{\textcolor{red}{GL: #1}}
 
@@
 
-\begin{frontmatter}
-%\title{Approximating Generalized Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
-%\title{Approximating  Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
-\title{Approximating  Likelihood Ratios with Calibrated Discriminative Classifiers}
-%\title{Approximating Likelihood Ratio Tests with Calibrated Classifiers}
-%\title{A Sample Document\thanksref{T1}}
-\runtitle{Approximating Likelihood Ratios with Classifiers}
-%\thankstext{T1}{Footnote to the title with the ``thankstext'' command.}
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
 
-\begin{aug}
-\author{\fnms{Kyle} \snm{ Cranmer}\thanksref{t1}\ead[label=e1]{kyle.cranmer@nyu.edu}}
-%,
-%\author{\fnms{Second} \snm{Author}\thanksref{t3,m1,m2}\ead[label=e2]{second@somewhere.com}}
-%\and
-%\author{\fnms{Third} \snm{Author}\thanksref{t1,m2}
-%\ead[label=e3]{third@somewhere.com}
-%\ead[label=u1,url]{http://www.foo.com}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\thankstext{t1}{US National Science Foundation grants PHY-0854724 and PHY-0955626.}
-%\thankstext{t2}{First supporter of the project}
-%\thankstext{t3}{Second supporter of the project}
-\runauthor{K. Cranmer}
+\if1\blind
+{
+  \title{\bf Approximating Likelihood Ratios with Calibrated Classifiers}
+  \author{Kyle Cranmer and Gilles Louppe\\
+          New York University}
+  \maketitle
+} \fi
 
-\affiliation{Center for Cosmology \and Particle Physics\thanksmark{m1} \\
-       \and Center for Data Science\thanksmark{m2}
-       New York University}
+\if0\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
+  \begin{center}
+    {\LARGE\bf Title}
+\end{center}
+  \medskip
+} \fi
 
-\address{Center for Cosmology and Particle Physics\\
-       New York University \\
-       New York, NY 10003, USA
-\phantom{E-mail:\ }\printead*{e1}}
-\end{aug}
+\bigskip
+\begin{abstract}
 
-\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-In particle physics likelihood ratio tests are established tools for statistical inference. 
-These tests are complicated by the fact that computer simulators are used as a generative model for 
-the data, but they do not provide a way to evaluate the likelihood function. 
-We demonstrate how discriminative classifiers can be used to approximate the likelihood function when 
-a generative model for the data is available for training and calibration.
-This offers an approach to parametric inference when simulators are used that is complementary to approximate Bayesian computation.
+% In particle physics likelihood ratio tests are established tools for statistical
+% inference.  These tests are complicated by the fact that computer simulators are
+% used as a generative model for the data, but they do not provide a way to
+% evaluate the likelihood function. We demonstrate how discriminative classifiers
+% can be used to approximate the likelihood function when a generative model for
+% the data is available for training and calibration.  This offers an approach to
+% parametric inference when simulators are used that is complementary to
+% approximate Bayesian computation.
+
+In particle physics, likelihood ratio tests are established tools for
+statistical inference. In many cases, these tests are complicated by the fact
+that computer simulators are used as a generative model for the data, which does
+not provide a way to evaluate the likelihood function. In this paper, we
+demonstrate that likelihood ratios are invariant under dimensionality reduction,
+provided the transformation is monotonic with the likelihood ratio. As a direct
+consequence, we show why and how discriminative classifiers can be used to
+approximate the likelihood ratio when only a generative model for the data is
+available. In particular, the proposed method offers a machine learning-based
+approach to statistical inference that is complementary to likelihood-free
+Bayesian inference algorithms (e.g., ABC), as it does not require the definition
+of a prior over model parameters.
+
 \end{abstract}
 
-\begin{keyword}[class=MSC]
-\kwd[Primary ]{62P35}
-\kwd{62F99}
-\kwd[; secondary ]{62H30}
-\end{keyword}
+\noindent%
+{\it Keywords:}  likelihood ratio, statistical inference, classification, particle physics
+\vfill
 
-\begin{keyword}
-\kwd{likelihood ratio}
-\kwd{classification}
-\kwd{particle physics}
-\end{keyword}
-
-\end{frontmatter}
+\newpage
+\spacingset{1.45} % DON'T change the spacing!
 
 \section{Introduction}
-%\subsection{}
 
-The likelihood function is the central object that summarizes the information from an experiment needed for inference of model parameters.
-The likelihood function is key to Bayesian inference, and many areas of science that report the results of classical hypothesis tests or confidence intervals use the (generalized or profile) likelihood ratio as a test statistic.
-It is increasingly common that a simulator (or generative model) is used to describe complex processes that tie 
-parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
-Directly evaluating the likelihood ratio in these cases is often impossible or is computationally impractical. 
-Approximate Bayesian Computation (ABC) is one approach to parameter inference in this simulation-based or likelihood-free setting~\citek{Rubin1984,Tavare1997,Marin2011}.
-Here we consider an alternative approach that can also be used in a classical setting where a prior over the parameters is not available.
-In particular, we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when a generative model for the data is available for training and calibration.
+The likelihood function is the central object that summarizes the information
+from an experiment needed for inference of model parameters. The likelihood
+function is key to Bayesian inference, and many areas of science that report the
+results of classical hypothesis tests or confidence intervals use the
+(generalized or profile) likelihood ratio as a test statistic. It is
+increasingly common that a simulator (or generative model) is used to describe
+complex processes that tie parameters $\theta$ of an underlying theory and
+measurement apparatus to high-dimensional observations $x$. Directly evaluating
+the likelihood ratio in these cases is often impossible or is computationally
+impractical. Approximate Bayesian Computation (ABC) is one approach to parameter
+inference in this simulation-based or likelihood-free
+setting~\citek{Rubin1984,Tavare1997,Marin2011}. Here we consider an alternative
+approach that can also be used in a classical setting where a prior over the
+parameters is not available. In particular, we demonstrate how discriminative
+classifiers can be used to construct equivalent likelihood ratio tests when a
+generative model for the data is available for training and calibration.
 
-As a concrete example, consider searches for new particles at the Large Hadron Collider. 
-The simulator that is sampling from $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.
+As a concrete example, consider searches for new particles at the Large Hadron
+Collider. The simulator that is sampling from $p(x|\theta)$ is based on quantum
+field theory, a detailed simulation of the particle detector, and data
+processing algorithms that transform raw sensor data into the feature vector
+$x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.
 
-The ATLAS and CMS experiments have published  hundreds papers where the 
-final  result was formulated as a hypothesis test or confidence interval 
-using a generalized  likelihood ratio test~\citek{Cowan:2010js}. This includes
-the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa} and 
+The ATLAS and CMS experiments have published  hundreds papers where the final
+result was formulated as a hypothesis test or confidence interval using a
+generalized  likelihood ratio test~\citek{Cowan:2010js}. This includes the
+discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa} and
 subsequent measurement of its properties.
 
-The bulk of the likelihood ratio tests at the LHC are based on the distribution of a single event-level feature
-that discriminates between a hypothesized process of interest (labeled \textit{signal}) and various other processes 
-(labeled \textit{background}). Typically,  pseudo-data from the simulator are used to approximate the density
-at various parameter points, and an interpolation algorithm is used to approximate the parametrized model~\citek{Cranmer:2012sba}.
+The bulk of the likelihood ratio tests at the LHC are based on the distribution
+of a single event-level feature that discriminates between a hypothesized
+process of interest (labeled \textit{signal}) and various other processes
+(labeled \textit{background}). Typically,  pseudo-data from the simulator are
+used to approximate the density at various parameter points, and an
+interpolation algorithm is used to approximate the parametrized
+model~\citek{Cranmer:2012sba}.
 
-To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers that take advantage of a high dimensional feature vector $x$.
-Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citek{Hocker:2007ht} are commonly used.
-Recently, there has been progress in using deep networks~\citek{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citek{HepML}, the largest Kaggle challenge in history.
+To improve the statistical power of these tests, hundreds of these searches have
+utilized supervised learning to train discriminative classifiers that take
+advantage of a high dimensional feature vector $x$. Within High Energy Physics
+(HEP) libraries such as TMVA that implement conventional techniques like the
+multi-layer perceptron and boosted decision trees~\citek{Hocker:2007ht} are
+commonly used. Recently, there has been progress in using deep
+networks~\citek{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons
+learned during HiggsML~\citek{HepML}, the largest Kaggle challenge in history.
 
-While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citek{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters.
-As noted in ~\citek{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
-The key distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g.
-the variance of an estimator for the mass parameter) is a per-experiment operation involving a data set with many events.
-They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process.
+While classification accuracy can lead to optimal approaches for simple
+hypothesis tests~\citek{Dempster1965}, that is no longer true in the context of
+parameter estimation or composite hypothesis tests with nuisance parameters. As
+noted in ~\citek{Whiteson:2006ws}, ``such methods are suboptimal because they
+assume that the selector with the highest classification accuracy will yield a
+mass measurement with the smallest statistical uncertainty.'' The key
+distinction is that evaluating the loss for classification is composed of many
+per-event operations, while evaluating the loss for a mass measurement (e.g. the
+variance of an estimator for the mass parameter) is a per-experiment operation
+involving a data set with many events. They went on to demonstrate a
+computationally intensive stochastic optimization technique based on the
+per-experiment loss out performed the two stage selection-estimation process.
 
-The initial motivation for this work was to extend the typical usage of discriminative classifiers in HEP to be robust to 
-nuisance parameters in the simulators.
-The scope of the result expanded once it became clear that this offers a way to approximate the likelihood function $p(x|\theta)$ in what is typically considered the likelihood-free setting.
-This approach is complementary to ABC as it does not require a prior over the parameters and can also be used in the classical (frequentist) setting.
-A strength of this approach is that it separates the quality of the approximation of the target likelihood from the quality of the calibration.
-In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.
+The initial motivation for this work was to extend the typical usage of
+discriminative classifiers in HEP to be robust to nuisance parameters in the
+simulators. The scope of the result expanded once it became clear that this
+offers a way to approximate the likelihood function $p(x|\theta)$ in what is
+typically considered the likelihood-free setting. This approach is complementary
+to ABC as it does not require a prior over the parameters and can also be used
+in the classical (frequentist) setting. A strength of this approach is that it
+separates the quality of the approximation of the target likelihood from the
+quality of the calibration.  In Section~\ref{S:Related} we discuss the scheme
+sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a
+dimensionality reduction map to aid in the estimation of the likelihood
+function.
 
@@
 
-\section*{Acknowldgements}
-KC would like to thank  
-Daniel Whiteson for encouragement throughout the project and
-Alex Ihler for challenging discussions that 
-led to a reformulation of the initial idea.
-KC would also like to thank Shimon Whiteson, Babak Shahbaba for advice in the presentation, Radford Neal for discussion of his earlier work, Yann LeCun, Philip Stark, and Pierre Baldi for their feedback on the
-project early in its conception, Bal\'azs K\'egl for discussions about the Kaggle challenge and feedback to the draft, 
-and Yuri Shirman for reassuring cross checks of the Theorem. 
-KC is supported by the US National Science Foundation grants PHY-0854724 and PHY-0955626. 
-KC is grateful to UC-Irvine for their hospitality while this research was carried out and the 
-Moore and Sloan foundations for their generous support of the data science environment at NYU.
-
-\newpage
-
-%\bibliographystyle{natbib}
-\bibliographystyle{imsart-nameyear}
+\bibliographystyle{apalike}
 \bibliography{learning.bib}
-%\begin{thebibliography}
-%\parametrized-learning}
-%\end{thebibliography
 

### 2016-01-18 jasa/paper-jasa.tex
Introduction section

@@
 \usepackage{hyperref}
+\usepackage{color}
 
@@
 
-\newcommand{\glnote}[1]{\textcolor{red}{GL: #1}}
+\newcommand{\glnote}[1]{\textcolor{red}{[GL: #1]}}
 
@@
 of a prior over model parameters.
+\glnote{Also mention conclusions on experimental results, as required by JASA guidelines.}
 
@@
 \section{Introduction}
+\label{sec:introduction}
+
+% The likelihood function is the central object that summarizes the information
+% from an experiment needed for inference of model parameters. The likelihood
+% function is key to Bayesian inference, and many areas of science that report the
+% results of classical hypothesis tests or confidence intervals use the
+% (generalized or profile) likelihood ratio as a test statistic. It is
+% increasingly common that a simulator (or generative model) is used to describe
+% complex processes that tie parameters $\theta$ of an underlying theory and
+% measurement apparatus to high-dimensional observations $x$. Directly evaluating
+% the likelihood ratio in these cases is often impossible or is computationally
+% impractical. Approximate Bayesian Computation (ABC) is one approach to parameter
+% inference in this simulation-based or likelihood-free
+% setting~\citep{Rubin1984,Tavare1997,Marin2011}. Here we consider an alternative
+% approach that can also be used in a classical setting where a prior over the
+% parameters is not available. In particular, we demonstrate how discriminative
+% classifiers can be used to construct equivalent likelihood ratio tests when a
+% generative model for the data is available for training and calibration.
 
@@
 from an experiment needed for inference of model parameters. The likelihood
-function is key to Bayesian inference, and many areas of science that report the
-results of classical hypothesis tests or confidence intervals use the
-(generalized or profile) likelihood ratio as a test statistic. It is
-increasingly common that a simulator (or generative model) is used to describe
-complex processes that tie parameters $\theta$ of an underlying theory and
-measurement apparatus to high-dimensional observations $x$. Directly evaluating
-the likelihood ratio in these cases is often impossible or is computationally
-impractical. Approximate Bayesian Computation (ABC) is one approach to parameter
-inference in this simulation-based or likelihood-free
-setting~\citek{Rubin1984,Tavare1997,Marin2011}. Here we consider an alternative
-approach that can also be used in a classical setting where a prior over the
-parameters is not available. In particular, we demonstrate how discriminative
-classifiers can be used to construct equivalent likelihood ratio tests when a
-generative model for the data is available for training and calibration.
+function is key to many areas of science that report the results of classical
+hypothesis tests or confidence intervals using the (generalized or profile)
+likelihood ratio as a test statistic. With the advance of computing technology,
+it has become increasingly common that a simulator (or generative model) is used
+to describe complex processes that tie parameters $\theta$ of an underlying
+theory and measurement apparatus to high-dimensional observations $x$. However,
+directly evaluating the likelihood function in these cases is often impossible
+or is computationally impractical. In this likelihood-free setting, various
+methods for statistical inference have been proposed, including most notably the
+Approximate Bayesian Computation (ABC) class of algorithms aiming at sampling
+from the posterior distribution of model parameters \glnote{Add references}.
 
-As a concrete example, consider searches for new particles at the Large Hadron
-Collider. The simulator that is sampling from $p(x|\theta)$ is based on quantum
-field theory, a detailed simulation of the particle detector, and data
-processing algorithms that transform raw sensor data into the feature vector
-$x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.
+The main result of this paper is to show that the likelihood ratio is invariant
+under dimensionality reduction, under the assumption that the transformation is
+monotonic with the likelihood ratio. As a direct consequence, we propose an
+alternative machine learning-based approach for likelihood-free inference that
+can also be used in a classical (frequentist) setting where a prior over the
+model parameters is not available. More specifically, we demonstrate why and how
+discriminative classifiers can be used to construct equivalent likelihood ratio
+tests when only a generative model for the data is available for training and
+calibration.
 
-The ATLAS and CMS experiments have published  hundreds papers where the final
-result was formulated as a hypothesis test or confidence interval using a
-generalized  likelihood ratio test~\citek{Cowan:2010js}. This includes the
-discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa} and
-subsequent measurement of its properties.
+% As a concrete example, consider searches for new particles at the Large Hadron
+% Collider. The simulator that is sampling from $p(x|\theta)$ is based on quantum
+% field theory, a detailed simulation of the particle detector, and data
+% processing algorithms that transform raw sensor data into the feature vector
+% $x$~\citep{Sjostrand:2006za,Agostinelli:2002hh}.
+%
+% The ATLAS and CMS experiments have published  hundreds papers where the final
+% result was formulated as a hypothesis test or confidence interval using a
+% generalized  likelihood ratio test~\citep{Cowan:2010js}. This includes the
+% discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and
+% subsequent measurement of its properties.
+%
+% The bulk of the likelihood ratio tests at the LHC are based on the distribution
+% of a single event-level feature that discriminates between a hypothesized
+% process of interest (labeled \textit{signal}) and various other processes
+% (labeled \textit{background}). Typically,  pseudo-data from the simulator are
+% used to approximate the density at various parameter points, and an
+% interpolation algorithm is used to approximate the parametrized
+% model~\citep{Cranmer:2012sba}.
+%
+% To improve the statistical power of these tests, hundreds of these searches have
+% utilized supervised learning to train discriminative classifiers that take
+% commonly used. Recently, there has been progress in using deep
+% networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons
+% learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history.
 
-The bulk of the likelihood ratio tests at the LHC are based on the distribution
-of a single event-level feature that discriminates between a hypothesized
-process of interest (labeled \textit{signal}) and various other processes
-(labeled \textit{background}). Typically,  pseudo-data from the simulator are
-used to approximate the density at various parameter points, and an
-interpolation algorithm is used to approximate the parametrized
-model~\citek{Cranmer:2012sba}.
+As a concrete example, let us consider searches for new particles at the Large
+Hadron Collider (LHC). The simulator that is sampling from $p(x|\theta)$ is
+based on quantum field theory, a detailed simulation of the particle detector,
+and data processing algorithms that transform raw sensor data into the feature
+vector $x$~\citep{Sjostrand:2006za,Agostinelli:2002hh}. The ATLAS and CMS
+experiments have published  hundreds papers where the final result was
+formulated as a hypothesis test or confidence interval using a generalized
+likelihood ratio test~\citep{Cowan:2010js}, including most notably the discovery
+of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and subsequent
+measurement of its properties. The bulk of the likelihood ratio tests at the LHC
+are based on the distribution of a single event-level feature that discriminates
+between a hypothesized process of interest (labeled \textit{signal}) and various
+other processes (labeled \textit{background}). Typically, pseudo-data from the
+simulator are used to approximate the density at various parameter points, and
+an interpolation algorithm is used to approximate the parametrized
+model~\citep{Cranmer:2012sba}. In particular, to improve the statistical power
+of these tests, hundreds of these searches have already been using supervised
+learning to train discriminative classifiers that take advantage of a high
+dimensional feature vector $x$.
 
-To improve the statistical power of these tests, hundreds of these searches have
-utilized supervised learning to train discriminative classifiers that take
-advantage of a high dimensional feature vector $x$. Within High Energy Physics
-(HEP) libraries such as TMVA that implement conventional techniques like the
-multi-layer perceptron and boosted decision trees~\citek{Hocker:2007ht} are
-commonly used. Recently, there has been progress in using deep
-networks~\citek{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons
-learned during HiggsML~\citek{HepML}, the largest Kaggle challenge in history.
+% While classification accuracy can lead to optimal approaches for simple
+% hypothesis tests~\citep{Dempster1965}, that is no longer true in the context of
+% parameter estimation or composite hypothesis tests with nuisance parameters. As
+% noted in ~\citep{Whiteson:2006ws}, ``such methods are suboptimal because they
+% assume that the selector with the highest classification accuracy will yield a
+% mass measurement with the smallest statistical uncertainty.'' The key
+% distinction is that evaluating the loss for classification is composed of many
+% per-event operations, while evaluating the loss for a mass measurement (e.g. the
+% variance of an estimator for the mass parameter) is a per-experiment operation
+% involving a data set with many events. They went on to demonstrate a
+% computationally intensive stochastic optimization technique based on the
+% per-experiment loss out performed the two stage selection-estimation process.
+%
+% The initial motivation for this work was to extend the typical usage of
+% discriminative classifiers in HEP to be robust to nuisance parameters in the
+% simulators. The scope of the result expanded once it became clear that this
+% offers a way to approximate the likelihood function $p(x|\theta)$ in what is
+% typically considered the likelihood-free setting. This approach is complementary
+% to ABC as it does not require a prior over the parameters and can also be used
+% in the classical (frequentist) setting. A strength of this approach is that it
+% separates the quality of the approximation of the target likelihood from the
+% quality of the calibration.  In Section~\ref{S:Related} we discuss the scheme
+% sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a
+% dimensionality reduction map to aid in the estimation of the likelihood
+% function.
 
-While classification accuracy can lead to optimal approaches for simple
-hypothesis tests~\citek{Dempster1965}, that is no longer true in the context of
-parameter estimation or composite hypothesis tests with nuisance parameters. As
-noted in ~\citek{Whiteson:2006ws}, ``such methods are suboptimal because they
-assume that the selector with the highest classification accuracy will yield a
-mass measurement with the smallest statistical uncertainty.'' The key
-distinction is that evaluating the loss for classification is composed of many
-per-event operations, while evaluating the loss for a mass measurement (e.g. the
-variance of an estimator for the mass parameter) is a per-experiment operation
-involving a data set with many events. They went on to demonstrate a
-computationally intensive stochastic optimization technique based on the
-per-experiment loss out performed the two stage selection-estimation process.
+The rest of this paper is organized as follows. In Section...
 
-The initial motivation for this work was to extend the typical usage of
-discriminative classifiers in HEP to be robust to nuisance parameters in the
-simulators. The scope of the result expanded once it became clear that this
-offers a way to approximate the likelihood function $p(x|\theta)$ in what is
-typically considered the likelihood-free setting. This approach is complementary
-to ABC as it does not require a prior over the parameters and can also be used
-in the classical (frequentist) setting. A strength of this approach is that it
-separates the quality of the approximation of the target likelihood from the
-quality of the calibration.  In Section~\ref{S:Related} we discuss the scheme
-sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a
-dimensionality reduction map to aid in the estimation of the likelihood
-function.
+\section{Likelihood ratio tests}
+\label{sec:likelihood-ratio-tests}
 
@@
 
-%\newpage
 \subsection{Prelude}
@@
 
-Vast literature exists around generative and discriminative classifiers~\citek{AndrewY.Ng}.
+Vast literature exists around generative and discriminative classifiers~\citep{AndrewY.Ng}.
 Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule.
@@
 This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
-In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$~\citek{Cranmer:2012sba}.
+In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$~\citep{Cranmer:2012sba}.
 
@@
 Thus one must implement the embedding of the classifier as in Eq.~\ref{eq:embedding}.
-A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
+A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
@@
 New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
-generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\citek{Cowan:2010js}.
-This approach was used for the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa}.
+generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\citep{Cowan:2010js}.
+This approach was used for the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
 
@@
 This formalism represents a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
-The work of  ~\citek{Whiteson:2006ws} is similar as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
+The work of  ~\citep{Whiteson:2006ws} is similar as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
 This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
 
-Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citek{Volobouev:2011vb}.
+Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citep{Volobouev:2011vb}.
 In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method.
@@
 It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage.
-In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citek{Aaltonen:2010yz}).
+In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citep{Aaltonen:2010yz}).
 
-Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citek{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citek{Allanach:2000kt}.
+Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citep{Allanach:2000kt}.
 Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
-The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citek{Gao:2010qx,Bolognesi:2012mm}.
+The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citep{Gao:2010qx,Bolognesi:2012mm}.
 

### 2016-01-20 jasa/paper-jasa.tex
save (work in progress, still a lot of things to move around)

@@
 \usepackage{amsmath}
-\usepackage{theorem}
+\usepackage{amsthm}
 \usepackage{xfrac}
@@
 \theoremstyle{plain}
-\newtheorem{thm}{Theorem}[section]
+\newtheorem{theorem}{Theorem}
+\newtheorem{proposition}[theorem]{Proposition}
 
@@
 {
-  \title{\bf Approximating Likelihood Ratios with Calibrated Classifiers}
+  \title{\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers}
   \author{Kyle Cranmer and Gilles Louppe\\
@@
 approach to statistical inference that is complementary to likelihood-free
-Bayesian inference algorithms (e.g., ABC), as it does not require the definition
-of a prior over model parameters.
-\glnote{Also mention conclusions on experimental results, as required by JASA guidelines.}
+Bayesian inference algorithms, like Approximate Bayesian Computation, as it does
+not require the definition of a prior over model parameters. \glnote{Also
+mention conclusions on experimental results, as required by JASA guidelines.}
 
@@
 \noindent%
-{\it Keywords:}  likelihood ratio, statistical inference, classification, particle physics
+{\it Keywords:}  likelihood ratio, likelihood-free inference, classification, particle physics
 \vfill
@@
 to describe complex processes that tie parameters $\theta$ of an underlying
-theory and measurement apparatus to high-dimensional observations $x$. However,
+theory and measurement apparatus to high-dimensional observations $\mathbf{x}$. However,
 directly evaluating the likelihood function in these cases is often impossible
@@
 under dimensionality reduction, under the assumption that the transformation is
-monotonic with the likelihood ratio. As a direct consequence, we propose an
+monotonic with the likelihood ratio. As a direct consequence, we derive and propose an
 alternative machine learning-based approach for likelihood-free inference that
@@
 As a concrete example, let us consider searches for new particles at the Large
-Hadron Collider (LHC). The simulator that is sampling from $p(x|\theta)$ is
+Hadron Collider (LHC). The simulator that is sampling from $p(\mathbf{x}|\theta)$ is
 based on quantum field theory, a detailed simulation of the particle detector,
@@
 learning to train discriminative classifiers that take advantage of a high
-dimensional feature vector $x$.
+dimensional feature vector $\mathbf{x}$. \glnote{We should then elaborate on how the
+proposed method could help these searches.}
 
@@
 
-\subsection{Notation and Assumptions}
+% 2) Likelihood ratio tests
+%
+% Introduce the mathematical problem with notations (was 1.1 and 1.2 before)
+% I would keep some of the comments of 1.3, but I feel some are not necessary
 
-We use the following notation:
-\begin{itemize}
- \item $x$: a vector of features for an event
- \item $D$: a data set of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
- \item $\theta$: parameters of a statistical model
-\item $p(x| \theta)$:  probability density  (simulation-based model) for $x$ given $\theta$
-%\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
-\item $y$: a class label used for training a classifier.
-\item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
-%\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
-\item $p( s_{\theta_0, \theta_1} | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$
-\end{itemize}
-We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.
+% \subsection{Notation and Assumptions}
+%
+% We use the following notation:
+% \begin{itemize}
+%  \item $x$: a vector of features for an event
+%  \item $D$: a data set of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
+%  \item $\theta$: parameters of a statistical model
+% \item $p(x| \theta)$:  probability density  (simulation-based model) for $x$ given $\theta$
+% %\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
+% \item $y$: a class label used for training a classifier.
+% \item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
+% %\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
+% \item $p( s_{\theta_0, \theta_1} | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$
+% \end{itemize}
+% We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.
 
-\subsection{Prelude}
-
-In the setting where one is interested in simple hypothesis testing between a null $\theta=\theta_0$ against an alternate $\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio
-\begin{equation}
-T(D; \theta_0, \theta_1) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
-\end{equation}
-is the most powerful test statistic. In order to evaluate $T(D)$, one must be able to evaluate the probability density
-$p(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that
-can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case
-high energy physics where the simulation of particle detectors can only be done in the `forward mode'.
+% \subsection{Prelude}
+%
+% In the setting where one is interested in simple hypothesis testing between a null $\theta=\theta_0$ against an alternate $\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio
+% \begin{equation}
+% T(D; \theta_0, \theta_1) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
+% \end{equation}
+% is the most powerful test statistic. In order to evaluate $T(D)$, one must be able to evaluate the probability density
+% $p(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that
+% can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case
+% high energy physics where the simulation of particle detectors can only be done in the `forward mode'.
 This same setting has been considered by \cite{ClaytonScott}, \cite{JMLR:v14:tong13a}, and \cite{Neal:2007zz}.
 
-The main result of this paper is to generalize the observation that one can form an equivalent test based on
-%\begin{equation}
-%T'(D) = \prod_{e=1}^n \frac{ p(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ p(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
-%\end{equation}
-\begin{equation}\label{eq:equivLRtest}
-T'(D; \theta_0, \theta_1) = \prod_{e=1}^n \frac{ p(s_e | \theta_0)}{ p(s_e | \theta_1)}
+Let $\mathbf{X}$ be a random vector with values $\mathbf{x} \in {\cal X}
+\subseteq \mathbb{R}^p$ and let $p_\mathbf{X}(\mathbf{x}|\theta)$ denote the
+density probability of $\mathbf{X}$ at value $\mathbf{x}$ under the
+parametrization $\theta$. Let also assume some i.i.d. observed data ${\cal D} = \{
+\mathbf{x}_1, \dots, \mathbf{x}_n \}$. In the setting where one is interested in
+simple hypothesis testing between a null $\theta=\theta_0$ against an alternate
+$\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio
+\begin{equation}\label{eqn:likelihood-ratio-test}
+T({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{X}(\mathbf{x}|\theta_0)}{ p_\mathbf{X}(\mathbf{x}|\theta_1)}
 \end{equation}
-if
-\begin{equation}\label{eq:montonic}
-s_e = s(x_e; \theta_0, \theta_1) = m\left(\, p(x_e|\theta_0) / p(x_e|\theta_1) \,\right) \;
-%s_e = s(x_e; \theta_0, \theta_1) = m\left(\frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)} \right) \;
-\end{equation}
-where $m$ is any strictly increasing or decreasing function. This result will be proven below.
-This allows us to recast the original likelihood ratio test into an alternate form in which supervised learning is used to train the discriminative classifier $s(x; \theta_0, \theta_1)$.
+is the most powerful test statistic. In order to evaluate $T({\cal D})$, one
+must be able to evaluate the probability density $p_\mathbf{X}(\mathbf{x}|
+\theta)$ at any value $\mathbf{x}$. However, it is increasingly common in
+science that one has a complex simulation that can act as generative model for
+$p_\mathbf{X}(\mathbf{x}|\theta)$, but one cannot evaluate the density directly.
+For instance, this is the case in high energy physics~\citep{Neal:2007zz} where
+the simulation of particle detectors can only be done in the forward mode.
+\glnote{Reinsert citations for \cite{ClaytonScott} and \cite{JMLR:v14:tong13a}.}
+
+% The main result of this paper is to generalize the observation that one can form an equivalent test based on
+% %\begin{equation}
+% %T'(D) = \prod_{e=1}^n \frac{ p(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ p(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
+% %\end{equation}
+% \begin{equation}\label{eq:equivLRtest}
+% T'(D; \theta_0, \theta_1) = \prod_{e=1}^n \frac{ p(s_e | \theta_0)}{ p(s_e | \theta_1)}
+% \end{equation}
+% if
+% \begin{equation}\label{eq:montonic}
+% s_e = s(x_e; \theta_0, \theta_1) = m\left(\, p(x_e|\theta_0) / p(x_e|\theta_1) \,\right) \;
+% %s_e = s(x_e; \theta_0, \theta_1) = m\left(\frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)} \right) \;
+% \end{equation}
+% where $m$ is any strictly increasing or decreasing function. This result will be proven below.
+% This allows us to recast the original likelihood ratio test into an alternate form in which supervised learning is used to train the discriminative classifier $s(x; \theta_0, \theta_1)$.
 The discriminative classifier can be trained with data $(x,y=0)$ generated
-from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$.
+% from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$.
 In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the classifier  parametrized in terms of $(\theta_0, \theta_1)$.
-
-Here we see that the original goal for simple hypothesis testing (i.e.
+%
+% Here we see that the original goal for simple hypothesis testing (i.e.
 to make a decision to accept or reject the null hypothesis based on the entire data set $D$) has been reformulated into a per-event classification problem.
@@
 
-\subsection{Comments on classification and frequentist hypothesis tests}
+% \subsection{Comments on classification and frequentist hypothesis tests}
+%
+% Vast literature exists around generative and discriminative
+% classifiers~\citep{AndrewY.Ng}. Typically, generative classifiers learn a model
+% for the joint probability $p(x,y)$, of the inputs $x$ and the classification
+% label $y$, and predict $p(y|x)$ via Bayes rule. In contrast, discriminative
+% classifiers model the posterior $p(y|x)$ directly. For classification tasks, one
+% then thresholds on $p(y|x)$. In both cases this description in terms of a
+% posterior requires a prior distribution for $p(y)$, which is either modeled
+% explicitly or learned from the training data. This familiar formulation of
+% classification may lead to some confusion in the setting of the current work.
+%
+% The first possible source of confusion we wish to avoid is that here
+% $p(x|\theta)$  is a  \textit{generative statistical model} for the features $x$,
+% not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a
+% traditional scientific theory or simulator, able to make predictions about $x$
+% and being motivated by domain-specific considerations.
+%
+% The second possible source of confusion is that we are not directly interested
+% in calibrating the classification score in terms of a per-event posterior
+% probability $p(y|x)$. Instead, we are interested in the approximation of the
+% per-experiment likelihood function or likelihood ratio, which might be used for
+% several purposes, including the calculation of p-values.
+%
+% Lastly, in the setting of frequentist hypothesis tests and confidence intervals,
+% we do not have a prior $\pi(\theta)$. While we can use the generative models to
+% produce training data $(x,y=0)$ generated from $p(x|\theta_0)$ and $(x,y=1)$
+% generated from $p(x|\theta_1)$, the relative mix $p(y)$ is arbitrary.  When
+% $p(y=0)=p(y=1)=\sfrac{1}{2}$, then
+% \begin{equation}
+% p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{p(x|\theta_1)}{p(x|\theta_0)+p(x|\theta_1))} \;,
+% \end{equation}
+% which is monotonic with the desired likelihood ratio
+% $p(x|\theta_1)/p(x|\theta_0)$. Since the prior $p(y)$ is not needed for the
+% target likelihood ratio test and because the classifier score $p(y|x)$ may not
+% be well calibrated, we choose to denote the classifier score $s(x)$ and simply
+% think of it as a deterministic dimensionality reduction map $s: X \to
+% \mathbb{R}$.  Similar points have been made by~\cite{ClaytonScott} and
+% \cite{Neal:2007zz}.
 
-Vast literature exists around generative and discriminative classifiers~\citep{AndrewY.Ng}.
-Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule.
-In contrast, discriminative classifiers model the posterior $p(y|x)$ directly.
-For classification tasks, one then thresholds on $p(y|x)$.
-In both cases this description in terms of a posterior requires a prior distribution for $p(y)$, which is either modeled explicitly or learned from the training data.
-This familiar formulation of classification may lead to some confusion in the setting of the current work.
+\section{Approximating likelihood ratios with classifiers}
+\label{sec:approx}
 
-The first possible source of confusion we wish to avoid is that here $p(x|\theta)$  is a  \textit{generative statistical model} for the features $x$, not a generative classifier.
-We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory or simulator, able to make predictions about $x$ and being motivated by domain-specific considerations.
+\glnote{Think of a better section title? The equivalence is strict, not approximate.}
 
-The second possible source of confusion is that
-we are not directly interested in calibrating the classification score in terms of a per-event posterior probability $p(y|x)$.
-Instead, we are interested in the approximation of the per-experiment likelihood function or likelihood ratio, which might be used for several purposes, including the calculation of p-values.
+The main result of this paper is to generalize the observation that one can
+form a test statistic
+\begin{equation}\label{eqn:likelihood-ratio-test-equiv}
+T'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{Y}(y=s(\mathbf{x}; \theta_0, \theta_1) | \theta_0)}{ p_\mathbf{Y}(y=s(\mathbf{x}; \theta_0, \theta_1) | \theta_1)}
+\end{equation}
+that is strictly equivalent to \ref{eqn:likelihood-ratio-test}, provided the change
+of variable $\mathbf{Y} = s(\mathbf{X}; \theta_0, \theta_1)$ is based
+on a parametrized function $s$ that is strictly monotonic with the density ratio
+\begin{equation}
+r(\mathbf{x};\theta_0, \theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.
+\end{equation}
+As derived below, this allows to recast the original likelihood ratio test into an alternate
+form in which supervised learning can be used to build
+$s(\mathbf{x}; \theta_0, \theta_1)$ as a discriminative classifier.  In
+Section~\ref{sec:generalized-likelihood-ratio} we extend this result to generalized likelihood ratio tests,
+where it will be useful to have the classifier  parametrized in terms of
+$(\theta_0, \theta_1)$.
 
-Lastly, in the setting of frequentist hypothesis tests and confidence intervals, we do not have a prior $\pi(\theta)$.
-While we can use the generative models to produce training data $(x,y=0)$ generated
-from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$, the relative mix $p(y)$
-is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then
+\subsection{Likelihood ratios under change of variable}
+
+\begin{theorem}
+    \label{thm:ratio-equivalence}
+    Let $\mathbf{X}$ be a random vector vector with values in ${\cal X} \subseteq \mathbb{R}^p$ and parametrized probability
+    density $p_{\mathbf{X}}(\mathbf{x} = (x_1, ..., x_p)|\theta)$ and let
+    $s : \mathbb{R}^p \mapsto \mathbb{R}$ be a function monotonic with the density ratio
+    $r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$,
+    for given parameters $\theta_0$ and $\theta_1$. In these conditions,
     \begin{equation}
-p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{p(x|\theta_1)}{p(x|\theta_0)+p(x|\theta_1))} \;,
+        r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} = \frac{p_\mathbf{Y}(y=s(\mathbf{x};\theta_0,\theta_1)|\theta_0)}{p_\mathbf{Y}(y=s(\mathbf{x};\theta_0,\theta_1)|\theta_1)},
     \end{equation}
-which is monotonic with the desired likelihood ratio $p(x|\theta_1)/p(x|\theta_0)$.
-Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.
-Similar points have been made by~\cite{ClaytonScott} and \cite{Neal:2007zz}.
+    where $p_\mathbf{Y}(y=s(\mathbf{x};\theta_0,\theta_1)|\theta)$ is the induced probability density of
+    $\mathbf{Y} = s(\mathbf{X};\theta_0,\theta_1)$.
+\end{theorem}
 
-\section{Dimensionality reduction and calibration}
+\begin{proof}
+Omitting for clarity the parametrization $(\theta_0,\theta_1)$ of $s(\mathbf{X};\theta_0,\theta_1)$, we have
+\begin{align}
+p_Y(y=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
+&= \frac{d}{dy}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq y\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
+&= \frac{d}{dy} \int_{\mathbb{R}^p} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
+&= \int_{\mathbb{R}^p} \frac{d}{dy} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
+&= \int_{\mathbb{R}^p} \delta(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime
+\end{align}
+where $H$ and $\delta$ are respectively the Heaviside step and the Dirac delta
+functions. Intuitively, this last expression can be understood as the integral
+over all $\mathbf{x}^\prime \in \mathbb{R}^p$ such that $s(\mathbf{x}^\prime) = y$, as picked
+by the Dirac delta function. Given Theorem 6.1.5 of \citet{Hrmander1990},
+it further comes
+\begin{align}
+p_Y(y=s(\mathbf{x})|\theta_0) &= \int_{\mathbb{R}^p} \delta(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
+&= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
+\end{align}
+where $|\nabla s(\mathbf{x}^\prime)| = \sqrt{\sum_{i=1}^p |\frac{\partial}{\partial x_i} s(\mathbf{x}^\prime)|^2}$
+and where $dS_{\mathbf{x}^\prime}$ is the Euclidean surface measure on $\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}$.
+Also, since $s(\mathbf{x})$ is monotonic with
+$\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$,
+it exists an invertible function $m:\mathbb{R}^+ \mapsto \mathbb{R}$ such
+that $s(\mathbf{x}) = m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$.
+In particular, we have
+\begin{align}
+\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} &= m^{-1}(s(\mathbf{x})) \nonumber \\
+p_\mathbf{X}(\mathbf{x}|\theta_0) &= m^{-1}(s(\mathbf{x})) p_\mathbf{X}(\mathbf{x}|\theta_1) \label{eqn:mapping}
+\end{align}
+Combining equations \ref{eqn:hormander} and \ref{eqn:mapping}, we have
+\begin{align}
+p_Y(y=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{p_\mathbf{X}(\mathbf{x}^\prime|\theta_0)}{|\nabla s(\mathbf{x}^\prime)|} dS_{\mathbf{x}^\prime} \nonumber \\
+&= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+&= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(y) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+&= m^{-1}(s(\mathbf{x})) \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}
+\end{align}
+Similarly, Equation~\ref{eqn:hormander} can be used to derive $p_Y(y=s(\mathbf{x})|\theta_1)$, finally yielding
+\begin{align}
+\frac{p_Y(y=s(\mathbf{x})|\theta_0)}{p_Y(y=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
+&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.
+\end{align}
+\end{proof}
 
-We are interested in reformulating the target likelihood ratio
+\subsection{Probabilistic classification for approximating ratios}
+
+\begin{proposition} \label{thm:best-classifier}
+Let $\mathbf{X} = (X_1, ..., X_p)$ and $Y$ be random input and output variables
+with values in ${\cal X} \subseteq \mathbb{R}^p$\footnote{Without loss of
+generality, the discussion assumes continuous random variables, but can be
+extended to random variables of mixed types with the Dirac delta function.}
+and ${\cal Y} = \{0, 1\}$ and mixed joint probability density  function
+$p_{\mathbf{X},Y}(\mathbf{x}, y)$\footnote{Defined as
+$p_{\mathbf{X},Y}(\mathbf{x}, y) = p_{\mathbf{X}|Y}(\mathbf{x}|y) P(Y=y) =
+P(Y=y|X=x) p_{\mathbf{X}}(\mathbf{x})$.}. For the squared error loss, the best
+regression function $s : {\cal X} \mapsto [0, 1]$, or equivalently the best
+probabilistic classifier, is
 \begin{equation}
-\ln T(D; \theta_0, \theta_1) =   \sum_{e=1}^n \underbrace{\log \left[ \frac {p(x_e | \theta_0) }{ p(x_e | \theta_1) } \right]}_{q(x_e)} \;.
+s^*(\mathbf{x}) = \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{P(Y=0) p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
 \end{equation}
-Here we see that the test statistic $T$ for the experiment is composed of a sum over events of the per-event function $q(x)$.
+\end{proposition}
+
+\begin{proof}
+For the squared error loss,
+\begin{align}
+s^*(\mathbf{x}) &= \argmin_{s(\mathbf{x})} \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ (Y - s(\mathbf{x}))^2 \} \nonumber \\
+&=  \argmin_{s(\mathbf{x})} \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y^2 \} - 2s(\mathbf{x}) \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + s(\mathbf{x})^2 \nonumber \\
+&=  \argmin_{s(\mathbf{x})} -2s(\mathbf{x}) \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + s(\mathbf{x})^2
+\end{align}
+The last expression is minimized when $\frac{d}{ds(\mathbf{x})} (-2s(\mathbf{x}) \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + s(\mathbf{x})^2) = 0$,
+that is when $-2 \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + 2 s(\mathbf{x}) = 0$, hence
+\begin{equation}
+s^*(\mathbf{x}) = \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \}.
+\end{equation}
+For ${\cal Y} = \{ 0, 1 \}$,
+\begin{align}
+\mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} &= P(Y=0|\mathbf{X}=\mathbf{x}) \times 0 +  P(Y=1|\mathbf{X}=\mathbf{x}) \times 1 \nonumber \\
+&= \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{p_{\mathbf{X}}(\mathbf{x})} \nonumber \\
+&= \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{P(Y=0) p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
+\end{align}
+\end{proof}
+
+For $P(Y=0)=P(Y=1)=\frac{1}{2}$, the best regression function $s^*$ simplifies
+to
+\begin{equation}
+s^*(\mathbf{x}) = \frac{p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
+\end{equation}
+If we further assume that samples for $Y=0$ are drawn from some parametrized
+distribution with probability density $p_{\mathbf{X}}(\mathbf{x}|\theta_0)$
+and similarly for $Y=1$, then the best regression function can be rewritten
+as
+\begin{equation}
+s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
+\end{equation}
+
+In particular, this regression function satisfies conditions of
+Theorem~\ref{thm:ratio-equivalence} since $s^*(\mathbf{x}) =
+m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$,
+for $m(r(\mathbf{x})) = \frac{1}{1 + r(\mathbf{x})}$, is monotonic
+with
+$\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$.
+In other words, Proposition~\ref{thm:best-classifier} yields a sufficient
+procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any {\it
+universally strongly consistent} algorithm can be used for learning $s^*$.
+Note however,
+that it is not a necessary procedure since Theorem~\ref{thm:ratio-equivalence} holds
+for any monotonic function $m$ of the density ratio, i.e., not for
+$m(r(\mathbf{x})) = \frac{r(\mathbf{x})}{1 + r(\mathbf{x})}$ only.
+
+Equivalently, since $m^{-1}(s^*(\mathbf{x})) = \frac{1 - s^*(\mathbf{x})}{s^*(\mathbf{x})} = r(\mathbf{x})$,
+Proposition~\ref{thm:best-classifier} shows that the density ratio estimation problem is strictly
+equivalent to probabilistic classification, and that algorithms proposed to
+solve one can be used for solving the other. In this context, Theorem~\ref{thm:ratio-equivalence}
+shows that in case we learn a probabilistic classifier $s(\mathbf{x})$
+which is imperfect up to a monotonic transformation of $r(\mathbf{x})$, then
+one can still resort to calibration (i.e., modelling $p_Y(y=s(\mathbf{x}))$)
+to compute $r(\mathbf{x})$ exactly.
+
+\glnote{Discuss that training on signal vs background instead of signal+background vs background is OK because this learns a function which is 1:1.}
+
+\glnote{Discuss the capacity of the classifier; Generalization to decomposed ratios
+Discuss about the need of calibration in case of imperfect classifiers (was first paragraphs of section 5)}
+
+% \section{Dimensionality reduction and calibration}
+%
+% We are interested in reformulating the target likelihood ratio
+% \begin{equation}
+% \ln T(D; \theta_0, \theta_1) =   \sum_{e=1}^n \underbrace{\log \left[ \frac {p(x_e | \theta_0) }{ p(x_e | \theta_1) } \right]}_{q(x_e)} \;.
+% \end{equation}
+% Here we see that the test statistic $T$ for the experiment is composed of a sum over events of the per-event function $q(x)$.
 A sum over a monotonic, but non-linear function of $q(x)$ would not lead to an equivalent statistic.
-
-The important part of the per-event function $q(x)$ is that it defines iso-contours in the feature space $x$.
+%
+% The important part of the per-event function $q(x)$ is that it defines iso-contours in the feature space $x$.
 As we will show, our goal is to learn a monotonic function of $p(x|\theta_0)/p(x|\theta_1)$, which will share the same iso-contours.
@@
 Our claim is that the generative model $p(x|\theta)$ can be used to calibrate the density $p(s|\theta)$ and that
-\begin{equation}\label{eq:enveloping}
-\ln T'(D; \theta_0, \theta_1) = \sum_{e=1}^n \underbrace{\log \left[ \frac {p(s_e | \theta_0) }{ p(s_e | \theta_1) } \right]}_{q(s_e)} \;,
-\end{equation}
-leads to an equivalent statistic.
+% \begin{equation}\label{eq:enveloping}
+% \ln T'(D; \theta_0, \theta_1) = \sum_{e=1}^n \underbrace{\log \left[ \frac {p(s_e | \theta_0) }{ p(s_e | \theta_1) } \right]}_{q(s_e)} \;,
+% \end{equation}
+% leads to an equivalent statistic.
+%
+% For notational simplicity, let $p_0(x) = p(x|\theta_0)$, $p_1(x) = p(x|\theta_1)$, and $\sfunc(x)=s(x; \theta_0, \theta_1)$.
+% The distribution of $x$ totally determines the distribution of $s$.
+% In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
+% Let $\Omega_{\sstar}$ be the level set $\{x \mid s(x; \theta_0, \theta_1) = \sstar \}$ and \mbox{$\hat{n}=\nabla s(x) / |\nabla s(x)|$} be the orthonormal vector to $\Omega_{\sstar}$ at the point $x$.
+%
+% We need to show that for all $x$, the density
+% \begin{equation}
+% p(q_x|\theta) = \int dx \delta(q_x-q_x(x)) p(x|\theta)  / | \hat{n} \cdot \nabla q_x  |
+% \end{equation}
+% is equal to the density
+% \begin{equation}
+% p(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, p(x|\theta) \, / | \hat{n} \cdot \nabla q_s  | \; .
+% \end{equation}
+% It is sufficient to show that $q_x(x) = q_s(s(x))$.
+% %$ \forall x\in\Omega_{\sstar}$.
+% The function $q_s(s)$ is based on the induced densities $p_0(s)$ and $p_1(s)$.  The induced density $p_1(s)$ is given by
+% \begin{equation}
+% p_1(\sstar) = \int dx \delta(\sstar-s(x)) p_1(x) = \int d\Omega_{\sstar} p_1(x)  / | \hat{n} \cdot \nabla s  |
+% \end{equation}
+% and a similar equation for $p_0(s)$.
+% %\textbf{Do we need Jacobian for x $\to$ s independent of delta function part, I think that's double counting?}
+%
+% \textbf{\flushleft Theorem 1:}
+% We have the following equality
+% \begin{equation}
+% \frac{p_1(s(x))}{p_0(s(x))} =  \frac{p_1(x)}{p_0(x)}  \; . %\;\hspace{3em} \forall x\in\Omega_{\sstar}\; .
+% \end{equation}
+% \textbf{Proof}
+% For $x\in \Omega_{\sstar}$, we can factor out of the integral the constant $p_1(x)/p_0(x)$.
+% Thus
+% \begin{equation}
+% p_1(\sstar) =  \int d\Omega_{\sstar} p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_{\sstar} p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
+% %p_1(\sstar) = \int dx \delta(\sstar-s(x)) p_1(x) = \int d\Omega_{\sstar} p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_{\sstar} p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
+% \end{equation}
+% and the integrals cancel in the likelihood ratio
+% \begin{equation}
+% \frac{p_1(\sstar)}{p_0(\sstar)} = \frac{p_1(x)}{p_0(x)} \frac{\int d\Omega_{\sstar} p_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_{\sstar} p_0(x) / | \hat{n} \cdot \nabla s  |} = \frac{p_1(x)}{p_0(x)}  \;\hspace{3em} \forall x\in\Omega_{\sstar}.
+% \end{equation}
+%
+% One can think of the ratio $p_1(s)/p_0(s)$ as a way of calibrating the the discriminative classifier and correcting for the monotonic transformation $m$ of the desired likelihood ratio as in Eq.~\ref{eq:montonic}.
 
-For notational simplicity, let $p_0(x) = p(x|\theta_0)$, $p_1(x) = p(x|\theta_1)$, and $\sfunc(x)=s(x; \theta_0, \theta_1)$.
-The distribution of $x$ totally determines the distribution of $s$.
-In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
-Let $\Omega_{\sstar}$ be the level set $\{x \mid s(x; \theta_0, \theta_1) = \sstar \}$ and \mbox{$\hat{n}=\nabla s(x) / |\nabla s(x)|$} be the orthonormal vector to $\Omega_{\sstar}$ at the point $x$.
+% \section{Embedding the classifier into the likelihood}
+%
+% Thus far we have shown that the target likelihood ratio $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via the univariate densities $p(s|\theta_0)/p(s|\theta_1)$ if the classifier $s(x|\theta_0, \theta_1)$ is a strictly increasing function of $p(x|\theta_0)/p(x|\theta_1)$.
+We now generalize from the ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the case where $\theta$ are continuous model parameters.
+%  We postpone the practicalities of training the classifier and estimating the density to Section~\ref{S:classifier} and continue in the likelihood-free setting with idealized classifiers and their densities.
+%
+% In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$ for the observed data and never refer back to the original features $x_e$.
+In the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0, \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
+%
+% The critical observation is that  if we postpone the evaluation of the classifier to the stage of evaluating the enveloping likelihood ratio, then we can identify the value of the parameters  that are being compared in the likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
+% \begin{equation}\label{eq:embedding}
+% T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
+% {p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
+% \end{equation}
+% This is equivalent to approximating the likelihood function for $\theta_0$  when $\theta_1$ is held fixed.
 
-We need to show that for all $x$, the density
-\begin{equation}
-p(q_x|\theta) = \int dx \delta(q_x-q_x(x)) p(x|\theta)  / | \hat{n} \cdot \nabla q_x  |
-\end{equation}
-is equal to the density
-\begin{equation}
-p(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, p(x|\theta) \, / | \hat{n} \cdot \nabla q_s  | \; .
-\end{equation}
-It is sufficient to show that $q_x(x) = q_s(s(x))$.
-%$ \forall x\in\Omega_{\sstar}$.
-The function $q_s(s)$ is based on the induced densities $p_0(s)$ and $p_1(s)$.  The induced density $p_1(s)$ is given by
-\begin{equation}
-p_1(\sstar) = \int dx \delta(\sstar-s(x)) p_1(x) = \int d\Omega_{\sstar} p_1(x)  / | \hat{n} \cdot \nabla s  |
-\end{equation}
-and a similar equation for $p_0(s)$.
-%\textbf{Do we need Jacobian for x $\to$ s independent of delta function part, I think that's double counting?}
+\section{Generalized likelihood ratio tests}
+\label{sec:generalized-likelihood-ratio}
 
-\textbf{\flushleft Theorem 1:}
-We have the following equality
-\begin{equation}
-\frac{p_1(s(x))}{p_0(s(x))} =  \frac{p_1(x)}{p_0(x)}  \; . %\;\hspace{3em} \forall x\in\Omega_{\sstar}\; .
-\end{equation}
-\textbf{Proof}
-For $x\in \Omega_{\sstar}$, we can factor out of the integral the constant $p_1(x)/p_0(x)$.
-Thus
-\begin{equation}
-p_1(\sstar) =  \int d\Omega_{\sstar} p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_{\sstar} p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
-%p_1(\sstar) = \int dx \delta(\sstar-s(x)) p_1(x) = \int d\Omega_{\sstar} p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_{\sstar} p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
-\end{equation}
-and the integrals cancel in the likelihood ratio
-\begin{equation}
-\frac{p_1(\sstar)}{p_0(\sstar)} = \frac{p_1(x)}{p_0(x)} \frac{\int d\Omega_{\sstar} p_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_{\sstar} p_0(x) / | \hat{n} \cdot \nabla s  |} = \frac{p_1(x)}{p_0(x)}  \;\hspace{3em} \forall x\in\Omega_{\sstar}.
-\end{equation}
-
-One can think of the ratio $p_1(s)/p_0(s)$ as a way of calibrating the the discriminative classifier and correcting for the monotonic transformation $m$ of the desired likelihood ratio as in Eq.~\ref{eq:montonic}.
-
-\section{Embedding the classifier into the likelihood}
-
-Thus far we have shown that the target likelihood ratio $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via the univariate densities $p(s|\theta_0)/p(s|\theta_1)$ if the classifier $s(x|\theta_0, \theta_1)$ is a strictly increasing function of $p(x|\theta_0)/p(x|\theta_1)$.
-We now generalize from the ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the case where $\theta$ are continuous model parameters.
- We postpone the practicalities of training the classifier and estimating the density to Section~\ref{S:classifier} and continue in the likelihood-free setting with idealized classifiers and their densities.
-
-In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$ for the observed data and never refer back to the original features $x_e$.
-In the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0, \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
-
-The critical observation is that  if we postpone the evaluation of the classifier to the stage of evaluating the enveloping likelihood ratio, then we can identify the value of the parameters  that are being compared in the likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
-\begin{equation}\label{eq:embedding}
-T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
-{p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
-\end{equation}
-This is equivalent to approximating the likelihood function for $\theta_0$  when $\theta_1$ is held fixed.
-
-\section{Composite hypotheses and the generalized likelihood ratio}\label{S:GLR}
+% \section{Composite hypotheses and the generalized likelihood ratio}\label{S:GLR}
 
@@
 
-In order for this approach to be useful in the likelihood-free setting, where it is not possible to evaluate the density $p(x|\theta)$, we need to be able to approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples $\{x_i\}$ drawn from the generative model $p(x|\theta)$.
-Denote the  dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{p}(\hat{s}|\theta)$.
+In order for this approach to be useful in the likelihood-free setting, where it
+is not possible to evaluate the density $p(x|\theta)$, we need to be able to
+approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples
+$\{x_i\}$ drawn from the generative model $p(x|\theta)$.  Denote the
+dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its
+distribution $\hat{p}(\hat{s}|\theta)$.
 
-One strength of this approach is that it factorizes the approximation of the per-event likelihood ratio ($\hat{s}(x; \theta_0, \theta_1) \approx s(x; \theta_0, \theta_1)$) from the calibration procedure ($\hat p(\hat s| \theta) \approx p(\hat{s}|\theta)$).
-Thus, even if the classifier does a poor job of reproducing the level sets of the per-event likelihood ratio, the density of $\hat{s}$ can still be well calibrated.
-In that case, one might loose power, but the resulting inference will still be valid.
-This point was made by \cite{Neal:2007zz} and is well appreciated by the particle physics community that typically takes a conservative attitude towards the use of machine learning classifiers precisely due to concerns about the calibration $p$-values in the face of nuisance parameters associated to the simulator.
+One strength of this approach is that it factorizes the approximation of the
+per-event likelihood ratio ($\hat{s}(x; \theta_0, \theta_1) \approx s(x;
+\theta_0, \theta_1)$) from the calibration procedure ($\hat p(\hat s| \theta)
+\approx p(\hat{s}|\theta)$). Thus, even if the classifier does a poor job of
+reproducing the level sets of the per-event likelihood ratio, the density of
+$\hat{s}$ can still be well calibrated. In that case, one might loose power, but
+the resulting inference will still be valid. This point was made by
+\cite{Neal:2007zz} and is well appreciated by the particle physics community
+that typically takes a conservative attitude towards the use of machine learning
+classifiers precisely due to concerns about the calibration $p$-values in the
+face of nuisance parameters associated to the simulator.
 
-\subsection{The fixed discriminative classification setting}
-For fixed $\theta_0$ and $\theta_1$ we can generate
-large samples from each model and train a classifier. To be concrete, let us use $p(x|\theta_0)$ to generate training
-data $(x_i,  y_i=0)$ and $p(x|\theta_1)$ to generate training data $(x_i , y_i=1)$.
-With balanced training data   \mbox{($p(y=1)=p(y=0)=\sfrac{1}{2}$)} a quadratic loss function will lead to classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) = p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the desired per-event likelihood ratio $q(x)$.
-Thus, standard supervised learning algorithms with various surrogate loss functions lead to discriminative classifiers that approximate a monotonic function of per-event likelihood ratio $q(x)$.
+\subsection{The fixed discriminative classification setting} For fixed
+$\theta_0$ and $\theta_1$ we can generate large samples from each model and
+train a classifier. To be concrete, let us use $p(x|\theta_0)$ to generate
+training data $(x_i,  y_i=0)$ and $p(x|\theta_1)$ to generate training data
+$(x_i , y_i=1)$. With balanced training data
+\mbox{($p(y=1)=p(y=0)=\sfrac{1}{2}$)} a quadratic loss function will lead to
+classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) =
+p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the
+desired per-event likelihood ratio $q(x)$. Thus, standard supervised learning
+algorithms with various surrogate loss functions lead to discriminative
+classifiers that approximate a monotonic function of per-event likelihood ratio
+$q(x)$.
 

### 2016-01-20 jasa/paper-jasa.tex
formatting of the tex file

@@
 \spacingset{1.45} % DON'T change the spacing!
+
+% Introduction =================================================================
+
+% - Introduction section, mostly the same as before:
+%    * likelihood free setup
+%    * searches for particles at LHC as an example
+%    * brief and clear description of the contribution of this paper
+%      (i.e., density ratio estimation with calibrated classifiers,
+%             as an alternative to other methods for parameter inference)
+% - outline
 
@@
 
+% Likelihood ratio tests =======================================================
+
+% - Introduce the mathematical problem with notations (was 1.1 and 1.2 before)
+% - I would keep some of the comments of 1.3, but I feel some are not necessary
+
 \section{Likelihood ratio tests}
@@
 
+% Approximating likelihood ratios with classifiers =============================
+
+% - Core of the paper (was section 2 + my note on slack)
+% - Derive theorem for the ratio of transformed densitites
+% - Show that consistent regressors minimizing the squared error loss fulfill conditions of the theorem
+% - Discuss that training on signal vs background instead of signal+background vs background is OK because this learns a function which is 1:1.
+=> focus the capacity of the classifier; Generalization to decomposed ratios
+% - Discuss about the need of calibration in case of imperfect classifiers (was first paragraphs of section 5)
+
 \section{Approximating likelihood ratios with classifiers}
@@
 \end{equation}
-As derived below, this allows to recast the original likelihood ratio test into an alternate
-form in which supervised learning can be used to build
+As derived below, this allows to recast the original likelihood ratio test into
+an alternate form in which supervised learning can be used to build
 $s(\mathbf{x}; \theta_0, \theta_1)$ as a discriminative classifier.  In
-Section~\ref{sec:generalized-likelihood-ratio} we extend this result to generalized likelihood ratio tests,
-where it will be useful to have the classifier  parametrized in terms of
-$(\theta_0, \theta_1)$.
+Section~\ref{sec:generalized-likelihood-ratio} we extend this result to
+generalized likelihood ratio tests, where it will be useful to have the
+classifier  parametrized in terms of $(\theta_0, \theta_1)$.
 
@@
 Let $\mathbf{X} = (X_1, ..., X_p)$ and $Y$ be random input and output variables
-with values in ${\cal X} \subseteq \mathbb{R}^p$\footnote{Without loss of
-generality, the discussion assumes continuous random variables, but can be
-extended to random variables of mixed types with the Dirac delta function.}
+with values in ${\cal X} \subseteq \mathbb{R}^p$
 and ${\cal Y} = \{0, 1\}$ and mixed joint probability density  function
-$p_{\mathbf{X},Y}(\mathbf{x}, y)$\footnote{Defined as
-$p_{\mathbf{X},Y}(\mathbf{x}, y) = p_{\mathbf{X}|Y}(\mathbf{x}|y) P(Y=y) =
-P(Y=y|X=x) p_{\mathbf{X}}(\mathbf{x})$.}. For the squared error loss, the best
+$p_{\mathbf{X},Y}(\mathbf{x}, y)$. For the squared error loss, the best
 regression function $s : {\cal X} \mapsto [0, 1]$, or equivalently the best
@@
 
-Equivalently, since $m^{-1}(s^*(\mathbf{x})) = \frac{1 - s^*(\mathbf{x})}{s^*(\mathbf{x})} = r(\mathbf{x})$,
-Proposition~\ref{thm:best-classifier} shows that the density ratio estimation problem is strictly
-equivalent to probabilistic classification, and that algorithms proposed to
-solve one can be used for solving the other. In this context, Theorem~\ref{thm:ratio-equivalence}
-shows that in case we learn a probabilistic classifier $s(\mathbf{x})$
-which is imperfect up to a monotonic transformation of $r(\mathbf{x})$, then
-one can still resort to calibration (i.e., modelling $p_Y(y=s(\mathbf{x}))$)
-to compute $r(\mathbf{x})$ exactly.
+Equivalently, since $m^{-1}(s^*(\mathbf{x})) = \frac{1 -
+s^*(\mathbf{x})}{s^*(\mathbf{x})} = r(\mathbf{x})$,
+Proposition~\ref{thm:best-classifier} shows that the likelihood ratio estimation
+problem is directly related to probabilistic classification, and that algorithms
+proposed to solve one can be used for solving the other. In this context,
+Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
+classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
+of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modelling
+$p_Y(y=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.
 
-\glnote{Discuss that training on signal vs background instead of signal+background vs background is OK because this learns a function which is 1:1.}
+\glnote{Discuss that training on signal vs background instead of
+signal+background vs background is OK because this learns a function which is
+1:1.}
 
-\glnote{Discuss the capacity of the classifier; Generalization to decomposed ratios
-Discuss about the need of calibration in case of imperfect classifiers (was first paragraphs of section 5)}
+\glnote{Discuss the capacity of the classifier; Generalization to decomposed
+ratios Discuss about the need of calibration in case of imperfect classifiers
+(was first paragraphs of section 5)}
 
@@
 % This is equivalent to approximating the likelihood function for $\theta_0$  when $\theta_1$ is held fixed.
+
+% Generalized likelihood ratio tests ===========================================
+
+% - mathematical problem (was section 4)
+% - maximum likelihood estimator inference from ratios
+% - discussion on parameterized classifier and practicalities regarding embedding (was sections 5.2 and 5.3)
 
@@
 
-\bigskip
-
-While the function $p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ will minimize the expected squared loss based on
-training data produced according to Algorithm~\ref{alg:training}, it is not clear how training data from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real world classifier with finite capacity.
-This is left as an area for future work.
+While the function $p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ will minimize
+the expected squared loss based on training data produced according to
+Algorithm~\ref{alg:training}, it is not clear how training data from $\theta'_0
+\ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real world
+classifier with finite capacity. This is left as an area for future work.
 
@@
 
-Once the classifier is trained, we can use the generative model together with a univariate density estimation technique (e.g.
-histograms or kernel density estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter points.
-For an single parameter point, this is a tractable univariate density estimation problem. The challenge comes from the need to estimate this density for all values of $\theta$.
-A straight forward approach would be to run the generative model on demand for any particular value of $\theta$.
-In the context of a likelihood fit this would mean that the optimization algorithm that is trying to maximize the likelihood with respect to $\theta$ needs access to the generative model $p(x|\theta)$.
-This can be  impractical when the generative model is computationally expensive or has high-latency (for instance some human intervention is required to reconfigure the generative model).
-In  HEP with a fixed classifier, it has become common  to interpolate the distribution between discrete values of $\theta$ in order to produce a continuous parametrization for $\hat p(\hat s | \theta)$~\citep{Cranmer:2012sba}.
+Once the classifier is trained, we can use the generative model together with a
+univariate density estimation technique (e.g. histograms or kernel density
+estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter
+points. For an single parameter point, this is a tractable univariate density
+estimation problem. The challenge comes from the need to estimate this density
+for all values of $\theta$. A straight forward approach would be to run the
+generative model on demand for any particular value of $\theta$. In the context
+of a likelihood fit this would mean that the optimization algorithm that is
+trying to maximize the likelihood with respect to $\theta$ needs access to the
+generative model $p(x|\theta)$. This can be  impractical when the generative
+model is computationally expensive or has high-latency (for instance some human
+intervention is required to reconfigure the generative model).  In  HEP with a
+fixed classifier, it has become common  to interpolate the distribution between
+discrete values of $\theta$ in order to produce a continuous parametrization for
+$\hat p(\hat s | \theta)$~\citep{Cranmer:2012sba}.
 
-There is a mild technical challenge in embedding the classifier into the likelihood.
-In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$.
-In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ since one does not know the true values of  $\theta_0$ and $\theta_1$.
-Thus one must implement the embedding of the classifier as in Eq.~\ref{eq:embedding}.
-A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
+There is a mild technical challenge in embedding the classifier into the
+likelihood. In the case of a fixed classifier $\hat s(x)$ it is possible to
+pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features
+$x_e$. In the parametrized setting it is not possible to pre-compute $\hat
+s(x_e; \theta_0, \theta_1)$ since one does not know the true values of
+$\theta_0$ and $\theta_1$. Thus one must implement the embedding of the
+classifier as in Eq.~\ref{eq:embedding}.  A concrete realization of this has
+been performed for probability models implemented with the \texttt{RooFit}
+probabilistic programing language and  classifiers implemented with
+\texttt{scikit-learn} and
+\texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
-One can easily imagine a number of approaches to embedding the classifier and estimating the density $\hat p(\hat s|\theta)$ and the relative merits of those approaches will depend critically on the dimensionality of $\theta$ and the computational cost of the generative model.
-We leave a more general strategy for this overarching optimization problem as an area of future work.
+One can easily imagine a number of approaches to embedding the classifier and
+estimating the density $\hat p(\hat s|\theta)$ and the relative merits of those
+approaches will depend critically on the dimensionality of $\theta$ and the
+computational cost of the generative model. We leave a more general strategy for
+this overarching optimization problem as an area of future work.
 
-\section{Applications in HEP}
+% Applications =================================================================
 
-In HEP we are often searching for some
-class of events, generically referred to as \textit{signal}, in the presence of a separate class
-of \textit{background} events.  For each event we measure some quantities $x$ that have corresponding distributions
-$p_b(x|\nu)$ for background and $p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing
-uncertainties in the underlying physics prediction or response of the measurement device. The
-total model is a mixture of the signal and background, and $\mu$ is the mixture coefficient associate
-to the signal component.
+% - typical usage in HEP with a concrete experimental example.
+%   * illustrate the method for several classifiers (new)
+%    * illustrate the method for several calibration routines (new)
+%    * compare with other methods (new)
+% - ratio of mixtures and decomposition (special case of generalized likelihood rario tests when parameters only affect the mixture coefficients => no need for parametrized classifiers)
+% - measure of particles properties, with a concrete example. Comparison with other likelihood-free inference methods (new)
+
+\section{Applications}
+
+In HEP we are often searching for some class of events, generically referred to
+as \textit{signal}, in the presence of a separate class of \textit{background}
+events.  For each event we measure some quantities $x$ that have corresponding
+distributions $p_b(x|\nu)$ for background and $p_s(x|\nu)$ for signal, where
+$\nu$ are nuisance parameters describing uncertainties in the underlying physics
+prediction or response of the measurement device. The total model is a mixture
+of the signal and background, and $\mu$ is the mixture coefficient associate to
+the signal component.
 \begin{equation}\label{eq:hepGen}
@@
 \end{equation}
-New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
-generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\citep{Cowan:2010js}.
-This approach was used for the discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
+New particle searches at the LHC are typically framed as hypothesis test where
+the null corresponds to $\mu=0$, and the generalized likelihood ratio (profiling
+over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test
+statistic~\citep{Cowan:2010js}. This approach was used for the discovery of the
+Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
 
@@
 
-In this setting, large samples of pseudo-data $\{x_i, y_i\}$ generated with some nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density $p_s(x|\nu_0)$.
-Importantly, the $y=1$ label corresponds to the signal only, and not to the alternate signal-plus-background hypothesis.
-The resulting classifier approximates the regression function $p_s(x|\nu_0)/(p_s(x|\nu_0)+p_b(x|\nu_0))$, which is one to one with the likelihood ratio of the null to the alternate $p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$.
-Associating the $y=1$ label to the signal component has advantages because it helps the classifier focus its capacity on the relevant regions in the feature space, particularly when the signal is a very small perturbation to the background (i.e.
-$\mu \ll 1$).
+In this setting, large samples of pseudo-data $\{x_i, y_i\}$ generated with some
+nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the
+background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density
+$p_s(x|\nu_0)$. Importantly, the $y=1$ label corresponds to the signal only, and
+not to the alternate signal-plus-background hypothesis. The resulting classifier
+approximates the regression function $p_s(x|\nu_0)/(p_s(x|\nu_0)+p_b(x|\nu_0))$,
+which is one to one with the likelihood ratio of the null to the alternate
+$p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$. Associating the $y=1$ label to
+the signal component has advantages because it helps the classifier focus its
+capacity on the relevant regions in the feature space, particularly when the
+signal is a very small perturbation to the background (i.e. $\mu \ll 1$).
 
- Once the classifier is trained, large samples of pseudo-data are drawn from $p_s(x | \nu)$ and $p_b(x | \nu)$ and we estimate the distributions  $\hat{p}_s(\hat s | \nu)$ and $\hat{p}_b(\hat s | \nu)$ continuously parametrized in $\nu$.
-An example of the distributions of $\hat s$ for the signal and background events with $\nu=\nu_0$ is shown in Figure~\ref{fig:tmva}.
+ Once the classifier is trained, large samples of pseudo-data are drawn from
+ $p_s(x | \nu)$ and $p_b(x | \nu)$ and we estimate the distributions
+ $\hat{p}_s(\hat s | \nu)$ and $\hat{p}_b(\hat s | \nu)$ continuously
+ parametrized in $\nu$. An example of the distributions of $\hat s$ for the
+ signal and background events with $\nu=\nu_0$ is shown in
+ Figure~\ref{fig:tmva}.
 
@@
 \end{figure}
-%\bigskip
 
 These steps feed into a subsequent statistical test based on the observed data
-${D=(x_1, \dots, x_n)}$.
-For each event, the classifier is evaluated and one performs inference on a parameter $\mu$ related to the presence of the signal contribution.
-In particular, one forms the statistical model\footnote{Sometimes there is an additional Poisson term when expected number of signal and background events is known, which is referred to as an extended likelihood or marked Poisson model.}
+${D=(x_1, \dots, x_n)}$. For each event, the classifier is evaluated and one
+performs inference on a parameter $\mu$ related to the presence of the signal
+contribution. In particular, one forms the statistical model\footnote{Sometimes
+there is an additional Poisson term when expected number of signal and
+background events is known, which is referred to as an extended likelihood or
+marked Poisson model.}
 \begin{equation}\label{eq:typicalML}
@@
 
-Nuisance parameters are an after thought in the typical usage of machine learning in HEP.
-In fact, most discussions would related to the training and optimizing the classifier only consider $p_b(x)$ and $p_s(x)$ with $\nu=\nu_0$ being implicit.
-However, as experimentalists we know that we must account for various forms of systematic uncertainty, parametrized by nuisance parameters $\nu$.
-In practice, we take the classifier as fixed and then propagate uncertainty through the classifier as in Eq.~\ref{eq:typicalML}.
-Building the distribution $p(\hat s|\nu)$ for values of $\nu$ other than the nominal $\nu_0$ used to train the classifier can be thought of as a calibration necessary for classical statistical inference; however, this classifier is clearly not optimal for $\nu \ne \nu_0$.
+Nuisance parameters are an after thought in the typical usage of machine
+learning in HEP. In fact, most discussions would related to the training and
+optimizing the classifier only consider $p_b(x)$ and $p_s(x)$ with $\nu=\nu_0$
+being implicit. However, as experimentalists we know that we must account for
+various forms of systematic uncertainty, parametrized by nuisance parameters
+$\nu$. In practice, we take the classifier as fixed and then propagate
+uncertainty through the classifier as in Eq.~\ref{eq:typicalML}. Building the
+distribution $p(\hat s|\nu)$ for values of $\nu$ other than the nominal $\nu_0$
+used to train the classifier can be thought of as a calibration necessary for
+classical statistical inference; however, this classifier is clearly not optimal
+for $\nu \ne \nu_0$.
 
@@
 
-The standard use of machine learning in HEP can be improved by training a parametrized classifier corresponding to the generalized likelihood ratio test
+The standard use of machine learning in HEP can be improved by training a
+parametrized classifier corresponding to the generalized likelihood ratio test
 \begin{equation}
@@
 
-There is an interesting distinction between this approach and the standard use in which the classifier is trained for a fixed $\nu_0$.
-In the standard use one trains a classifier for signal vs.
-background, which is equivalent (in an ideal setting) to training a classifier for  null (background-only) vs.
-alternate (signal-plus-backgound) since the resulting regression functions are one-to-one with each other.
-In contrast, in the case of the generalized likelihood ratio test
+There is an interesting distinction between this approach and the standard use
+in which the classifier is trained for a fixed $\nu_0$. In the standard use one
+trains a classifier for signal vs. background, which is equivalent (in an ideal
+setting) to training a classifier for  null (background-only) vs. alternate
+(signal-plus-backgound) since the resulting regression functions are one-to-one
+with each other. In contrast, in the case of the generalized likelihood ratio
+test
 \begin{equation}\label{eq:hep_improved}
@@
 \end{equation}
-the background components don't cancel and there is an additional term $p_b(x| \hat{\hat{ \nu}})/p_b(x| {\hat{ \nu}})$.
-In practice, with classifiers of finite capacity, there will be some tradeoff between taking into account this additional term and the more challenging learning problem when $\mu$ is very small.
+the background components don't cancel and there is an additional term $p_b(x|
+\hat{\hat{ \nu}})/p_b(x| {\hat{ \nu}})$. In practice, with classifiers of finite
+capacity, there will be some tradeoff between taking into account this
+additional term and the more challenging learning problem when $\mu$ is very
+small.
 
@@
 
-In this section we generalize the capacity focusing technique of training classifiers to discriminate between components of a mixture model.
-First, we generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
+In this section we generalize the capacity focusing technique of training
+classifiers to discriminate between components of a mixture model. First, we
+generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
 \begin{equation}
@@
 
-While the original motivation for this work was to improve the treatment of systematic uncertainties in new particle searches by parametrizing the classifier in terms of the nuisance parameters $\nu$, the same approach can be used for parameter inference.
-In the case of new particle searches the parameter of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
-When measuring particle properties the distribution of the features also depend on parameters such as a particle's mass and quantum numbers.
-This is easily accommodated by extending $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$ includes both parameters of interest and nuisance parameters.
+While the original motivation for this work was to improve the treatment of
+systematic uncertainties in new particle searches by parametrizing the
+classifier in terms of the nuisance parameters $\nu$, the same approach can be
+used for parameter inference. In the case of new particle searches the parameter
+of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
+When measuring particle properties the distribution of the features also depend
+on parameters such as a particle's mass and quantum numbers. This is easily
+accommodated by extending $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$
+includes both parameters of interest and nuisance parameters.
 
-This formalism represents a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure.
-The work of  ~\citep{Whiteson:2006ws} is similar as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed.
-This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.
+This formalism represents a significant step forward in the usage of machine
+learning in HEP, where classifiers have always been used between two static
+classes of events and not parametrized explicitly in terms of the physical
+quantities we wish to measure. The work of  ~\citep{Whiteson:2006ws} is similar
+as the stochastic optimization was directly trying to minimize the measurement
+uncertainty of a particle's mass; however, the resulting classifier was fixed.
+This approach also offers the advantage that it explicitly reformulates the
+per-experiment optimization to the per-event optimization, which is less
+computationally intensive.
 
-Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citep{Volobouev:2011vb}.
-In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method.
-Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time to calculate the likelihood ratio $q(x)$ for a single event.
-The work here can be seen as aiming at the same conceptual target, but utilizing machine learning to overcome the complexity of the detector simulation.
-It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage.
-In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citep{Aaltonen:2010yz}).
+Another approach that is similar in spirit is the so-called matrix element
+method, in which one  directly computes an approximate likelihood ratio by
+performing a computationally intensive integral associated to the detector
+response~\citep{Volobouev:2011vb}. In the approach considered in this paper, the
+detector response is naturally handled by the Monte Carlo sampling used in the
+simulation of the detector; however, that integral is intractable for the matrix
+element method. Even with drastic simplifications of the detector response, the
+matrix element method can take several minutes of CPU time to calculate the
+likelihood ratio $q(x)$ for a single event. The work here can be seen as aiming
+at the same conceptual target, but utilizing machine learning to overcome the
+complexity of the detector simulation. It also offers enormous speed increase
+for evaluating the likelihood at the cost of an initial training stage. In
+practice, the matrix element method has only been used for searches and
+measurement of a single physical parameter (sometimes with a single nuisance
+parameter as in~\citep{Aaltonen:2010yz}).
 
-Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citep{Allanach:2000kt}.
-Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters.
-The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citep{Gao:2010qx,Bolognesi:2012mm}.
+Contemporary examples where the technique presented here could have major impact
+on HEP include the measurement of coefficients to quantum mechanical operators
+describing the decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so
+lucky, measurement of the mass of supersymmetric particles in cascade
+decays~\citep{Allanach:2000kt}.  Both of these examples involve data sets with
+many events, each with a feature vector $x$ that has on the order of 10
+components, and a parameter vector $\theta$ with 5-10 parameters of interest and
+possibly many more nuisance parameters. The state of the art for the operator
+coefficients of the Higgs decay uses the so-called matrix element likelihood
+analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is
+approximated by neglecting detector effects~\citep{Gao:2010qx,Bolognesi:2012mm}.
 
-\bigskip
+% Related works ================================================================
 
-\section{Related work}\label{S:Related}
+\section{Related works}\label{S:Related}
 
-\cite{ClaytonScott} and \cite{JMLR:v14:tong13a} consider the machine learning problem associated to Neyman-Pearson hypothesis testing.
-As in this work, they consider the situation where one does not have access to the underlying distributions, but only has i.i.d.
-samples from each hypothesis.
-This work generalizes that goal from the Neyman-Pearson setting to generalized likelihood ratio tests and emphasizes the connection with classification.
-Perhaps a  formal treatment similar to the Neyman-Pearson case can be brought to bear in this more general setting.
-In a similarly titled work, \cite{Gutmann2014} advocate using the cross-validated classification accuracy as the similarity metric used in ABC.
-While the goal there is also parameter inference in the likelihood-free setting,  ``classifier ABC'' is very different than the approach presented here.
-\cite{TommiJaakkola} explore a way of leveraging generative models to derive kernel functions for use in discriminative methods.
-This interesting work is distinct from the point made here in which the generative model is being used for the purpose of providing training data and calibration.
+\cite{ClaytonScott} and \cite{JMLR:v14:tong13a} consider the machine learning
+problem associated to Neyman-Pearson hypothesis testing. As in this work, they
+consider the situation where one does not have access to the underlying
+distributions, but only has i.i.d. samples from each hypothesis. This work
+generalizes that goal from the Neyman-Pearson setting to generalized likelihood
+ratio tests and emphasizes the connection with classification. Perhaps a  formal
+treatment similar to the Neyman-Pearson case can be brought to bear in this more
+general setting. In a similarly titled work, \cite{Gutmann2014} advocate using
+the cross-validated classification accuracy as the similarity metric used in
+ABC. While the goal there is also parameter inference in the likelihood-free
+setting,  ``classifier ABC'' is very different than the approach presented here.
+\cite{TommiJaakkola} explore a way of leveraging generative models to derive
+kernel functions for use in discriminative methods. This interesting work is
+distinct from the point made here in which the generative model is being used
+for the purpose of providing training data and calibration.
 %\cite{McCallum} consider a hybrid generative/discriminative classifier; however, the goal of that work is not to leverage a generative model for the data, but to use both approaches to learn different subsets of the parameters in a single hybrid classifier.
-\cite{BiancaZadrozny} emphasize the importance of calibrated probability estimates from decision trees and naive Bayesian classifiers and investigate various approaches to achieve this.
-In contrast to that work, we are not interested in calibrated probability estimates for $p(y|x)$ for individual events, but instead we use the calibration to correct for non-linear transformations of the target likelihood ratio and, perhaps, to provide calibrated p-values based on those likelihood ratio tests.
-\cite{Ihler2004} take on a different problem (tests of statistical independence) by using machine learning algorithms to find  scalar maps from the high-dimensional feature space that achieve the desired statistical goal when the fundamental high-dimensional test is intractable.
+\cite{BiancaZadrozny} emphasize the importance of calibrated probability
+estimates from decision trees and naive Bayesian classifiers and investigate
+various approaches to achieve this. In contrast to that work, we are not
+interested in calibrated probability estimates for $p(y|x)$ for individual
+events, but instead we use the calibration to correct for non-linear
+transformations of the target likelihood ratio and, perhaps, to provide
+calibrated p-values based on those likelihood ratio tests. \cite{Ihler2004} take
+on a different problem (tests of statistical independence) by using machine
+learning algorithms to find  scalar maps from the high-dimensional feature space
+that achieve the desired statistical goal when the fundamental high-dimensional
+test is intractable.
 
-\cite{Neal:2007zz} also considered the problem of approximating the likelihood function when only a generative model is available.
-That work sketches a scheme in which one uses a classifier with both $x$ and $\theta$ as an input to serve as a dimensionality reduction map.
-The key distinction comes in the handling of $\theta$.
-Neal says  ``we cannot use the classifier on real data, since we don't know the correct value for [$\theta]$'' and goes on to outline an approach where one uses regression on a per-event basis to estimate $\hat{\theta}(x)$ and perform the composition $s(x; \hat{\theta}(x))$ (much like profiling).\footnote{Neal considers a lower-dimensional `bottlenecks' $\theta^*=g(\theta)$, which are not essential to the discussion here.}  This can lead to a significant loss of information since (at least in most particle physics examples) a single event carries little information about the true value of $\theta$, though the full data set $D$ may be informative -- for instance, a single observation would not be sufficient to estimate the variance of a distribution, though repeated observations would.
-The work of Neal correctly identifies this as an approximation of the target likelihood even in the case of a ideal classifier.
-In contrast, the approach described here does not eliminate the dependence of the classifier on $\theta$.%
-\footnote{As a technical point, in Neal's work, the focus is on approximating the likelihood function (up to a multiplicative constant), which is equivalent to evaluating the ratio  with respect to a fixed $\theta_1$ as in Eq.~\ref{eq:mle_withs}.
-In Neal's case, the dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is constant; however, in this approach the map ratio is explicitly parametrized in terms of $\theta$, so the ratio is important for canceling the corresponding Jacobian factors.} Instead, we embed a parametrized classifier into the likelihood and postpone the evaluation of the classifier to the point of evaluation of the likelihood when $\theta$ is explicitly being tested.
-This avoids the loss of information that occurs from the regression step $\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact result in the case of an ideal classifier.
-In both cases, the quality of the classifier is factorized from the calibration of its density, which allows for valid inference even if there is a loss of power due to a non ideal classifier.
+\cite{Neal:2007zz} also considered the problem of approximating the likelihood
+function when only a generative model is available. That work sketches a scheme
+in which one uses a classifier with both $x$ and $\theta$ as an input to serve
+as a dimensionality reduction map. The key distinction comes in the handling of
+$\theta$.  Neal says  ``we cannot use the classifier on real data, since we
+don't know the correct value for [$\theta]$'' and goes on to outline an approach
+where one uses regression on a per-event basis to estimate $\hat{\theta}(x)$ and
+perform the composition $s(x; \hat{\theta}(x))$ (much like
+profiling).\footnote{Neal considers a lower-dimensional `bottlenecks'
+$\theta^*=g(\theta)$, which are not essential to the discussion here.}  This can
+lead to a significant loss of information since (at least in most particle
+physics examples) a single event carries little information about the true value
+of $\theta$, though the full data set $D$ may be informative -- for instance, a
+single observation would not be sufficient to estimate the variance of a
+distribution, though repeated observations would.  The work of Neal correctly
+identifies this as an approximation of the target likelihood even in the case of
+a ideal classifier. In contrast, the approach described here does not eliminate
+the dependence of the classifier on $\theta$.\footnote{As a technical point,
+in Neal's work, the focus is on approximating the likelihood function (up to a
+multiplicative constant), which is equivalent to evaluating the ratio  with
+respect to a fixed $\theta_1$ as in Eq.~\ref{eq:mle_withs}. In Neal's case, the
+dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is
+constant; however, in this approach the map ratio is explicitly parametrized in
+terms of $\theta$, so the ratio is important for canceling the corresponding
+Jacobian factors.} Instead, we embed a parametrized classifier into the
+likelihood and postpone the evaluation of the classifier to the point of
+evaluation of the likelihood when $\theta$ is explicitly being tested. This
+avoids the loss of information that occurs from the regression step
+$\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact
+result in the case of an ideal classifier. In both cases, the quality of the
+classifier is factorized from the calibration of its density, which allows for
+valid inference even if there is a loss of power due to a non ideal classifier.
+
+% Conclusions ==================================================================
 
@@
 
-We have outlined a technique to reformulate generalized likelihood ratio test over a high-dimensional data set with multiple events in terms of a univariate density of a classifier score.
-We have shown that a parametrized family of discriminative classifiers $\hat s(x; \theta_0, \theta_1)$ trained and calibrated with a simulator  $p(x|\theta)$ can be used to approximate the  likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not possible to directly evaluate the likelihood $p(x|\theta)$.
+We have outlined a technique to reformulate generalized likelihood ratio test
+over a high-dimensional data set with multiple events in terms of a univariate
+density of a classifier score. We have shown that a parametrized family of
+discriminative classifiers $\hat s(x; \theta_0, \theta_1)$ trained and
+calibrated with a simulator  $p(x|\theta)$ can be used to approximate the
+likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not
+possible to directly evaluate the likelihood $p(x|\theta)$.
 %A technique for decomposing this ratio when the generative model is a mixture of components was presented with the aim to help focus capacity of the classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small mixture coefficient.
 This approach leverages the power of machine learning in a classical statistical setting.
-It offers an alternative to approximate Bayesian computation for parameter inference in the likelihood-free setting that can also be used in the frequentist formalism without specifying a prior over the parameters.
-A strength of this approach is that it separates the quality of the approximation of the target likelihood from the quality of the calibration.
-The former is related to the ability of supervised learning approaches to  classification, which will continue to improve.
-The calibration procedure for a particular parameter point is fairly straight forward since it involves estimating a univariate density using a generative model of the data.
-The difficulty of the calibration stage is performing this calibration continuously in $\theta$.
-Different strategies to this calibration are anticipated depending on the dimensionality of $\theta$, the complexity of the resulting likelihood function, and the difficulties associated to running the simulator.
+It offers an alternative to approximate Bayesian computation for parameter
+inference in the likelihood-free setting that can also be used in the
+frequentist formalism without specifying a prior over the parameters. A strength
+of this approach is that it separates the quality of the approximation of the
+target likelihood from the quality of the calibration. The former is related to
+the ability of supervised learning approaches to  classification, which will
+continue to improve. The calibration procedure for a particular parameter point
+is fairly straight forward since it involves estimating a univariate density
+using a generative model of the data. The difficulty of the calibration stage is
+performing this calibration continuously in $\theta$. Different strategies to
+this calibration are anticipated depending on the dimensionality of $\theta$,
+the complexity of the resulting likelihood function, and the difficulties
+associated to running the simulator.
 

### 2016-01-21 jasa/paper-jasa.tex
continued

@@
 
-% 2) Likelihood ratio tests
-%
-% Introduce the mathematical problem with notations (was 1.1 and 1.2 before)
-% I would keep some of the comments of 1.3, but I feel some are not necessary
-
 % \subsection{Notation and Assumptions}
@@
 
+% Approximating likelihood ratios with classifiers =============================
+
+% - Core of the paper (was section 2 + my note on slack)
+% - Derive theorem for the ratio of transformed densitites
+% - Show that consistent regressors minimizing the squared error loss fulfill conditions of the theorem
+% - Discuss that training on signal vs background instead of signal+background vs background is OK because this learns a function which is 1:1.
+=> focus the capacity of the classifier; Generalization to decomposed ratios
+% - Discuss about the need of calibration in case of imperfect classifiers (was first paragraphs of section 5)
+
+\section{Approximating likelihood ratios with classifiers}
+\label{sec:approx}
+
+\glnote{Think of a better section title? The equivalence is strict, not approximate.}
+
 % The main result of this paper is to generalize the observation that one can form an equivalent test based on
@@
 
-% \subsection{Comments on classification and frequentist hypothesis tests}
-%
-% Vast literature exists around generative and discriminative
-% classifiers~\citep{AndrewY.Ng}. Typically, generative classifiers learn a model
-% for the joint probability $p(x,y)$, of the inputs $x$ and the classification
-% label $y$, and predict $p(y|x)$ via Bayes rule. In contrast, discriminative
-% classifiers model the posterior $p(y|x)$ directly. For classification tasks, one
-% then thresholds on $p(y|x)$. In both cases this description in terms of a
-% posterior requires a prior distribution for $p(y)$, which is either modeled
-% explicitly or learned from the training data. This familiar formulation of
-% classification may lead to some confusion in the setting of the current work.
-%
-% The first possible source of confusion we wish to avoid is that here
-% $p(x|\theta)$  is a  \textit{generative statistical model} for the features $x$,
-% not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a
-% traditional scientific theory or simulator, able to make predictions about $x$
-% and being motivated by domain-specific considerations.
-%
-% The second possible source of confusion is that we are not directly interested
-% in calibrating the classification score in terms of a per-event posterior
-% probability $p(y|x)$. Instead, we are interested in the approximation of the
-% per-experiment likelihood function or likelihood ratio, which might be used for
-% several purposes, including the calculation of p-values.
-%
-% Lastly, in the setting of frequentist hypothesis tests and confidence intervals,
-% we do not have a prior $\pi(\theta)$. While we can use the generative models to
-% produce training data $(x,y=0)$ generated from $p(x|\theta_0)$ and $(x,y=1)$
-% generated from $p(x|\theta_1)$, the relative mix $p(y)$ is arbitrary.  When
-% $p(y=0)=p(y=1)=\sfrac{1}{2}$, then
-% \begin{equation}
-% p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{p(x|\theta_1)}{p(x|\theta_0)+p(x|\theta_1))} \;,
-% \end{equation}
-% which is monotonic with the desired likelihood ratio
-% $p(x|\theta_1)/p(x|\theta_0)$. Since the prior $p(y)$ is not needed for the
-% target likelihood ratio test and because the classifier score $p(y|x)$ may not
-% be well calibrated, we choose to denote the classifier score $s(x)$ and simply
-% think of it as a deterministic dimensionality reduction map $s: X \to
-% \mathbb{R}$.  Similar points have been made by~\cite{ClaytonScott} and
-% \cite{Neal:2007zz}.
-
-% Approximating likelihood ratios with classifiers =============================
-
-% - Core of the paper (was section 2 + my note on slack)
-% - Derive theorem for the ratio of transformed densitites
-% - Show that consistent regressors minimizing the squared error loss fulfill conditions of the theorem
-% - Discuss that training on signal vs background instead of signal+background vs background is OK because this learns a function which is 1:1.
-=> focus the capacity of the classifier; Generalization to decomposed ratios
-% - Discuss about the need of calibration in case of imperfect classifiers (was first paragraphs of section 5)
-
-\section{Approximating likelihood ratios with classifiers}
-\label{sec:approx}
-
-\glnote{Think of a better section title? The equivalence is strict, not approximate.}
-
-The main result of this paper is to generalize the observation that one can
-form a test statistic
+The main result of this paper (see Theorem~\ref{thm:ratio-equivalence}) is to
+generalize the observation that one can form a test statistic
 \begin{equation}\label{eqn:likelihood-ratio-test-equiv}
@@
 
-\subsection{Likelihood ratios under change of variable}
+\subsection{Likelihood ratios under change of variables}
+
+% \section{Dimensionality reduction and calibration}
+%
+% We are interested in reformulating the target likelihood ratio
+% \begin{equation}
+% \ln T(D; \theta_0, \theta_1) =   \sum_{e=1}^n \underbrace{\log \left[ \frac {p(x_e | \theta_0) }{ p(x_e | \theta_1) } \right]}_{q(x_e)} \;.
+% \end{equation}
+% Here we see that the test statistic $T$ for the experiment is composed of a sum over events of the per-event function $q(x)$.
+A sum over a monotonic, but non-linear function of $q(x)$ would not lead to an equivalent statistic.
+%
+% The important part of the per-event function $q(x)$ is that it defines iso-contours in the feature space $x$.
+As we will show, our goal is to learn a monotonic function of $p(x|\theta_0)/p(x|\theta_1)$, which will share the same iso-contours.
+Then the remaining challenge is to find the appropriate monotonic function that gives back a linear function of $q(x)$.
+Our claim is that the generative model $p(x|\theta)$ can be used to calibrate the density $p(s|\theta)$ and that
+% \begin{equation}\label{eq:enveloping}
+% \ln T'(D; \theta_0, \theta_1) = \sum_{e=1}^n \underbrace{\log \left[ \frac {p(s_e | \theta_0) }{ p(s_e | \theta_1) } \right]}_{q(s_e)} \;,
+% \end{equation}
+% leads to an equivalent statistic.
+%
+% For notational simplicity, let $p_0(x) = p(x|\theta_0)$, $p_1(x) = p(x|\theta_1)$, and $\sfunc(x)=s(x; \theta_0, \theta_1)$.
+% The distribution of $x$ totally determines the distribution of $s$.
+% In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
+% Let $\Omega_{\sstar}$ be the level set $\{x \mid s(x; \theta_0, \theta_1) = \sstar \}$ and \mbox{$\hat{n}=\nabla s(x) / |\nabla s(x)|$} be the orthonormal vector to $\Omega_{\sstar}$ at the point $x$.
+%
+% We need to show that for all $x$, the density
+% \begin{equation}
+% p(q_x|\theta) = \int dx \delta(q_x-q_x(x)) p(x|\theta)  / | \hat{n} \cdot \nabla q_x  |
+% \end{equation}
+% is equal to the density
+% \begin{equation}
+% p(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, p(x|\theta) \, / | \hat{n} \cdot \nabla q_s  | \; .
+% \end{equation}
+% It is sufficient to show that $q_x(x) = q_s(s(x))$.
+% %$ \forall x\in\Omega_{\sstar}$.
+% The function $q_s(s)$ is based on the induced densities $p_0(s)$ and $p_1(s)$.  The induced density $p_1(s)$ is given by
+% \begin{equation}
+% p_1(\sstar) = \int dx \delta(\sstar-s(x)) p_1(x) = \int d\Omega_{\sstar} p_1(x)  / | \hat{n} \cdot \nabla s  |
+% \end{equation}
+% and a similar equation for $p_0(s)$.
+% %\textbf{Do we need Jacobian for x $\to$ s independent of delta function part, I think that's double counting?}
+%
+% \textbf{\flushleft Theorem 1:}
+% We have the following equality
+% \begin{equation}
+% \frac{p_1(s(x))}{p_0(s(x))} =  \frac{p_1(x)}{p_0(x)}  \; . %\;\hspace{3em} \forall x\in\Omega_{\sstar}\; .
+% \end{equation}
+% \textbf{Proof}
+% For $x\in \Omega_{\sstar}$, we can factor out of the integral the constant $p_1(x)/p_0(x)$.
+% Thus
+% \begin{equation}
+% p_1(\sstar) =  \int d\Omega_{\sstar} p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_{\sstar} p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
+% %p_1(\sstar) = \int dx \delta(\sstar-s(x)) p_1(x) = \int d\Omega_{\sstar} p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_{\sstar} p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
+% \end{equation}
+% and the integrals cancel in the likelihood ratio
+% \begin{equation}
+% \frac{p_1(\sstar)}{p_0(\sstar)} = \frac{p_1(x)}{p_0(x)} \frac{\int d\Omega_{\sstar} p_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_{\sstar} p_0(x) / | \hat{n} \cdot \nabla s  |} = \frac{p_1(x)}{p_0(x)}  \;\hspace{3em} \forall x\in\Omega_{\sstar}.
+% \end{equation}
+%
+% One can think of the ratio $p_1(s)/p_0(s)$ as a way of calibrating the the discriminative classifier and correcting for the monotonic transformation $m$ of the desired likelihood ratio as in Eq.~\ref{eq:montonic}.
 
@@
 
+% \subsection{The fixed discriminative classification setting} For fixed
+% $\theta_0$ and $\theta_1$ we can generate large samples from each model and
+% train a classifier. To be concrete, let us use $p(x|\theta_0)$ to generate
+% training data $(x_i,  y_i=0)$ and $p(x|\theta_1)$ to generate training data
+% $(x_i , y_i=1)$. With balanced training data
+% \mbox{($p(y=1)=p(y=0)=\sfrac{1}{2}$)} a quadratic loss function will lead to
+% classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) =
+% p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the
+% desired per-event likelihood ratio $q(x)$. Thus, standard supervised learning
+% algorithms with various surrogate loss functions lead to discriminative
+% classifiers that approximate a monotonic function of per-event likelihood ratio
+% $q(x)$.
+
 \begin{proposition} \label{thm:best-classifier}
@@
 m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$,
-for $m(r(\mathbf{x})) = \frac{1}{1 + r(\mathbf{x})}$, is monotonic
-with
+for $m(r(\mathbf{x})) = \frac{1}{1 + r(\mathbf{x})}$, is monotonic with
 $\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$.
 In other words, Proposition~\ref{thm:best-classifier} yields a sufficient
-procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any {\it
-universally strongly consistent} algorithm can be used for learning $s^*$.
-Note however,
-that it is not a necessary procedure since Theorem~\ref{thm:ratio-equivalence} holds
-for any monotonic function $m$ of the density ratio, i.e., not for
-$m(r(\mathbf{x})) = \frac{r(\mathbf{x})}{1 + r(\mathbf{x})}$ only.
+procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
+{\it universally strongly consistent} algorithm can be used for learning $s^*$.
+Note however, that it is not a necessary procedure since
+Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
+density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{r(\mathbf{x})}{1 +
+r(\mathbf{x})}$ only.
 
@@
 
-\glnote{Discuss that training on signal vs background instead of
-signal+background vs background is OK because this learns a function which is
-1:1.}
-
-\glnote{Discuss the capacity of the classifier; Generalization to decomposed
-ratios Discuss about the need of calibration in case of imperfect classifiers
-(was first paragraphs of section 5)}
-
-% \section{Dimensionality reduction and calibration}
-%
-% We are interested in reformulating the target likelihood ratio
-% \begin{equation}
-% \ln T(D; \theta_0, \theta_1) =   \sum_{e=1}^n \underbrace{\log \left[ \frac {p(x_e | \theta_0) }{ p(x_e | \theta_1) } \right]}_{q(x_e)} \;.
-% \end{equation}
-% Here we see that the test statistic $T$ for the experiment is composed of a sum over events of the per-event function $q(x)$.
-A sum over a monotonic, but non-linear function of $q(x)$ would not lead to an equivalent statistic.
-%
-% The important part of the per-event function $q(x)$ is that it defines iso-contours in the feature space $x$.
-As we will show, our goal is to learn a monotonic function of $p(x|\theta_0)/p(x|\theta_1)$, which will share the same iso-contours.
-Then the remaining challenge is to find the appropriate monotonic function that gives back a linear function of $q(x)$.
-Our claim is that the generative model $p(x|\theta)$ can be used to calibrate the density $p(s|\theta)$ and that
-% \begin{equation}\label{eq:enveloping}
-% \ln T'(D; \theta_0, \theta_1) = \sum_{e=1}^n \underbrace{\log \left[ \frac {p(s_e | \theta_0) }{ p(s_e | \theta_1) } \right]}_{q(s_e)} \;,
-% \end{equation}
-% leads to an equivalent statistic.
-%
-% For notational simplicity, let $p_0(x) = p(x|\theta_0)$, $p_1(x) = p(x|\theta_1)$, and $\sfunc(x)=s(x; \theta_0, \theta_1)$.
-% The distribution of $x$ totally determines the distribution of $s$.
-% In the application at hand, the function $s$ maps a high-dimensional feature vector $x$ to $\mathbb{R}^+$.
-% Let $\Omega_{\sstar}$ be the level set $\{x \mid s(x; \theta_0, \theta_1) = \sstar \}$ and \mbox{$\hat{n}=\nabla s(x) / |\nabla s(x)|$} be the orthonormal vector to $\Omega_{\sstar}$ at the point $x$.
-%
-% We need to show that for all $x$, the density
-% \begin{equation}
-% p(q_x|\theta) = \int dx \delta(q_x-q_x(x)) p(x|\theta)  / | \hat{n} \cdot \nabla q_x  |
-% \end{equation}
-% is equal to the density
-% \begin{equation}
-% p(q_s|\theta) = \int dx \delta(q_s-q_s(s(x))) \, p(x|\theta) \, / | \hat{n} \cdot \nabla q_s  | \; .
-% \end{equation}
-% It is sufficient to show that $q_x(x) = q_s(s(x))$.
-% %$ \forall x\in\Omega_{\sstar}$.
-% The function $q_s(s)$ is based on the induced densities $p_0(s)$ and $p_1(s)$.  The induced density $p_1(s)$ is given by
-% \begin{equation}
-% p_1(\sstar) = \int dx \delta(\sstar-s(x)) p_1(x) = \int d\Omega_{\sstar} p_1(x)  / | \hat{n} \cdot \nabla s  |
-% \end{equation}
-% and a similar equation for $p_0(s)$.
-% %\textbf{Do we need Jacobian for x $\to$ s independent of delta function part, I think that's double counting?}
-%
-% \textbf{\flushleft Theorem 1:}
-% We have the following equality
-% \begin{equation}
-% \frac{p_1(s(x))}{p_0(s(x))} =  \frac{p_1(x)}{p_0(x)}  \; . %\;\hspace{3em} \forall x\in\Omega_{\sstar}\; .
-% \end{equation}
-% \textbf{Proof}
-% For $x\in \Omega_{\sstar}$, we can factor out of the integral the constant $p_1(x)/p_0(x)$.
-% Thus
-% \begin{equation}
-% p_1(\sstar) =  \int d\Omega_{\sstar} p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_{\sstar} p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
-% %p_1(\sstar) = \int dx \delta(\sstar-s(x)) p_1(x) = \int d\Omega_{\sstar} p_1(x) / | \hat{n} \cdot \nabla s  |= \frac{p_1(x)}{p_0(x)} \int d\Omega_{\sstar} p_0(x)  / | \hat{n} \cdot \nabla s  | \;,
-% \end{equation}
-% and the integrals cancel in the likelihood ratio
-% \begin{equation}
-% \frac{p_1(\sstar)}{p_0(\sstar)} = \frac{p_1(x)}{p_0(x)} \frac{\int d\Omega_{\sstar} p_0(x)/ | \hat{n} \cdot \nabla s  |}{ \int d\Omega_{\sstar} p_0(x) / | \hat{n} \cdot \nabla s  |} = \frac{p_1(x)}{p_0(x)}  \;\hspace{3em} \forall x\in\Omega_{\sstar}.
-% \end{equation}
-%
-% One can think of the ratio $p_1(s)/p_0(s)$ as a way of calibrating the the discriminative classifier and correcting for the monotonic transformation $m$ of the desired likelihood ratio as in Eq.~\ref{eq:montonic}.
-
-% \section{Embedding the classifier into the likelihood}
-%
-% Thus far we have shown that the target likelihood ratio $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be reproduced via the univariate densities $p(s|\theta_0)/p(s|\theta_1)$ if the classifier $s(x|\theta_0, \theta_1)$ is a strictly increasing function of $p(x|\theta_0)/p(x|\theta_1)$.
-We now generalize from the ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the case where $\theta$ are continuous model parameters.
-%  We postpone the practicalities of training the classifier and estimating the density to Section~\ref{S:classifier} and continue in the likelihood-free setting with idealized classifiers and their densities.
-%
-% In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$ for the observed data and never refer back to the original features $x_e$.
-In the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0, \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
-%
-% The critical observation is that  if we postpone the evaluation of the classifier to the stage of evaluating the enveloping likelihood ratio, then we can identify the value of the parameters  that are being compared in the likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
-% \begin{equation}\label{eq:embedding}
-% T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
-% {p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
-% \end{equation}
-% This is equivalent to approximating the likelihood function for $\theta_0$  when $\theta_1$ is held fixed.
-
-% Generalized likelihood ratio tests ===========================================
-
-% - mathematical problem (was section 4)
-% - maximum likelihood estimator inference from ratios
-% - discussion on parameterized classifier and practicalities regarding embedding (was sections 5.2 and 5.3)
-
-\section{Generalized likelihood ratio tests}
-\label{sec:generalized-likelihood-ratio}
-
-% \section{Composite hypotheses and the generalized likelihood ratio}\label{S:GLR}
-
-In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative $\theta \in \Theta_0^C$, the generalized likelihood ratio\footnote{Also known as the profile likelihood ratio.} test is commonly used
-\begin{equation}
-\lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p(D | \theta)}{ \sup_{\theta \in \Theta} p(D | \theta)} \; .
-\end{equation}
-This generalized likelihood ratio can be used both for hypothesis tests in the presence of nuisance parameters or to create confidence intervals with or without nuisance parameters.
-Often, the parameter vector is broken into two components $\theta=(\mu,\nu)$, where the $\mu$ components are considered parameters of interest while the $\nu$ components are considered nuisance parameters.
-In that case $\Theta_0$ corresponds to all values of $\nu$ with $\mu$ fixed.
-
-Denote the maximum likelihood estimator
-\begin{equation}\label{eq:mle}
-\hat{\theta} = \argmax_\theta  p(D | \theta)
-\end{equation}
-and the conditional maximum likelihood estimator
-\begin{equation}\label{eq:cmle}
-\hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p(D | \theta) \; .
-\end{equation}
-
-It is not obvious that if we are working with the distributions $p(s|\theta)$ (for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the same estimators.
-Fortunately, there is a construction based on $p(s|\theta)$ that works.
-The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as the value that maximizes the likelihood ratio with respect to $p(D|\theta_1)$ for some fixed value of $\theta_1$.
-This allows us to use Theorem~1 to reformulate the maximum likelihood estimate
-\begin{equation}\label{eq:mle_withs}
-\hat{\theta} = %\argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} =
-\argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
-\end{equation}
-It is important that we include the denominator $p(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.
-
-\section{Learning the correct mapping and its distribution}\label{S:classifier}
+\subsection{Learning the correct mapping and its distribution}
+
+\glnote{This section needs to be improved.}
 
 In order for this approach to be useful in the likelihood-free setting, where it
-is not possible to evaluate the density $p(x|\theta)$, we need to be able to
-approximate both $s(x|\theta_0, \theta_1)$ and $p(s|\theta)$ based on samples
-$\{x_i\}$ drawn from the generative model $p(x|\theta)$.  Denote the
-dimensionality reduction map $\hat{s}(x; \theta_0, \theta_1)$ and its
-distribution $\hat{p}(\hat{s}|\theta)$.
+is not possible to evaluate the density $p(\mathbf{x}|\theta)$, we need to be
+able to approximate both $s(\mathbf{x};\theta_0, \theta_1)$ and
+$p(s(\mathbf{x})|\theta)$ based on samples $\{x_i\}$ drawn from the generative
+model $p(\mathbf{x}|\theta)$.  Denote the dimensionality reduction map
+$\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{p}(\hat{s}|\theta)$.
 
 One strength of this approach is that it factorizes the approximation of the
-per-event likelihood ratio ($\hat{s}(x; \theta_0, \theta_1) \approx s(x;
-\theta_0, \theta_1)$) from the calibration procedure ($\hat p(\hat s| \theta)
-\approx p(\hat{s}|\theta)$). Thus, even if the classifier does a poor job of
-reproducing the level sets of the per-event likelihood ratio, the density of
-$\hat{s}$ can still be well calibrated. In that case, one might loose power, but
-the resulting inference will still be valid. This point was made by
+per-event likelihood ratio ($\hat{s}(\mathbf{x}; \theta_0, \theta_1) \approx
+s(\mathbf{x}; \theta_0, \theta_1)$) from the calibration procedure ($\hat p(\hat
+s| \theta) \approx p(\hat{s}|\theta)$). Thus, even if the classifier does a poor
+job of reproducing the level sets of the per-event likelihood ratio, the density
+of $\hat{s}$ can still be well calibrated. In that case, one might loose power,
+but the resulting inference will still be valid. This point was made by
 \cite{Neal:2007zz} and is well appreciated by the particle physics community
@@
 
-\subsection{The fixed discriminative classification setting} For fixed
-$\theta_0$ and $\theta_1$ we can generate large samples from each model and
-train a classifier. To be concrete, let us use $p(x|\theta_0)$ to generate
-training data $(x_i,  y_i=0)$ and $p(x|\theta_1)$ to generate training data
-$(x_i , y_i=1)$. With balanced training data
-\mbox{($p(y=1)=p(y=0)=\sfrac{1}{2}$)} a quadratic loss function will lead to
-classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) =
-p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$, which is  monotonic with the
-desired per-event likelihood ratio $q(x)$. Thus, standard supervised learning
-algorithms with various surrogate loss functions lead to discriminative
-classifiers that approximate a monotonic function of per-event likelihood ratio
-$q(x)$.
+\subsection{Classification and frequentist hypothesis tests}
+
+\glnote{Shall we keep everything from this?}
+
+Vast literature exists around generative and discriminative
+classifiers~\citep{AndrewY.Ng}. Typically, generative classifiers learn a model
+for the joint probability $p(\mathbf{x}, y)$, of the inputs $x$ and the classification
+label $y$, and predict $p(y|\mathbf{x})$ via Bayes rule. In contrast, discriminative
+classifiers model the posterior $p(y|\mathbf{x})$ directly. For classification tasks, one
+then thresholds on $p(y|\mathbf{x})$. In both cases this description in terms of a
+posterior requires a prior distribution for $p(y)$, which is either modeled
+explicitly or learned from the training data. This familiar formulation of
+classification may lead to some confusion in the setting of the current work.
+
+The first possible source of confusion we wish to avoid is that here
+$p(\mathbf{x}|\theta)$  is a  \textit{generative statistical model} for the features $\mathbf{x}$,
+not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a
+traditional scientific theory or simulator, able to make predictions about $\mathbf{x}$
+and being motivated by domain-specific considerations.
+
+The second possible source of confusion is that we are not directly interested
+in calibrating the classification score in terms of a per-event posterior
+probability $p(y|\mathbf{x})$. Instead, we are interested in the approximation of the
+per-experiment likelihood function or likelihood ratio, which might be used for
+several purposes, including the calculation of p-values.
+
+Lastly, in the setting of frequentist hypothesis tests and confidence intervals,
+we do not have a prior $\pi(\theta)$. While we can use the generative models to
+produce training data $(\mathbf{x} ,y=0)$ generated from
+$p(\mathbf{x}|\theta_0)$ and $(\mathbf{x}, y=1)$ generated from
+$p(\mathbf{x}|\theta_1)$, the relative mix $p(y)$ is arbitrary. Since the prior
+$p(y)$ is not needed for the target likelihood ratio test and because the
+classifier score $p(y|\mathbf{x})$ may not be well calibrated, we choose to denote the
+classifier score $s(\mathbf{x})$ and simply think of it as a deterministic dimensionality
+reduction map $s: \mathbb{R}^p \to \mathbb{R}$.  Similar points have been made
+by~\cite{ClaytonScott} and \cite{Neal:2007zz}.
+
+% Generalized likelihood ratio tests ===========================================
+
+% - mathematical problem (was section 4)
+% - maximum likelihood estimator inference from ratios
+% - discussion on parameterized classifier and practicalities regarding embedding (was sections 5.2 and 5.3)
+
+\section{Generalized likelihood ratio tests}
+\label{sec:generalized-likelihood-ratio}
+
+% Thus far we have shown that the target likelihood ratio
+% $p(x|\theta_0)/p(x|\theta_1)$ with high dimensional features $x$ can be
+% reproduced via the univariate densities $p(s|\theta_0)/p(s|\theta_1)$ if the
+% classifier $s(x|\theta_0, \theta_1)$ is a strictly increasing function of
+% $p(x|\theta_0)/p(x|\theta_1)$. We now generalize from the ratio of two simple
+% hypotheses specified by $\theta_0$ and $\theta_1$ to the case where $\theta$ are
+% continuous model parameters. We postpone the practicalities of training the
+% classifier and estimating the density to Section~\ref{S:classifier} and continue
+% in the likelihood-free setting with idealized classifiers and their densities.
+%
+% In the case of a fixed classifier $s(x)$ it is possible to compute $s_e=s(x_e)$
+% for the observed data and never refer back to the original features $x_e$. In
+% the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0,
+% \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
+%
+% The critical observation is that  if we postpone the evaluation of the
+% classifier to the stage of evaluating the enveloping likelihood ratio, then we
+% can identify the value of the parameters  that are being compared in the
+% likelihood ratio with the values used as input to $s(x;\theta_0, \theta_1)$.
+%
+% \begin{equation}\label{eq:embedding}
+% T(D; \theta_0, \theta_1) = \prod_e \frac{p(x_e|\theta_0)}{p(x_e|\theta_1)} = \prod_e  \frac{ p (s(x_e; \theta_0, \theta_1) | \theta_0)}
+% {p (s(x_e; \theta_0,  \theta_1) | \theta_1) } \; .
+% \end{equation}
+%
+% This is equivalent to approximating the likelihood function for $\theta_0$  when
+% $\theta_1$ is held fixed.
+
+% \section{Composite hypotheses and the generalized likelihood ratio}\label{S:GLR}
+
+In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
+$\theta \in \Theta_0^C$, the generalized likelihood ratio\footnote{Also known as
+the profile likelihood ratio.} test is commonly used
+\begin{equation}
+\lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p(D | \theta)}{ \sup_{\theta \in \Theta} p(D | \theta)} \; .
+\end{equation}
+This generalized likelihood ratio can be used both for hypothesis tests in the
+presence of nuisance parameters or to create confidence intervals with or
+without nuisance parameters.  Often, the parameter vector is broken into two
+components $\theta=(\mu,\nu)$, where the $\mu$ components are considered
+parameters of interest while the $\nu$ components are considered nuisance
+parameters. In that case $\Theta_0$ corresponds to all values of $\nu$ with
+$\mu$ fixed.
+
+Denote the maximum likelihood estimator
+\begin{equation}\label{eq:mle}
+\hat{\theta} = \argmax_\theta  p(D | \theta)
+\end{equation}
+and the conditional maximum likelihood estimator
+\begin{equation}\label{eq:cmle}
+\hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p(D | \theta) \; .
+\end{equation}
+
+It is not obvious that if we are working with the distributions $p(s|\theta)$
+(for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the
+same estimators. Fortunately, there is a construction based on $p(s|\theta)$
+that works. The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as
+the value that maximizes the likelihood ratio with respect to $p(D|\theta_1)$
+for some fixed value of $\theta_1$. This allows us to use Theorem~1 toreformulate the maximum likelihood estimate
+
+\begin{equation}\label{eq:mle_withs}
+\hat{\theta} = %\argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} =
+\argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
+\end{equation}
+It is important that we include the denominator $p(s(x_e; \theta, \theta_1)
+|\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.
 
@@
 
-In order to provide parameter inference in the likelihood-free setting, we must train a family of  classifiers parametrized by $\theta_0$ and $\theta_1$, the
-parameters associated to the null and alternate hypotheses, respectively. While this could be done independently
-for all $\theta_0$ and $\theta_1$, it is desirable and convenient to have a smooth evolution of the classification score as a function of the parameters.
-Thus, we anticipate a single learning stage based on training data with input $(x, \theta_0, \theta_1)_i$ and target $y_i$.
-Somewhat unusually, the unknown values of the parameters are taken
-as input to the classifier; their values will be specified via the enveloping (generalized) likelihood ratio of Eq.~\ref{eq:enveloping}.
-We denote the learned family of classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based roughly on the following algorithmic flow.
+In order to provide parameter inference in the likelihood-free setting, we must
+train a family of  classifiers parametrized by $\theta_0$ and $\theta_1$, the
+parameters associated to the null and alternate hypotheses, respectively. While
+this could be done independently for all $\theta_0$ and $\theta_1$, it is
+desirable and convenient to have a smooth evolution of the classification score
+as a function of the parameters. Thus, we anticipate a single learning stage
+based on training data with input $(x, \theta_0, \theta_1)_i$ and target $y_i$.
+Somewhat unusually, the unknown values of the parameters are taken as input to
+the classifier; their values will be specified via the enveloping (generalized)
+likelihood ratio of Eq.~\ref{eq:enveloping}. We denote the learned family of
+classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based
+roughly on the following algorithmic flow.
 % \begin{algorithm}[ht]
@@
 % - ratio of mixtures and decomposition (special case of generalized likelihood rario tests when parameters only affect the mixture coefficients => no need for parametrized classifiers)
+%   discuss about focusing the capacity of the classifier
 % - measure of particles properties, with a concrete example. Comparison with other likelihood-free inference methods (new)

### 2016-01-21 jasa/paper-jasa.tex
Continued

@@
 
+\glnote{Needs some work.}
+
 % Thus far we have shown that the target likelihood ratio
@@
 % in the likelihood-free setting with idealized classifiers and their densities.
+
+Thus far we have shown that the target likelihood ratio
+$p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$ with high dimensional features
+$\mathbf{x}$ can be reproduced via the univariate densities
+$p(s(\mathbf{x};\theta_0,\theta_1)|\theta_0)/p(s(\mathbf{x};\theta_0,\theta_1)|\theta_1)$
+if the reduction $s(\mathbf{x}|\theta_0, \theta_1)$ is monotonic with
+$p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$. We now generalize from the
+ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the
+case of composite hypothesis testing where $\theta$ are continuous model
+parameters.
 %
@@
 In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
-$\theta \in \Theta_0^C$, the generalized likelihood ratio\footnote{Also known as
-the profile likelihood ratio.} test is commonly used
-\begin{equation}
-\lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p(D | \theta)}{ \sup_{\theta \in \Theta} p(D | \theta)} \; .
-\end{equation}
+$\theta \in \Theta_0^C$, the generalized likelihood ratio test, also known as
+the profile likelihood ratio test, is commonly used
+\begin{equation}\label{eqn:generalized-lr}
+\lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
+\end{equation}
+\glnote{For the denominator, shouldnt we have $\theta \in \Theta_0^C$ instead?}
 This generalized likelihood ratio can be used both for hypothesis tests in the
@@
 
-Denote the maximum likelihood estimator
+Let denote the maximum likelihood estimator
 \begin{equation}\label{eq:mle}
-\hat{\theta} = \argmax_\theta  p(D | \theta)
-\end{equation}
-and the conditional maximum likelihood estimator
-\begin{equation}\label{eq:cmle}
-\hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p(D | \theta) \; .
-\end{equation}
-
+\hat{\theta} = \argmax_\theta  p({\cal D} | \theta) \; .
+\end{equation}
+% and the conditional maximum likelihood estimator
+% \begin{equation}\label{eq:cmle}
+% \hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p({\cal D} | \theta) \; .
+% \end{equation}
 It is not obvious that if we are working with the distributions $p(s|\theta)$
-(for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the
+(for some particular $s(\mathbf{x}; \theta_0, \theta_1)$ comparison) that we can find the
 same estimators. Fortunately, there is a construction based on $p(s|\theta)$
 that works. The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as
-the value that maximizes the likelihood ratio with respect to $p(D|\theta_1)$
-for some fixed value of $\theta_1$. This allows us to use Theorem~1 toreformulate the maximum likelihood estimate
-
-\begin{equation}\label{eq:mle_withs}
-\hat{\theta} = %\argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} =
-\argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
-\end{equation}
-It is important that we include the denominator $p(s(x_e; \theta, \theta_1)
+the value that maximizes the likelihood ratio with respect to $p({\cal
+D}|\theta_1)$ for some fixed value of $\theta_1$. This allows us to use
+Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood estimate:
+\begin{align}\label{eq:mle_withs}
+\hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
+&= \argmax_\theta  \prod \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
+&= \argmax_\theta  \prod \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \; .
+\end{align}
+Note that it is important that we include the denominator $p(s(\mathbf{x}; \theta, \theta_1)
 |\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.
 
-\subsection{Training a parametrized classifier}
+\subsection{Parametrized classifier}
+
+\glnote{Needs to be clarified.}
+\glnote{Do we put back the algorithm snippet? I think this could be directly
+explained in words.}
 
@@
 as a function of the parameters. Thus, we anticipate a single learning stage
-based on training data with input $(x, \theta_0, \theta_1)_i$ and target $y_i$.
+based on training data with input $(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$.
 Somewhat unusually, the unknown values of the parameters are taken as input to
 the classifier; their values will be specified via the enveloping (generalized)
-likelihood ratio of Eq.~\ref{eq:enveloping}. We denote the learned family of
-classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based
+likelihood ratio of Eqn.~\ref{eqn:generalized-lr}. We denote the learned family of
+classifiers $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$, and anticipate the training based
 roughly on the following algorithmic flow.
+
 % \begin{algorithm}[ht]
@@
 
-While the function $p(x|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ will minimize
+While the function $p(\mathbf{x}|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ will minimize
 the expected squared loss based on training data produced according to
@@
 
-\subsection{Practicalities of embedding the classifier into the likelihood}
+\subsection{Parametrized calibration}
+
+\glnote{Needs to be clarified as well.}
 
@@
 
-There is a mild technical challenge in embedding the classifier into the
-likelihood. In the case of a fixed classifier $\hat s(x)$ it is possible to
-pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features
-$x_e$. In the parametrized setting it is not possible to pre-compute $\hat
-s(x_e; \theta_0, \theta_1)$ since one does not know the true values of
-$\theta_0$ and $\theta_1$. Thus one must implement the embedding of the
-classifier as in Eq.~\ref{eq:embedding}.  A concrete realization of this has
-been performed for probability models implemented with the \texttt{RooFit}
-probabilistic programing language and  classifiers implemented with
-\texttt{scikit-learn} and
-\texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
+% There is a mild technical challenge in embedding the classifier into the
+% likelihood. In the case of a fixed classifier $\hat s(x)$ it is possible to
+% pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features
+% $x_e$. In the parametrized setting it is not possible to pre-compute $\hat
+% s(x_e; \theta_0, \theta_1)$ since one does not know the true values of
+% $\theta_0$ and $\theta_1$. Thus one must implement the embedding of the
+% classifier as in Eq.~\ref{eq:embedding}.
+%  A concrete realization of this has
+% been performed for probability models implemented with the \texttt{RooFit}
+% probabilistic programing language and  classifiers implemented with
+% \texttt{scikit-learn} and
+% \texttt{TMVA}~\citep{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.
 
@@
 
-In HEP we are often searching for some class of events, generically referred to
-as \textit{signal}, in the presence of a separate class of \textit{background}
-events.  For each event we measure some quantities $x$ that have corresponding
-distributions $p_b(x|\nu)$ for background and $p_s(x|\nu)$ for signal, where
-$\nu$ are nuisance parameters describing uncertainties in the underlying physics
-prediction or response of the measurement device. The total model is a mixture
-of the signal and background, and $\mu$ is the mixture coefficient associate to
-the signal component.
+\subsection{High energy physics}
+
+\glnote{Add a concrete HEP example where we illustrate and compare
+the method for several classifiers and calibration algorithms.}
+\glnote{Notations need to be adapted.}
+\glnote{Rethink the outline? Start with comparison of mixtures (5.1.4) and then
+specialize to background vs background+signal, and thereby discuss that we can
+focus the capacity of the classifier.}
+
+In high energy physics, we are often searching for some class of events,
+generically referred to as \textit{signal}, in the presence of a separate class
+of \textit{background} events.  For each event we measure some quantities $x$
+that have corresponding distributions $p_b(x|\nu)$ for background and
+$p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing
+uncertainties in the underlying physics prediction or response of the
+measurement device. The total model is a mixture of the signal and background,
+and $\mu$ is the mixture coefficient associate to the signal component.
 \begin{equation}\label{eq:hepGen}
@@
 
-\subsection{Typical usage pattern}
+\subsubsection{Typical usage pattern}
 
@@
 
-\begin{figure}[htbp]
+\begin{figure}[t]
 \begin{center}
@@
 
-\subsection{Comments on typical usage of machine learning in HEP}
+\subsubsection{Comments on typical usage of machine learning in HEP}
 
@@
 
-\subsection{A more powerful  approach}
+\subsubsection{A more powerful  approach}
 
@@
 
-\subsection{Decomposing tests between mixture models into their components}
+\subsubsection{Decomposing tests between mixture models into their components}
 
@@
 
-\subsection{Measuring particle properties}
+\subsection{Likelihood-free inference}
+
+\glnote{Rework this section, add a concrete example and compare with other
+likelihood-free inference algorithms.}
 
@@
 
-\section{Related works}\label{S:Related}
+\section{Related works}
+\label{sec:related}
+
+\glnote{Add related works on density ratio estimation.}
+\glnote{Remove others that are less directly related.}
 
@@
 \section{Conclusions}
+\label{sec:conclusions}
 

### 2016-01-21 jasa/paper-jasa.tex
Section 1: nitpicks

@@
 hypothesis tests or confidence intervals using the (generalized or profile)
-likelihood ratio as a test statistic. With the advance of computing technology,
-it has become increasingly common that a simulator (or generative model) is used
-to describe complex processes that tie parameters $\theta$ of an underlying
-theory and measurement apparatus to high-dimensional observations $\mathbf{x}$. However,
-directly evaluating the likelihood function in these cases is often impossible
-or is computationally impractical. In this likelihood-free setting, various
-methods for statistical inference have been proposed, including most notably the
-Approximate Bayesian Computation (ABC) class of algorithms aiming at sampling
-from the posterior distribution of model parameters \glnote{Add references}.
+likelihood ratio as a test statistic. At the same time, with the advance of
+computing technology, it has become increasingly common that a simulator (or
+generative model) is used to describe complex processes that tie parameters
+$\theta$ of an underlying theory and measurement apparatus to high-dimensional
+observations $\mathbf{x}$. However, directly evaluating the likelihood function
+in these cases is often impossible or is computationally impractical. In this
+likelihood-free setting, various methods for statistical inference have been
+proposed, including most notably the Approximate Bayesian Computation (ABC)
+class of algorithms aiming at sampling from the posterior distribution of model
+parameters \glnote{Add references}.
 
 The main result of this paper is to show that the likelihood ratio is invariant
-under dimensionality reduction, under the assumption that the transformation is
-monotonic with the likelihood ratio. As a direct consequence, we derive and propose an
-alternative machine learning-based approach for likelihood-free inference that
-can also be used in a classical (frequentist) setting where a prior over the
-model parameters is not available. More specifically, we demonstrate why and how
-discriminative classifiers can be used to construct equivalent likelihood ratio
-tests when only a generative model for the data is available for training and
-calibration.
+under dimensionality reduction, under the assumption that the corresponding
+transformation is monotonic with the likelihood ratio. As a direct consequence,
+we derive and propose an alternative machine learning-based approach for
+likelihood-free inference that can also be used in a classical (frequentist)
+setting where a prior over the model parameters is not available. More
+specifically, we demonstrate why and how discriminative classifiers can be used
+to construct equivalent likelihood ratio tests when only a generative model for
+the data is available for training and calibration.
 
@@
 and data processing algorithms that transform raw sensor data into the feature
-vector $x$~\citep{Sjostrand:2006za,Agostinelli:2002hh}. The ATLAS and CMS
+vector $\mathbf{x}$~\citep{Sjostrand:2006za,Agostinelli:2002hh}. The ATLAS and CMS
 experiments have published  hundreds papers where the final result was
@@
 
-The rest of this paper is organized as follows. In Section...
+The rest of this paper is organized as follows. In Section... \glnote{todo}
 

### 2016-01-21 jasa/paper-jasa.tex
Section 2: line break
Section: postpone parametrization of s(x) until section 4

@@
 \end{equation}
-is the most powerful test statistic. In order to evaluate $T({\cal D})$, one
-must be able to evaluate the probability density $p_\mathbf{X}(\mathbf{x}|
-\theta)$ at any value $\mathbf{x}$. However, it is increasingly common in
-science that one has a complex simulation that can act as generative model for
-$p_\mathbf{X}(\mathbf{x}|\theta)$, but one cannot evaluate the density directly.
-For instance, this is the case in high energy physics~\citep{Neal:2007zz} where
-the simulation of particle detectors can only be done in the forward mode.
-\glnote{Reinsert citations for \cite{ClaytonScott} and \cite{JMLR:v14:tong13a}.}
+is the most powerful test statistic.
+
+In order to evaluate $T({\cal D})$, one must be able to evaluate the probability
+density $p_\mathbf{X}(\mathbf{x}| \theta)$ at any value $\mathbf{x}$. However,
+it is increasingly common in science that one has a complex simulation that can
+act as generative model for $p_\mathbf{X}(\mathbf{x}|\theta)$, but one cannot
+evaluate the density directly. For instance, this is the case in high energy
+physics~\citep{Neal:2007zz} where the simulation of particle detectors can only
+be done in the forward mode. \glnote{Reinsert citations for \cite{ClaytonScott}
+and \cite{JMLR:v14:tong13a}.}
 
@@
 \begin{equation}\label{eqn:likelihood-ratio-test-equiv}
-T'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{Y}(y=s(\mathbf{x}; \theta_0, \theta_1) | \theta_0)}{ p_\mathbf{Y}(y=s(\mathbf{x}; \theta_0, \theta_1) | \theta_1)}
+T'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_0)}{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_1)}
 \end{equation}
 that is strictly equivalent to \ref{eqn:likelihood-ratio-test}, provided the change
-of variable $\mathbf{Y} = s(\mathbf{X}; \theta_0, \theta_1)$ is based
-on a parametrized function $s$ that is strictly monotonic with the density ratio
+of variable $\mathbf{Y} = s(\mathbf{X})$ is based
+on a (parametrized) function $s$ that is strictly monotonic with the density ratio
 \begin{equation}
@@
 an alternate form in which supervised learning can be used to build
-$s(\mathbf{x}; \theta_0, \theta_1)$ as a discriminative classifier.  In
+$s(\mathbf{x})$ as a discriminative classifier.  In
 Section~\ref{sec:generalized-likelihood-ratio} we extend this result to
@@
     \begin{equation}
-        r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} = \frac{p_\mathbf{Y}(y=s(\mathbf{x};\theta_0,\theta_1)|\theta_0)}{p_\mathbf{Y}(y=s(\mathbf{x};\theta_0,\theta_1)|\theta_1)},
+        r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} = \frac{p_\mathbf{Y}(y=s(\mathbf{x})|\theta_0)}{p_\mathbf{Y}(y=s(\mathbf{x})|\theta_1)},
     \end{equation}
@@
 \begin{proof}
-Omitting for clarity the parametrization $(\theta_0,\theta_1)$ of $s(\mathbf{X};\theta_0,\theta_1)$, we have
+Starting from the definition of the probability density function, we have
 \begin{align}
@@
 \end{equation}
-
 In particular, this regression function satisfies conditions of
@@
 
-\glnote{This section needs to be improved.}
+\glnote{This section needs to be improved. Not sure we need a section on its own for that either.}
 
@@
 is not possible to evaluate the density $p(\mathbf{x}|\theta)$, we need to be
-able to approximate both $s(\mathbf{x};\theta_0, \theta_1)$ and
+able to approximate both $s(\mathbf{x})$ and
 $p(s(\mathbf{x})|\theta)$ based on samples $\{x_i\}$ drawn from the generative
 model $p(\mathbf{x}|\theta)$.  Denote the dimensionality reduction map
-$\hat{s}(x; \theta_0, \theta_1)$ and its distribution $\hat{p}(\hat{s}|\theta)$.
+$\hat{s}(x)$ and its distribution $\hat{p}(\hat{s}|\theta)$.
 
 One strength of this approach is that it factorizes the approximation of the
-per-event likelihood ratio ($\hat{s}(\mathbf{x}; \theta_0, \theta_1) \approx
-s(\mathbf{x}; \theta_0, \theta_1)$) from the calibration procedure ($\hat p(\hat
+per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx
+s(\mathbf{x})$) from the calibration procedure ($\hat p(\hat
 s| \theta) \approx p(\hat{s}|\theta)$). Thus, even if the classifier does a poor

### 2016-01-21 jasa/paper-jasa.tex
Section 4: introduce parametrization of s

@@
 Thus far we have shown that the target likelihood ratio
-$p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$ with high dimensional features
+$r(\mathbf{x};\theta_0,\theta_1)=p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$ with high dimensional features
 $\mathbf{x}$ can be reproduced via the univariate densities
-$p(s(\mathbf{x};\theta_0,\theta_1)|\theta_0)/p(s(\mathbf{x};\theta_0,\theta_1)|\theta_1)$
-if the reduction $s(\mathbf{x}|\theta_0, \theta_1)$ is monotonic with
-$p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$. We now generalize from the
+$p(s(\mathbf{x})|\theta_0)$ and $p(s(\mathbf{x})|\theta_1)$
+if the reduction $s(\mathbf{x})$ is monotonic with
+$r(\mathbf{x};\theta_0,\theta_1)$. We now generalize from the
 ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the
@@
 \begin{equation}\label{eqn:generalized-lr}
-\lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
-\end{equation}
-\glnote{For the denominator, shouldnt we have $\theta \in \Theta_0^C$ instead?}
+\Lambda({\cal D}; \Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
+\end{equation}
 This generalized likelihood ratio can be used both for hypothesis tests in the
@@
 % \end{equation}
-It is not obvious that if we are working with the distributions $p(s|\theta)$
-(for some particular $s(\mathbf{x}; \theta_0, \theta_1)$ comparison) that we can find the
-same estimators. Fortunately, there is a construction based on $p(s|\theta)$
-that works. The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as
+It is not obvious that if we are working instead with distributions $p(s(\mathbf{x})|\theta)$
+that we can find the
+same estimators. Fortunately, there is a construction based on $s$
+that works. The maximum likelihood estimate of Eqn.~\ref{eq:mle} is the same as
 the value that maximizes the likelihood ratio with respect to $p({\cal
@@
 &= \argmax_\theta  \prod \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
-&= \argmax_\theta  \prod \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \; .
+&= \argmax_\theta  \prod \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;
 \end{align}
-Note that it is important that we include the denominator $p(s(\mathbf{x}; \theta, \theta_1)
+where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parametrized} transformation $s$
+of $X$ in terms of $(\theta,\theta_1)$. Note that it is important that we include the denominator $p(s(\mathbf{x}; \theta, \theta_1)
 |\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.

### 2016-01-21 jasa/paper-jasa.tex

@@
 
-Let denote the maximum likelihood estimator
+Evaluating Eqn.~\ref{eqn:generalized-lr} requires finding the maximum likelihood
+estimator
 \begin{equation}\label{eq:mle}
-\hat{\theta} = \argmax_\theta  p({\cal D} | \theta) \; .
-\end{equation}
-% and the conditional maximum likelihood estimator
-% \begin{equation}\label{eq:cmle}
-% \hat{\hat{\theta}} = \argmax_{\theta \in \Theta_0}  p({\cal D} | \theta) \; .
-% \end{equation}
-It is not obvious that if we are working instead with distributions $p(s(\mathbf{x})|\theta)$
-that we can find the
-same estimators. Fortunately, there is a construction based on $s$
-that works. The maximum likelihood estimate of Eqn.~\ref{eq:mle} is the same as
-the value that maximizes the likelihood ratio with respect to $p({\cal
-D}|\theta_1)$ for some fixed value of $\theta_1$. This allows us to use
-Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood estimate:
+    \hat{\theta} = \argmax_\theta p({\cal D} | \theta),
+\end{equation}
+which is again
+made difficult in the likelihood-free setting.  Also, it is not obvious that if
+we are working instead with distributions $p(s(\mathbf{x})|\theta)$ that we can
+find the same estimators. Fortunately, there is a construction based on $s$ that
+works: the maximum likelihood estimate of Eqn.~\ref{eq:mle} is the same as the
+value that maximizes the likelihood ratio with respect to $p({\cal D}|\theta_1)$,
+for some fixed value of $\theta_1$. This allows us to use
+Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
+estimate as
 \begin{align}\label{eq:mle_withs}
@@
 &= \argmax_\theta  \prod \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
-&= \argmax_\theta  \prod \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;
+&= \argmax_\theta  \prod \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;,
 \end{align}
-where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parametrized} transformation $s$
-of $X$ in terms of $(\theta,\theta_1)$. Note that it is important that we include the denominator $p(s(\mathbf{x}; \theta, \theta_1)
-|\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.
+where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parametrized}
+transformation $s$ of $X$ in terms of $(\theta,\theta_1)$ that is monotonic
+with $r(\mathbf{x};\theta,\theta_1)$. Note that it is important that we include
+the denominator $p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)$ because this
+cancels Jacobian factors that vary with $\theta$.
 

### 2016-01-22 jasa/paper-jasa.tex
Section 4: further improvements

@@
 \usepackage{color}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 \subsection{Probabilistic classification for approximating ratios}
+\label{sec:clf-for-ratios}
 
@@
 
-\glnote{Needs some work.}
-
 % Thus far we have shown that the target likelihood ratio
@@
 Thus far we have shown that the target likelihood ratio
-$r(\mathbf{x};\theta_0,\theta_1)=p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$ with high dimensional features
-$\mathbf{x}$ can be reproduced via the univariate densities
-$p(s(\mathbf{x})|\theta_0)$ and $p(s(\mathbf{x})|\theta_1)$
-if the reduction $s(\mathbf{x})$ is monotonic with
-$r(\mathbf{x};\theta_0,\theta_1)$. We now generalize from the
-ratio of two simple hypotheses specified by $\theta_0$ and $\theta_1$ to the
-case of composite hypothesis testing where $\theta$ are continuous model
-parameters.
+$r(\mathbf{x};\theta_0,\theta_1)=p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$
+with high dimensional features $\mathbf{x}$ can be reproduced via the univariate
+densities $p(s(\mathbf{x})|\theta_0)$ and $p(s(\mathbf{x})|\theta_1)$ if the
+reduction $s(\mathbf{x})$ is monotonic with $r(\mathbf{x};\theta_0,\theta_1)$.
+We now generalize from the ratio of two simple hypotheses specified by
+$\theta_0$ and $\theta_1$ to the case of composite hypothesis testing where
+$\theta$ are continuous model parameters.
 %
@@
 In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
-$\theta \in \Theta_0^C$, the generalized likelihood ratio test, also known as
-the profile likelihood ratio test, is commonly used
+$\theta \in \Theta_1$ (possibly such that $\Theta_0 \subseteq \Theta_1$), the
+generalized likelihood ratio test, also known as the profile likelihood ratio
+test, is commonly used
 \begin{equation}\label{eqn:generalized-lr}
-\Lambda({\cal D}; \Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
+\Lambda({\cal D}; \Theta_0; \Theta_1) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta_1} p({\cal D} | \theta)} \; .
 \end{equation}
@@
 
-Evaluating Eqn.~\ref{eqn:generalized-lr} requires finding the maximum likelihood
-estimator
+Evaluating the generalized likelihood ratio as defined by
+Eqn.~\ref{eqn:generalized-lr} requires finding for both the numerator and the
+denominator the maximum likelihood estimator
 \begin{equation}\label{eq:mle}
-    \hat{\theta} = \argmax_\theta p({\cal D} | \theta),
-\end{equation}
-which is again
-made difficult in the likelihood-free setting.  Also, it is not obvious that if
-we are working instead with distributions $p(s(\mathbf{x})|\theta)$ that we can
-find the same estimators. Fortunately, there is a construction based on $s$ that
-works: the maximum likelihood estimate of Eqn.~\ref{eq:mle} is the same as the
-value that maximizes the likelihood ratio with respect to $p({\cal D}|\theta_1)$,
-for some fixed value of $\theta_1$. This allows us to use
-Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
+    \hat{\theta} = \argmax_\theta p({\cal D} | \theta).
+\end{equation}
+Again, this is made difficult in the likelihood-free setting and it is not
+obvious that we can find the same estimators if we are working instead with
+$p(s(\mathbf{x})|\theta)$. Fortunately, there is a construction
+based on $s$ that works: the maximum likelihood estimate of Eqn.~\ref{eq:mle} is
+the same as the value that maximizes the likelihood ratio with respect to
+$p({\cal D}|\theta_1)$, for some fixed value of $\theta_1$. This allows us to
+use Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
 estimate as
@@
 \hat{\theta} &= \argmax_\theta  p({\cal D} | \theta) \nonumber \\
-&= \argmax_\theta  \prod \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
-&= \argmax_\theta  \prod \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;,
+&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(\mathbf{x}| \theta)}{p(\mathbf{x}|\theta_1)} \nonumber \\
+&= \argmax_\theta  \prod_{\mathbf{x} \in {\cal D}} \frac{p(s(\mathbf{x}; \theta, \theta_1) | \theta)}{p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)} \;,
 \end{align}
 where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parametrized}
-transformation $s$ of $X$ in terms of $(\theta,\theta_1)$ that is monotonic
+transformation $s$ of $\mathbf{X}$ in terms of $(\theta,\theta_1)$ that is monotonic
 with $r(\mathbf{x};\theta,\theta_1)$. Note that it is important that we include
@@
 
-\glnote{Needs to be clarified.}
-\glnote{Do we put back the algorithm snippet? I think this could be directly
-explained in words.}
-
-In order to provide parameter inference in the likelihood-free setting, we must
-train a family of classifiers parametrized by $\theta_0$ and $\theta_1$, the
-parameters associated to the null and alternate hypotheses, respectively. While
-this could be done independently for all $\theta_0$ and $\theta_1$, it is
-desirable and convenient to have a smooth evolution of the classification score
-as a function of the parameters. Thus, we anticipate a single learning stage
-based on training data with input $(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$.
-Somewhat unusually, the unknown values of the parameters are taken as input to
-the classifier; their values will be specified via the enveloping (generalized)
-likelihood ratio of Eqn.~\ref{eqn:generalized-lr}. We denote the learned family of
-classifiers $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$, and anticipate the training based
-roughly on the following algorithmic flow.
-
-% \begin{algorithm}[ht]
-% \caption{Training of the parametrized classifier.}\label{alg:training}
-% \begin{algorithmic}
-% \STATE initialize trainingData to \{\}
-% \FOR{ $\theta_0$ in $\Theta$ }
-% 	\FOR{ $\theta_1$ in $\Theta$ }
-% 		\STATE generate $x_i^0 \sim p(x|\theta_0)$
-% 		\STATE append $\{ (x_i^0, \theta_0, \theta_1, y=0) \}$ to trainingData
-% 		\STATE generate $x_i^1 \sim p(x|\theta_1)$
-% 		\STATE append $\{ (x_i^1, \theta_0, \theta_1, y=1) \}$ to trainingData
-% 	\ENDFOR
-% \ENDFOR
-% \STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_1)$
-% \end{algorithmic}
-% \end{algorithm}%\vspace{-1.5em}
-
-While the function $p(\mathbf{x}|\theta_1)/(p(x|\theta_0)+p(x|\theta_1))$ will minimize
-the expected squared loss based on training data produced according to
-Algorithm~\ref{alg:training}, it is not clear how training data from $\theta'_0
-\ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real world
-classifier with finite capacity. This is left as an area for future work.
+In order to provide parameter inference in the likelihood-free setting as
+described above, we must train a family $s(\mathbf{x};\theta,\theta_1)$ of classifiers parametrized by
+$\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
+hypotheses, respectively. While this could be done independently for all
+$\theta_0$ and $\theta_1$, using the procedure outlined in
+Section~\ref{sec:clf-for-ratios}, it is desirable and convenient to have a
+smooth evolution of the classification score as a function of the parameters.
+For this reason, we anticipate a single learning stage based on training data with input
+$(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
+unknown values of the parameters are taken as input to the classifier; their
+values will be specified via the enveloping (generalized) likelihood ratio of
+Eqn.~\ref{eqn:generalized-lr}.
+
+\begin{algorithm}[t]
+\caption{Training of a parametrized classifier.}\label{alg:training}
+    \begin{algorithmic}
+        \State ${\cal L := \{ \}}$
+        \For{ $\theta_0$ in $\Theta_0$ }
+        	\For{ $\theta_1$ in $\Theta_1$ }
+        		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$
+        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, y=0) \}$ to ${\cal L}$
+        		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$
+        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, y=1) \}$ to ${\cal L}$
+        	\EndFor
+        \EndFor
+        \State Learn $s(\mathbf{x}; \theta_0, \theta_1)$ from ${\cal L}$
+    \end{algorithmic}
+\end{algorithm}
+
+While the function
+$p(\mathbf{x}|\theta_1)/(p(\mathbf{x}|\theta_0)+p(\mathbf{x}|\theta_1))$ will
+minimize the expected squared error loss based on training data produced
+according to Algorithm~\ref{alg:training}, it is not clear how training data
+from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real
+world classifier with finite capacity. This is left as an area for future work.
+
+\glnote{Shall we keep Algorithm~\ref{alg:training} or rather give a sketch in
+words? In its current state, the algorithm does not explain how to build ${\cal
+L}$ when the parameter spaces are continuous. It also does not specify how many
+samples should be generated per parameter value.}
 
 \subsection{Parametrized calibration}
-
-\glnote{Needs to be clarified as well.}
 
@@
 univariate density estimation technique (e.g. histograms or kernel density
-estimation) to approximate $\hat{p}(\hat{s}|\theta)$ for specific parameter
-points. For an single parameter point, this is a tractable univariate density
-estimation problem. The challenge comes from the need to estimate this density
+estimation) to approximate $p(s|\theta)$ for specific parameter
+points. For a single parameter point, this is a tractable univariate density
+estimation problem. The challenge comes from the need to calibrate this density
 for all values of $\theta$. A straight forward approach would be to run the
@@
 trying to maximize the likelihood with respect to $\theta$ needs access to the
-generative model $p(x|\theta)$. This can be  impractical when the generative
+generative model $p(\mathbf{x}|\theta)$. This can be  impractical when the generative
 model is computationally expensive or has high-latency (for instance some human
-intervention is required to reconfigure the generative model).  In  HEP with a
+intervention is required to reconfigure the generative model).  In HEP, with a
 fixed classifier, it has become common  to interpolate the distribution between
 discrete values of $\theta$ in order to produce a continuous parametrization for
-$\hat p(\hat s | \theta)$~\citep{Cranmer:2012sba}.
+$p(s | \theta)$~\citep{Cranmer:2012sba}.
+One can easily imagine a number of approaches to embedding the classifier and
+estimating the density $p(s|\theta)$ and the relative merits of those
+approaches will depend critically on the dimensionality of $\theta$ and the
+computational cost of the generative model. We leave a more general strategy for
+this overarching optimization problem as an area of future work.
+
+\glnote{Sub-sections 4.1 and 4.2 could directly be inlined within section 4. What do you think?}
 
@@
 
-One can easily imagine a number of approaches to embedding the classifier and
-estimating the density $\hat p(\hat s|\theta)$ and the relative merits of those
-approaches will depend critically on the dimensionality of $\theta$ and the
-computational cost of the generative model. We leave a more general strategy for
-this overarching optimization problem as an area of future work.
-
 % Applications =================================================================
@@
 New particle searches at the LHC are typically framed as hypothesis test where
-the null corresponds to $\mu=0$, and the generalized likelihood ratio (profiling
-over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test
-statistic~\citep{Cowan:2010js}. This approach was used for the discovery of the
-Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa}.
+the null corresponds to $\mu=0$, and the generalized likelihood ratio is used as a test
+statistic.
 
@@
 It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems.
-\begin{eqnarray}\label{eq:decompose}
-\frac{p(x|\theta_0)}{p(x|\theta_1)} & =& \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)}
-\\
-%&=& \sum_c \frac{ w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \\
-%&=& \sum_c \left[ \frac{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)}{ w_c(\theta_0) p_c(x| \theta_0)}  \right]^{-1} \\
-&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \\ \label{eq:decomposedResult}
-&=& \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1}
-\end{eqnarray}
+\begin{align}
+\frac{p(x|\theta_0)}{p(x|\theta_1)} &= \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \nonumber \\
+&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \nonumber \\
+&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} \label{eq:decomposedResult}
+\end{align}
 The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.

### 2016-01-22 jasa/paper-jasa.tex
Section 3.3: added some more inline comments

@@
 simulator are used to approximate the density at various parameter points, and
-an interpolation algorithm is used to approximate the parametrized
+an interpolation algorithm is used to approximate the parameterized
 model~\citep{Cranmer:2012sba}. In particular, to improve the statistical power
@@
 density probability of $\mathbf{X}$ at value $\mathbf{x}$ under the
-parametrization $\theta$. Let also assume some i.i.d. observed data ${\cal D} = \{
+parameterization $\theta$. Let also assume some i.i.d. observed data ${\cal D} = \{
 \mathbf{x}_1, \dots, \mathbf{x}_n \}$. In the setting where one is interested in
@@
 of variable $\mathbf{Y} = s(\mathbf{X})$ is based
-on a (parametrized) function $s$ that is strictly monotonic with the density ratio
+on a (parameterized) function $s$ that is strictly monotonic with the density ratio
 \begin{equation}
@@
 generalized likelihood ratio tests, where it will be useful to have the
-classifier  parametrized in terms of $(\theta_0, \theta_1)$.
+classifier  parameterized in terms of $(\theta_0, \theta_1)$.
 
@@
     \label{thm:ratio-equivalence}
-    Let $\mathbf{X}$ be a random vector vector with values in ${\cal X} \subseteq \mathbb{R}^p$ and parametrized probability
+    Let $\mathbf{X}$ be a random vector vector with values in ${\cal X} \subseteq \mathbb{R}^p$ and parameterized probability
     density $p_{\mathbf{X}}(\mathbf{x} = (x_1, ..., x_p)|\theta)$ and let
@@
 
-\subsection{Probabilistic classification for approximating ratios}
+\subsection{Probabilistic classification for likelihood ratios}
 \label{sec:clf-for-ratios}
@@
 \end{equation}
-If we further assume that samples for $Y=0$ are drawn from some parametrized
+If we further assume that samples for $Y=0$ are drawn from some parameterized
 distribution with probability density $p_{\mathbf{X}}(\mathbf{x}|\theta_0)$
@@
 classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
-of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modelling
+of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
 $p_Y(y=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.
 
-\subsection{Learning the correct mapping and its distribution}
-
-\glnote{This section needs to be improved. Not sure we need a section on its own for that either.}
-
-In order for this approach to be useful in the likelihood-free setting, where it
-is not possible to evaluate the density $p(\mathbf{x}|\theta)$, we need to be
-able to approximate both $s(\mathbf{x})$ and
-$p(s(\mathbf{x})|\theta)$ based on samples $\{x_i\}$ drawn from the generative
-model $p(\mathbf{x}|\theta)$.  Denote the dimensionality reduction map
-$\hat{s}(x)$ and its distribution $\hat{p}(\hat{s}|\theta)$.
+\subsection{Approximating the reduction map and its distribution}
+
+In order for this approach to be useful in the likelihood-free setting, we need
+to be able to approximate both $s(\mathbf{x})$ and $p(s(\mathbf{x})|\theta)$
+based on a finite number of samples $\{\mathbf{x}_i\}$ drawn from the generative
+model $p(\mathbf{x}|\theta)$.
+Let denote the approximated dimensionality
+reduction map $\hat{s}(\mathbf{x})$ and its approximated distribution
+$\hat{p}(\hat{s}(\mathbf{x})|\theta)$. \glnote{Explain how to approximate $p(s)$ in practice?}
 
 One strength of this approach is that it factorizes the approximation of the
-per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx
-s(\mathbf{x})$) from the calibration procedure ($\hat p(\hat
-s| \theta) \approx p(\hat{s}|\theta)$). Thus, even if the classifier does a poor
-job of reproducing the level sets of the per-event likelihood ratio, the density
-of $\hat{s}$ can still be well calibrated. In that case, one might loose power,
-but the resulting inference will still be valid. This point was made by
+per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from the
+calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx p(\hat{s}(\mathbf{x})|\theta)$).
+Thus, even if the classifier does a poor job at reproducing the level sets of
+the per-event likelihood ratio, the density of $\hat{s}$ can still be well
+calibrated. In that case, one might loose power, but the resulting inference
+will still be valid.\glnote{Shouldn't we elaborate?} This point was made by
 \cite{Neal:2007zz} and is well appreciated by the particle physics community
@@
 face of nuisance parameters associated to the simulator.
+
+\glnote{We should comment as to what happens if $\hat{s}(x)$ is not monotonic with $r$}
+
+\glnote{Add pseudo-code to summarize the whole pipeline?}
 
@@
 \end{align}
-where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parametrized}
+where $s(\mathbf{x};\theta,\theta_1)$ denotes a \textit{parameterized}
 transformation $s$ of $\mathbf{X}$ in terms of $(\theta,\theta_1)$ that is monotonic
@@
 
-\subsection{Parametrized classifier}
+\subsection{Parameterized classifier}
 
 In order to provide parameter inference in the likelihood-free setting as
-described above, we must train a family $s(\mathbf{x};\theta,\theta_1)$ of classifiers parametrized by
+described above, we must train a family $s(\mathbf{x};\theta,\theta_1)$ of classifiers parameterized by
 $\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
@@
 \begin{algorithm}[t]
-\caption{Training of a parametrized classifier.}\label{alg:training}
+\caption{Training of a parameterized classifier.}\label{alg:training}
     \begin{algorithmic}
@@
 
-\subsection{Parametrized calibration}
+\subsection{Parameterized calibration}
 
@@
 fixed classifier, it has become common  to interpolate the distribution between
-discrete values of $\theta$ in order to produce a continuous parametrization for
+discrete values of $\theta$ in order to produce a continuous parameterization for
 $p(s | \theta)$~\citep{Cranmer:2012sba}.
@@
  $\hat{p}_s(\hat s | \nu)$ and $\hat{p}_b(\hat s | \nu)$ continuously
- parametrized in $\nu$. An example of the distributions of $\hat s$ for the
+ parameterized in $\nu$. An example of the distributions of $\hat s$ for the
  signal and background events with $\nu=\nu_0$ is shown in
@@
 being implicit. However, as experimentalists we know that we must account for
-various forms of systematic uncertainty, parametrized by nuisance parameters
+various forms of systematic uncertainty, parameterized by nuisance parameters
 $\nu$. In practice, we take the classifier as fixed and then propagate
@@
 The standard use of machine learning in HEP can be improved by training a
-parametrized classifier corresponding to the generalized likelihood ratio test
+parameterized classifier corresponding to the generalized likelihood ratio test
 \begin{equation}
@@
 setting) to training a classifier for  null (background-only) vs. alternate
-(signal-plus-backgound) since the resulting regression functions are one-to-one
+(signal-plus-background) since the resulting regression functions are one-to-one
 with each other. In contrast, in the case of the generalized likelihood ratio
@@
 \hat{\hat{ \nu}})/p_b(x| {\hat{ \nu}})$. In practice, with classifiers of finite
-capacity, there will be some tradeoff between taking into account this
+capacity, there will be some trade-off between taking into account this
 additional term and the more challenging learning problem when $\mu$ is very
@@
 In the situation where the only free parameters of the  model are the mixture coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio).
-Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parametrized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
+Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parameterized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
 
@@
 While the original motivation for this work was to improve the treatment of
-systematic uncertainties in new particle searches by parametrizing the
+systematic uncertainties in new particle searches by parameterizing the
 classifier in terms of the nuisance parameters $\nu$, the same approach can be
@@
 learning in HEP, where classifiers have always been used between two static
-classes of events and not parametrized explicitly in terms of the physical
+classes of events and not parameterized explicitly in terms of the physical
 quantities we wish to measure. The work of  ~\citep{Whiteson:2006ws} is similar
@@
 dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is
-constant; however, in this approach the map ratio is explicitly parametrized in
+constant; however, in this approach the map ratio is explicitly parameterized in
 terms of $\theta$, so the ratio is important for canceling the corresponding
-Jacobian factors.} Instead, we embed a parametrized classifier into the
+Jacobian factors.} Instead, we embed a parameterized classifier into the
 likelihood and postpone the evaluation of the classifier to the point of
@@
 over a high-dimensional data set with multiple events in terms of a univariate
-density of a classifier score. We have shown that a parametrized family of
+density of a classifier score. We have shown that a parameterized family of
 discriminative classifiers $\hat s(x; \theta_0, \theta_1)$ trained and

### 2016-01-22 jasa/paper-jasa.tex
Section 6: related work

@@
 
+\glnote{Add reference to \cite{BiancaZadrozny} here.}
+
 \glnote{Add pseudo-code to summarize the whole pipeline?}
@@
 
-\section{Related works}
+\section{Related work}
 \label{sec:related}
 
-\glnote{Add related works on density ratio estimation.}
-\glnote{Remove others that are less directly related.}
-
-\cite{ClaytonScott} and \cite{JMLR:v14:tong13a} consider the machine learning
-problem associated to Neyman-Pearson hypothesis testing. As in this work, they
-consider the situation where one does not have access to the underlying
-distributions, but only has i.i.d. samples from each hypothesis. This work
-generalizes that goal from the Neyman-Pearson setting to generalized likelihood
-ratio tests and emphasizes the connection with classification. Perhaps a  formal
-treatment similar to the Neyman-Pearson case can be brought to bear in this more
-general setting. In a similarly titled work, \cite{Gutmann2014} advocate using
-the cross-validated classification accuracy as the similarity metric used in
-ABC. While the goal there is also parameter inference in the likelihood-free
-setting,  ``classifier ABC'' is very different than the approach presented here.
-\cite{TommiJaakkola} explore a way of leveraging generative models to derive
-kernel functions for use in discriminative methods. This interesting work is
-distinct from the point made here in which the generative model is being used
-for the purpose of providing training data and calibration.
-%\cite{McCallum} consider a hybrid generative/discriminative classifier; however, the goal of that work is not to leverage a generative model for the data, but to use both approaches to learn different subsets of the parameters in a single hybrid classifier.
-\cite{BiancaZadrozny} emphasize the importance of calibrated probability
-estimates from decision trees and naive Bayesian classifiers and investigate
-various approaches to achieve this. In contrast to that work, we are not
-interested in calibrated probability estimates for $p(y|x)$ for individual
-events, but instead we use the calibration to correct for non-linear
-transformations of the target likelihood ratio and, perhaps, to provide
-calibrated p-values based on those likelihood ratio tests. \cite{Ihler2004} take
-on a different problem (tests of statistical independence) by using machine
-learning algorithms to find  scalar maps from the high-dimensional feature space
-that achieve the desired statistical goal when the fundamental high-dimensional
-test is intractable.
-
-\cite{Neal:2007zz} also considered the problem of approximating the likelihood
-function when only a generative model is available. That work sketches a scheme
-in which one uses a classifier with both $x$ and $\theta$ as an input to serve
-as a dimensionality reduction map. The key distinction comes in the handling of
-$\theta$.  Neal says  ``we cannot use the classifier on real data, since we
-don't know the correct value for [$\theta]$'' and goes on to outline an approach
-where one uses regression on a per-event basis to estimate $\hat{\theta}(x)$ and
-perform the composition $s(x; \hat{\theta}(x))$ (much like
-profiling).\footnote{Neal considers a lower-dimensional `bottlenecks'
-$\theta^*=g(\theta)$, which are not essential to the discussion here.}  This can
-lead to a significant loss of information since (at least in most particle
-physics examples) a single event carries little information about the true value
-of $\theta$, though the full data set $D$ may be informative -- for instance, a
-single observation would not be sufficient to estimate the variance of a
-distribution, though repeated observations would.  The work of Neal correctly
-identifies this as an approximation of the target likelihood even in the case of
-a ideal classifier. In contrast, the approach described here does not eliminate
-the dependence of the classifier on $\theta$.\footnote{As a technical point,
-in Neal's work, the focus is on approximating the likelihood function (up to a
-multiplicative constant), which is equivalent to evaluating the ratio  with
-respect to a fixed $\theta_1$ as in Eq.~\ref{eq:mle_withs}. In Neal's case, the
-dependence on $\theta$ is eliminated via $s(x; \hat{\theta}(x))$ and the map is
-constant; however, in this approach the map ratio is explicitly parameterized in
-terms of $\theta$, so the ratio is important for canceling the corresponding
-Jacobian factors.} Instead, we embed a parameterized classifier into the
-likelihood and postpone the evaluation of the classifier to the point of
-evaluation of the likelihood when $\theta$ is explicitly being tested. This
+The closest work to the proposed method is due to \cite{Neal:2007zz}, who
+similarly considers the problem of approximating the likelihood function when
+only a generative model is available. That work sketches a scheme in which one
+uses a classifier with both $\mathbf{x}$ and $\theta$ as an input to serve as a
+dimensionality reduction map. The key distinction comes in the handling of
+$\theta$.  Neal argues that a classifier cannot be used on real data, since we
+do not know the correct value for $\theta$, and goes on to outline an approach
+where one uses regression on a per-event basis to estimate
+$\hat{\theta}(\mathbf{x})$ and perform the composition $s(\mathbf{x};
+\hat{\theta}(\mathbf{x}))$. As pointed out by the author, this can lead to a
+significant loss of information since a single observation $\mathbf{x}$ may
+carry little information about the true value of $\theta$, though a full data
+set ${\cal D}$ may be informative -- for instance, a single observation would
+not be sufficient to estimate the variance of a distribution, though repeated
+observations would. The work of \cite{Neal:2007zz} correctly identifies this as
+an approximation of the target likelihood even in the case of a ideal
+classifier. In contrast, the approach described here does not eliminate the
+dependence of the classifier on $\theta$\footnote{As a technical point, in
+\citep{Neal:2007zz}, the focus is on approximating the likelihood function (up
+to a multiplicative constant), which is equivalent to evaluating the ratio  with
+respect to a fixed $\theta_1$ as in Eqn.~\ref{eq:mle_withs}. In
+\citep{Neal:2007zz}, the dependence on $\theta$ is eliminated via $s(\mathbf{x};
+\hat{\theta}(\mathbf{x}))$ and the map is constant; however, in this approach
+the map ratio is explicitly parameterized in terms of $\theta$, so the ratio is
+important for canceling the corresponding Jacobian factors. \glnote{Shall we
+keep this footenote? Would be better to give details in the text. JASA
+recommends not using footnotes.}}. Instead, we embed a parameterized classifier
+into the likelihood and postpone the evaluation of the classifier to the point
+of evaluation of the likelihood when $\theta$ is explicitly being tested. This
 avoids the loss of information that occurs from the regression step
-$\hat{\theta}(x)$ proposed by Neal and leads to Theorem~1, which is an exact
-result in the case of an ideal classifier. In both cases, the quality of the
-classifier is factorized from the calibration of its density, which allows for
-valid inference even if there is a loss of power due to a non ideal classifier.
+$\hat{\theta}(\mathbf{x})$ proposed by \cite{Neal:2007zz} and leads to
+Theorem~\ref{thm:ratio-equivalence}, which is an exact result in the case of an
+ideal classifier. In both cases, the quality of the classifier is factorized
+from the calibration of its density, which allows for valid inference even if
+there is a loss of power due to a non ideal classifier. \glnote{We could shorten this paragraph.}
+
+Also close to our work, \cite{ClaytonScott} and \cite{JMLR:v14:tong13a} consider
+the machine learning problem associated to Neyman-Pearson hypothesis testing. In
+a similar setup, they consider the situation where one does not have access to
+the underlying distributions, but only has i.i.d. samples from each hypothesis.
+This work generalizes that goal from the Neyman-Pearson setting to generalized
+likelihood ratio tests and emphasizes the connection with classification.
+% Perhaps a  formal treatment similar to the Neyman-Pearson case can be brought to
+% bear in this more general setting.
+\cite{Ihler2004} take on a different problem
+(tests of statistical independence) by using machine learning algorithms to find
+scalar maps from the high-dimensional feature space that achieve the desired
+statistical goal when the fundamental high-dimensional test is intractable.
+\glnote{Add other references on hypothesis testing with ML?}
+
+% In a similarly titled work, \cite{Gutmann2014} advocate using
+% the cross-validated classification accuracy as the similarity metric used in
+% ABC. While the goal there is also parameter inference in the likelihood-free
+% setting,  their method is very different than the approach presented here.
+% \cite{TommiJaakkola} explore a way of leveraging generative models to derive
+% kernel functions for use in discriminative methods. This interesting work is
+% distinct from the point made here in which the generative model is being used
+% for the purpose of providing training data and calibration.
+% \cite{McCallum} consider a hybrid generative/discriminative classifier;
+% however, the goal of that work is not to leverage a generative model for the
+% data, but to use both approaches to learn different subsets of the parameters in
+% a single hybrid classifier.
+% \cite{BiancaZadrozny} emphasize the importance of calibrated probability
+% estimates from decision trees and naive Bayesian classifiers and investigate
+% various approaches to achieve this. In contrast to that work, we are not
+% interested in calibrated probability estimates for $p(y|x)$ for individual
+% events, but instead we use the calibration to correct for non-linear
+% transformations of the target likelihood ratio and, perhaps, to provide
+% calibrated p-values based on those likelihood ratio tests.
+
+More generally, likelihood ratio testing directly relates to the density ratio
+estimation problem, which consists in estimating the ratio of two probability
+densities from given and finite collections of observations ${\cal D}_0$ and
+${\cal D}_1$. Density ratio estimation is connected to many machine learning
+fundamental problems, including domain adaptation~\citep{sugiyama2012machine}, probabilistic
+classification and regression~\citep{vapnik1998statistical}, outlier
+detection~\citep{hido2011statistical}, and many others. In contrast with the
+proposed method which decouples reduction from calibration, most of the
+approaches proposed within the literature (see
+\cite{sugiyama2012density,gretton2009covariate,nguyen2010estimating,vapnik2013constructive}
+and references therein) provide solutions for estimating
+$r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$. Under some
+assumptions, the convergence of the obtained estimates was also proven for some
+of these approaches.
 

### 2016-01-22 jasa/paper-jasa.tex

@@
 calibrated. In that case, one might loose power, but the resulting inference
-will still be valid.\glnote{Shouldn't we elaborate?} This point was made by
+will still be valid.\glnote{Shouldn't we elaborate? Since this is one of our selling points, we should give more details in my opinion.} This point was made by
 \cite{Neal:2007zz} and is well appreciated by the particle physics community
@@
 
-We have outlined a technique to reformulate generalized likelihood ratio test
-over a high-dimensional data set with multiple events in terms of a univariate
-density of a classifier score. We have shown that a parameterized family of
-discriminative classifiers $\hat s(x; \theta_0, \theta_1)$ trained and
-calibrated with a simulator  $p(x|\theta)$ can be used to approximate the
-likelihood ratio  $\prod_i p(x_i|\theta_0)/p(x_i|\theta_1)$ even when it is not
-possible to directly evaluate the likelihood $p(x|\theta)$.
-%A technique for decomposing this ratio when the generative model is a mixture of components was presented with the aim to help focus capacity of the classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small mixture coefficient.
-This approach leverages the power of machine learning in a classical statistical setting.
-It offers an alternative to approximate Bayesian computation for parameter
-inference in the likelihood-free setting that can also be used in the
-frequentist formalism without specifying a prior over the parameters. A strength
-of this approach is that it separates the quality of the approximation of the
-target likelihood from the quality of the calibration. The former is related to
-the ability of supervised learning approaches to  classification, which will
-continue to improve. The calibration procedure for a particular parameter point
-is fairly straight forward since it involves estimating a univariate density
-using a generative model of the data. The difficulty of the calibration stage is
-performing this calibration continuously in $\theta$. Different strategies to
-this calibration are anticipated depending on the dimensionality of $\theta$,
-the complexity of the resulting likelihood function, and the difficulties
-associated to running the simulator.
+In this work, we have outlined an approach to reformulate generalized likelihood
+ratio testing over a high-dimensional data set in terms of a univariate density
+of a classifier score. We have shown that a parameterized family of
+discriminative classifiers $\hat s(\mathbf{x}; \theta_0, \theta_1)$ trained and
+calibrated with a simulator can be used to approximate the likelihood ratio,
+even when it is not possible to directly evaluate the likelihood
+$p(\mathbf{x}|\theta)$.
+% A technique for decomposing this ratio when the generative model is a mixture
+% of components was presented with the aim to help focus capacity of the
+% classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small
+% mixture coefficient.  This approach leverages the power of machine learning in a
+% classical statistical setting.
+The proposed method offers an alternative to approximate Bayesian computation
+for parameter inference in the likelihood-free setting that can also be used in
+the frequentist formalism without specifying a prior over the parameters. A
+strength of this approach is that it separates the quality of the approximation
+of the target likelihood from the quality of the calibration. The former is
+related to the ability of supervised learning approaches to  classification,
+which will continue to improve. The calibration procedure for a particular
+parameter point is fairly straight forward since it involves estimating a
+univariate density using a generative model of the data. The difficulty of the
+calibration stage is performing this calibration continuously in $\theta$.
+Different strategies to this calibration are anticipated depending on the
+dimensionality of $\theta$, the complexity of the resulting likelihood function,
+and the difficulties associated to running the simulator.
 

### 2016-01-25 jasa/paper-jasa.tex
Minor improvements

@@
 
-In particle physics, likelihood ratio tests are established tools for
-statistical inference. In many cases, these tests are complicated by the fact
+In many fields of science, generalized likelihood ratio tests are established tools for
+statistical inference. In practice, these tests are often complicated by the fact
 that computer simulators are used as a generative model for the data, which does
@@
 consequence, we show why and how discriminative classifiers can be used to
-approximate the likelihood ratio when only a generative model for the data is
+approximate the generalized likelihood ratio statistic when only a generative model for the data is
 available. In particular, the proposed method offers a machine learning-based
@@
 The likelihood function is the central object that summarizes the information
-from an experiment needed for inference of model parameters. The likelihood
-function is key to many areas of science that report the results of classical
+from an experiment needed for inference of model parameters. In particular, it
+is key to many areas of science that report the results of classical
 hypothesis tests or confidence intervals using the (generalized or profile)
@@
 observations $\mathbf{x}$. However, directly evaluating the likelihood function
-in these cases is often impossible or is computationally impractical. In this
-likelihood-free setting, various methods for statistical inference have been
-proposed, including most notably the Approximate Bayesian Computation (ABC)
-class of algorithms aiming at sampling from the posterior distribution of model
-parameters \glnote{Add references}.
+in these cases is often impossible or is computationally impractical.
 
@@
 specifically, we demonstrate why and how discriminative classifiers can be used
-to construct equivalent likelihood ratio tests when only a generative model for
+to construct equivalent generalized likelihood ratio test statistics when only a generative model for
 the data is available for training and calibration.
@@
 \begin{equation}\label{eqn:likelihood-ratio-test}
-T({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{X}(\mathbf{x}|\theta_0)}{ p_\mathbf{X}(\mathbf{x}|\theta_1)}
+\lambda({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{X}(\mathbf{x}|\theta_0)}{ p_\mathbf{X}(\mathbf{x}|\theta_1)}
 \end{equation}
@@
 
-In order to evaluate $T({\cal D})$, one must be able to evaluate the probability
+In order to evaluate $\lambda({\cal D})$, one must be able to evaluate the probability
 density $p_\mathbf{X}(\mathbf{x}| \theta)$ at any value $\mathbf{x}$. However,
@@
 \begin{equation}\label{eqn:likelihood-ratio-test-equiv}
-T'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_0)}{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_1)}
+\lambda'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_0)}{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_1)}
 \end{equation}
@@
 generalized likelihood ratio tests, where it will be useful to have the
-classifier  parameterized in terms of $(\theta_0, \theta_1)$.
+classifier decision function $s$ parameterized in terms of $(\theta_0, \theta_1)$.
 
@@
 Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
-density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{r(\mathbf{x})}{1 +
+density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{1}{1 +
 r(\mathbf{x})}$ only.
@@
 Thus far we have shown that the target likelihood ratio
-$r(\mathbf{x};\theta_0,\theta_1)=p(\mathbf{x}|\theta_0)/p(\mathbf{x}|\theta_1)$
+$r(\mathbf{x};\theta_0,\theta_1)=\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$
 with high dimensional features $\mathbf{x}$ can be reproduced via the univariate
@@
 the same as the value that maximizes the likelihood ratio with respect to
-$p({\cal D}|\theta_1)$, for some fixed value of $\theta_1$. This allows us to
+$p({\cal D}|\theta_1)$, for some fixed value of $\theta_1$\glnote{We should mention that the support of $p(\mathbf{x}|\theta_1)$ should cover the support of $p(\mathbf{x}|\theta)$.}.
+This allows us to
 use Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
@@
 $r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$. Under some
-assumptions, the convergence of the obtained estimates was also proven for some
+assumptions, the convergence of the obtained estimates is also proven for some
 of these approaches.

### 2016-01-27 jasa/paper-jasa.tex
Section 4: use standard definition of GLRT

@@
 In the case of composite hypotheses $\theta \in \Theta_0$ against an alternative
-$\theta \in \Theta_1$ (possibly such that $\Theta_0 \subseteq \Theta_1$), the
+$\theta \in \Theta_1$ (such that $\Theta_0 \cap \Theta_1 = \emptyset$ and $\Theta_0 \cup \Theta_1 = \Theta$), the
 generalized likelihood ratio test, also known as the profile likelihood ratio
@@
 \begin{equation}\label{eqn:generalized-lr}
-\Lambda({\cal D}; \Theta_0; \Theta_1) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta_1} p({\cal D} | \theta)} \; .
+\Lambda({\cal D}; \Theta_0, \Theta) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
 \end{equation}

### 2016-02-01 jasa/paper-jasa.tex
Section 6: related work

@@
 performs inference on a parameter $\mu$ related to the presence of the signal
-contribution. In particular, one forms the statistical model\footnote{Sometimes
-there is an additional Poisson term when expected number of signal and
-background events is known, which is referred to as an extended likelihood or
-marked Poisson model.}
+contribution. In particular, one forms the statistical model.
+% \footnote{Sometimes
+% there is an additional Poisson term when expected number of signal and
+% background events is known, which is referred to as an extended likelihood or
+% marked Poisson model.}
 \begin{equation}\label{eq:typicalML}
@@
 carry little information about the true value of $\theta$, though a full data
-set ${\cal D}$ may be informative -- for instance, a single observation would
-not be sufficient to estimate the variance of a distribution, though repeated
-observations would. The work of \cite{Neal:2007zz} correctly identifies this as
+set ${\cal D}$ may be informative.
+% for instance, a single observation would
+% not be sufficient to estimate the variance of a distribution, though repeated
+% observations would.
+The work of \cite{Neal:2007zz} correctly identifies this as
 an approximation of the target likelihood even in the case of a ideal
 classifier. In contrast, the approach described here does not eliminate the
-dependence of the classifier on $\theta$\footnote{As a technical point, in
-\citep{Neal:2007zz}, the focus is on approximating the likelihood function (up
-to a multiplicative constant), which is equivalent to evaluating the ratio  with
-respect to a fixed $\theta_1$ as in Eqn.~\ref{eq:mle_withs}. In
-\citep{Neal:2007zz}, the dependence on $\theta$ is eliminated via $s(\mathbf{x};
-\hat{\theta}(\mathbf{x}))$ and the map is constant; however, in this approach
-the map ratio is explicitly parameterized in terms of $\theta$, so the ratio is
-important for canceling the corresponding Jacobian factors. \glnote{Shall we
-keep this footenote? Would be better to give details in the text. JASA
-recommends not using footnotes.}}. Instead, we embed a parameterized classifier
+dependence of the classifier on $\theta$.
+% \footnote{As a technical point, in
+% \citep{Neal:2007zz}, the focus is on approximating the likelihood function (up
+% to a multiplicative constant), which is equivalent to evaluating the ratio  with
+% respect to a fixed $\theta_1$ as in Eqn.~\ref{eq:mle_withs}. In
+% \citep{Neal:2007zz}, the dependence on $\theta$ is eliminated via $s(\mathbf{x};
+% \hat{\theta}(\mathbf{x}))$ and the map is constant; however, in this approach
+% the map ratio is explicitly parameterized in terms of $\theta$, so the ratio is
+% important for canceling the corresponding Jacobian factors. \glnote{Shall we
+% keep this footenote? Would be better to give details in the text. JASA
+% recommends not using footnotes.}}.
+Instead, we embed a parameterized classifier
 into the likelihood and postpone the evaluation of the classifier to the point
@@
 from the calibration of its density, which allows for valid inference even if
-there is a loss of power due to a non ideal classifier. \glnote{We could shorten this paragraph.}
+there is a loss of power due to a non ideal classifier.
 
@@
 statistical goal when the fundamental high-dimensional test is intractable.
-\glnote{Add other references on hypothesis testing with ML?}
 
@@
 ${\cal D}_1$. Density ratio estimation is connected to many machine learning
-fundamental problems, including domain adaptation~\citep{sugiyama2012machine}, probabilistic
-classification and regression~\citep{vapnik1998statistical}, outlier
-detection~\citep{hido2011statistical}, and many others. In contrast with the
-proposed method which decouples reduction from calibration, most of the
+fundamental problems, including transfer learning~\citep{sugiyama2012machine},
+probabilistic classification and regression~\citep{vapnik1998statistical},
+outlier detection~\citep{hido2011statistical}, and many others. For learning
+under covariate shift, \cite{shimodaira2000improving} and \cite{sugiyama2005input} estimate
+the density ratio $r(\mathbf{x};\theta_0,\theta_1) =
+\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ from straightforward
+approximations $\hat{p}(\mathbf{x}|\theta_0)$ and $\hat{p}(\mathbf{x}|\theta_1)$
+separately obtained using kernel density estimation. Despite its theoretical
+consistency, this approach is known to be ineffective in
+practice~\citep{sugiyama2007covariate,bickel2009discriminative}, in particular
+as it relies on solving a problem (i.e., modelling distributions, possibly for high dimensional cases) that is
+more difficult than required (i.e., modelling a ratio between distributions).
+While the proposed method also proceeds in two similar steps, estimating
+$p(s(\mathbf{x}))$ is a much simpler problem than estimating $p(\mathbf{x})$,
+since $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
+the informative content of $r(\mathbf{x})$ is preserved.
+In contrast with the
+proposed method which decouples reduction from calibration, other
 approaches proposed within the literature (see
 \cite{sugiyama2012density,gretton2009covariate,nguyen2010estimating,vapnik2013constructive}
-and references therein) provide solutions for estimating
+and references therein) also provide solutions for estimating
 $r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$. Under some
@@
 
+\glnote{Improve given new draft.}
 In this work, we have outlined an approach to reformulate generalized likelihood

### 2016-02-01 jasa/paper-jasa.tex
Section 2: Rework, move things around and simplify

@@
 
+\subsection{Framework}
+
 % \subsection{Notation and Assumptions}
@@
 
-\section{Approximating likelihood ratios with classifiers}
+\subsection{Approximating likelihood ratios with classifiers}
 \label{sec:approx}
-
-\glnote{Think of a better section title? The equivalence is strict, not approximate.}
 
@@
 
-\subsection{Likelihood ratios under change of variables}
+%\subsection{Likelihood ratios under change of variables}
 
@@
 \begin{align}
-p_Y(y=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
-&= \frac{d}{dy}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq y\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
-&= \frac{d}{dy} \int_{\mathbb{R}^p} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
-&= \int_{\mathbb{R}^p} \frac{d}{dy} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
+% p_Y(y=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
+p_Y(y=s(\mathbf{x})|\theta_0) &= \frac{d}{dy}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq y\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
+% &= \frac{d}{dy} \int_{\mathbb{R}^p} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
+% &= \int_{\mathbb{R}^p} \frac{d}{dy} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
 &= \int_{\mathbb{R}^p} \delta(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime
 \end{align}
-where $H$ and $\delta$ are respectively the Heaviside step and the Dirac delta
-functions. Intuitively, this last expression can be understood as the integral
+Intuitively, this expression can be understood as the integral
 over all $\mathbf{x}^\prime \in \mathbb{R}^p$ such that $s(\mathbf{x}^\prime) = y$, as picked
-by the Dirac delta function. Given Theorem 6.1.5 of \citet{Hrmander1990},
+by the Dirac $\delta$ function. Given Theorem 6.1.5 of \citet{Hrmander1990},
 it further comes
-\begin{align}
-p_Y(y=s(\mathbf{x})|\theta_0) &= \int_{\mathbb{R}^p} \delta(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
-&= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
-\end{align}
+\begin{equation}
+p_Y(y=s(\mathbf{x})|\theta_0) = \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
+\end{equation}
 where $|\nabla s(\mathbf{x}^\prime)| = \sqrt{\sum_{i=1}^p |\frac{\partial}{\partial x_i} s(\mathbf{x}^\prime)|^2}$
@@
 \end{align}
-Combining equations \ref{eqn:hormander} and \ref{eqn:mapping}, we have
+Combining equations \ref{eqn:hormander} and \ref{eqn:mapping}, the density ratio $\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$ can be pulled out of the integral, resulting in
 \begin{align}
-p_Y(y=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{p_\mathbf{X}(\mathbf{x}^\prime|\theta_0)}{|\nabla s(\mathbf{x}^\prime)|} dS_{\mathbf{x}^\prime} \nonumber \\
-&= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+% p_Y(y=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \nonumber \\
+p_Y(y=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
 &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(y) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
 &= m^{-1}(s(\mathbf{x})) \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
-&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}
+&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}.
 \end{align}
@@
 
-\subsection{Probabilistic classification for likelihood ratios}
-\label{sec:clf-for-ratios}
+% \subsection{Probabilistic classification for likelihood ratios}
+% \label{sec:clf-for-ratios}
 
@@
 
-\begin{proposition} \label{thm:best-classifier}
-Let $\mathbf{X} = (X_1, ..., X_p)$ and $Y$ be random input and output variables
-with values in ${\cal X} \subseteq \mathbb{R}^p$
-and ${\cal Y} = \{0, 1\}$ and mixed joint probability density  function
-$p_{\mathbf{X},Y}(\mathbf{x}, y)$. For the squared error loss, the best
-regression function $s : {\cal X} \mapsto [0, 1]$, or equivalently the best
-probabilistic classifier, is
-\begin{equation}
-s^*(\mathbf{x}) = \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{P(Y=0) p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
-\end{equation}
-\end{proposition}
-
-\begin{proof}
-For the squared error loss,
-\begin{align}
-s^*(\mathbf{x}) &= \argmin_{s(\mathbf{x})} \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ (Y - s(\mathbf{x}))^2 \} \nonumber \\
-&=  \argmin_{s(\mathbf{x})} \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y^2 \} - 2s(\mathbf{x}) \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + s(\mathbf{x})^2 \nonumber \\
-&=  \argmin_{s(\mathbf{x})} -2s(\mathbf{x}) \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + s(\mathbf{x})^2
-\end{align}
-The last expression is minimized when $\frac{d}{ds(\mathbf{x})} (-2s(\mathbf{x}) \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + s(\mathbf{x})^2) = 0$,
-that is when $-2 \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + 2 s(\mathbf{x}) = 0$, hence
-\begin{equation}
-s^*(\mathbf{x}) = \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \}.
-\end{equation}
-For ${\cal Y} = \{ 0, 1 \}$,
-\begin{align}
-\mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} &= P(Y=0|\mathbf{X}=\mathbf{x}) \times 0 +  P(Y=1|\mathbf{X}=\mathbf{x}) \times 1 \nonumber \\
-&= \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{p_{\mathbf{X}}(\mathbf{x})} \nonumber \\
-&= \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{P(Y=0) p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
-\end{align}
-\end{proof}
-
-For $P(Y=0)=P(Y=1)=\frac{1}{2}$, the best regression function $s^*$ simplifies
-to
-\begin{equation}
-s^*(\mathbf{x}) = \frac{p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
-\end{equation}
-If we further assume that samples for $Y=0$ are drawn from some parameterized
-distribution with probability density $p_{\mathbf{X}}(\mathbf{x}|\theta_0)$
-and similarly for $Y=1$, then the best regression function can be rewritten
-as
+In light of this result, the likelihood ratio estimation problem can now
+be recast as a (probabilistic) classification problem, by noticing that the regression
+function
 \begin{equation}
@@
 \end{equation}
-In particular, this regression function satisfies conditions of
-Theorem~\ref{thm:ratio-equivalence} since $s^*(\mathbf{x}) =
-m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$,
-for $m(r(\mathbf{x})) = \frac{1}{1 + r(\mathbf{x})}$, is monotonic with
-$\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$.
-In other words, Proposition~\ref{thm:best-classifier} yields a sufficient
+modelled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
+from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies conditions of
+Theorem~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s} for further details).
+In other words, supervised learning yields a sufficient
 procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
@@
 r(\mathbf{x})}$ only.
-
-Equivalently, since $m^{-1}(s^*(\mathbf{x})) = \frac{1 -
-s^*(\mathbf{x})}{s^*(\mathbf{x})} = r(\mathbf{x})$,
-Proposition~\ref{thm:best-classifier} shows that the likelihood ratio estimation
-problem is directly related to probabilistic classification, and that algorithms
-proposed to solve one can be used for solving the other. In this context,
+Equivalently,
 Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
@@
 
-\subsection{Approximating the reduction map and its distribution}
+\subsection{Approximating the reduction map}
 
@@
 One strength of this approach is that it factorizes the approximation of the
-per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from the
-calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx p(\hat{s}(\mathbf{x})|\theta)$).
-Thus, even if the classifier does a poor job at reproducing the level sets of
-the per-event likelihood ratio, the density of $\hat{s}$ can still be well
-calibrated. In that case, one might loose power, but the resulting inference
-will still be valid.\glnote{Shouldn't we elaborate? Since this is one of our selling points, we should give more details in my opinion.} This point was made by
-\cite{Neal:2007zz} and is well appreciated by the particle physics community
-that typically takes a conservative attitude towards the use of machine learning
-classifiers precisely due to concerns about the calibration $p$-values in the
-face of nuisance parameters associated to the simulator.
+per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from
+the calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx
+p(\hat{s}(\mathbf{x})|\theta)$). Thus, even if the classifier does a poor job at
+reproducing the level sets of the per-event likelihood ratio, the density of
+$\hat{s}$ can still be well calibrated. In that case, one might loose power, but
+the resulting inference will still be valid.\glnote{Shouldn't we elaborate?
+Since this is one of our selling points, we should give more details in my
+opinion.} \glnote{Some of this is also already outlined in the previous section.}
+This point was made by \cite{Neal:2007zz} and is well appreciated by the
+particle physics community that typically takes a conservative attitude towards
+the use of machine learning classifiers precisely due to concerns about the
+calibration $p$-values in the face of nuisance parameters associated to the
+simulator.
 
@@
 
-\subsection{Classification and frequentist hypothesis tests}
-
-\glnote{Shall we keep everything from this?}
-
-Vast literature exists around generative and discriminative
-classifiers~\citep{AndrewY.Ng}. Typically, generative classifiers learn a model
-for the joint probability $p(\mathbf{x}, y)$, of the inputs $x$ and the classification
-label $y$, and predict $p(y|\mathbf{x})$ via Bayes rule. In contrast, discriminative
-classifiers model the posterior $p(y|\mathbf{x})$ directly. For classification tasks, one
-then thresholds on $p(y|\mathbf{x})$. In both cases this description in terms of a
-posterior requires a prior distribution for $p(y)$, which is either modeled
-explicitly or learned from the training data. This familiar formulation of
-classification may lead to some confusion in the setting of the current work.
-
-The first possible source of confusion we wish to avoid is that here
-$p(\mathbf{x}|\theta)$  is a  \textit{generative statistical model} for the features $\mathbf{x}$,
-not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a
-traditional scientific theory or simulator, able to make predictions about $\mathbf{x}$
-and being motivated by domain-specific considerations.
-
-The second possible source of confusion is that we are not directly interested
-in calibrating the classification score in terms of a per-event posterior
-probability $p(y|\mathbf{x})$. Instead, we are interested in the approximation of the
-per-experiment likelihood function or likelihood ratio, which might be used for
-several purposes, including the calculation of p-values.
-
-Lastly, in the setting of frequentist hypothesis tests and confidence intervals,
-we do not have a prior $\pi(\theta)$. While we can use the generative models to
-produce training data $(\mathbf{x} ,y=0)$ generated from
-$p(\mathbf{x}|\theta_0)$ and $(\mathbf{x}, y=1)$ generated from
-$p(\mathbf{x}|\theta_1)$, the relative mix $p(y)$ is arbitrary. Since the prior
-$p(y)$ is not needed for the target likelihood ratio test and because the
-classifier score $p(y|\mathbf{x})$ may not be well calibrated, we choose to denote the
-classifier score $s(\mathbf{x})$ and simply think of it as a deterministic dimensionality
-reduction map $s: \mathbb{R}^p \to \mathbb{R}$.  Similar points have been made
-by~\cite{ClaytonScott} and \cite{Neal:2007zz}.
+% \subsection{Classification and frequentist hypothesis tests}
+%
+% \glnote{Shall we keep everything from this?}
+%
+% Vast literature exists around generative and discriminative
+% classifiers~\citep{AndrewY.Ng}. Typically, generative classifiers learn a model
+% for the joint probability $p(\mathbf{x}, y)$, of the inputs $x$ and the classification
+% label $y$, and predict $p(y|\mathbf{x})$ via Bayes rule. In contrast, discriminative
+% classifiers model the posterior $p(y|\mathbf{x})$ directly. For classification tasks, one
+% then thresholds on $p(y|\mathbf{x})$. In both cases this description in terms of a
+% posterior requires a prior distribution for $p(y)$, which is either modeled
+% explicitly or learned from the training data. This familiar formulation of
+% classification may lead to some confusion in the setting of the current work.
+%
+% The first possible source of confusion we wish to avoid is that here
+% $p(\mathbf{x}|\theta)$  is a  \textit{generative statistical model} for the features $\mathbf{x}$,
+% not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a
+% traditional scientific theory or simulator, able to make predictions about $\mathbf{x}$
+% and being motivated by domain-specific considerations.
+%
+% The second possible source of confusion is that we are not directly interested
+% in calibrating the classification score in terms of a per-event posterior
+% probability $p(y|\mathbf{x})$. Instead, we are interested in the approximation of the
+% per-experiment likelihood function or likelihood ratio, which might be used for
+% several purposes, including the calculation of p-values.
+%
+% Lastly, in the setting of frequentist hypothesis tests and confidence intervals,
+% we do not have a prior $\pi(\theta)$. While we can use the generative models to
+% produce training data $(\mathbf{x} ,y=0)$ generated from
+% $p(\mathbf{x}|\theta_0)$ and $(\mathbf{x}, y=1)$ generated from
+% $p(\mathbf{x}|\theta_1)$, the relative mix $p(y)$ is arbitrary. Since the prior
+% $p(y)$ is not needed for the target likelihood ratio test and because the
+% classifier score $p(y|\mathbf{x})$ may not be well calibrated, we choose to denote the
+% classifier score $s(\mathbf{x})$ and simply think of it as a deterministic dimensionality
+% reduction map $s: \mathbb{R}^p \to \mathbb{R}$.  Similar points have been made
+% by~\cite{ClaytonScott} and \cite{Neal:2007zz}.
 
@@
 
-\section{Applications}
+\section{Examples and applications}
+
+\subsection{Toy example}
+
+\glnote{todo}
 
@@
 
+\appendix
+
+\section{Probabilistic classification for building $s$}
+\label{app:clf-for-s}
+
+In this appendix, we show that the probabilistic classification framework
+yields a reduction $s$ which satisfies conditions of Theorem~\ref{thm:ratio-equivalence}.
+
+\begin{proposition} \label{thm:best-classifier}
+Let $\mathbf{X} = (X_1, ..., X_p)$ and $Y$ be random input and output variables
+with values in ${\cal X} \subseteq \mathbb{R}^p$
+and ${\cal Y} = \{0, 1\}$ and mixed joint probability density  function
+$p_{\mathbf{X},Y}(\mathbf{x}, y)$. For the squared error loss, the best
+regression function $s : {\cal X} \mapsto [0, 1]$, or equivalently the best
+probabilistic classifier, is
+\begin{equation}
+s^*(\mathbf{x}) = \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{P(Y=0) p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
+\end{equation}
+\end{proposition}
+
+\begin{proof}
+For the squared error loss,
+\begin{align}
+s^*(\mathbf{x}) &= \argmin_{s(\mathbf{x})} \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ (Y - s(\mathbf{x}))^2 \} \nonumber \\
+&=  \argmin_{s(\mathbf{x})} \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y^2 \} - 2s(\mathbf{x}) \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + s(\mathbf{x})^2 \nonumber \\
+&=  \argmin_{s(\mathbf{x})} -2s(\mathbf{x}) \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + s(\mathbf{x})^2
+\end{align}
+The last expression is minimized when $\frac{d}{ds(\mathbf{x})} (-2s(\mathbf{x}) \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + s(\mathbf{x})^2) = 0$,
+that is when $-2 \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} + 2 s(\mathbf{x}) = 0$, hence
+\begin{equation}
+s^*(\mathbf{x}) = \mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \}.
+\end{equation}
+For ${\cal Y} = \{ 0, 1 \}$,
+\begin{align}
+\mathbb{E}_{Y|\mathbf{X}=\mathbf{x}} \{ Y \} &= P(Y=0|\mathbf{X}=\mathbf{x}) \times 0 +  P(Y=1|\mathbf{X}=\mathbf{x}) \times 1 \nonumber \\
+&= \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{p_{\mathbf{X}}(\mathbf{x})} \nonumber \\
+&= \frac{P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{P(Y=0) p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + P(Y=1) p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
+\end{align}
+\end{proof}
+
+For $P(Y=0)=P(Y=1)=\frac{1}{2}$, the best regression function $s^*$ simplifies
+to
+\begin{equation}
+s^*(\mathbf{x}) = \frac{p_{\mathbf{X}|Y}(\mathbf{x}|Y=1)}{p_{\mathbf{X}|Y}(\mathbf{x} | Y=0) + p_{\mathbf{X}|Y}(\mathbf{x} | Y=1)}.
+\end{equation}
+If we further assume that samples for $Y=0$ are drawn from some parameterized
+distribution with probability density $p_{\mathbf{X}}(\mathbf{x}|\theta_0)$
+and similarly for $Y=1$, then the best regression function can be rewritten
+as
+\begin{equation}
+s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
+\end{equation}
+In particular, this regression function satisfies conditions of
+Theorem~\ref{thm:ratio-equivalence} since $s^*(\mathbf{x}) =
+m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$,
+for $m(r(\mathbf{x})) = \frac{1}{1 + r(\mathbf{x})}$, is monotonic with
+$\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$.
+
 \end{document}

### 2016-02-01 jasa/paper-jasa.tex
Notes and small changes

@@
 not provide a way to evaluate the likelihood function. In this paper, we
-demonstrate that likelihood ratios are invariant under dimensionality reduction,
-provided the transformation is monotonic with the likelihood ratio. As a direct
+demonstrate that likelihood ratios are invariant under dimensionality reductions $\mathbb{R}^p \mapsto \mathbb{R}$,
+provided the transformation is itself monotonic with the likelihood ratio. As a direct
 consequence, we show why and how discriminative classifiers can be used to
@@
 approach to statistical inference that is complementary to likelihood-free
-Bayesian inference algorithms, like Approximate Bayesian Computation, as it does
+Bayesian inference algorithms, such as Approximate Bayesian Computation, as it does
 not require the definition of a prior over model parameters. \glnote{Also
@@
 The main result of this paper is to show that the likelihood ratio is invariant
-under dimensionality reduction, under the assumption that the corresponding
-transformation is monotonic with the likelihood ratio. As a direct consequence,
+under dimensionality reductions $\mathbb{R}^p \mapsto \mathbb{R}$, under the assumption that the corresponding
+transformation is itself monotonic with the likelihood ratio. As a direct consequence,
 we derive and propose an alternative machine learning-based approach for
@@
 between a hypothesized process of interest (labeled \textit{signal}) and various
-other processes (labeled \textit{background}). Typically, pseudo-data from the
+other processes (labeled \textit{background}). Typically, data generated from the
 simulator are used to approximate the density at various parameter points, and
@@
 In order to evaluate $\lambda({\cal D})$, one must be able to evaluate the probability
-density $p_\mathbf{X}(\mathbf{x}| \theta)$ at any value $\mathbf{x}$. However,
+densities $p_\mathbf{X}(\mathbf{x}| \theta_0)$ and $p_\mathbf{X}(\mathbf{x}| \theta_1)$ at any value $\mathbf{x}$. However,
 it is increasingly common in science that one has a complex simulation that can
@@
 
-The main result of this paper (see Theorem~\ref{thm:ratio-equivalence}) is to
+The main result of this paper is to
 generalize the observation that one can form a test statistic
@@
 that is strictly equivalent to \ref{eqn:likelihood-ratio-test}, provided the change
-of variable $\mathbf{Y} = s(\mathbf{X})$ is based
+of variable $\mathbf{Y} = s(\mathbf{X})$ \glnote{Use another symbol instead $\mathbf{Y}$} is based
 on a (parameterized) function $s$ that is strictly monotonic with the density ratio
@@
 the denominator $p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)$ because this
-cancels Jacobian factors that vary with $\theta$.
+cancels Jacobian factors that vary with $\theta$. \glnote{Say that if $s$ is not monotonic with $r$, then it might be bias $\hat\theta$.}
 
@@
 \caption{Training of a parameterized classifier.}\label{alg:training}
+\glnote{Outline the full generalized likelihood ratio estimation pipeline instead}
     \begin{algorithmic}
@@
 
-\glnote{Sub-sections 4.1 and 4.2 could directly be inlined within section 4. What do you think?}
-
 % There is a mild technical challenge in embedding the classifier into the
@@
 
+\section{Mixture models}
+
+\glnote{todo}
+
 \section{Examples and applications}
@@
 \glnote{todo}
+\glnote{+ diagnostics}
 
@@
 
-\glnote{Add a concrete HEP example where we illustrate and compare
-the method for several classifiers and calibration algorithms.}
-\glnote{Notations need to be adapted.}
-\glnote{Rethink the outline? Start with comparison of mixtures (5.1.4) and then
-specialize to background vs background+signal, and thereby discuss that we can
-focus the capacity of the classifier.}
+\glnote{Rewrite and shorten.}
 
@@
 
-\subsection{Likelihood-free inference}
+\subsubsection{Measuring particle properties}
 
@@
 the informative content of $r(\mathbf{x})$ is preserved.
-In contrast with the
+Finally, in contrast with the
 proposed method which decouples reduction from calibration, other
@@
 and references therein) also provide solutions for estimating
-$r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$. Under some
+$r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$, in one step. Under some
 assumptions, the convergence of the obtained estimates is also proven for some
@@
 \end{equation}
-If we further assume that samples for $Y=0$ are drawn from some parameterized
-distribution with probability density $p_{\mathbf{X}}(\mathbf{x}|\theta_0)$
-and similarly for $Y=1$, then the best regression function can be rewritten
+If we further assume that samples for $Y=0$ (resp. $Y=1$) are drawn from some parameterized
+distribution with probability density $p_{\mathbf{X}}(\mathbf{x}|\theta_0)$ (resp. $p_{\mathbf{X}}(\mathbf{x}|\theta_1)$), then the best regression function can be rewritten
 as

### 2016-02-01 jasa/paper-jasa.tex
Section 4: mixture models

@@
 provided the transformation is itself monotonic with the likelihood ratio. As a direct
-consequence, we show why and how discriminative classifiers can be used to
+consequence, we show that discriminative classifiers can be used to
 approximate the generalized likelihood ratio statistic when only a generative model for the data is
@@
 setting where a prior over the model parameters is not available. More
-specifically, we demonstrate why and how discriminative classifiers can be used
+specifically, we demonstrate that discriminative classifiers can be used
 to construct equivalent generalized likelihood ratio test statistics when only a generative model for
@@
 density probability of $\mathbf{X}$ at value $\mathbf{x}$ under the
-parameterization $\theta$. Let also assume some i.i.d. observed data ${\cal D} = \{
+parameterization $\theta$. Let also assume i.i.d. observed data ${\cal D} = \{
 \mathbf{x}_1, \dots, \mathbf{x}_n \}$. In the setting where one is interested in
@@
 physics~\citep{Neal:2007zz} where the simulation of particle detectors can only
-be done in the forward mode. \glnote{Reinsert citations for \cite{ClaytonScott}
-and \cite{JMLR:v14:tong13a}.}
+be done in the forward mode.
 
@@
 
-\glnote{todo}
+% In this section we generalize the capacity focusing technique of training
+% classifiers to discriminate between components of a mixture model. First, we
+% generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
+% \begin{equation}
+% p(x|\theta)=\sum_c w_c(\theta) p_c(x| \theta) \;.
+% \end{equation}
+% It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems.
+% \begin{align}
+% \frac{p(x|\theta_0)}{p(x|\theta_1)} &= \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \nonumber \\
+% &= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \nonumber \\
+% &= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} \label{eq:decomposedResult}
+% \end{align}
+% The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.
+The third line uses Theorem~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
+In the situation where the only free parameters of the  model are the mixture coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio).
+Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parameterized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
+
+In the setting of (simple or composite) hypothesis testing between
+models defined as mixtures of several components, i.e. when
+\begin{equation}
+p(\mathbf{x}|\theta)=\sum_c w_c(\theta) p_c(\mathbf{x}| \theta),
+\end{equation}
+the target likelihood ratio can be rewritten in terms of pairwise
+classification problems. We indeed have
+\begin{align}
+\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)} &= \frac{\sum_c w_c(\theta_0) p_c(\mathbf{x}| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(\mathbf{x}| \theta_1)} \nonumber \\
+&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(\mathbf{x}| \theta_1)}{  p_c(\mathbf{x}| \theta_0)}  \right]^{-1} \nonumber \\
+&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta_1)}{ p_c(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta_0)}  \right]^{-1}.
+\label{eq:decomposedResult}
+\end{align}
+The second line is a trivial, but a useful decomposition into pairwise
+density ratio sub-problems between $p_{c'}(\mathbf{x}|\theta_1)$ and
+$p_c(\mathbf{x}|\theta_0)$.  The third line uses
+Theorem~\ref{thm:ratio-equivalence} to relate the high-dimensional likelihood
+ratio into an equivalent calibrated likelihood ratio based on the univariate
+density of the corresponding classifier.
+
+In applications where mixture models are commonly used, e.g. as in high energy
+physics, this decomposition allows one to construct better likelihood ratio
+estimates since it forces classifiers $s_{c,c'}$ to focus on simpler
+sub-problems, for which higher accuracy is expected.
+
+Finally, as a technical point, in the situation where the only free parameters
+of the  model are the mixture coefficients $w_c$, the distributions
+$p_{c}(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta)$ are independent of
+$\theta$. For this reason, sub-ratios $\frac{
+p_{c'}(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)|\theta_1)}{
+p_c(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)|\theta_0)}$ simplify to $\frac{
+p_{c'}(s_{c,c'}(\mathbf{x}))}{ p_c(s_{c,c'}(\mathbf{x}))}$, and can all be
+pre-computed without having to resort to parameterized classification and
+calibration.
 
@@
 small.
-
-\subsubsection{Decomposing tests between mixture models into their components}
-
-In this section we generalize the capacity focusing technique of training
-classifiers to discriminate between components of a mixture model. First, we
-generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
-\begin{equation}
-p(x|\theta)=\sum_c w_c(\theta) p_c(x| \theta) \;.
-\end{equation}
-It is possible to re-write the target likelihood ratio between two mixture models in terms of pairwise classification problems.
-\begin{align}
-\frac{p(x|\theta_0)}{p(x|\theta_1)} &= \frac{\sum_c w_c(\theta_0) p_c(x| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(x| \theta_1)} \nonumber \\
-&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(x| \theta_1)}{  p_c(x| \theta_0)}  \right]^{-1} \nonumber \\
-&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c',\theta_0, \theta_1}| \theta_1)}{ p_c(s_{c,c',\theta_0, \theta_1}| \theta_0)}  \right]^{-1} \label{eq:decomposedResult}
-\end{align}
-The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.
-The third line uses Theorem~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
-In the situation where the only free parameters of the  model are the mixture coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio).
-Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parameterized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
 

### 2016-02-01 jasa/paper-jasa.tex
Section 2: change Y to U

@@
   \title{\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers}
-  \author{Kyle Cranmer and Gilles Louppe\\
+  \author{Kyle Cranmer, Juan Pavez and Gilles Louppe\\
           New York University}
@@
 
-\subsection{Framework}
+\subsection{Simple hypothesis testing}
 
@@
 % This allows us to recast the original likelihood ratio test into an alternate form in which supervised learning is used to train the discriminative classifier $s(x; \theta_0, \theta_1)$.
-The discriminative classifier can be trained with data $(x,y=0)$ generated
-% from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$.
+The discriminative classifier can be trained with data $(x,u=0)$ generated
+% from $p(x|\theta_0)$ and $(x,u=1)$ generated from $p(x|\theta_1)$.
 In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the classifier  parametrized in terms of $(\theta_0, \theta_1)$.
@@
 \begin{equation}\label{eqn:likelihood-ratio-test-equiv}
-\lambda'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_0)}{ p_\mathbf{Y}(y=s(\mathbf{x}) | \theta_1)}
+\lambda'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{U}(u=s(\mathbf{x}) | \theta_0)}{ p_\mathbf{U}(u=s(\mathbf{x}) | \theta_1)}
 \end{equation}
 that is strictly equivalent to \ref{eqn:likelihood-ratio-test}, provided the change
-of variable $\mathbf{Y} = s(\mathbf{X})$ \glnote{Use another symbol instead $\mathbf{Y}$} is based
+of variable $\mathbf{U} = s(\mathbf{X})$ is based
 on a (parameterized) function $s$ that is strictly monotonic with the density ratio
@@
     \begin{equation}
-        r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} = \frac{p_\mathbf{Y}(y=s(\mathbf{x})|\theta_0)}{p_\mathbf{Y}(y=s(\mathbf{x})|\theta_1)},
-    \end{equation}
-    where $p_\mathbf{Y}(y=s(\mathbf{x};\theta_0,\theta_1)|\theta)$ is the induced probability density of
-    $\mathbf{Y} = s(\mathbf{X};\theta_0,\theta_1)$.
+        r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} = \frac{p_\mathbf{U}(u=s(\mathbf{x})|\theta_0)}{p_\mathbf{U}(u=s(\mathbf{x})|\theta_1)},
+    \end{equation}
+    where $p_\mathbf{U}(u=s(\mathbf{x};\theta_0,\theta_1)|\theta)$ is the induced probability density of
+    $\mathbf{U} = s(\mathbf{X};\theta_0,\theta_1)$.
 \end{theorem}
@@
 \begin{align}
-% p_Y(y=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
-p_Y(y=s(\mathbf{x})|\theta_0) &= \frac{d}{dy}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq y\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
+% p_U(u=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
+p_U(u=s(\mathbf{x})|\theta_0) &= \frac{d}{du}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq u\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
 % &= \frac{d}{dy} \int_{\mathbb{R}^p} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
 % &= \int_{\mathbb{R}^p} \frac{d}{dy} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
-&= \int_{\mathbb{R}^p} \delta(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime
+&= \int_{\mathbb{R}^p} \delta(u - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime
 \end{align}
 Intuitively, this expression can be understood as the integral
-over all $\mathbf{x}^\prime \in \mathbb{R}^p$ such that $s(\mathbf{x}^\prime) = y$, as picked
+over all $\mathbf{x}^\prime \in \mathbb{R}^p$ such that $s(\mathbf{x}^\prime) = u$, as picked
 by the Dirac $\delta$ function. Given Theorem 6.1.5 of \citet{Hrmander1990},
@@
 \begin{equation}
-p_Y(y=s(\mathbf{x})|\theta_0) = \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
+p_U(u=s(\mathbf{x})|\theta_0) = \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
 \end{equation}
 where $|\nabla s(\mathbf{x}^\prime)| = \sqrt{\sum_{i=1}^p |\frac{\partial}{\partial x_i} s(\mathbf{x}^\prime)|^2}$
-and where $dS_{\mathbf{x}^\prime}$ is the Euclidean surface measure on $\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}$.
+and where $dS_{\mathbf{x}^\prime}$ is the Euclidean surface measure on $\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}$.
 Also, since $s(\mathbf{x})$ is monotonic with
@@
 \begin{align}
-% p_Y(y=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \nonumber \\
-p_Y(y=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
-&= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(y) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
-&= m^{-1}(s(\mathbf{x})) \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
-&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}.
+% p_U(u=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \nonumber \\
+p_U(u=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+&= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(u) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+&= m^{-1}(s(\mathbf{x})) \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}.
 \end{align}
-Similarly, Equation~\ref{eqn:hormander} can be used to derive $p_Y(y=s(\mathbf{x})|\theta_1)$, finally yielding
+Similarly, Equation~\ref{eqn:hormander} can be used to derive $p_U(u=s(\mathbf{x})|\theta_1)$, finally yielding
 \begin{align}
-\frac{p_Y(y=s(\mathbf{x})|\theta_0)}{p_Y(y=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
+\frac{p_U(u=s(\mathbf{x})|\theta_0)}{p_U(u=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
 &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.
@@
 % $(x_i , y_i=1)$. With balanced training data
-% \mbox{($p(y=1)=p(y=0)=\sfrac{1}{2}$)} a quadratic loss function will lead to
+% \mbox{($p(u=1)=p(u=0)=\sfrac{1}{2}$)} a quadratic loss function will lead to
 % classifiers that approximate the regression function  $\hat{s}(x) \approx p(y|x) =
@@
 \end{equation}
-modelled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
+modeled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
 from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies conditions of
@@
 of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
-$p_Y(y=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.
+$p_U(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.
 
@@
 % we do not have a prior $\pi(\theta)$. While we can use the generative models to
-% produce training data $(\mathbf{x} ,y=0)$ generated from
-% $p(\mathbf{x}|\theta_0)$ and $(\mathbf{x}, y=1)$ generated from
+% produce training data $(\mathbf{x} ,u=0)$ generated from
+% $p(\mathbf{x}|\theta_0)$ and $(\mathbf{x}, u=1)$ generated from
 % $p(\mathbf{x}|\theta_1)$, the relative mix $p(y)$ is arbitrary. Since the prior
@@
         		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$
-        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, y=0) \}$ to ${\cal L}$
+        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, u=0) \}$ to ${\cal L}$
         		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$
-        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, y=1) \}$ to ${\cal L}$
+        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, u=1) \}$ to ${\cal L}$
         	\EndFor
@@
 
-\section{Mixture models}
+\subsection{Mixture models}
 
@@
 
-In the setting of (simple or composite) hypothesis testing between
+In the special case of (simple or composite) hypothesis testing between
 models defined as mixtures of several components, i.e. when
@@
 In this setting, large samples of pseudo-data $\{x_i, y_i\}$ generated with some
-nominal values of the parameters $\nu_0$, where $y=0$ corresponds to the
-background density $p_b(x|\nu_0)$  and $y=1$ corresponds to signal density
-$p_s(x|\nu_0)$. Importantly, the $y=1$ label corresponds to the signal only, and
+nominal values of the parameters $\nu_0$, where $u=0$ corresponds to the
+background density $p_b(x|\nu_0)$  and $u=1$ corresponds to signal density
+$p_s(x|\nu_0)$. Importantly, the $u=1$ label corresponds to the signal only, and
 not to the alternate signal-plus-background hypothesis. The resulting classifier
@@
 which is one to one with the likelihood ratio of the null to the alternate
-$p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$. Associating the $y=1$ label to
+$p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$. Associating the $u=1$ label to
 the signal component has advantages because it helps the classifier focus its
@@
 practice~\citep{sugiyama2007covariate,bickel2009discriminative}, in particular
-as it relies on solving a problem (i.e., modelling distributions, possibly for high dimensional cases) that is
-more difficult than required (i.e., modelling a ratio between distributions).
+as it relies on modeling numerator and denominator high-dimensional densities,
+which is a harder problem than modeling their ratio only.
 While the proposed method also proceeds in two similar steps, estimating

### 2016-02-01 jasa/paper-jasa.tex
Section 2.3

@@
 function
-\begin{equation}
+\begin{equation}\label{eqn:best-s-clf}
 s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
@@
 
-\subsection{Approximating the reduction map}
-
-In order for this approach to be useful in the likelihood-free setting, we need
-to be able to approximate both $s(\mathbf{x})$ and $p(s(\mathbf{x})|\theta)$
-based on a finite number of samples $\{\mathbf{x}_i\}$ drawn from the generative
-model $p(\mathbf{x}|\theta)$.
-Let denote the approximated dimensionality
-reduction map $\hat{s}(\mathbf{x})$ and its approximated distribution
-$\hat{p}(\hat{s}(\mathbf{x})|\theta)$. \glnote{Explain how to approximate $p(s)$ in practice?}
-
-One strength of this approach is that it factorizes the approximation of the
-per-event likelihood ratio ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from
+\subsection{Learning and calibrating $s$}
+
+In order for the proposed approach to be useful in the likelihood-free setting,
+we need to be able to approximate both $s(\mathbf{x})$ and
+$p(s(\mathbf{x})|\theta)$ based on a finite number of samples $\{\mathbf{x}_i\}$
+drawn from the generative model $p(\mathbf{x}|\theta)$.
+
+As outlined above, any consistent probalilistic classification algorithm can be
+used for learning an approximate  map $\hat{s}(\mathbf{x})$ of
+Eqn.~\ref{eqn:best-s-clf}. In the common case where the density ratio is
+expected to smoothly vary around $\mathbf{x}$, we would however recommend
+learning models whose output value $\hat{s}(\mathbf{x})$ also smoothly varies
+around $\mathbf{x}$, such as neural networks. In particular, tree-based
+methods are not expected to work so well for this use case, since they usually
+model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
+\glnote{We should comment as to what happens if $\hat{s}(x)$ is not monotonic with $r$}
+
+Given a reduction map $s$, our results show that a statistic equivalent to the
+likelihood ratio can be constructed, provided $p(s(\mathbf{x})|\theta)$ can be
+evaluated. Again, we do not have a direct and exact way for evaluating this
+density, but an approximation $\hat p(\hat s(\mathbf{x})| \theta)$ can be built
+instead, e.g. using density estimation or calibration algorithms, such as
+histograms, KDE or isotonic regression, as applied on $\{\hat
+s(\mathbf{x}_i)\}$, for samples $\{\mathbf{x}_i\}$  drawn from the generative
+model. Most notably, learning such an approximation of $p(s(\mathbf{x})|\theta)$
+is a much simpler problem than learning $p(\mathbf{x}|\theta)$, since the
+reduction $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
+the (simpler) informative content of $r(\mathbf{x})$ is preserved.
+
+One strength of the proposed approach is that it factorizes the approximation of the
+per-sample reduction ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from
 the calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx
@@
 $\hat{s}$ can still be well calibrated. In that case, one might loose power, but
-the resulting inference will still be valid.\glnote{Shouldn't we elaborate?
-Since this is one of our selling points, we should give more details in my
-opinion.} \glnote{Some of this is also already outlined in the previous section.}
+the resulting inference will still be valid.
 This point was made by \cite{Neal:2007zz} and is well appreciated by the
@@
 the use of machine learning classifiers precisely due to concerns about the
-calibration $p$-values in the face of nuisance parameters associated to the
+calibration of $p$-values in the face of nuisance parameters associated to the
 simulator.
-
-\glnote{We should comment as to what happens if $\hat{s}(x)$ is not monotonic with $r$}
-
-\glnote{Add reference to \cite{BiancaZadrozny} here.}
 
@@
 the same as the value that maximizes the likelihood ratio with respect to
-$p({\cal D}|\theta_1)$, for some fixed value of $\theta_1$\glnote{We should mention that the support of $p(\mathbf{x}|\theta_1)$ should cover the support of $p(\mathbf{x}|\theta)$.}.
+$p({\cal D}|\theta_1)$, for some fixed value of $\theta_1$ chosen such that the support of $p(\mathbf{x}|\theta_1)$ covers the support of $p(\mathbf{x}|\theta)$.
 This allows us to
@@
 the denominator $p(s(\mathbf{x}; \theta, \theta_1) |\theta_1)$ because this
-cancels Jacobian factors that vary with $\theta$. \glnote{Say that if $s$ is not monotonic with $r$, then it might be bias $\hat\theta$.}
+cancels Jacobian factors that vary with $\theta$.
+
+Finally, once the maximum likelihood estimates have been found for both the numerator
+and denominator of Eqn.~\ref{eqn:generalized-lr}, the generalized likelihood
+ratio can be estimated as outlined in Section~\ref{sec:approx}
+for simple hypothesis testing.
 
@@
 $\theta_0$ and $\theta_1$, using the procedure outlined in
-Section~\ref{sec:clf-for-ratios}, it is desirable and convenient to have a
+Section~\ref{sec:likelihood-ratio-tests}, it is desirable and convenient to have a
 smooth evolution of the classification score as a function of the parameters.

### 2016-02-02 jasa/paper-jasa.tex
Many small changes

@@
 The likelihood function is the central object that summarizes the information
-from an experiment needed for inference of model parameters. In particular, it
+from an experiment needed for inference of model parameters. It
 is key to many areas of science that report the results of classical
@@
 \begin{align}
-% p_U(u=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
-p_U(u=s(\mathbf{x})|\theta_0) &= \frac{d}{du}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq u\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
+% p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
+p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{du}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq u\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
 % &= \frac{d}{dy} \int_{\mathbb{R}^p} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
@@
 \begin{equation}
-p_U(u=s(\mathbf{x})|\theta_0) = \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
+p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) = \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
 \end{equation}
@@
 \begin{align}
-% p_U(u=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \nonumber \\
-p_U(u=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
 &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(u) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
@@
 \end{align}
-Similarly, Equation~\ref{eqn:hormander} can be used to derive $p_U(u=s(\mathbf{x})|\theta_1)$, finally yielding
+Similarly, Equation~\ref{eqn:hormander} can be used to derive $p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_1)$, finally yielding
 \begin{align}
-\frac{p_U(u=s(\mathbf{x})|\theta_0)}{p_U(u=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
-&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.
+\frac{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0)}{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
+&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}. \label{eqn:jacob}
 \end{align}
@@
 In light of this result, the likelihood ratio estimation problem can now
-be recast as a (probabilistic) classification problem, by noticing that the regression
+be recast as a (probabilistic) classification problem, by noticing that the decision
 function
@@
 of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
-$p_U(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.
+$p_{\mathbf{U}}(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.
 
@@
 learning models whose output value $\hat{s}(\mathbf{x})$ also smoothly varies
-around $\mathbf{x}$, such as neural networks. In particular, tree-based
+around $\mathbf{x}$, such as neural networks. For small training sets, tree-based
 methods are not expected to work so well for this use case, since they usually
 model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
-\glnote{We should comment as to what happens if $\hat{s}(x)$ is not monotonic with $r$}
+In particular, in such cases where $s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$, Jacobian
+factors in Eqn.~\ref{eqn:jacob} do not cancel, resulting in a biased
+approximation of the density ratio.
 
@@
 instead, e.g. using density estimation or calibration algorithms, such as
-histograms, KDE or isotonic regression, as applied on $\{\hat
+histograms, KDE or isotonic regression, and as applied on $\{\hat
 s(\mathbf{x}_i)\}$, for samples $\{\mathbf{x}_i\}$  drawn from the generative
@@
 
-One strength of the proposed approach is that it factorizes the approximation of the
-per-sample reduction ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from
-the calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx
-p(\hat{s}(\mathbf{x})|\theta)$). Thus, even if the classifier does a poor job at
-reproducing the level sets of the per-event likelihood ratio, the density of
+One strength of the proposed approach is that it factorizes the approximation of
+the per-sample reduction ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from the
+calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx
+p(\hat{s}(\mathbf{x})|\theta)$). Thus, even if the classifier does a poor at
+learning the optimal decision function~\ref{eqn:best-s-clf} and therefore at
+reproducing the level sets of the per-sample likelihood ratio, the density of
 $\hat{s}$ can still be well calibrated. In that case, one might loose power, but
-the resulting inference will still be valid.
-This point was made by \cite{Neal:2007zz} and is well appreciated by the
-particle physics community that typically takes a conservative attitude towards
-the use of machine learning classifiers precisely due to concerns about the
-calibration of $p$-values in the face of nuisance parameters associated to the
-simulator.
-
-\glnote{Add pseudo-code to summarize the whole pipeline?}
+the resulting inference will still be valid. This point was made by
+\cite{Neal:2007zz} and is well appreciated by the particle physics community
+that typically takes a conservative attitude towards the use of machine learning
+classifiers precisely due to concerns about the calibration of $p$-values in the
+face of nuisance parameters associated to the simulator.
 
@@
 In order to provide parameter inference in the likelihood-free setting as
-described above, we must train a family $s(\mathbf{x};\theta,\theta_1)$ of classifiers parameterized by
+described above, we must train a family $s(\mathbf{x};\theta_0,\theta_1)$ of classifiers parameterized by
 $\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
@@
 Eqn.~\ref{eqn:generalized-lr}.
+While the function
+$p(\mathbf{x}|\theta_1)/(p(\mathbf{x}|\theta_0)+p(\mathbf{x}|\theta_1))$ will
+minimize the expected squared error loss based on training data produced
+according to Algorithm~\ref{alg:training}, it is not clear how training data
+from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real
+world classifier with finite capacity. This is left as an area for future work.
 
@@
 
-While the function
-$p(\mathbf{x}|\theta_1)/(p(\mathbf{x}|\theta_0)+p(\mathbf{x}|\theta_1))$ will
-minimize the expected squared error loss based on training data produced
-according to Algorithm~\ref{alg:training}, it is not clear how training data
-from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real
-world classifier with finite capacity. This is left as an area for future work.
-
-\glnote{Shall we keep Algorithm~\ref{alg:training} or rather give a sketch in
-words? In its current state, the algorithm does not explain how to build ${\cal
-L}$ when the parameter spaces are continuous. It also does not specify how many
-samples should be generated per parameter value.}
-
 \subsection{Parameterized calibration}
@@
 estimation) to approximate $p(s|\theta)$ for specific parameter
-points. For a single parameter point, this is a tractable univariate density
+points. For a single parameter point $\theta$, this is a tractable univariate density
 estimation problem. The challenge comes from the need to calibrate this density
-for all values of $\theta$. A straight forward approach would be to run the
+for all values of $\theta$. A straightforward approach would be to run the
 generative model on demand for any particular value of $\theta$. In the context
@@
 model is computationally expensive or has high-latency (for instance some human
-intervention is required to reconfigure the generative model).  In HEP, with a
+intervention is required to reconfigure the generative model).
+
+In high energy physics, with a
 fixed classifier, it has become common  to interpolate the distribution between
@@
 In the special case of (simple or composite) hypothesis testing between
-models defined as mixtures of several components, i.e. when
+models defined as mixtures of several components, i.e. when $p(\mathbf{x}|\theta$ can be written as
 \begin{equation}
@@
 \end{equation}
-the target likelihood ratio can be rewritten in terms of pairwise
+the target likelihood ratio can be formulated in terms of pairwise
 classification problems. We indeed have
@@
 
-In applications where mixture models are commonly used, e.g. as in high energy
-physics, this decomposition allows one to construct better likelihood ratio
-estimates since it forces classifiers $s_{c,c'}$ to focus on simpler
-sub-problems, for which higher accuracy is expected.
+In applications where mixture models are commonly used this decomposition allows
+one to construct better likelihood ratio estimates since it forces classifiers
+$s_{c,c'}$ to focus on simpler sub-problems, for which higher accuracy is
+expected.
 
@@
 consistency, this approach is known to be ineffective in
-practice~\citep{sugiyama2007covariate,bickel2009discriminative}, in particular
-as it relies on modeling numerator and denominator high-dimensional densities,
+practice~\citep{sugiyama2007covariate,bickel2009discriminative}, since
+it relies on modeling numerator and denominator high-dimensional densities,
 which is a harder problem than modeling their ratio only.
 While the proposed method also proceeds in two similar steps, estimating
-$p(s(\mathbf{x}))$ is a much simpler problem than estimating $p(\mathbf{x})$,
+$p(s(\mathbf{x}))$ is much easier than estimating $p(\mathbf{x})$,
 since $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
@@
 \cite{sugiyama2012density,gretton2009covariate,nguyen2010estimating,vapnik2013constructive}
-and references therein) also provide solutions for estimating
+and references therein) provide solutions for estimating
 $r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$, in one step. Under some
@@
 
-\glnote{Improve given new draft.}
 In this work, we have outlined an approach to reformulate generalized likelihood
@@
 which will continue to improve. The calibration procedure for a particular
-parameter point is fairly straight forward since it involves estimating a
+parameter point is fairly straightforward since it involves estimating a
 univariate density using a generative model of the data. The difficulty of the
@@
 dimensionality of $\theta$, the complexity of the resulting likelihood function,
-and the difficulties associated to running the simulator.
+or the practical issues associated to running the simulator.
 

### 2016-02-02 jasa/paper-jasa.tex
Section 3.2: parameterized classification

@@
 learning to train discriminative classifiers that take advantage of a high
-dimensional feature vector $\mathbf{x}$. \glnote{We should then elaborate on how the
-proposed method could help these searches.}
+dimensional feature vector $\mathbf{x}$.
 
@@
 
+Thus far we have shown that the target likelihood ratio
+$r(\mathbf{x};\theta_0,\theta_1)=\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$
+with high dimensional features $\mathbf{x}$ can be reproduced via the univariate
+densities $p(s(\mathbf{x})|\theta_0)$ and $p(s(\mathbf{x})|\theta_1)$ if the
+reduction $s(\mathbf{x})$ is monotonic with $r(\mathbf{x};\theta_0,\theta_1)$.
+We now generalize from the ratio of two simple hypotheses specified by
+$\theta_0$ and $\theta_1$ to the case of composite hypothesis testing where
+$\theta$ are continuous model parameters.
+
+\subsection{Composite hypothesis testing}
+
 % Thus far we have shown that the target likelihood ratio
@@
 
-Thus far we have shown that the target likelihood ratio
-$r(\mathbf{x};\theta_0,\theta_1)=\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$
-with high dimensional features $\mathbf{x}$ can be reproduced via the univariate
-densities $p(s(\mathbf{x})|\theta_0)$ and $p(s(\mathbf{x})|\theta_1)$ if the
-reduction $s(\mathbf{x})$ is monotonic with $r(\mathbf{x};\theta_0,\theta_1)$.
-We now generalize from the ratio of two simple hypotheses specified by
-$\theta_0$ and $\theta_1$ to the case of composite hypothesis testing where
-$\theta$ are continuous model parameters.
 %
@@
 
-\subsection{Parameterized classifier}
+% For completeness, the proposed method is summarized in
+% Algorithm~\ref{alg:training}, while further details for training and calibrating
+% efficiently a family $s(\mathbf{x};\theta_0,\theta_1)$ of classifiers are
+% discussed in sections \ref{sec:param-clf} and \ref{sec:param-calibration}.
+
+\subsection{Parameterized classification}
+\label{sec:param-clf}
 
 In order to provide parameter inference in the likelihood-free setting as
-described above, we must train a family $s(\mathbf{x};\theta_0,\theta_1)$ of classifiers parameterized by
-$\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
-hypotheses, respectively. While this could be done independently for all
-$\theta_0$ and $\theta_1$, using the procedure outlined in
-Section~\ref{sec:likelihood-ratio-tests}, it is desirable and convenient to have a
-smooth evolution of the classification score as a function of the parameters.
-For this reason, we anticipate a single learning stage based on training data with input
-$(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
+described above, we must train a family $s(\mathbf{x};\theta_0,\theta_1)$ of
+classifiers parameterized by $\theta_0$ and $\theta_1$, the parameters
+associated to the null and alternate hypotheses, respectively. While this could
+be done independently for all $\theta_0$ and $\theta_1$, using the procedure
+outlined in Section~\ref{sec:likelihood-ratio-tests}, it is desirable and
+convenient to have a smooth evolution of the classification score as a function
+of the parameters. For this reason, we anticipate a single learning stage based
+on training data with input $(\mathbf{x}, \theta_0, \theta_1)_i$ and target
+$y_i$, as outlined in Algorithm~\ref{alg:training}.
+Somewhat unusually, the
 unknown values of the parameters are taken as input to the classifier; their
 values will be specified via the enveloping (generalized) likelihood ratio of
-Eqn.~\ref{eqn:generalized-lr}.
-While the function
-$p(\mathbf{x}|\theta_1)/(p(\mathbf{x}|\theta_0)+p(\mathbf{x}|\theta_1))$ will
-minimize the expected squared error loss based on training data produced
-according to Algorithm~\ref{alg:training}, it is not clear how training data
-from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real
-world classifier with finite capacity. This is left as an area for future work.
+Eqn.~\ref{eqn:generalized-lr}.  In this way, the parameterized classifier
+now models the distribution of the output $y$ conditional to $(\mathbf{x}, \theta_0, \theta_1)$,
+for any $\mathbf{x}$ and any combination of parameter values $\theta_0, \theta_1$.
+
+While the optimal decision function \ref{eqn:best-s-clf} is expected to be
+learned for the parameter values $\theta_0$ and $\theta_1$ selected in
+Algorithm~\ref{alg:training}, it is not clear whether the optimal decision
+function can be expected for data generated from  $\theta'_0$ and $\theta'_1$
+never jointly encountered at learning. Similarly, it is not clear how the
+limited capacity of the classifier may impact the performance the resulting
+parameterized decision function. These issues are left as an area for future
+work, but preliminary exploration by \cite{Baldi:2016fzo} show that
+Algorithm~\ref{alg:training} is an effective practical approach.
 
 \begin{algorithm}[t]
-\caption{Training of a parameterized classifier.}\label{alg:training}
-\glnote{Outline the full generalized likelihood ratio estimation pipeline instead}
+\caption{Learning a parametrized classifier.}\label{alg:training}
 \begin{algorithmic}
-        \State ${\cal L := \{ \}}$
-        \For{ $\theta_0$ in $\Theta_0$ }
-        	\For{ $\theta_1$ in $\Theta_1$ }
-        		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$
-        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, u=0) \}$ to ${\cal L}$
-        		\State generate $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$
-        		\State append $\{ (\mathbf{x}, \theta_0, \theta_1, u=1) \}$ to ${\cal L}$
-        	\EndFor
-        \EndFor
-        \State Learn $s(\mathbf{x}; \theta_0, \theta_1)$ from ${\cal L}$
+    \State ${\cal L := \{ \}}$;
+    \While{ $\text{size}({\cal L}) < N $ }
+        \State Select or draw $\theta_0$, $\theta_1$ from $\Theta_0$, $\Theta_1$;
+	    \State Draw $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$;
+		\State ${\cal L} := {\cal L} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=0) \}$;
+		\State Draw $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$;
+		\State ${\cal L} := {\cal L} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=1) \}$;
+    \EndWhile
+    \State Learn a single classifier $s(\mathbf{x}; \theta_0, \theta_1)$ from ${\cal L}$.
 \end{algorithmic}
@@
 \subsection{Parameterized calibration}
+\label{sec:param-calibration}
 

### 2016-02-02 jasa/paper-jasa.tex
Section 4.2: HEP

@@
 
-\glnote{Rewrite and shorten.}
-
 In high energy physics, we are often searching for some class of events,
 generically referred to as \textit{signal}, in the presence of a separate class
-of \textit{background} events.  For each event we measure some quantities $x$
-that have corresponding distributions $p_b(x|\nu)$ for background and
-$p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing
-uncertainties in the underlying physics prediction or response of the
-measurement device. The total model is a mixture of the signal and background,
-and $\mu$ is the mixture coefficient associate to the signal component.
+of \textit{background} events.  For each event we measure some quantities
+$\mathbf{x}$, with corresponding distributions $p_s(\mathbf{x}|\nu)$ for signal
+and $p_b(\mathbf{x}|\nu)$ for background, and where $\nu$ are nuisance
+parameters describing uncertainties in the underlying physics prediction or
+response of the measurement device. The total model is a mixture of the signal
+and background, and $\mu$ is the mixture coefficient associated to the signal
+component, that is
 \begin{equation}\label{eq:hepGen}
-p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( x_e \, |\,  \nu)  + (1-\mu)\, p_b( x_e \,|\, \nu) \,\right] \; ,
-\end{equation}
-New particle searches at the LHC are typically framed as hypothesis test where
-the null corresponds to $\mu=0$, and the generalized likelihood ratio is used as a test
-statistic.
-
-\subsubsection{Typical usage pattern}
-
-In this setting, large samples of pseudo-data $\{x_i, y_i\}$ generated with some
-nominal values of the parameters $\nu_0$, where $u=0$ corresponds to the
-background density $p_b(x|\nu_0)$  and $u=1$ corresponds to signal density
-$p_s(x|\nu_0)$. Importantly, the $u=1$ label corresponds to the signal only, and
-not to the alternate signal-plus-background hypothesis. The resulting classifier
-approximates the regression function $p_s(x|\nu_0)/(p_s(x|\nu_0)+p_b(x|\nu_0))$,
-which is one to one with the likelihood ratio of the null to the alternate
-$p(x|\mu=0,\nu_0)/p(x|\mu,\nu_0)$ for all $\mu$. Associating the $u=1$ label to
-the signal component has advantages because it helps the classifier focus its
-capacity on the relevant regions in the feature space, particularly when the
-signal is a very small perturbation to the background (i.e. $\mu \ll 1$).
-
- Once the classifier is trained, large samples of pseudo-data are drawn from
- $p_s(x | \nu)$ and $p_b(x | \nu)$ and we estimate the distributions
- $\hat{p}_s(\hat s | \nu)$ and $\hat{p}_b(\hat s | \nu)$ continuously
- parameterized in $\nu$. An example of the distributions of $\hat s$ for the
- signal and background events with $\nu=\nu_0$ is shown in
- Figure~\ref{fig:tmva}.
-
-\begin{figure}[t]
-\begin{center}
- \includegraphics[height=1.7in]{figures/example-TMVA-BDT.pdf}
- \includegraphics[height=1.7in]{figures/example-TMVA-ROC.pdf}
-\caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).
+p( {\cal D} | \mu, \nu) = \prod_{\mathbf{x} \in {\cal D}} \left[ \mu p_s( \mathbf{x} |  \nu)  + (1-\mu) p_b( \mathbf{x} | \nu) \right] \;.
+\end{equation}
+Accordingly, new particle searches at the LHC are typically framed as
+hypothesis tests where the null corresponds to $\mu=0$, and the generalized
+likelihood ratio is used as a test statistic.
+
+In this setting, a classifier is usually built from a large sample of data
+$\{\mathbf{x}_i, y_i\}$ generated with some nominal values of the parameters
+$\nu=\nu_0$, with $y=0$ for samples drawn from the background component and
+$y=1$ for samples drawn from the signal component. Importantly, the $y=1$ label
+corresponds to the signal only hypothesis, and not to the alternate
+signal-plus-background hypothesis. The resulting classifier approximates the
+regression function
+$\frac{p_s(\mathbf{x}|\nu_0)}{p_s(\mathbf{x}|\nu_0)+p_b(\mathbf{x}|\nu_0)}$, which is
+one to one with the likelihood ratio of the null to the alternate
+$\frac{p(\mathbf{x}|\mu=0,\nu_0)}{p(\mathbf{x}|\mu,\nu_0)}$ for all $\mu$ and therefore
+satisifies conditions of Theorem~\ref{thm:ratio-equivalence}. Associating the
+$y=1$ label to the signal component has advantages because it helps the
+classifier focus its capacity on the relevant regions in the feature space,
+particularly when the signal is a very small perturbation to the background
+(i.e. $\mu \ll 1$). Once the classifier is trained, large samples of data are
+generated from $p_s(\mathbf{x} | \nu)$ and $p_b(\mathbf{x} | \nu)$ and we
+estimate the distributions $\hat{p}_s(\hat s(\mathbf{x}) | \nu)$ and
+$\hat{p}_b(\hat s(\mathbf{x}) | \nu)$ continuously parameterized in $\nu$.
+
+ % An example of the distributions of $\hat s$ for the
+ % signal and background events with $\nu=\nu_0$ is shown in
+ % Figure~\ref{fig:tmva}.
+
+% \begin{figure}[t]
+% \begin{center}
+%  \includegraphics[height=1.7in]{figures/example-TMVA-BDT.pdf}
+%  \includegraphics[height=1.7in]{figures/example-TMVA-ROC.pdf}
+% \caption{Left: an example of the distributions $p_b(\hat s|\nu)$ and $p_s(\hat s|\nu)$ when the classifier $s$ is a boosted-decision tree (BDT).
 Right: the corresponding ROC curve (right) for this and other classifiers.
 (Figures taken from TMVA manual.)}
-\label{fig:tmva}
-\end{center}
-\end{figure}
-
-These steps feed into a subsequent statistical test based on the observed data
-${D=(x_1, \dots, x_n)}$. For each event, the classifier is evaluated and one
-performs inference on a parameter $\mu$ related to the presence of the signal
-contribution. In particular, one forms the statistical model.
-% \footnote{Sometimes
-% there is an additional Poisson term when expected number of signal and
-% background events is known, which is referred to as an extended likelihood or
-% marked Poisson model.}
-\begin{equation}\label{eq:typicalML}
-p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu \hat{p}_s( \hat s(x_e) \, |\,  \nu)  + (1-\mu)\, \hat{p}_b( \hat s(x_e) \,|\, \nu) \,\right] \; .
-\end{equation}
-
-\subsubsection{Comments on typical usage of machine learning in HEP}
+% \label{fig:tmva}
+% \end{center}
+% \end{figure}
+
+% These steps feed into a subsequent statistical test based on the observed data
+% ${D=(x_1, \dots, x_n)}$. For each event, the classifier is evaluated and one
+% performs inference on a parameter $\mu$ related to the presence of the signal
+% contribution. In particular, one forms the statistical model.
+% % \footnote{Sometimes
+% % there is an additional Poisson term when expected number of signal and
+% % background events is known, which is referred to as an extended likelihood or
+% % marked Poisson model.}
+% \begin{equation}\label{eq:typicalML}
+% p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu \hat{p}_s( \hat s(x_e) \, |\,  \nu)  + (1-\mu)\, \hat{p}_b( \hat s(x_e) \,|\, \nu) \,\right] \; .
+% \end{equation}
+
+% \subsubsection{Comments on typical usage of machine learning in HEP}
 
 Nuisance parameters are an after thought in the typical usage of machine
-learning in HEP. In fact, most discussions would related to the training and
-optimizing the classifier only consider $p_b(x)$ and $p_s(x)$ with $\nu=\nu_0$
+learning in high energy physics. In fact, most discussions related to the training and
+optimizing the classifier only consider $p_b(\mathbf{x})$ and $p_s(\mathbf{x})$, with $\nu=\nu_0$
 being implicit. However, as experimentalists we know that we must account for
 various forms of systematic uncertainty, parameterized by nuisance parameters
-$\nu$. In practice, we take the classifier as fixed and then propagate
-uncertainty through the classifier as in Eq.~\ref{eq:typicalML}. Building the
-distribution $p(\hat s|\nu)$ for values of $\nu$ other than the nominal $\nu_0$
-used to train the classifier can be thought of as a calibration necessary for
-classical statistical inference; however, this classifier is clearly not optimal
-for $\nu \ne \nu_0$.
-
-\subsubsection{A more powerful  approach}
-
-The standard use of machine learning in HEP can be improved by training a
-parameterized classifier corresponding to the generalized likelihood ratio test
-\begin{equation}
-\lambda(\mu) = \frac{p(D|\mu, \hat{\hat{\nu}})}{p(D|\hat \mu, {\hat{\nu}})} \;,
-\end{equation}
-following the approach outlined in Section~\ref{S:GLR}.
-
-There is an interesting distinction between this approach and the standard use
-in which the classifier is trained for a fixed $\nu_0$. In the standard use one
-trains a classifier for signal vs. background, which is equivalent (in an ideal
-setting) to training a classifier for  null (background-only) vs. alternate
-(signal-plus-background) since the resulting regression functions are one-to-one
-with each other. In contrast, in the case of the generalized likelihood ratio
-test
-\begin{equation}\label{eq:hep_improved}
- \frac{p(x| 0, \hat{\hat{ \nu}})}{p(x|\hat \mu, \hat\nu)} =  \frac{p_b(x| \hat{\hat{ \nu}})}{ \hat \mu p_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, p_b( x_e \,|\, \hat \nu)} \; ,
-\end{equation}
-the background components don't cancel and there is an additional term $p_b(x|
-\hat{\hat{ \nu}})/p_b(x| {\hat{ \nu}})$. In practice, with classifiers of finite
-capacity, there will be some trade-off between taking into account this
-additional term and the more challenging learning problem when $\mu$ is very
-small.
-
-\subsubsection{Measuring particle properties}
-
-\glnote{Rework this section, add a concrete example and compare with other
-likelihood-free inference algorithms.}
+$\nu$. In practice, we take the classifier $\hat s$ as fixed and then propagate
+uncertainty by estimating $\hat{p}_s(\hat s(\mathbf{x}) | \nu)$. Building such
+distributions for values of $\nu$ other than the nominal $\nu_0$
+used at training can be thought of as a calibration necessary for
+classical statistical inference. However, this classifier is clearly not optimal
+for $\nu \ne \nu_0$. For the this reason, it is expected that the integrated parameterized approach proposed in this work
+would yield more accurate estimates of the generalized likelihood ratio.
+
+% \subsubsection{A more powerful  approach}
+%
+% The standard use of machine learning in HEP can be improved by training a
+% parameterized classifier corresponding to the generalized likelihood ratio test
+% \begin{equation}
+% \lambda(\mu) = \frac{p(D|\mu, \hat{\hat{\nu}})}{p(D|\hat \mu, {\hat{\nu}})} \;,
+% \end{equation}
+% following the approach outlined in Section~\ref{S:GLR}.
+%
+% There is an interesting distinction between this approach and the standard use
+% in which the classifier is trained for a fixed $\nu_0$. In the standard use one
+% trains a classifier for signal vs. background, which is equivalent (in an ideal
+% setting) to training a classifier for  null (background-only) vs. alternate
+% (signal-plus-background) since the resulting regression functions are one-to-one
+% with each other. In contrast, in the case of the generalized likelihood ratio
+% test
+% \begin{equation}\label{eq:hep_improved}
+%  \frac{p(x| 0, \hat{\hat{ \nu}})}{p(x|\hat \mu, \hat\nu)} =  \frac{p_b(x| \hat{\hat{ \nu}})}{ \hat \mu p_s( x_e \, |\,  \hat\nu)  + (1- \hat \mu )\, p_b( x_e \,|\, \hat \nu)} \; ,
+% \end{equation}
+% the background components don't cancel and there is an additional term $p_b(x|
+% \hat{\hat{ \nu}})/p_b(x| {\hat{ \nu}})$. In practice, with classifiers of finite
+% capacity, there will be some trade-off between taking into account this
+% additional term and the more challenging learning problem when $\mu$ is very
+% small.
 
@@
 classifier in terms of the nuisance parameters $\nu$, the same approach can be
-used for parameter inference. In the case of new particle searches the parameter
+used for parameter inference. In the case of new particle searches, the parameter
 of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
-When measuring particle properties the distribution of the features also depend
+When measuring particle properties, the distribution of the features also depend
 on parameters such as a particle's mass and quantum numbers. This is easily
-accommodated by extending $p_s(x|\nu) \to p_s(x|\theta)$, where $\theta$
+accommodated by extending $p_s(\mathbf{x}|\nu) \to p_s(\mathbf{x}|\theta)$, where $\theta$
 includes both parameters of interest and nuisance parameters.
-
 This formalism represents a significant step forward in the usage of machine
-learning in HEP, where classifiers have always been used between two static
+learning in high energy physics, where classifiers have always been used between two static
 classes of events and not parameterized explicitly in terms of the physical
-quantities we wish to measure. The work of  ~\citep{Whiteson:2006ws} is similar
+quantities we wish to measure.
+The work of \cite{Whiteson:2006ws} is similar,
 as the stochastic optimization was directly trying to minimize the measurement
@@
 matrix element method can take several minutes of CPU time to calculate the
-likelihood ratio $q(x)$ for a single event. The work here can be seen as aiming
-at the same conceptual target, but utilizing machine learning to overcome the
+likelihood ratio for a single event. The work here can be seen as aiming
+at the same conceptual target, but relying on machine learning to overcome the
 complexity of the detector simulation. It also offers enormous speed increase
@@
 Contemporary examples where the technique presented here could have major impact
-on HEP include the measurement of coefficients to quantum mechanical operators
+include the measurement of coefficients to quantum mechanical operators
 describing the decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so
@@
 decays~\citep{Allanach:2000kt}.  Both of these examples involve data sets with
-many events, each with a feature vector $x$ that has on the order of 10
+many events, each with a feature vector $\mathbf{x}$ that has on the order of 10
 components, and a parameter vector $\theta$ with 5-10 parameters of interest and
@@
 coefficients of the Higgs decay uses the so-called matrix element likelihood
-analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is
+analysis (MELA) in which the equivalent of $s(\mathbf{x}; \theta_0, \theta_1)$ is
 approximated by neglecting detector effects~\citep{Gao:2010qx,Bolognesi:2012mm}.

### 2016-02-04 jasa/paper-jasa.tex
Section 2.3: Juan comments

@@
 &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}.
+\label{eqn:factorization}
 \end{align}
@@
 model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
-In particular, in such cases where $s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$, Jacobian
-factors in Eqn.~\ref{eqn:jacob} do not cancel, resulting in a biased
+In particular, in such cases where $s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$,
+the induced probability does not factorize as in Eqn.~\ref{eqn:factorization}, resulting in a biased 
 approximation of the density ratio.

### 2016-02-05 jasa/paper-jasa.tex
Fix typos

@@
 
-As outlined above, any consistent probalilistic classification algorithm can be
+As outlined above, any consistent probabilistic classification algorithm can be
 used for learning an approximate  map $\hat{s}(\mathbf{x})$ of
@@
 $\frac{p(\mathbf{x}|\mu=0,\nu_0)}{p(\mathbf{x}|\mu,\nu_0)}$ for all $\mu$ and therefore
-satisifies conditions of Theorem~\ref{thm:ratio-equivalence}. Associating the
+satisfies conditions of Theorem~\ref{thm:ratio-equivalence}. Associating the
 $y=1$ label to the signal component has advantages because it helps the

### 2016-02-05 jasa/paper-jasa.tex
Section 1: outline

@@
 Bayesian inference algorithms, such as Approximate Bayesian Computation, as it does
-not require the definition of a prior over model parameters. \glnote{Also
-mention conclusions on experimental results, as required by JASA guidelines.}
+not require the definition of a prior over model parameters. Experimental results
+on artificial problems illustrate the potential of the proposed method.
 
@@
 
-The rest of this paper is organized as follows. In Section... \glnote{todo}
+The rest of the paper is organized as follows. In
+Section~\ref{sec:likelihood-ratio-tests}, we first introduce the likelihood
+ratio test statistic in the setting of simple hypothesis testing, and then
+outline how it can be computed exactly using calibrated classifiers.
+In Section~\ref{sec:generalized-likelihood-ratio}, we generalize the proposed
+approach to the case of composite hypothesis testing and discuss directions for
+approximating the statistic efficiently. We then illustrate the proposed
+method in Section~\ref{sec:examples} and outline how it could improve
+statistical analysis within the field of high energy physics. Related work
+and conclusions are finally presented in sections~\ref{sec:related} and \ref{sec:conclusions}.
 
@@
 \section{Examples and applications}
+\label{sec:examples}
 

### 2016-02-05 jasa/paper-jasa.tex
Cleanup bibliography

@@
 discrete values of $\theta$ in order to produce a continuous parameterization for
-$p(s | \theta)$~\citep{Cranmer:2012sba}.
+$p(s | \theta)$~\citep{read1999linear,Cranmer:2012sba,baak2015interpolation}.
 One can easily imagine a number of approaches to embedding the classifier and

### 2016-02-05 jasa/paper-jasa.tex
Fix typos

@@
 vector $\mathbf{x}$~\citep{Sjostrand:2006za,Agostinelli:2002hh}. The ATLAS and CMS
-experiments have published  hundreds papers where the final result was
+experiments have published  hundreds of papers where the final result was
 formulated as a hypothesis test or confidence interval using a generalized
@@
 calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx
-p(\hat{s}(\mathbf{x})|\theta)$). Thus, even if the classifier does a poor at
+p(\hat{s}(\mathbf{x})|\theta)$). Thus, even if the classifier does a poor job at
 learning the optimal decision function~\ref{eqn:best-s-clf} and therefore at
@@
 In the special case of (simple or composite) hypothesis testing between
-models defined as mixtures of several components, i.e. when $p(\mathbf{x}|\theta$ can be written as
+models defined as mixtures of several components, i.e. when $p(\mathbf{x}|\theta)$ can be written as
 \begin{equation}

### 2016-02-05 jasa/paper-jasa.tex
Appendix: note for other losses

@@
 
-In this appendix, we show that the probabilistic classification framework
+In this appendix, we show for completeness that the probabilistic classification framework
 yields a reduction $s$ which satisfies conditions of Theorem~\ref{thm:ratio-equivalence}.
@@
 
+Proposition~\ref{thm:best-classifier} holds for the squared error loss, but it
+can be similarly shown that classifiers minimizing the exponential loss, the
+binomial log-likelihood (or cross-entropy) or the squared hinge loss are also
+monotonic with the density ratio~\citep{friedman2000additive,lin2002support}.
+However, a classifier with discrete outputs and minimizing the zero-one loss
+does not satisfy conditions of the theorem.
+
 \end{document}

### 2016-02-08 jasa/paper-jasa.tex

@@
 \usepackage{algpseudocode}
+\usepackage{caption}
+\usepackage{subcaption}
 
@@
 \subsection{Mixture models}
+\label{sec:mixture}
 
@@
 
-\glnote{todo}
-\glnote{+ diagnostics}
+As a simple but illustrative example, let us first consider the approximation of
+the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the mixtures
+$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
+Figure~\ref{fig:1} and defined as
+\begin{align}
+p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - g) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - g) p_{c_1}(\mathbf{x}) + g p_{c_2}(\mathbf{x}), \\
+p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
+\end{align}
+where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
+$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $g$ is a parameter of $p(\mathbf{x}|\theta_0)$
+set to $0.05$.
+
+Using the method outlined in Section~\ref{sec:approx},
+figures~\ref{fig:2}, \ref{fig:3} and \ref{fig:4} show the approximated
+ratio when using respectively a linear model, a neural network or a random
+forest. The blue curve corresponds to the true target log-ratio, the green curve is
+the direct approximation $\log \left( \frac{\hat s(\mathbf{x})}{1 - \hat s(\mathbf{x})} \right)$
+without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
+s(\mathbf{x}) | \theta_0)}{\hat p(\hat s(\mathbf{x}) | \theta_1)}$ as calibrated
+using kernel density estimation for a fixed bandwidth set to $0.2$. Finally, the
+cyan curve shows the approximated ratio when decomposing the mixtures, as
+described in Section~\ref{sec:mixture}.
+
+\begin{figure}
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{figures/fig1.pdf}
+        \caption{$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$}
+        \label{fig:1}
+    \end{subfigure}
+    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
+      %(or a blank line to force the subfigure onto a new line)
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{figures/fig2-a.pdf}
+        \caption{Linear model}
+        \label{fig:2}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{figures/fig2-b.pdf}
+        \caption{Neural network}
+        \label{fig:3}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{figures/fig2-c.pdf}
+        \caption{Random forest}
+        \label{fig:4}
+    \end{subfigure}
+    \caption{Approximation of the log-likelihood ratio $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ with calibrated classifiers.
+\glnote{I need to regenerate these figures using cross-validation for training and calibrating!}}\label{fig:animals}
+
+\end{figure}
+
+\glnote{Diagnostics.}
+
+\glnote{Extend to illustrate a generalized likelihood ratio test with
+parameterized classifier / calibration by assuming that the true value of $g$ is unknown.}
 

### 2016-02-09 jasa/paper-jasa.tex
Section 4.1: regenerate figures

@@
 figures~\ref{fig:2}, \ref{fig:3} and \ref{fig:4} show the approximated
-ratio when using respectively a linear model, a neural network or a random
+ratio when using respectively a linear model, a 2-layer neural network or a random
 forest. The blue curve corresponds to the true target log-ratio, the green curve is
@@
 s(\mathbf{x}) | \theta_0)}{\hat p(\hat s(\mathbf{x}) | \theta_1)}$ as calibrated
-using kernel density estimation for a fixed bandwidth set to $0.2$. Finally, the
+using histograms. Finally, the
 cyan curve shows the approximated ratio when decomposing the mixtures, as
@@
 
+    \vspace{1em}
+
     \begin{subfigure}[b]{0.45\textwidth}
@@
     \end{subfigure}
-    \caption{Approximation of the log-likelihood ratio $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ with calibrated classifiers.
-\glnote{I need to regenerate these figures using cross-validation for training and calibrating!}}\label{fig:animals}
+    \caption{Approximation of the log-likelihood ratio $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ with calibrated classifiers.}
 

### 2016-02-09 jasa/paper-jasa.tex

@@
 \begin{algorithm}[t]
-\caption{Learning a parametrized classifier.}\label{alg:training}
+\caption{Learning a parameterized classifier.}\label{alg:training}
 \begin{algorithmic}
@@
 Using the method outlined in Section~\ref{sec:approx},
-figures~\ref{fig:2}, \ref{fig:3} and \ref{fig:4} show the approximated
+figures~\ref{fig:2a}, \ref{fig:2b} and \ref{fig:2c} show the approximated
 ratio when using respectively a linear model, a 2-layer neural network or a random
@@
 described in Section~\ref{sec:mixture}.
+Hyper-parameters of the classifiers are all tuned on a finite grid of values
+in order to minimize the mean squared error, as estimated by 3-fold cross-validation.
 
@@
         \caption{Linear model}
-        \label{fig:2}
+        \label{fig:2a}
     \end{subfigure}
@@
         \caption{Neural network}
-        \label{fig:3}
+        \label{fig:2b}
     \end{subfigure}
@@
         \caption{Random forest}
-        \label{fig:4}
+        \label{fig:2c}
     \end{subfigure}
-    \caption{Approximation of the log-likelihood ratio $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ with calibrated classifiers.}
-
+    \caption{Approximation of the log-likelihood ratio
+             $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$
+             with calibrated classifiers.}
 \end{figure}
 
-\glnote{Diagnostics.}
-
-\glnote{Extend to illustrate a generalized likelihood ratio test with
-parameterized classifier / calibration by assuming that the true value of $g$ is unknown.}
+The first fact to highlight is that, in the finite sample setting, uncalibrated
+direct approximations $\log \left( \frac{\hat s(\mathbf{x})}{1 - \hat
+s(\mathbf{x})} \right)$, as shown in green, do not yield very accurate estimates
+of the ratio, even when using universal approximators such as a neural network
+or a random forest. This observation is at the origin of this work, and led us
+to show that the ratio $r(\mathbf{x})$ can still be computed exactly by forming
+the equivalent statistic
+$\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
+transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
+constructing $\frac{\hat p(\hat s(\mathbf{x})|\theta_0)}{\hat p(\hat
+s(\mathbf{x})|\theta_1)}$ as we propose, estimates appear much closer to the
+true ratio than their direct counterparts, as illustrated in red. The
+approximated ratios show to be accurate \glnote{Would be nice to be quantitative
+here.} for all three classifiers, in particular in regions with enough density
+(i.e., from $\mathbf{x}=-3$ to $\mathbf{x}=2$) and where the capacity of the
+classifier is focused. By contrast, in regions where the densities are low
+(i.e., at the boundaries) the approximated ratio show to have high variance
+since the underlying classifier cannot make very accurate predictions, because
+of the few training samples it is trained on in these parts of the input space.
+Accordingly, when leveraging the fact that densities are mixtures and decomposing their
+ratio, the capacity of the underlying classifiers can be focused on easier
+supervised learning tasks, resulting as expected in even more accurate approximations, as
+shown in cyan for all three classifiers.
+
+As shown in the figures, qualitatively different results are obtained depending
+the classification algorithm embedded within the approximation of the ratio. In
+this case, the neural network-based approximation appears to work best, due to
+the capacity of the classifier to model a decision function $s(\mathbf{x})$
+continuous and almost monotonic with the ratio. By contrast, the random forest
+yields a decision function which is piece-wise constant, therefore non-continuous
+and not strictly monotonic with the ratio in the finite sample regime, which
+appears to result in more variance in the approximated ratio. Finally, the
+linear model produces a decision function that is continuous but only piece-wise
+monotonic with the ratio. Due to the symmetry of the ratio around
+$\mathbf{x}=1$, this still results in an accurate approximation, but poorer
+results should be expected on more complex problems. Similarly, calibrating
+$\hat s(\mathbf{x})$ to construct $\hat p(\hat s(\mathbf{x}))$ is also key to
+obtain accurate results. While standard histograms with fixed bins have been
+used here for illustrative purposes, we anticipate that the embedded density
+estimation algorithm should be carefully chosen and tuned, on a case by case
+basis.
+
+Let us now consider the composite hypothesis testing setting, by comparing
+$H_0: g = 0$ versus the alternative $H_1: g > 0$. That is, we want to evaluate
+\begin{equation}
+\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | g = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | g=\theta)} \; .
+\end{equation}
+Following the approach outlined in Section~\ref{sec:generalized-likelihood-ratio},
+computing the generalized likelihood ratio statistic requires finding
+the maximum likelihood estimator for the denominator, which can be carried out using
+approximated likelihood ratios. Reusing the decomposed neural network-based
+approximation (or equivalently, using a parameterized classifier) and minimizing the mean negative log-ratio of some observed data
+generated for $g=0.05$,
+we successfully find as estimator $\hat g=0.049$, as shown in Figure~\ref{fig:3}.
+From there, the generalized likelihood ratio can be approximated as in the simple
+hypothesis setting.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figures/fig3.pdf}
+    \caption{Using approximated likelihood ratios for parameter inference.}
+    \label{fig:3}
+\end{figure}
+
+\glnote{Smoothly interpolate between histogram bins, in order to get a smooth ratio as well. Worth mentioning in the paper!}
+\glnote{Discuss diagnostics?}
+\glnote{Better highlight advantages of the method?}
+\glnote{Conduct more thorough and convincing experiments?}
 

### 2016-02-10 jasa/paper-jasa.tex
Regenerate figures using same data for all classifiers

@@
 $p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $g$ is a parameter of $p(\mathbf{x}|\theta_0)$
-set to $0.05$.
+set to $0.05$. Samples drawn from $p(\mathbf{x}|\theta_0)$ are shown in the
+figure and will be used later for inference.
 
@@
 approximation (or equivalently, using a parameterized classifier) and minimizing the mean negative log-ratio of some observed data
-generated for $g=0.05$,
+generated for $g=0.05$ (as illustrated in Figure~\ref{fig:1}),
 we successfully find as estimator $\hat g=0.049$, as shown in Figure~\ref{fig:3}.
@@
 
-\glnote{Smoothly interpolate between histogram bins, in order to get a smooth ratio as well. Worth mentioning in the paper!}
+\glnote{Smoothly interpolate between histogram bins, in order to get a smooth ratio as well?}
 \glnote{Discuss diagnostics?}

### 2016-02-11 jasa/paper-jasa.tex
Acknowledgments

@@
 
+\section*{Acknowledgements} KC would like to thank Daniel Whiteson for
+encouragement throughout the project and Alex Ihler for challenging discussions
+that  led to a reformulation of the initial idea. KC would also like to thank
+Shimon Whiteson, Babak Shahbaba for advice in the presentation, Radford Neal for
+discussion of his earlier work, Yann LeCun, Philip Stark, and Pierre Baldi for
+their feedback on the project early in its conception, Bal\'azs K\'egl for
+feedback to the draft,  and Yuri
+Shirman for reassuring cross checks of the Theorem.  KC is supported by the US
+National Science Foundation grants PHY-0854724 and PHY-0955626.  KC is grateful
+to UC-Irvine for their hospitality while this research was carried out and the
+Moore and Sloan foundations for their generous support of the data science
+environment at NYU.
+\glnote{Update the acknowlegdments? What do we need to mention regarding my financial support?}
+
 \bibliographystyle{apalike}

### 2016-02-13 jasa/paper-jasa.tex
Regenerate figures

@@
 true ratio than their direct counterparts, as illustrated in red. The
-approximated ratios show to be accurate \glnote{Would be nice to be quantitative
-here.} for all three classifiers, in particular in regions with enough density
+approximated ratios show to be accurate for all three classifiers, in particular in regions with enough density
 (i.e., from $\mathbf{x}=-3$ to $\mathbf{x}=2$) and where the capacity of the
@@
 As shown in the figures, qualitatively different results are obtained depending
-the classification algorithm embedded within the approximation of the ratio. In
-this case, the neural network-based approximation appears to work best, due to
-the capacity of the classifier to model a decision function $s(\mathbf{x})$
-continuous and almost monotonic with the ratio. By contrast, the random forest
-yields a decision function which is piece-wise constant, therefore non-continuous
-and not strictly monotonic with the ratio in the finite sample regime, which
-appears to result in more variance in the approximated ratio. Finally, the
-linear model produces a decision function that is continuous but only piece-wise
-monotonic with the ratio. Due to the symmetry of the ratio around
-$\mathbf{x}=1$, this still results in an accurate approximation, but poorer
-results should be expected on more complex problems. Similarly, calibrating
-$\hat s(\mathbf{x})$ to construct $\hat p(\hat s(\mathbf{x}))$ is also key to
-obtain accurate results. While standard histograms with fixed bins have been
-used here for illustrative purposes, we anticipate that the embedded density
-estimation algorithm should be carefully chosen and tuned, on a case by case
-basis.
+the classification algorithm embedded within the approximation of the ratio. The
+neural network-based approximation appears to work well, due to the capacity of
+the classifier to model a decision function $s(\mathbf{x})$ continuous and
+almost monotonic with the ratio. The random forest yields a decision function
+which is piece-wise constant, therefore non-continuous and not strictly
+monotonic with the ratio in the finite sample regime. On this problem, the
+random forest still approximates the ratio correctly, but more variance can be
+observed. Finally, the linear model produces a decision function that is
+continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
+the ratio around $\mathbf{x}=1$, this still results in an accurate
+approximation, but poorer results should be expected on more complex problems.
+Similarly, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
+s(\mathbf{x}))$ is also key to obtain accurate results. While standard
+histograms with fixed bins have been used here for illustrative purposes, we
+anticipate that the embedded density estimation algorithm should be carefully
+chosen and tuned, on a case by case basis.
 
@@
 generated for $g=0.05$ (as illustrated in Figure~\ref{fig:1}),
-we successfully find as estimator $\hat g=0.049$, as shown in Figure~\ref{fig:3}.
+we successfully find as estimator $\hat g=0.048$, as shown in Figure~\ref{fig:3}.
 From there, the generalized likelihood ratio can be approximated as in the simple
@@
 
-\glnote{Smoothly interpolate between histogram bins, in order to get a smooth ratio as well?}
 \glnote{Discuss diagnostics?}

### 2016-02-16 jasa/paper-jasa.tex

@@
 $p_{\mathbf{U}}(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.
+For this reason, the proposed method is expected to be more robust than
+directly using $\frac{1 - s(\mathbf{x})}{s(\mathbf{x})}$ as an approximate of
+$r(\mathbf{x})$.
 
@@
 model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
-In particular, in such cases where $s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$,
+In particular, in such cases where $\hat s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$,
 the induced probability does not factorize as in Eqn.~\ref{eqn:factorization}, resulting in a biased
@@
 figures~\ref{fig:2a}, \ref{fig:2b} and \ref{fig:2c} show the approximated
-ratio when using respectively a linear model, a 2-layer neural network or a random
+ratio when using respectively a linear model, a 2-layer neural network and a random
 forest. The blue curve corresponds to the true target log-ratio, the green curve is
-the direct approximation $\log \left( \frac{\hat s(\mathbf{x})}{1 - \hat s(\mathbf{x})} \right)$
+the direct approximation $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$
 without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
@@
 The first fact to highlight is that, in the finite sample setting, uncalibrated
-direct approximations $\log \left( \frac{\hat s(\mathbf{x})}{1 - \hat
-s(\mathbf{x})} \right)$, as shown in green, do not yield very accurate estimates
+direct approximations $\log \left( \frac{1 - \hat
+s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$, as shown in green, do not yield very accurate estimates
 of the ratio, even when using universal approximators such as a neural network
@@
 neural network-based approximation appears to work well, due to the capacity of
-the classifier to model a decision function $s(\mathbf{x})$ continuous and
+the classifier to model a smooth decision function $s(\mathbf{x})$
 almost monotonic with the ratio. The random forest yields a decision function
@@
 observed. Finally, the linear model produces a decision function that is
-continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
+smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
 the ratio around $\mathbf{x}=1$, this still results in an accurate
@@
 
-\glnote{Discuss diagnostics?}
-\glnote{Better highlight advantages of the method?}
-\glnote{Conduct more thorough and convincing experiments?}
+\subsection{Advanced example}
+
+\glnote{todo}
 

### 2016-02-26 jasa/paper-jasa.tex

@@
 
-\subsection{Toy example}
+\subsection{Mixtures of 1D normals}
 
@@
 approximated likelihood ratios. Reusing the decomposed neural network-based
-approximation (or equivalently, using a parameterized classifier) and minimizing the mean negative log-ratio of some observed data
-generated for $g=0.05$ (as illustrated in Figure~\ref{fig:1}),
-we successfully find as estimator $\hat g=0.048$, as shown in Figure~\ref{fig:3}.
+approximation and minimizing the mean negative log-ratio of some observed data
+generated for $g=0.05$,
+Figure~\ref{fig:3} shows the empirical distribution of $\hat g$ when repeating the generation and optimization process.
+The figure clearly suggests that the proposed inference procedure yields in this case
+an unbiased estimator ($\mathbb{E} [\hat g] = 0.0502$, as empirically estimated over $200$ trials).
 From there, the generalized likelihood ratio can be approximated as in the simple
@@
     \includegraphics[width=0.5\textwidth]{figures/fig3.pdf}
-    \caption{Using approximated likelihood ratios for parameter inference.}
+    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat g$.}
     \label{fig:3}

### 2016-03-01 jasa/paper-jasa.tex
Final draft

@@
 \subsection{Composite hypothesis testing}
+\label{sec:composite-hypothesis-testing}
 
@@
 
-\subsection{Mixtures of 1D normals}
-
-As a simple but illustrative example, let us first consider the approximation of
-the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the mixtures
+In this section, we illustrate the proposed method on two representative
+examples and then discuss its application to high energy physics. For reproducibility and extended details, the source
+code is available online\footnote{\url{https://github.com/diana-hep/carl}}.
+
+\subsection{Likelihood ratios of mixtures of normals}
+
+As a simple and illustrative example, let us first consider the approximation of
+the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the 1D mixtures
 $p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\textwidth]{figures/fig3.pdf}
+    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat g$.}
+    \label{fig:3}
+\end{figure}
+
 Let us now consider the composite hypothesis testing setting, by comparing
@@
 
+\subsection{Parameterized inference from multidimensional data}
+
+Let us now consider the more challenging problem of maximum likelihood
+inference from multidimensional data generated by a black-box simulator.
+For the sake of the illustration, we will assume 5-dimensional observed samples $\mathbf{x}$ generated
+from the following process $p_0$:
+    \begin{enumerate}
+        \item $\mathbf{z} := (z_0, z_1, z_2, z_3, z_4)$, such that
+            $z_0 \sim {\cal N}(\mu=A, \sigma=1)$,
+            $z_1 \sim {\cal N}(\mu=B, \sigma=3)$,
+            $z_2 \sim {\text{Mixture}}({\cal N}(\mu=-2, \sigma=1), {\cal N}(\mu=2, \sigma=0.5))$,
+            $z_3 \sim {\text{Exponential}(\lambda=3)}$ and
+            $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
+        \item $\mathbf{x} := R  \mathbf{z}$, where $R$ is a fixed
+        semi-positive definite $5 \times 5$ matrix defining a fixed projection
+        of $\mathbf{z}$ into the observed space.
+    \end{enumerate}
+Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
+our goal is to infer the nominal values $A=1$ and $B=-1$ that
+were used for generating this data.
+
 \begin{figure}
     \centering
-    \includegraphics[width=0.5\textwidth]{figures/fig3.pdf}
-    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat g$.}
-    \label{fig:3}
+    \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
+    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $A=1$ and $B=-1$.}
+    \label{fig:4}
 \end{figure}
 
-\subsection{Advanced example}
-
-\glnote{todo}
+\begin{figure}
+    \centering
+    \includegraphics[width=1.0\textwidth]{figures/fig5.pdf}
+    \caption{Parameter inference from parameterized approximated likelihood ratios.
+             The constructed objective is minimized with Bayesian optimization.
+             Red dots correspond to sampled candidate pairs $(A,B)$.
+             The blue dot corresponds to the minimum found after $50$ iterations. Nominal values are $A=1$ and $B=-1$.}
+    \label{fig:5}
+\end{figure}
+
+As previously, we make use of Eqn.~\ref{eq:mle_withs} for parameter inference
+with approximated likelihood ratios and fix the alternate to $\theta_1:=(A=0,
+B=0)$. Following recommendations from Section~\ref{sec:param-clf} to reduce
+computational requirements, we train a single parameterized classifier
+$s(\mathbf{x};\theta:=(A,B),\theta_1)$ (i.e., a neural network with 2 hidden layers of 40 nodes) and whose output is calibrated on-the-fly
+with isotonic regression for every candidate parameter pair $\theta:=(A,B)$.
+Minimizing the mean negative log of the constructed objective with Bayesian
+optimization~\citep{brochu2010tutorial} eventually leads to the estimated
+posterior mean and variance shown in Figure~\ref{fig:5}. The objective is minimized after $50$ iterations
+at $(\hat A=1.00,\hat B=-1.15)$. Again, repeated experiments (omitted here) confirm that
+the proposed procedure yields an unbiased maximum likelihood estimator.
 
@@
 More generally, likelihood ratio testing directly relates to the density ratio
-estimation problem, which consists in estimating the ratio of two probability
-densities from given and finite collections of observations ${\cal D}_0$ and
+estimation problem, which consists in estimating the ratio of two
+densities from finite collections of observations ${\cal D}_0$ and
 ${\cal D}_1$. Density ratio estimation is connected to many machine learning
@@
 
-\section*{Acknowledgements} KC would like to thank Daniel Whiteson for
-encouragement throughout the project and Alex Ihler for challenging discussions
+\section*{Acknowledgements} KC and GL are both supported through NSF ACI-1450310.
+KC would like to thank Daniel Whiteson for
+encouragement and Alex Ihler for challenging discussions
 that  led to a reformulation of the initial idea. KC would also like to thank
@@
 feedback to the draft,  and Yuri
-Shirman for reassuring cross checks of the Theorem.  KC is supported by the US
-National Science Foundation grants PHY-0854724 and PHY-0955626.  KC is grateful
+Shirman for reassuring cross checks of the Theorem.   KC is grateful
 to UC-Irvine for their hospitality while this research was carried out and the
@@
 environment at NYU.
-\glnote{Update the acknowlegdments? What do we need to mention regarding my financial support?}
 

### 2016-03-01 jasa/paper-jasa.tex
Minor edits

@@
 In the special case of (simple or composite) hypothesis testing between
-models defined as mixtures of several components, i.e. when $p(\mathbf{x}|\theta)$ can be written as
+models defined as known mixtures of several components, i.e. when $p(\mathbf{x}|\theta)$ can be written as
 \begin{equation}
@@
 
-In applications where mixture models are commonly used this decomposition allows
+In applications where mixture models are commonly used, this decomposition allows
 one to construct better likelihood ratio estimates since it forces classifiers
@@
 computational requirements, we train a single parameterized classifier
-$s(\mathbf{x};\theta:=(A,B),\theta_1)$ (i.e., a neural network with 2 hidden layers of 40 nodes) and whose output is calibrated on-the-fly
+$s(\mathbf{x};\theta:=(A,B),\theta_1)$; in this case, a 2-layer neural network.
+Since the generative model is not expensive, the classifier output is calibrated on-the-fly
 with isotonic regression for every candidate parameter pair $\theta:=(A,B)$.

### 2016-03-01 jasa/paper-jasa.tex
Update MLE histogram figure

@@
 The figure clearly suggests that the proposed inference procedure yields in this case
-an unbiased estimator ($\mathbb{E} [\hat g] = 0.0502$, as empirically estimated over $200$ trials).
+an unbiased estimator ($\mathbb{E} [\hat g] = 0.04995$, as empirically estimated over $1000$ trials).
 From there, the generalized likelihood ratio can be approximated as in the simple

### 2016-03-02 jasa/paper-jasa.tex
Compare with true contours

@@
     \centering
-    \includegraphics[width=1.0\textwidth]{figures/fig5.pdf}
-    \caption{Parameter inference from parameterized approximated likelihood ratios.
-             The constructed objective is minimized with Bayesian optimization.
-             Red dots correspond to sampled candidate pairs $(A,B)$.
-             The blue dot corresponds to the minimum found after $50$ iterations. Nominal values are $A=1$ and $B=-1$.}
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[height=12.5em]{figures/fig5a.pdf}
+        \caption{True mean negative log-ratio}
+        \label{fig:5a}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[height=12.5em]{figures/fig5b.pdf}
+        \caption{Estimated mean negative log-ratio}
+        \label{fig:5b}
+    \end{subfigure}
+    \caption{Inference from parameterized approximated likelihood ratios.
+             (\ref{fig:5a}) True mean negative log-ratio for the observed data ${\cal D}$.
+The red dot corresponds to the nominal values $A=1, B=-1$, while the green dot is the maximum likelihood estimator $\hat A=1.006, \hat B=-1.012$.
+             (\ref{fig:5a}) Estimated mean negative log-ratio, as approximated from the constructed objective by the gaussian process within the Bayesian optimization process.
+The blue dot is the estimated maximum likelihood estimator $\hat A=0.956, \hat B=-0.869$ while the white dots correspond to the points sampled during optimization.}
     \label{fig:5}
@@
 with isotonic regression for every candidate parameter pair $\theta:=(A,B)$.
-Minimizing the mean negative log of the constructed objective with Bayesian
+Minimizing the mean negative log of the constructed ratio with Bayesian
 optimization~\citep{brochu2010tutorial} eventually leads to the estimated
-posterior mean and variance shown in Figure~\ref{fig:5}. The objective is minimized after $50$ iterations
-at $(\hat A=1.00,\hat B=-1.15)$. Again, repeated experiments (omitted here) confirm that
+posterior mean shown in Figure~\ref{fig:5b}, which is close to the true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations,
+the estimated minimizer is found at $(\hat A=0.956, \hat B=-0.869$). 
+Again, repeated experiments (omitted here) confirm that
 the proposed procedure yields an unbiased maximum likelihood estimator.

### 2016-03-02 jasa/paper-jasa.tex
Some more changes

@@
     \label{thm:ratio-equivalence}
-    Let $\mathbf{X}$ be a random vector vector with values in ${\cal X} \subseteq \mathbb{R}^p$ and parameterized probability
+    Let $\mathbf{X}$ be a random vector with values in ${\cal X} \subseteq \mathbb{R}^p$ and parameterized probability
     density $p_{\mathbf{X}}(\mathbf{x} = (x_1, ..., x_p)|\theta)$ and let
@@
 modeled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
-from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies conditions of
+from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies the conditions of
 Theorem~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s} for further details).
@@
 Equivalently,
-Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
+Theorem~\ref{thm:ratio-equivalence} shows that in the case that we learn a probabilistic
 classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
@@
 directly using $\frac{1 - s(\mathbf{x})}{s(\mathbf{x})}$ as an approximate of
-$r(\mathbf{x})$.
+$r(\mathbf{x})$ (which indeed converges towards $r(x)$ when $s(x)$ tends to $s^*(x)$).
 
@@
 model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
-In particular, in such cases where $\hat s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$,
+In such cases where $\hat s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$,
 the induced probability does not factorize as in Eqn.~\ref{eqn:factorization}, resulting in a biased
 approximation of the density ratio.
+Provided enough training data, accurate results could however still be achieved,
+given the universal approximator capacity of tree-based models.
 
@@
     \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
-    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $A=1$ and $B=-1$.}
+    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $(A=1, B=-1)$.}
     \label{fig:4}
@@
              (\ref{fig:5a}) True mean negative log-ratio for the observed data ${\cal D}$.
-The red dot corresponds to the nominal values $A=1, B=-1$, while the green dot is the maximum likelihood estimator $\hat A=1.006, \hat B=-1.012$.
-             (\ref{fig:5a}) Estimated mean negative log-ratio, as approximated from the constructed objective by the gaussian process within the Bayesian optimization process.
-The blue dot is the estimated maximum likelihood estimator $\hat A=0.956, \hat B=-0.869$ while the white dots correspond to the points sampled during optimization.}
+The red dot corresponds to the nominal values $(A=1, B=-1)$, while the green dot is the maximum likelihood estimator $(\hat A=1.006, \hat B=-1.012)$.
+             (\ref{fig:5b}) Estimated mean negative log-ratio, as approximated from the constructed objective by the gaussian process within the Bayesian optimization procedure.
+The blue dot is the estimated maximum likelihood estimator $(\hat A=0.956, \hat B=-0.869)$ while the white dots correspond to the points sampled during optimization.}
     \label{fig:5}

### 2016-03-02 jasa/paper-jasa.tex
Spelling

@@
 The red dot corresponds to the nominal values $(A=1, B=-1)$, while the green dot is the maximum likelihood estimator $(\hat A=1.006, \hat B=-1.012)$.
-             (\ref{fig:5b}) Estimated mean negative log-ratio, as approximated from the constructed objective by the gaussian process within the Bayesian optimization procedure.
+             (\ref{fig:5b}) Estimated mean negative log-ratio, as approximated from the constructed objective by the Gaussian process within the Bayesian optimization procedure.
 The blue dot is the estimated maximum likelihood estimator $(\hat A=0.956, \hat B=-0.869)$ while the white dots correspond to the points sampled during optimization.}
@@
 
-\section*{Acknowledgements} KC and GL are both supported through NSF ACI-1450310.
+\section*{Acknowledgments} KC and GL are both supported through NSF ACI-1450310.
 KC would like to thank Daniel Whiteson for

### 2016-03-06 jasa/paper-jasa.tex
Added Juan affiliation
Update first example

@@
   \title{\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers}
-  \author{Kyle Cranmer, Juan Pavez and Gilles Louppe\\
-          New York University}
+  \author{Kyle Cranmer$^1$, Juan Pavez$^2$ and Gilles Louppe$^1$\hspace{.2cm}\\
+          $^1$New York University\\
+          $^2$Federico Santa María University}
   \maketitle
@@
 approximation of the density ratio.
-Provided enough training data, accurate results could however still be achieved,
+Provided enough training data, accurate results can however still be achieved,
 given the universal approximator capacity of tree-based models.
@@
 $p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
-Figure~\ref{fig:1} and defined as
+Figure~\ref{fig:1a} and defined as
 \begin{align}
-p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - g) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - g) p_{c_1}(\mathbf{x}) + g p_{c_2}(\mathbf{x}), \\
+p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - \gamma) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \gamma) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}), \\
 p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
@@
 where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
-$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $g$ is a parameter of $p(\mathbf{x}|\theta_0)$
+$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $\gamma$ is a parameter of $p(\mathbf{x}|\theta_0)$
 set to $0.05$. Samples drawn from $p(\mathbf{x}|\theta_0)$ are shown in the
@@
 Using the method outlined in Section~\ref{sec:approx},
-figures~\ref{fig:2a}, \ref{fig:2b} and \ref{fig:2c} show the approximated
+figures~\ref{fig:1b}, \ref{fig:1c} and \ref{fig:1d} show the approximated
 ratio when using respectively a linear model, a 2-layer neural network and a random
@@
 using histograms. Finally, the
-cyan curve shows the approximated ratio when decomposing the mixtures, as
+cyan curve shows the approximated ratio when both decomposing the mixtures and calibrating the underlying classifiers, as
 described in Section~\ref{sec:mixture}.
-Hyper-parameters of the classifiers are all tuned on a finite grid of values
-in order to minimize the mean squared error, as estimated by 3-fold cross-validation.
 
@@
     \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1.pdf}
+        \includegraphics[width=\textwidth]{figures/fig1a.pdf}
         \caption{$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$}
-        \label{fig:1}
+        \label{fig:1a}
     \end{subfigure}
@@
     \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig2-a.pdf}
+        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
         \caption{Linear model}
-        \label{fig:2a}
+        \label{fig:1b}
     \end{subfigure}
@@
     \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig2-b.pdf}
+        \includegraphics[width=\textwidth]{figures/fig1c.pdf}
         \caption{Neural network}
-        \label{fig:2b}
+        \label{fig:1c}
     \end{subfigure}
@@
     \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig2-c.pdf}
+        \includegraphics[width=\textwidth]{figures/fig1d.pdf}
         \caption{Random forest}
-        \label{fig:2c}
+        \label{fig:1d}
     \end{subfigure}
@@
 direct approximations $\log \left( \frac{1 - \hat
-s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$, as shown in green, do not yield very accurate estimates
+s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate estimates
 of the ratio, even when using universal approximators such as a neural network
@@
 true ratio than their direct counterparts, as illustrated in red. The
-approximated ratios show to be accurate for all three classifiers, in particular in regions with enough density
-(i.e., from $\mathbf{x}=-3$ to $\mathbf{x}=2$) and where the capacity of the
+approximated ratios show to be more accurate for all three classifiers, in particular in regions with enough density
+and where the capacity of the
 classifier is focused. By contrast, in regions where the densities are low
-(i.e., at the boundaries) the approximated ratio show to have high variance
-since the underlying classifier cannot make very accurate predictions, because
+(i.e., at the boundaries) the approximated ratio show to less acurate
+since the underlying classifier cannot make precise predictions, because
 of the few training samples it is trained on in these parts of the input space.
-Accordingly, when leveraging the fact that densities are mixtures and decomposing their
+When leveraging the fact that densities are mixtures and decomposing their
 ratio, the capacity of the underlying classifiers can be focused on easier
 supervised learning tasks, resulting as expected in even more accurate approximations, as
-shown in cyan for all three classifiers.
+shown in cyan.
 
@@
 the classification algorithm embedded within the approximation of the ratio. The
-neural network-based approximation appears to work well, due to the capacity of
-the classifier to model a smooth decision function $s(\mathbf{x})$
-almost monotonic with the ratio. The random forest yields a decision function
+neural network-based approximation appears to work best, due to the capacity of
+the classifier to model a smooth decision function $s(\mathbf{x})$ which is
+very close to the optimal decision function. The random forest yields a decision function
 which is piece-wise constant, therefore non-continuous and not strictly
@@
 approximation, but poorer results should be expected on more complex problems.
-Similarly, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
-s(\mathbf{x}))$ is also key to obtain accurate results. While standard
-histograms with fixed bins have been used here for illustrative purposes, we
+While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
+s(\mathbf{x}))$ is key to obtain accurate results. Standard
+histograms with fixed bins have been used here for illustrative purposes, but we
 anticipate that the embedded density estimation algorithm should be carefully
@@
     \centering
-    \includegraphics[width=0.5\textwidth]{figures/fig3.pdf}
-    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat g$.}
-    \label{fig:3}
+
+    \begin{subfigure}[t]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{figures/fig2a.pdf}
+        \caption{True MLEs vs. approximated MLEs.}
+        \label{fig:2a}
+    \end{subfigure}
+    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
+      %(or a blank line to force the subfigure onto a new line)
+    \begin{subfigure}[t]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{figures/fig2b.pdf}
+        \caption{$-2 \log \frac{p({\cal D}|\gamma=0.05)}{\sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)}$}
+        \label{fig:2b}
+    \end{subfigure}
+
+    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat \gamma$.}
+    \label{fig:2}
 \end{figure}
@@
 Let us now consider the composite hypothesis testing setting, by comparing
-$H_0: g = 0$ versus the alternative $H_1: g > 0$. That is, we want to evaluate
+$H_0: \gamma = 0$ versus the alternative $H_1: \gamma > 0$. That is, we want to evaluate
 \begin{equation}
-\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | g = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | g=\theta)} \; .
+\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | \gamma = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)} \; .
 \end{equation}
-Following the approach outlined in Section~\ref{sec:generalized-likelihood-ratio},
-computing the generalized likelihood ratio statistic requires finding
-the maximum likelihood estimator for the denominator, which can be carried out using
-approximated likelihood ratios. Reusing the decomposed neural network-based
-approximation and minimizing the mean negative log-ratio of some observed data
-generated for $g=0.05$,
-Figure~\ref{fig:3} shows the empirical distribution of $\hat g$ when repeating the generation and optimization process.
-The figure clearly suggests that the proposed inference procedure yields in this case
-an unbiased estimator ($\mathbb{E} [\hat g] = 0.04995$, as empirically estimated over $1000$ trials).
-From there, the generalized likelihood ratio can be approximated as in the simple
-hypothesis setting.
+Following the approach outlined in
+Section~\ref{sec:generalized-likelihood-ratio}, computing the generalized
+likelihood ratio statistic requires finding the maximum likelihood estimator
+(MLE) for the denominator, which can be carried out using approximated
+likelihood ratios. Reusing the decomposed neural network-based approximation and
+minimizing the mean negative log-ratio of some observed data generated for
+$\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs
+over resampled observed data,
+compared to the distribution of the true MLEs. The figure clearly suggests that
+the proposed inference procedure yields in this case an unbiased estimator
+($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
+trials). Note that had the underlying ratio not been accurate enough because of
+poor training or poor calibration, the two distributions would not be exactly
+overlapping, indicating estimator bias.  Figure \ref{fig:2b} also confirms
+that the generalized likelihood ratio statistic obtained with our method is $\chi^2$ distributed,
+as should be the true  generalized likelihood ratio statistic.
 

### 2016-03-06 jasa/paper-jasa.tex
Accent

@@
           $^1$New York University\\
-          $^2$Federico Santa María University}
+          $^2$Federico Santa Mar\'ia University}
   \maketitle
@@
 The first fact to highlight is that, in the finite sample setting, uncalibrated
-direct approximations $\log \left( \frac{1 - \hat
-s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate estimates
-of the ratio, even when using universal approximators such as a neural network
-or a random forest. This observation is at the origin of this work, and led us
-to show that the ratio $r(\mathbf{x})$ can still be computed exactly by forming
-the equivalent statistic
+direct approximations $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat
+s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate
+estimates of the ratio, even when using universal approximators such as a neural
+network or a random forest. This observation is at the origin of this work, and
+led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
+forming the equivalent statistic
 $\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
@@
 true ratio than their direct counterparts, as illustrated in red. The
-approximated ratios show to be more accurate for all three classifiers, in particular in regions with enough density
-and where the capacity of the
-classifier is focused. By contrast, in regions where the densities are low
-(i.e., at the boundaries) the approximated ratio show to less acurate
-since the underlying classifier cannot make precise predictions, because
-of the few training samples it is trained on in these parts of the input space.
-When leveraging the fact that densities are mixtures and decomposing their
-ratio, the capacity of the underlying classifiers can be focused on easier
-supervised learning tasks, resulting as expected in even more accurate approximations, as
-shown in cyan.
+approximated ratios show to be more accurate for all three classifiers, even
+when the direct approximation was not very good in the first place, e.g. as for the
+linear model. In regions where the densities are low (i.e., at the
+boundaries) the approximated ratio show to be less acurate since the underlying
+classifier cannot make precise predictions, because of the few training samples
+it is trained on in these parts of the input space. When leveraging the fact
+that densities are mixtures and decomposing their ratio, the capacity of the
+underlying classifiers can be focused on easier supervised learning tasks,
+resulting as expected in even more accurate approximations, as shown in cyan.
 
@@
 
-    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat \gamma$.}
+    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat \gamma$, as empirically estimated 1000 times over resampled observed data.}
     \label{fig:2}
@@
 minimizing the mean negative log-ratio of some observed data generated for
-$\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs
-over resampled observed data,
+$\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
 compared to the distribution of the true MLEs. The figure clearly suggests that
@@
 poor training or poor calibration, the two distributions would not be exactly
-overlapping, indicating estimator bias.  Figure \ref{fig:2b} also confirms
+overlapping, indicating estimator bias. As a second check, Figure \ref{fig:2b} also confirms
 that the generalized likelihood ratio statistic obtained with our method is $\chi^2$ distributed,
-as should be the true  generalized likelihood ratio statistic.
+as should be the true statistic.
 

### 2016-03-06 jasa/paper-jasa.tex
A, B -> \alpha, \beta

@@
         \item $\mathbf{z} := (z_0, z_1, z_2, z_3, z_4)$, such that
-            $z_0 \sim {\cal N}(\mu=A, \sigma=1)$,
-            $z_1 \sim {\cal N}(\mu=B, \sigma=3)$,
+            $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
+            $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
             $z_2 \sim {\text{Mixture}}({\cal N}(\mu=-2, \sigma=1), {\cal N}(\mu=2, \sigma=0.5))$,
@@
 Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
-our goal is to infer the nominal values $A=1$ and $B=-1$ that
+our goal is to infer the nominal values $\alpha=1$ and $\beta=-1$ that
 were used for generating this data.
@@
     \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
-    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $(A=1, B=-1)$.}
+    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $(\alpha=1, \beta=-1)$.}
     \label{fig:4}
@@
     \caption{Inference from parameterized approximated likelihood ratios.
-             (\ref{fig:5a}) True mean negative log-ratio for the observed data ${\cal D}$.
-The red dot corresponds to the nominal values $(A=1, B=-1)$, while the green dot is the maximum likelihood estimator $(\hat A=1.006, \hat B=-1.012)$.
-             (\ref{fig:5b}) Estimated mean negative log-ratio, as approximated from the constructed objective by the Gaussian process within the Bayesian optimization procedure.
-The blue dot is the estimated maximum likelihood estimator $(\hat A=0.956, \hat B=-0.869)$ while the white dots correspond to the points sampled during optimization.}
+             (\ref{fig:5a}) True mean negative log-ratio for the observed data ${\cal D}$. The red dot corresponds
+             to the nominal values $(\alpha=1, \beta=-1)$, while the green dot is the maximum likelihood estimator
+             $(\hat \alpha=1.006, \hat \beta=-1.012)$.
+             (\ref{fig:5b}) Estimated mean negative log-ratio, as approximated from the constructed objective by
+             the Gaussian process within the Bayesian optimization procedure. The blue dot is the estimated maximum
+             likelihood estimator $(\hat \alpha=0.956, \hat \beta=-0.869)$ while the white dots correspond to
+             the points sampled during optimization.}
     \label{fig:5}
@@
 As previously, we make use of Eqn.~\ref{eq:mle_withs} for parameter inference
-with approximated likelihood ratios and fix the alternate to $\theta_1:=(A=0,
-B=0)$. Following recommendations from Section~\ref{sec:param-clf} to reduce
-computational requirements, we train a single parameterized classifier
-$s(\mathbf{x};\theta:=(A,B),\theta_1)$; in this case, a 2-layer neural network.
-Since the generative model is not expensive, the classifier output is calibrated on-the-fly
-with isotonic regression for every candidate parameter pair $\theta:=(A,B)$.
+with approximated likelihood ratios and fix the alternate to
+$\theta_1:=(\alpha=0, \beta=0)$. Following recommendations from
+Section~\ref{sec:param-clf} to reduce computational requirements, we train a
+single parameterized classifier $s(\mathbf{x};\theta:=(\alpha,\beta),\theta_1)$;
+in this case, a 2-layer neural network. Since the generative model is not
+expensive, the classifier output is calibrated on-the-fly with isotonic
+regression for every candidate parameter pair $\theta:=(\alpha,\beta)$.
 Minimizing the mean negative log of the constructed ratio with Bayesian
 optimization~\citep{brochu2010tutorial} eventually leads to the estimated
-posterior mean shown in Figure~\ref{fig:5b}, which is close to the true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations,
-the estimated minimizer is found at $(\hat A=0.956, \hat B=-0.869$).
-Again, repeated experiments (omitted here) confirm that
-the proposed procedure yields an unbiased maximum likelihood estimator.
+posterior mean shown in Figure~\ref{fig:5b}, which is close to the true
+objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
+estimated minimizer is found at $(\hat \alpha=0.956, \hat \beta=-0.869$). Again,
+repeated experiments (omitted here) confirm that the proposed procedure yields
+an unbiased maximum likelihood estimator.
 

### 2016-03-07 jasa/paper-jasa.tex
minor changes

@@
 \begin{align}
-p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - \gamma) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \gamma) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}), \\
+p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - \frac{\gamma}{2}) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \frac{\gamma}{2}) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}), \\
 p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
@@
 anticipate that the embedded density estimation algorithm should be carefully
-chosen and tuned, on a case by case basis.
+chosen and tuned, on a case by case basis. Overall, this first example confirms
+that even if the classifier does a poor joob a modeling the optimal decision
+function, good approximations of the likelihood ratio can still be obtained
+with proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
+the mixture when possible.
 
@@
 Let us now consider the composite hypothesis testing setting, by comparing
-$H_0: \gamma = 0$ versus the alternative $H_1: \gamma > 0$. That is, we want to evaluate
+$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
 \begin{equation}
-\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | \gamma = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)} \; .
+\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)} \; .
 \end{equation}
@@
 (MLE) for the denominator, which can be carried out using approximated
-likelihood ratios. Reusing the decomposed neural network-based approximation and
-minimizing the mean negative log-ratio of some observed data generated for
+likelihood ratios. Reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c} and
+minimizing the mean negative log-ratio of observed data generated for
 $\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
@@
 that the generalized likelihood ratio statistic obtained with our method is $\chi^2$ distributed,
-as should be the true statistic.
+as should be the true statistic when the null is true.
 

### 2016-03-08 jasa/paper-jasa.tex
Improve first example

--- 2016-03-06 jasa/paper-jasa.tex
@@
 As a simple and illustrative example, let us first consider the approximation of
-the likelihood log-ratio $\log \left( r(\mathbf{x};\theta_0,\theta_1) \right)$ between the 1D mixtures
-$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$ illustrated in
+the likelihood log-ratio $\log \left( r(\mathbf{x};\gamma=0.05,\gamma=0) \right)$ between the 1D mixtures
+$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$ illustrated in
 Figure~\ref{fig:1a} and defined as
 \begin{align}
-p(\mathbf{x}|\theta_0) &= (\frac{1}{2} - \gamma) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \gamma) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}), \\
-p(\mathbf{x}|\theta_1) &= \frac{1}{2} p_{c_0}(\mathbf{x}) + \frac{1}{2} p_{c_1}(\mathbf{x}),
+p(\mathbf{x}|\gamma) &= (\frac{1}{2} - \frac{\gamma}{2}) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \frac{\gamma}{2}) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}),
 \end{align}
 where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
-$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$ and where $\gamma$ is a parameter of $p(\mathbf{x}|\theta_0)$
-set to $0.05$. Samples drawn from $p(\mathbf{x}|\theta_0)$ are shown in the
+$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$. Samples drawn for the nominal value $\gamma=0.05$ are shown in the
 figure and will be used later for inference.
@@
 without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
-s(\mathbf{x}) | \theta_0)}{\hat p(\hat s(\mathbf{x}) | \theta_1)}$ as calibrated
+s(\mathbf{x}) | \gamma=0.05)}{\hat p(\hat s(\mathbf{x}) | \gamma=0)}$ as calibrated
 using histograms. Finally, the
@@
         \includegraphics[width=\textwidth]{figures/fig1a.pdf}
-        \caption{$p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$}
+        \caption{$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$}
         \label{fig:1a}
@@
     \caption{Approximation of the log-likelihood ratio
-             $\log r(\mathbf{x};\theta_0,\theta_1) = \log \frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$
+             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
              with calibrated classifiers.}
@@
 forming the equivalent statistic
-$\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
+$\frac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
 transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
-constructing $\frac{\hat p(\hat s(\mathbf{x})|\theta_0)}{\hat p(\hat
-s(\mathbf{x})|\theta_1)}$ as we propose, estimates appear much closer to the
+constructing $\frac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat
+s(\mathbf{x})|\gamma=0)}$ as we propose, estimates appear much closer to the
 true ratio than their direct counterparts, as illustrated in red. The
@@
 anticipate that the embedded density estimation algorithm should be carefully
-chosen and tuned, on a case by case basis.
+chosen and tuned, on a case by case basis. Overall, this first example confirms
+that even if the classifier does a poor joob at modeling the optimal decision
+function, good approximations of the likelihood ratio can still be obtained
+with the proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
+the mixture when possible.
 
@@
         \includegraphics[width=\textwidth]{figures/fig2b.pdf}
-        \caption{$-2 \log \frac{p({\cal D}|\gamma=0.05)}{\sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)}$}
+        \caption{$-2 \log \frac{p({\cal D}|\gamma=0.05)}{\sup_{\gamma \in \Theta} p({\cal D} | \gamma)}$}
         \label{fig:2b}
@@
 
-Let us now consider the composite hypothesis testing setting, by comparing
-$H_0: \gamma = 0$ versus the alternative $H_1: \gamma > 0$. That is, we want to evaluate
+Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
+$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
 \begin{equation}
-\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | \gamma = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)} \; .
-\end{equation}
-Following the approach outlined in
-Section~\ref{sec:generalized-likelihood-ratio}, computing the generalized
-likelihood ratio statistic requires finding the maximum likelihood estimator
-(MLE) for the denominator, which can be carried out using approximated
-likelihood ratios. Reusing the decomposed neural network-based approximation and
-minimizing the mean negative log-ratio of some observed data generated for
-$\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
-compared to the distribution of the true MLEs. The figure clearly suggests that
+\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
+\end{equation}
+By exploiting the fact that
+\begin{equation}
+\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
+\end{equation}
+the generalized likelihood ratio statistic can be computed by evaluating
+both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
+In particular, the maximum likelihood estimator (MLE) in the second term
+can be found as outlined in
+Section~\ref{sec:generalized-likelihood-ratio}.
+Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
+compared to the distribution of the true MLEs. It clearly suggests that
 the proposed inference procedure yields in this case an unbiased estimator
@@
 poor training or poor calibration, the two distributions would not be exactly
-overlapping, indicating estimator bias. As a second check, Figure \ref{fig:2b} also confirms
-that the generalized likelihood ratio statistic obtained with our method is $\chi^2$ distributed,
-as should be the true statistic.
+overlapping, indicating bias. Figure \ref{fig:2b} also confirms
+that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
+as should be the true statistic when the null is true.
 

### 2016-03-10 jasa/paper-jasa.tex

@@
 linear model. In regions where the densities are low (i.e., at the
-boundaries) the approximated ratio show to be less acurate since the underlying
+boundaries) the approximated ratio show to be less accurate since the underlying
 classifier cannot make precise predictions, because of the few training samples
@@
 chosen and tuned, on a case by case basis. Overall, this first example confirms
-that even if the classifier does a poor joob at modeling the optimal decision
+that even if the classifier does a poor job at modeling the optimal decision
 function, good approximations of the likelihood ratio can still be obtained
@@
 
-    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat \gamma$, as empirically estimated 1000 times over resampled observed data.}
+    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator
+            $\hat \gamma$, as empirically estimated 1000 times over re-sampled observed data.}
     \label{fig:2}
@@
 poor training or poor calibration, the two distributions would not be exactly
-overlapping, indicating bias. Figure \ref{fig:2b} also confirms
+overlapping, indicating bias. From this, the initial $\Lambda$ test statistic can finally
+be evaluated to determine in this case that the null should not be rejected.  Figure \ref{fig:2b} also confirms
 that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
@@
     \end{enumerate}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
+    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $(\alpha=1, \beta=-1)$. The size of ${\cal D}$ is set to $500$.}
+    \label{fig:4}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \begin{subfigure}[t]{0.31\textwidth}
+        \includegraphics[width=\textwidth]{figures/fig5a.pdf}
+        \caption{  }
+        \label{fig:5a}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[t]{0.31\textwidth}
+        \includegraphics[width=\textwidth]{figures/fig5b.pdf}
+        \caption{ }
+        \label{fig:5b}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[t]{0.31\textwidth}
+        \includegraphics[width=\textwidth]{figures/fig5c.pdf}
+        \caption{ }
+        \label{fig:5c}
+    \end{subfigure}
+    \caption{Inference from parameterized approximated likelihood ratios. The red dot corresponds
+    to the nominal values $(\alpha=1, \beta=-1)$, the green dot is the maximum likelihood estimator
+    $(\hat \alpha=0.896, \hat \beta=-0.984)$ while the blue dot is the estimated maximum likelihood estimator $(\hat \alpha=0.893, \hat \beta=-1.135)$.
+    1, 2 and 3-$\sigma$ contours are shown in white.
+             (\ref{fig:5a}) Exact values of the objective for the observed data ${\cal D}$.
+             (\ref{fig:5b}) Sampled values of the approximated objective for a $15 \times 15$ fixed grid.
+             (\ref{fig:5c}) Posterior mean of the approximated objective, as estimated by the Gaussian process within the Bayesian optimization procedure.
+White dots show the sampled candidates during the optimization process.}
+    \label{fig:5}
+\end{figure}
+
 Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
@@
 were used for generating this data.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
-    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $(\alpha=1, \beta=-1)$.}
-    \label{fig:4}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[height=12.5em]{figures/fig5a.pdf}
-        \caption{True mean negative log-ratio}
-        \label{fig:5a}
-    \end{subfigure}
-    ~
-    \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[height=12.5em]{figures/fig5b.pdf}
-        \caption{Estimated mean negative log-ratio}
-        \label{fig:5b}
-    \end{subfigure}
-    \caption{Inference from parameterized approximated likelihood ratios.
-             (\ref{fig:5a}) True mean negative log-ratio for the observed data ${\cal D}$. The red dot corresponds
-             to the nominal values $(\alpha=1, \beta=-1)$, while the green dot is the maximum likelihood estimator
-             $(\hat \alpha=1.006, \hat \beta=-1.012)$.
-             (\ref{fig:5b}) Estimated mean negative log-ratio, as approximated from the constructed objective by
-             the Gaussian process within the Bayesian optimization procedure. The blue dot is the estimated maximum
-             likelihood estimator $(\hat \alpha=0.956, \hat \beta=-0.869)$ while the white dots correspond to
-             the points sampled during optimization.}
-    \label{fig:5}
-\end{figure}
-
-As previously, we make use of Eqn.~\ref{eq:mle_withs} for parameter inference
-with approximated likelihood ratios and fix the alternate to
-$\theta_1:=(\alpha=0, \beta=0)$. Following recommendations from
-Section~\ref{sec:param-clf} to reduce computational requirements, we train a
-single parameterized classifier $s(\mathbf{x};\theta:=(\alpha,\beta),\theta_1)$;
-in this case, a 2-layer neural network. Since the generative model is not
-expensive, the classifier output is calibrated on-the-fly with isotonic
-regression for every candidate parameter pair $\theta:=(\alpha,\beta)$.
-Minimizing the mean negative log of the constructed ratio with Bayesian
-optimization~\citep{brochu2010tutorial} eventually leads to the estimated
-posterior mean shown in Figure~\ref{fig:5b}, which is close to the true
-objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
-estimated minimizer is found at $(\hat \alpha=0.956, \hat \beta=-0.869$). Again,
-repeated experiments (omitted here) confirm that the proposed procedure yields
-an unbiased maximum likelihood estimator.
+We use the objective function
+\begin{equation}\label{eqn:obj-ex2}
+    -2 \log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }
+\end{equation}
+that we approximate by exploiting the fact that
+\begin{equation}\label{eqn:obj-ex2-expanded}
+\log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) } = \log \frac{p({\cal D} | \alpha, \beta)}{ p({\cal D} | \alpha=0, \beta=0) } - \log \frac{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }{ p({\cal D} | \alpha=0, \beta=0)}.
+\end{equation}
+Following recommendations from Section~\ref{sec:param-clf} to reduce
+computational requirements, both terms in the right hand side of Eqn.~\ref{eqn:obj-ex2-expanded} are evaluated using a single
+parameterized approximated  likelihood ratio, in this case based on a parameterized
+2-layer neural network where the alternate is fixed to $\alpha=0, \beta=0$.
+Since the generative model is not expensive, the classifier output is calibrated
+on-the-fly with histograms for  every candidate parameter pair $(\alpha,\beta)$.
+Minimizing the constructed approximation of the objective with
+Bayesian optimization~\citep{brochu2010tutorial} eventually leads to the
+estimated posterior mean shown in Figure~\ref{fig:5c}, which is close to the
+true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
+estimated minimizer is found at $(\hat \alpha=0.893, \hat \beta=-1.135$). For
+comparison, Figure~\ref{fig:5b} shows the objective value as directly
+approximated by the proposed method for a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
+roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.
+As discussed in Section~\ref{sec:param-calibration}, parameterized calibration
+is a solution to this issue, but is left for now as an area for future work.
+Overall, this example further illustrates and confirms the ability
+of the proposed method for inference, even in the case of multi-dimensional data where
+direct approximations $p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$
+are often difficult to construct.
 

### 2016-03-10 jasa/paper-jasa.tex
text changes

@@
 In many fields of science, generalized likelihood ratio tests are established tools for
-statistical inference. In practice, these tests are often complicated by the fact
-that computer simulators are used as a generative model for the data, which does
-not provide a way to evaluate the likelihood function. In this paper, we
-demonstrate that likelihood ratios are invariant under dimensionality reductions $\mathbb{R}^p \mapsto \mathbb{R}$,
-provided the transformation is itself monotonic with the likelihood ratio. As a direct
-consequence, we show that discriminative classifiers can be used to
+statistical inference. 
+ At the same time, it has become increasingly common that a simulator (or
+generative model) is used to describe complex processes that tie parameters
+$\theta$ of an underlying theory and measurement apparatus to high-dimensional
+observations $\mathbf{x}\in \mathbb{R}^p$. However, simulator often do not provide a way to evaluate 
+the likelihood function for a given observation $\mathbf{x}$, which motivates a new class of likelihood-free
+inference algorithms. 
+In this paper, we show that likelihood ratios are invariant under a specific class of dimensionality reduction maps $\mathbb{R}^p \mapsto \mathbb{R}$.
+As a direct consequence, we show that discriminative classifiers can be used to
 approximate the generalized likelihood ratio statistic when only a generative model for the data is
-available. In particular, the proposed method offers a machine learning-based
-approach to statistical inference that is complementary to likelihood-free
-Bayesian inference algorithms, such as Approximate Bayesian Computation, as it does
-not require the definition of a prior over model parameters. Experimental results
-on artificial problems illustrate the potential of the proposed method.
+available.
+This leads to a new machine learning-based approach to likelihood-free inference that is complementary to Approximate Bayesian Computation, and which does not require a prior on the model parameters.
+Experimental results
+on artificial problems with known exact likelihoods illustrate the potential of the proposed method.
 
@@
 between a hypothesized process of interest (labeled \textit{signal}) and various
-other processes (labeled \textit{background}). Typically, data generated from the
+other processes (labeled \textit{background}). 
+Typically, data generated from the
 simulator are used to approximate the density at various parameter points, and
 an interpolation algorithm is used to approximate the parameterized
-model~\citep{Cranmer:2012sba}. In particular, to improve the statistical power
+model~\citep{Cranmer:2012sba}. In order to improve the statistical power
 of these tests, hundreds of these searches have already been using supervised
-learning to train discriminative classifiers that take advantage of a high
-dimensional feature vector $\mathbf{x}$.
+learning to train classifiers to discriminate between two two discrete hypotheses based
+on a high dimensional feature vector $\mathbf{x}$. The results of this paper outline
+how to extend the use of discriminative classifiers for composite hypotheses (parametrized by $\theta$) 
+in a way
+that fits naturally into the established likelihood based inference techniques.
 
@@
 statistical analysis within the field of high energy physics. Related work
-and conclusions are finally presented in sections~\ref{sec:related} and \ref{sec:conclusions}.
+and conclusions are finally presented in Sections~\ref{sec:related} and \ref{sec:conclusions}.
 
@@
     $s : \mathbb{R}^p \mapsto \mathbb{R}$ be a function monotonic with the density ratio
-    $r(\mathbf{x};\theta_0,\theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$,
+    $r(\mathbf{x};\theta_0,\theta_1)$,
     for given parameters $\theta_0$ and $\theta_1$. In these conditions,
@@
 Also, since $s(\mathbf{x})$ is monotonic with
-$\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$,
-it exists an invertible function $m:\mathbb{R}^+ \mapsto \mathbb{R}$ such
-that $s(\mathbf{x}) = m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$.
+$r(\mathbf{X}; \theta_0, \theta_1)$,
+there exists an invertible function $m:\mathbb{R}^+ \mapsto \mathbb{R}$ such
+that $s(\mathbf{x}) = m(r(\mathbf{X}; \theta_0, \theta_1))$.
 In particular, we have
@@
 \end{align}
-Combining equations \ref{eqn:hormander} and \ref{eqn:mapping}, the density ratio $\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$ can be pulled out of the integral, resulting in
+Combining equations \ref{eqn:hormander} and \ref{eqn:mapping}, the density ratio $r(\mathbf{X}; \theta_0, \theta_1)$ can be pulled out of the integral, resulting in
 \begin{align}
@@
 Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
-density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{1}{1 +
-r(\mathbf{x})}$ only.
+density ratio,  not only for $m(r(\mathbf{x})) = (1 +
+r(\mathbf{x}))^{-1}$.
 Equivalently,
@@
 For this reason, the proposed method is expected to be more robust than
-directly using $\frac{1 - s(\mathbf{x})}{s(\mathbf{x})}$ as an approximate of
+directly using ${(1 - s(\mathbf{x}))}/{s(\mathbf{x})}$ as an approximate of
 $r(\mathbf{x})$ (which indeed converges towards $r(x)$ when $s(x)$ tends to $s^*(x)$).
@@
 Thus far we have shown that the target likelihood ratio
-$r(\mathbf{x};\theta_0,\theta_1)=\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$
+$r(\mathbf{x};\theta_0,\theta_1)$
 with high dimensional features $\mathbf{x}$ can be reproduced via the univariate
@@
 \begin{algorithmic}
-    \State ${\cal L := \{ \}}$;
+    \State ${\cal T := \{ \}}$;
     \While{ $\text{size}({\cal L}) < N $ }
@@
 	    \State Draw $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$;
-		\State ${\cal L} := {\cal L} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=0) \}$;
+		\State ${\cal T} := {\cal T} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=0) \}$;
 		\State Draw $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$;
-		\State ${\cal L} := {\cal L} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=1) \}$;
+		\State ${\cal T} := {\cal T} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=1) \}$;
     \EndWhile
-    \State Learn a single classifier $s(\mathbf{x}; \theta_0, \theta_1)$ from ${\cal L}$.
+    \State Learn a single classifier $s(\mathbf{x}; \theta_0, \theta_1)$ from ${\cal T}$.
 \end{algorithmic}
@@
 As a simple and illustrative example, let us first consider the approximation of
-the likelihood log-ratio $\log \left( r(\mathbf{x};\gamma=0.05,\gamma=0) \right)$ between the 1D mixtures
+the log-likelihood ratio $\log \left( r(\mathbf{x};\gamma=0.05,\gamma=0) \right)$ between the 1D mixtures
 $p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$ illustrated in
@@
 \begin{align}
-p(\mathbf{x}|\gamma) &= (\frac{1}{2} - \frac{\gamma}{2}) p_{c_0}(\mathbf{x}) + (\frac{1}{2} - \frac{\gamma}{2}) p_{c_1}(\mathbf{x}) + \gamma p_{c_2}(\mathbf{x}),
+p(\mathbf{x}|\gamma) &= (1-\gamma)\frac{p_{c_0}(\mathbf{x}) +  p_{c_1}(\mathbf{x})}{2}   + \gamma \, p_{c_2}(\mathbf{x}),
 \end{align}
@@
     \end{subfigure}
-    \caption{Inference from parameterized approximated likelihood ratios. The red dot corresponds
-    to the nominal values $(\alpha=1, \beta=-1)$, the green dot is the maximum likelihood estimator
-    $(\hat \alpha=0.896, \hat \beta=-0.984)$ while the blue dot is the estimated maximum likelihood estimator $(\hat \alpha=0.893, \hat \beta=-1.135)$.
+    \caption{Inference from exact and approximate likelihood ratios. The red dot corresponds
+    to the true values $(\alpha=1, \beta=-1)$ used to generate $\mathcal{D}$, the green dot is the MLE
+    $(\hat \alpha=0.896, \hat \beta=-0.984)$ from the exact likelihood, while the blue dot is the MLE $(\hat \alpha=0.893, \hat \beta=-1.135)$ from the approximate likelihood.
     1, 2 and 3-$\sigma$ contours are shown in white.
-             (\ref{fig:5a}) Exact values of the objective for the observed data ${\cal D}$.
-             (\ref{fig:5b}) Sampled values of the approximated objective for a $15 \times 15$ fixed grid.
-             (\ref{fig:5c}) Posterior mean of the approximated objective, as estimated by the Gaussian process within the Bayesian optimization procedure.
-White dots show the sampled candidates during the optimization process.}
+             (\ref{fig:5a}) The exact log-likelihood for the observed data ${\cal D}$.
+             (\ref{fig:5b}) The approximate log-likelihood  evaluated on a coarse $15 \times 15$ fixed grid.
+             (\ref{fig:5c}) A Gaussian Process surrogate of the approximate log-likelihood ratio estimated from a Bayesian optimization procedure.
+White dots show the  parameter points sampled during the optimization process.}
     \label{fig:5}

### 2016-03-11 jasa/paper-jasa.tex
update minimum for new figures

@@
 \end{align}
-where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.5625)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
+where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.25^2)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
 $p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$. Samples drawn for the nominal value $\gamma=0.05$ are shown in the
@@
     \end{subfigure}
+
     \caption{Approximation of the log-likelihood ratio
@@
 forming the equivalent statistic
-$\frac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
+$\sfrac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
 transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
-constructing $\frac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat
+constructing $\sfrac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat
 s(\mathbf{x})|\gamma=0)}$ as we propose, estimates appear much closer to the
@@
     to the true values $(\alpha=1, \beta=-1)$ used to generate $\mathcal{D}$, the green dot is the MLE
-    $(\hat \alpha=0.896, \hat \beta=-0.984)$ from the exact likelihood, while the blue dot is the MLE $(\hat \alpha=0.893, \hat \beta=-1.135)$ from the approximate likelihood.
+    $(\hat \alpha=1.012, \hat \beta=-0.9221)$  from the exact likelihood, while the blue dot is the MLE $(\hat \alpha=1.008, \hat \beta=-1.004)$ from the approximate likelihood.
     1, 2 and 3-$\sigma$ contours are shown in white.
@@
 true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
-estimated minimizer is found at $(\hat \alpha=0.893, \hat \beta=-1.135$). For
+estimated minimizer is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$. For
 comparison, Figure~\ref{fig:5b} shows the objective value as directly
@@
 regression function
-$\frac{p_s(\mathbf{x}|\nu_0)}{p_s(\mathbf{x}|\nu_0)+p_b(\mathbf{x}|\nu_0)}$, which is
+$\sfrac{p_s(\mathbf{x}|\nu_0)}{p_s(\mathbf{x}|\nu_0)+p_b(\mathbf{x}|\nu_0)}$, which is
 one to one with the likelihood ratio of the null to the alternate
@@
 under covariate shift, \cite{shimodaira2000improving} and \cite{sugiyama2005input} estimate
-the density ratio $r(\mathbf{x};\theta_0,\theta_1) =
-\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ from straightforward
+the density ratio $r(\mathbf{x};\theta_0,\theta_1)$ from straightforward
 approximations $\hat{p}(\mathbf{x}|\theta_0)$ and $\hat{p}(\mathbf{x}|\theta_1)$

### 2016-03-13 jasa/paper-jasa.tex
new figures
minor text changes

@@
 \begin{equation}\label{eqn:generalized-lr}
-\Lambda({\cal D}; \Theta_0, \Theta) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
-\end{equation}
+\Lambda(\Theta_0) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
+\end{equation}
+%\begin{equation}\label{eqn:generalized-lr}
+%\Lambda({\cal D}; \Theta_0, \Theta) =  \frac{ \sup_{\theta \in \Theta_0} p({\cal D} | \theta)}{ \sup_{\theta \in \Theta} p({\cal D} | \theta)} \; .
+%\end{equation}
 This generalized likelihood ratio can be used both for hypothesis tests in the
@@
       %(or a blank line to force the subfigure onto a new line)
+%    \begin{subfigure}[b]{0.45\textwidth}
+%        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
+%        \caption{Linear model}
+%        \label{fig:1b}
+%    \end{subfigure}
     \begin{subfigure}[b]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
-        \caption{Linear model}
-        \label{fig:1b}
+        \includegraphics[width=\textwidth]{figures/p_sij.png}
+        \caption{Pairwise classifier score distributions}
+        \label{fig:p_sij}
     \end{subfigure}
@@
         \label{fig:1d}
+    \end{subfigure}
+
+%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
+%      %(or a blank line to force the subfigure onto a new line)
+%    \begin{subfigure}[b]{0.3\textwidth}
+%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
+%        \caption{Likelihood ratio distributions}
+%        \label{fig:decomposed_ratio}
+%    \end{subfigure}
+    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
+      %(or a blank line to force the subfigure onto a new line)
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
+        \caption{Likelihood ratio distributions}
+        \label{fig:decomposed_ratio}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{figures/likelihood.png}
+        \caption{log-likelihood curve }
+        \label{fig:likelihood}
     \end{subfigure}
@@
 the mixture when possible.
+%
+%
+%\begin{figure}
+%    \centering
+%    \begin{subfigure}[b]{0.3\textwidth}
+%        \includegraphics[width=\textwidth]{figures/p_sij.png}
+%        \caption{Pairwise classifier score distributions}
+%        \label{fig:p_sij}
+%    \end{subfigure}
+%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
+%      %(or a blank line to force the subfigure onto a new line)
+%    \begin{subfigure}[b]{0.3\textwidth}
+%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
+%        \caption{Likelihood ratio distributions}
+%        \label{fig:decomposed_ratio}
+%    \end{subfigure}
+%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
+%      %(or a blank line to force the subfigure onto a new line)
+%    \begin{subfigure}[b]{0.3\textwidth}
+%        \includegraphics[width=\textwidth]{figures/likelihood.png}
+%        \caption{log-likelihood curve }
+%        \label{fig:likelihood}
+%    \end{subfigure}
+%
+%    \caption{Approximation of the log-likelihood ratio
+%             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
+%             with calibrated classifiers.}
+%\end{figure}
+%
+%%%%%%%%%%%%%
 
@@
 \begin{equation}
-\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
-\end{equation}
+\Lambda( \gamma = 0.05) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
+\end{equation}
+%\begin{equation}
+%\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
+%\end{equation}
 By exploiting the fact that
@@
 \begin{equation}\label{eqn:obj-ex2}
-    -2 \log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }
-\end{equation}
-that we approximate by exploiting the fact that
+    -2 \log \Lambda(\alpha, \beta ) = -2 \log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }
+\end{equation}
+that we calculate by exploiting the fact that
 \begin{equation}\label{eqn:obj-ex2-expanded}

### 2016-03-13 jasa/paper-jasa.tex
minor text changes

@@
 the direct approximation $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$
-without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
-s(\mathbf{x}) | \gamma=0.05)}{\hat p(\hat s(\mathbf{x}) | \gamma=0)}$ as calibrated
+without calibration, while the red curve is the approximation $\log \Lambda( \gamma=0.05)$ as calibrated
 using histograms. Finally, the
@@
     \centering
-    \begin{subfigure}[b]{0.45\textwidth}
+    \begin{subfigure}[b]{0.4\textwidth}
         \includegraphics[width=\textwidth]{figures/fig1a.pdf}
-        \caption{$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$}
+        \caption{$p(\mathbf{x}|\gamma)$  for $\gamma=0.05$ and $\gamma=0$}
         \label{fig:1a}
@@
 %    \end{subfigure}
-    \begin{subfigure}[b]{0.45\textwidth}
+    \begin{subfigure}[b]{0.4\textwidth}
         \includegraphics[width=\textwidth]{figures/p_sij.png}
-        \caption{Pairwise classifier score distributions}
+        \caption{$p(s_{ij}(\mathbf{x}) |\gamma)$ for $\gamma=0.05$ and $\gamma=0$}
         \label{fig:p_sij}
@@
 
-    \begin{subfigure}[b]{0.45\textwidth}
+    \begin{subfigure}[b]{0.4\textwidth}
         \includegraphics[width=\textwidth]{figures/fig1c.pdf}
@@
     ~
-    \begin{subfigure}[b]{0.45\textwidth}
+    \begin{subfigure}[b]{0.4\textwidth}
         \includegraphics[width=\textwidth]{figures/fig1d.pdf}
@@
       %(or a blank line to force the subfigure onto a new line)
-    \begin{subfigure}[b]{0.45\textwidth}
+    \begin{subfigure}[b]{0.4\textwidth}
         \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
-        \caption{Likelihood ratio distributions}
+        \caption{$p( \log r(s(\mathbf{x})) \, | \,\gamma = 0.05)$ }
         \label{fig:decomposed_ratio}
@@
     ~
-    \begin{subfigure}[b]{0.45\textwidth}
+    \begin{subfigure}[b]{0.4\textwidth}
         \includegraphics[width=\textwidth]{figures/likelihood.png}
@@
              $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
-             with calibrated classifiers.}
+             with calibrated classifiers. Words Words Words Words Words Words Words Words Words Words Words Words Words Words}
 \end{figure}

### 2016-03-13 jasa/paper-jasa.tex
making progress

@@
 
-\subsection{Learning and calibrating $s$}
+\subsection{Learning and calibrating $s$}\label{s:learning_calibration}
 
@@
 In such cases where $\hat s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$,
-the induced probability does not factorize as in Eqn.~\ref{eqn:factorization}, resulting in a biased
-approximation of the density ratio.
+the induced probability does not factorize as in Eqn.~\ref{eqn:factorization}, 
+leading to artifacts in the resulting approximation of the density ratio.
 Provided enough training data, accurate results can however still be achieved,
@@
 density, but an approximation $\hat p(\hat s(\mathbf{x})| \theta)$ can be built
-instead, e.g. using density estimation or calibration algorithms, such as
-histograms, KDE or isotonic regression, and as applied on $\{\hat
-s(\mathbf{x}_i)\}$, for samples $\{\mathbf{x}_i\}$  drawn from the generative
+instead, e.g. using density estimation algorithms, such as
+histograms or kernel density estimation applied to $\{\hat
+s(\mathbf{x}_i)\}$, where the $\{\mathbf{x}_i\}$  are drawn from the generative
 model. Most notably, learning such an approximation of $p(s(\mathbf{x})|\theta)$
@@
 
+An alternative approach for calibration is to approximate the density ratio 
+$r(\hat{s}(\mathbf{x}))$ directly. 
+%There are various strategies. 
+For instance, isotonic regression, which is commonly used to transform the classifier score  $\hat{s}(\mathbf{x})$ into $\hat{s}_\textrm{iso}(\mathbf{x})$ that more accurately reflect  the posterior probability $s^*(\mathbf{x})$
+of Eq.~\ref{eqn:best-s-clf}, can be used for calibration.
+This is done by inverting the relationship ${r}(\mathbf{x}) = (1-s^*(\mathbf{x}))/s^*(\mathbf{x})$ to obtain $\hat{r}(\mathbf{x}) = (1-\hat{s}_\textrm{iso}(\mathbf{x}))/\hat{s}_\textrm{iso}(\mathbf{x})$.
+\textbf{Gilles, should we add more about direct density ratio estimation (currently in 5. Related Work) here?}
+
+%An alternative approach for calibration is to approximate the density ratio 
+%$r(\hat{s}(\mathbf{x}); \theta_0, \theta_1)$ directly. 
+%%There are various strategies. 
+%For instance, isotonic regression, which is commonly used to transform the classifier score  $\hat{s}(\mathbf{x})$ into $\hat{s}_\textrm{iso}(\mathbf{x})$ that more accurately reflect  the posterior probability $s^*(\mathbf{x})$
+%of Eq.~\ref{eqn:best-s-clf}, can be used for calibration.
+This is done by inverting the relationship ${r}(\mathbf{x}; \theta_0, \theta_1) = (1-s^*(\mathbf{x}))/s^*(\mathbf{x})$ to obtain $\hat{r}(\mathbf{x}; \theta_0, \theta_1) = (1-\hat{s}_\textrm{iso}(\mathbf{x}))/\hat{s}_\textrm{iso}(\mathbf{x})$.
+%\textbf{Gilles, should we add more about direct density ratio estimation (currently in 5. Related Work) here?}
+
+%. , which for balanced training samples has the form 
+%$\hat{p}(\theta_1 | \hat{s}(\mathbf{x})) \approx p(\hat{s}(\mathbf{x}) | \theta_0) / (p(\hat{s}(\mathbf{x}) | \theta_1) + p(\hat{s}(\mathbf{x}) | \theta_1)) = 1/(1+r(\hat{s}(\mathbf{x}); \theta_0, \theta_1)))$.
+
 One strength of the proposed approach is that it factorizes the approximation of
-the per-sample reduction ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from the
+the dimensionality reduction ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from the
 calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx
-p(\hat{s}(\mathbf{x})|\theta)$). Thus, even if the classifier does a poor job at
+p(\hat{s}(\mathbf{x})|\theta)$ or $\hat{r}(\hat{s}(\mathbf{x})) \approx r(\hat{s}(\mathbf{x}))$). Thus, even if the classifier does a poor job at
 learning the optimal decision function~\ref{eqn:best-s-clf} and therefore at
@@
 function can be expected for data generated from  $\theta'_0$ and $\theta'_1$
-never jointly encountered at learning. Similarly, it is not clear how the
+never jointly encountered during learning. Similarly, it is not clear how the
 limited capacity of the classifier may impact the performance the resulting
@@
 
-Once the classifier is trained, we can use the generative model together with a
-univariate density estimation technique (e.g. histograms or kernel density
-estimation) to approximate $p(s|\theta)$ for specific parameter
-points. For a single parameter point $\theta$, this is a tractable univariate density
+Once the parametrized classifier $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$ is trained, 
+we can use the generative model together with one of the calibration strategies 
+discussed in Sec.~\ref{s:learning_calibration} for particular values of $\theta_0$ and $\theta_1$. 
+%For instance, 
+%together with a
+%univariate density estimation technique (e.g. histograms or kernel density
+%estimation) to approximate $p(s|\theta)$ for specific parameter
+%points. 
+For a single parameter point $\theta$, this is a tractable univariate density
 estimation problem. The challenge comes from the need to calibrate this density
@@
 trying to maximize the likelihood with respect to $\theta$ needs access to the
-generative model $p(\mathbf{x}|\theta)$. This can be  impractical when the generative
+generative model $p(\mathbf{x}|\theta)$.  This is the strategy used for the examples presented in Sec.~\ref{sec:examples}.
+
+Calibrating the density on-demand can be  impractical when the generative
 model is computationally expensive or has high-latency (for instance some human
-intervention is required to reconfigure the generative model).
-
-In high energy physics, with a
-fixed classifier, it has become common  to interpolate the distribution between
-discrete values of $\theta$ in order to produce a continuous parameterization for
+intervention is required to reconfigure the generative model). In high energy physics, where 
+it is common to calibrate the distribution of a fixed classifier. There the strategy is to interpolate 
+the distribution between discrete values of $\theta$ in order to produce a continuous parameterization for
 $p(s | \theta)$~\citep{read1999linear,Cranmer:2012sba,baak2015interpolation}.
-One can easily imagine a number of approaches to embedding the classifier and
-estimating the density $p(s|\theta)$ and the relative merits of those
-approaches will depend critically on the dimensionality of $\theta$ and the
+One can easily imagine a number of approaches to parametrized calibration and the 
+relative merits of those approaches will depend critically on the dimensionality of $\theta$ and the
 computational cost of the generative model. We leave a more general strategy for
@@
 the target likelihood ratio can be formulated in terms of pairwise
-classification problems. We indeed have
+classification problems. Specifically, we can write
 \begin{align}
@@
 In applications where mixture models are commonly used, this decomposition allows
-one to construct better likelihood ratio estimates since it forces classifiers
+one to construct better likelihood ratio estimates since it allows the classifiers
 $s_{c,c'}$ to focus on simpler sub-problems, for which higher accuracy is
@@
 $p_{c}(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta)$ are independent of
-$\theta$. For this reason, sub-ratios $\frac{
+$\theta$. For this reason, sub-ratios $r_{c, c'}(\mathbf{x}; \theta_0, \theta_1) = \frac{
 p_{c'}(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)|\theta_1)}{
 p_c(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)|\theta_0)}$ simplify to $\frac{
-p_{c'}(s_{c,c'}(\mathbf{x}))}{ p_c(s_{c,c'}(\mathbf{x}))}$, and can all be
-pre-computed without having to resort to parameterized classification and
+p_{c'}(s_{c,c'}(\mathbf{x}))}{ p_c(s_{c,c'}(\mathbf{x}))}$, which can be
+pre-computed without the need of parameterized classification or
 calibration.
@@
 In this section, we illustrate the proposed method on two representative
-examples and then discuss its application to high energy physics. For reproducibility and extended details, the source
-code is available online\footnote{\url{https://github.com/diana-hep/carl}}.
+examples where the exact likelihood is known and then discuss its application to high energy physics.
+The code used to produce the results and extended details for these examples is available in Ref.~\citep{carl}.%\footnote{\url{https://github.com/diana-hep/carl}}.
 
@@
 the log-likelihood ratio $\log \left( r(\mathbf{x};\gamma=0.05,\gamma=0) \right)$ between the 1D mixtures
-$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$ illustrated in
-Figure~\ref{fig:1a} and defined as
+$p(\mathbf{x}|\gamma=0.05)$ and $p(\mathbf{x}|\gamma=0)$ defined as
 \begin{align}
@@
 where $p_{c_0} := {\cal N}(\mu=-2, \sigma^2=0.25^2)$, $p_{c_1} := {\cal N}(\mu=0, \sigma^2=4)$,
-$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$. Samples drawn for the nominal value $\gamma=0.05$ are shown in the
-figure and will be used later for inference.
-
-Using the method outlined in Section~\ref{sec:approx},
-figures~\ref{fig:1b}, \ref{fig:1c} and \ref{fig:1d} show the approximated
-ratio when using respectively a linear model, a 2-layer neural network and a random
-forest. The blue curve corresponds to the true target log-ratio, the green curve is
-the direct approximation $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$
-without calibration, while the red curve is the approximation $\log \Lambda( \gamma=0.05)$ as calibrated
+$p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$. Samples drawn for the nominal value $\gamma=0.05$ are shown in
+Figure~\ref{fig:1a} and used later for inference.
+
+Figure~\ref{fig:p_sij} shows the intermediate stages for the decomposition described in Section~\ref{sec:mixture}.
+The blue and green curves show $p_{c'}(\hat{s}_{c,c'}(\mathbf{x}))$,  the distributions for the score for the sub-classifiers for the three pair-wise comparisons of the mixture components.
+The red curves in Fig.~\ref{fig:p_sij} show the approximation of the density ratio obtained from those distributions.
+
+%Using the method outlined in Section~\ref{sec:approx},
+
+Figures~\ref{fig:1c} and \ref{fig:1d} show the approximate $\log \left( \hat{r}(\mathbf{x}) \right)$ as a function of $\mathbf{x}$ using a 
+2-layer neural network and a random forest for the classifier $\hat{s}(\mathbf{x})$.
+%ratio when using respectively a linear model, a 2-layer neural network and a random
+%forest. 
+The blue curves corresponds to the exact log-ratio. The green curve shows 
+$\log \left( (1-\hat s(\mathbf{x}))/\hat s(\mathbf{x}) \right)$
+without calibration. The red curve is the approximation $\log \hat{r}(\mathbf{x})$ calibrated
 using histograms. Finally, the
-cyan curve shows the approximated ratio when both decomposing the mixtures and calibrating the underlying classifiers, as
-described in Section~\ref{sec:mixture}.
+cyan curve shows the approximated log-likelihood ratio when decomposing the mixture, as
+described in Section~\ref{sec:mixture},  and calibrating the underlying classifiers with histograms.
 
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/p_sij.png}
-        \caption{$p(s_{ij}(\mathbf{x}) |\gamma)$ for $\gamma=0.05$ and $\gamma=0$}
+        \includegraphics[width=\textwidth]{figures/fig-p_sij.pdf}
+        \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $1/(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))$}
         \label{fig:p_sij}
@@
         \includegraphics[width=\textwidth]{figures/fig1c.pdf}
-        \caption{Neural network}
+        \caption{$\log \hat r(\mathbf{x})$ using neural network}
         \label{fig:1c}
@@
         \includegraphics[width=\textwidth]{figures/fig1d.pdf}
-        \caption{Random forest}
+        \caption{$\log \hat r(\mathbf{x})$ using random forest}
         \label{fig:1d}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
-        \caption{$p( \log r(s(\mathbf{x})) \, | \,\gamma = 0.05)$ }
+        \includegraphics[width=\textwidth]{figures/fig-ratio-dist.pdf}
+        \caption{$p( \log \hat r(\mathbf{x})) \, | \,\gamma = 0.05)$ }
         \label{fig:decomposed_ratio}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/likelihood.png}
+        \includegraphics[width=\textwidth]{figures/fig-likelihood.pdf}
         \caption{log-likelihood curve }
@@
     \caption{Approximation of the log-likelihood ratio
-             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
-             with calibrated classifiers. Words Words Words Words Words Words Words Words Words Words Words Words Words Words}
+             $\log r(\mathbf{x};\gamma=0.05,\gamma=0)$
+             with calibrated classifiers. }
 \end{figure}
@@
 The first fact to highlight is that, in the finite sample setting, uncalibrated
-direct approximations $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat
-s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate
+direct approximations $\log \left( \hat{r}(\mathbf{x}) \right)$, as shown in green, may not always yield very accurate
 estimates of the ratio, even when using universal approximators such as a neural
@@
 s(\mathbf{x})|\gamma=0)}$ as we propose, estimates appear much closer to the
-true ratio than their direct counterparts, as illustrated in red. The
-approximated ratios show to be more accurate for all three classifiers, even
-when the direct approximation was not very good in the first place, e.g. as for the
-linear model. In regions where the densities are low (i.e., at the
-boundaries) the approximated ratio show to be less accurate since the underlying
-classifier cannot make precise predictions, because of the few training samples
-it is trained on in these parts of the input space. When leveraging the fact
+true ratio than their direct counterparts, as illustrated in red. 
+%The
+%approximated ratios show to be more accurate for all three classifiers, even
+%when the direct approximation was not very good in the first place, e.g. as for the
+%linear model. 
+In regions where the densities are low (i.e., at the
+boundaries) the approximated ratio is less accurate since the underlying
+classifier cannot make precise predictions due to the lack training samples
+ in these parts of the input space. When leveraging the fact
 that densities are mixtures and decomposing their ratio, the capacity of the
-underlying classifiers can be focused on easier supervised learning tasks,
+underlying classifiers can be more effectively focused on easier supervised learning tasks,
 resulting as expected in even more accurate approximations, as shown in cyan.
@@
 random forest still approximates the ratio correctly, but more variance can be
-observed. Finally, the linear model produces a decision function that is
-smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
-the ratio around $\mathbf{x}=1$, this still results in an accurate
-approximation, but poorer results should be expected on more complex problems.
+observed. 
+%Finally, the linear model produces a decision function that is
+%smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
+%the ratio around $\mathbf{x}=1$, this still results in an accurate
+%approximation, but poorer results should be expected on more complex problems.
 While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat

### 2016-03-13 jasa/paper-jasa.tex
removing redundancies

@@
 2-layer neural network and a random forest for the classifier $\hat{s}(\mathbf{x})$.
+The neural network provides a smoother approximation, while the random has some artifacts due to the fact that the
+decision function is piece-wise constant.
 %ratio when using respectively a linear model, a 2-layer neural network and a random
 %forest. 
-The blue curves corresponds to the exact log-ratio. The green curve shows 
+The blue curves show the exact log-ratio, the green curves show 
 $\log \left( (1-\hat s(\mathbf{x}))/\hat s(\mathbf{x}) \right)$
-without calibration. The red curve is the approximation $\log \hat{r}(\mathbf{x})$ calibrated
-using histograms. Finally, the
-cyan curve shows the approximated log-likelihood ratio when decomposing the mixture, as
-described in Section~\ref{sec:mixture},  and calibrating the underlying classifiers with histograms.
+without calibration, while the red curve is the improved approximation $\log \hat{r}(\mathbf{x})$ calibrated
+using histograms. Finally, the cyan curve shows the approximated log-likelihood ratio when decomposing the mixture, as
+seen in Fig.~\ref{fig:p_sij}.  By leveraging the fact that densities are mixtures, the capacity 
+of the underlying classifiers can be more effectively focused on easier classification tasks,
+resulting as expected in even more accurate approximations.
 
@@
 
-The first fact to highlight is that, in the finite sample setting, uncalibrated
-direct approximations $\log \left( \hat{r}(\mathbf{x}) \right)$, as shown in green, may not always yield very accurate
-estimates of the ratio, even when using universal approximators such as a neural
-network or a random forest. This observation is at the origin of this work, and
-led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
-forming the equivalent statistic
-$\sfrac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
-transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
-constructing $\sfrac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat
-s(\mathbf{x})|\gamma=0)}$ as we propose, estimates appear much closer to the
-true ratio than their direct counterparts, as illustrated in red. 
-%The
-%approximated ratios show to be more accurate for all three classifiers, even
-%when the direct approximation was not very good in the first place, e.g. as for the
-%linear model. 
-In regions where the densities are low (i.e., at the
-boundaries) the approximated ratio is less accurate since the underlying
-classifier cannot make precise predictions due to the lack training samples
- in these parts of the input space. When leveraging the fact
-that densities are mixtures and decomposing their ratio, the capacity of the
-underlying classifiers can be more effectively focused on easier supervised learning tasks,
-resulting as expected in even more accurate approximations, as shown in cyan.
-
-As shown in the figures, qualitatively different results are obtained depending
-the classification algorithm embedded within the approximation of the ratio. The
-neural network-based approximation appears to work best, due to the capacity of
-the classifier to model a smooth decision function $s(\mathbf{x})$ which is
-very close to the optimal decision function. The random forest yields a decision function
-which is piece-wise constant, therefore non-continuous and not strictly
-monotonic with the ratio in the finite sample regime. On this problem, the
-random forest still approximates the ratio correctly, but more variance can be
-observed. 
-%Finally, the linear model produces a decision function that is
-%smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
-%the ratio around $\mathbf{x}=1$, this still results in an accurate
-%approximation, but poorer results should be expected on more complex problems.
+%The first feature to highlight is that, in the finite sample setting, uncalibrated
+%direct approximations $\log \left(  (1-\hat s(\mathbf{x}))/\hat s(\mathbf{x})  \right)$, as shown in green, may not yield a very accurate
+%estimate of the ratio, even when using universal approximators such as a neural
+%network or a random forest. 
+%%This observation is at the origin of this work, and
+%%led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
+%%forming the equivalent statistic
+%%$\sfrac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
+%%transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. 
+%The approximation of the log-likelihood ratio is significantly improved when calibrating
+%the classifier score with univariate density estimation
+%% according to 
+%%$\hat{r}(\mathbf{x}) = \sfrac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat s(\mathbf{x})|\gamma=0)}$, 
+%as illustrated in red. 
+%%The
+%%approximated ratios show to be more accurate for all three classifiers, even
+%%when the direct approximation was not very good in the first place, e.g. as for the
+%%linear model. 
+%In regions where the densities are low (i.e., near the
+%boundaries) the approximated ratio is less accurate.
+%% since the underlying
+%%classifier cannot make precise predictions due to the lack training samples
+%% in these parts of the input space. 
+% When leveraging the fact
+%that densities are mixtures and decomposing their ratio as described in Section~\ref{sec:mixture}, 
+%the capacity of the underlying classifiers can be more effectively focused on easier supervised learning tasks,
+%resulting as expected in even more accurate approximations, as shown in cyan.
+
+%As shown in the figures, qualitatively different results are obtained depending
+%the classification algorithm embedded within the approximation of the ratio. The
+%neural network-based approximation appears to work best, due to the capacity of
+%the classifier to model a smooth decision function $s(\mathbf{x})$ which is
+%very close to the optimal decision function. The random forest yields a decision function
+%which is piece-wise constant, therefore non-continuous and not strictly
+%monotonic with the ratio in the finite sample regime. On this problem, the
+%random forest still approximates the ratio correctly, but more variance can be
+%observed. 
+%%Finally, the linear model produces a decision function that is
+%%smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
+%%the ratio around $\mathbf{x}=1$, this still results in an accurate
+%%approximation, but poorer results should be expected on more complex problems.
+
 While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat

### 2016-03-13 jasa/paper-jasa.tex
cleaning up example 1 text

@@
         \includegraphics[width=\textwidth]{figures/fig-likelihood.pdf}
-        \caption{log-likelihood curve }
+        \caption{$-\log \Lambda(\gamma)$ }
         \label{fig:likelihood}
@@
 
-While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
-s(\mathbf{x}))$ is key to obtain accurate results. Standard
-histograms with fixed bins have been used here for illustrative purposes, but we
-anticipate that the embedded density estimation algorithm should be carefully
-chosen and tuned, on a case by case basis. Overall, this first example confirms
-that even if the classifier does a poor job at modeling the optimal decision
-function, good approximations of the likelihood ratio can still be obtained
-with the proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
-the mixture when possible.
-%
-%
-%\begin{figure}
-%    \centering
-%    \begin{subfigure}[b]{0.3\textwidth}
-%        \includegraphics[width=\textwidth]{figures/p_sij.png}
-%        \caption{Pairwise classifier score distributions}
-%        \label{fig:p_sij}
-%    \end{subfigure}
-%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
-%      %(or a blank line to force the subfigure onto a new line)
-%    \begin{subfigure}[b]{0.3\textwidth}
-%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
-%        \caption{Likelihood ratio distributions}
-%        \label{fig:decomposed_ratio}
-%    \end{subfigure}
-%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
-%      %(or a blank line to force the subfigure onto a new line)
-%    \begin{subfigure}[b]{0.3\textwidth}
-%        \includegraphics[width=\textwidth]{figures/likelihood.png}
-%        \caption{log-likelihood curve }
-%        \label{fig:likelihood}
-%    \end{subfigure}
-%
-%    \caption{Approximation of the log-likelihood ratio
-%             $\log r(\mathbf{x};\gamma=0.05,\gamma=0) = \log \frac{p(\mathbf{x}|\gamma=0.05)}{p(\mathbf{x}|\gamma=0)}$
-%             with calibrated classifiers.}
-%\end{figure}
-%
-%%%%%%%%%%%%%
+As the results show, calibrating $\hat s(\mathbf{x})$ through univariate density estimation of 
+$\hat p(\hat s(\mathbf{x}))$ is key to obtaining accurate results. Standard
+histograms with uniform binning have been used here for illustrative purposes, but we
+anticipate that more sophisticated calibration strategies will be important in further development
+of this method. We leave this as an area for future study.
+
+Figure~\ref{fig:decomposed_ratio} shows the distribution of $ \log \hat r(\mathbf{x}))$ for $\gamma=0.05$
+using the decomposed approximation of $\hat r(\mathbf{x}))$ with neural networks and the distribution
+of the exact log-likelihood ratio.
+While there are some artifacts in the distribution in the low-probability regions and the maximum value of the log-likehood ratio is underestimated, the overall shape of the distribution is well approximated.
+
+Finally, we see the log-likelihood curve $-\log \Lambda(\gamma)$ for the dataset $\mathcal{D}$ shown in 
+Fig.~\ref{fig:1a}. The exact likelihood curve is very well approximated by the method, confirming that
+even when the raw classifier does a poor job at modeling the $s^*(\mathbf{x}$, a good approximations of the 
+likelihood ratio can still be obtained by calibrating $s(\mathbf{x})$ (and by decomposing
+the mixture, if possible).
+
+Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
+$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
+\begin{equation}
+\Lambda( \gamma = 0.05) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
+\end{equation}
+%\begin{equation}
+%\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
+%\end{equation}
+By exploiting the fact that
+\begin{equation}
+\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
+\end{equation}
+the generalized likelihood ratio statistic can be computed by evaluating
+both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
+In particular, the maximum likelihood estimator (MLE) in the second term
+can be found as outlined in
+Section~\ref{sec:generalized-likelihood-ratio}.
 
@@
 
-Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
-$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
-\begin{equation}
-\Lambda( \gamma = 0.05) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
-\end{equation}
-%\begin{equation}
-%\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
-%\end{equation}
-By exploiting the fact that
-\begin{equation}
-\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
-\end{equation}
-the generalized likelihood ratio statistic can be computed by evaluating
-both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
-In particular, the maximum likelihood estimator (MLE) in the second term
-can be found as outlined in
-Section~\ref{sec:generalized-likelihood-ratio}.
 Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,

### 2016-03-13 jasa/paper-jasa.tex
finishing example 1

@@
 
-Finally, we see the log-likelihood curve $-\log \Lambda(\gamma)$ for the dataset $\mathcal{D}$ shown in 
-Fig.~\ref{fig:1a}. The exact likelihood curve is very well approximated by the method, confirming that
-even when the raw classifier does a poor job at modeling the $s^*(\mathbf{x}$, a good approximations of the 
-likelihood ratio can still be obtained by calibrating $s(\mathbf{x})$ (and by decomposing
-the mixture, if possible).
-
-Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
-$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
+Finally, we come to the log-likelihood curve 
+%shown in Figure~\ref{fig:likelihood}
 \begin{equation}
-\Lambda( \gamma = 0.05) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
-\end{equation}
-%\begin{equation}
-%\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
-%\end{equation}
+\log \Lambda( \gamma ) =  \log \frac{ p({\cal D} | \gamma ) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
+\end{equation}
+for the dataset $\mathcal{D}$ shown in  Fig.~\ref{fig:1a}. 
 By exploiting the fact that
@@
 \end{equation}
-the generalized likelihood ratio statistic can be computed by evaluating
-both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
-In particular, the maximum likelihood estimator (MLE) in the second term
-can be found as outlined in
-Section~\ref{sec:generalized-likelihood-ratio}.
+the generalized likelihood ratio can be computed by evaluating
+both terms with respect to a common reference $\gamma=0$
+%reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
+%In particular, the maximum likelihood estimator (MLE) in the second term
+%can be found 
+as outlined in Section~\ref{sec:generalized-likelihood-ratio}.
+Figure~\ref{fig:likelihood} shows that the exact likelihood curve is very well 
+approximated by the method, confirming that
+even when the raw classifier does a poor job at modeling the $s^*(\mathbf{x})$, a good approximations of the 
+likelihood ratio can still be obtained by calibrating $s(\mathbf{x})$ (and by decomposing
+the mixture, if possible).
+
+%
+%Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
+%$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
+%\begin{equation}
+%\Lambda( \gamma = 0.05) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
+%\end{equation}
+%%\begin{equation}
+%%\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
+%%\end{equation}
+%By exploiting the fact that
+%\begin{equation}
+%\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
+%\end{equation}
+%the generalized likelihood ratio statistic can be computed by evaluating
+%both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
+%In particular, the maximum likelihood estimator (MLE) in the second term
+%can be found as outlined in
+%Section~\ref{sec:generalized-likelihood-ratio}.
 
@@
 
-Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
-compared to the distribution of the true MLEs. It clearly suggests that
-the proposed inference procedure yields in this case an unbiased estimator
-($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
-trials). Note that had the underlying ratio not been accurate enough because of
-poor training or poor calibration, the two distributions would not be exactly
-overlapping, indicating bias. From this, the initial $\Lambda$ test statistic can finally
-be evaluated to determine in this case that the null should not be rejected.  Figure \ref{fig:2b} also confirms
-that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
-as should be the true statistic when the null is true.
+\bigskip
+An advantage of this approach compared to ABC is that the 
+classifier and calibration -- computationally intensive 
+parts of the approximation  -- are independent of the of the dataset $\mathcal{D}$. 
+Thus once trained and calibrated, the approximation can be applied to any dataset $\mathcal{D}$. 
+This makes it computationally efficient to perform ensemble tests of the method.
+
+Figure~\ref{fig:2a} shows the empirical distribution of the maximum likelihood estimators (MLEs) from the 
+approximate likelihood compared to the distribution of the MLEs from the exact likelihood. 
+It clearly demonstraites that in this case the proposed inference procedure yields an unbiased estimator with the essentially the same variance as the exact MLE.
+%($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
+%trials). 
+%Note that had the underlying ratio not been accurate enough because of
+%poor training or poor calibration, the two distributions would not be exactly
+%overlapping, indicating bias. 
+In addition to the MLE we can consider the coverage of a confidence interval based on the likelihood ratio test statistic.
+This is done by evaluating $-2\log \Lambda(\gamma=0.05)$ for samples drawn from $p(\mathbf{x} | \gamma=0.05)$.
+Wilks's theorem states that the distribution of $-2\log \Lambda(\gamma=0.05)$ should follow a $chi^2_1$ distribution.
+%
+%From this, the initial $\Lambda$ test statistic can finally
+%be evaluated to determine in this case that the null should not be rejected.  
+Figure \ref{fig:2b} also confirms this behavior, supporting the initial motivation of this method for likelihood-based inference.
+% that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
+%as should be the true statistic when the null is true.
 

### 2016-03-14 jasa/paper-jasa.tex
example 2 and HEP section

@@
         \includegraphics[width=\textwidth]{figures/fig2b.pdf}
-        \caption{$-2 \log \frac{p({\cal D}|\gamma=0.05)}{\sup_{\gamma \in \Theta} p({\cal D} | \gamma)}$}
+        \caption{$p(-2 \log \Lambda(\gamma=0.05) \, | \, \gamma=0.05)$}
         \label{fig:2b}
@@
 approximate likelihood compared to the distribution of the MLEs from the exact likelihood. 
-It clearly demonstraites that in this case the proposed inference procedure yields an unbiased estimator with the essentially the same variance as the exact MLE.
+It clearly demonstrates that in this case the approximate likelihood yields an unbiased estimator with the essentially the same variance as the exact MLE.
 %($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
@@
 %overlapping, indicating bias. 
-In addition to the MLE we can consider the coverage of a confidence interval based on the likelihood ratio test statistic.
+In addition to the MLE, we can study the coverage of a confidence interval based on the likelihood ratio test statistic.
 This is done by evaluating $-2\log \Lambda(\gamma=0.05)$ for samples drawn from $p(\mathbf{x} | \gamma=0.05)$.
-Wilks's theorem states that the distribution of $-2\log \Lambda(\gamma=0.05)$ should follow a $chi^2_1$ distribution.
+Wilks's theorem states that the distribution of $-2\log \Lambda(\gamma=0.05)$ should follow a $\chi^2_1$ distribution.
 %
@@
 %be evaluated to determine in this case that the null should not be rejected.  
-Figure \ref{fig:2b} also confirms this behavior, supporting the initial motivation of this method for likelihood-based inference.
+Figure \ref{fig:2b} also confirms this behavior, supporting the applicability of this method for likelihood-based inference techniques in the likelihood-free setting.
 % that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
@@
 
-Let us now consider the more challenging problem of maximum likelihood
-inference from multidimensional data generated by a black-box simulator.
-For the sake of the illustration, we will assume 5-dimensional observed samples $\mathbf{x}$ generated
+Let us now consider the more challenging problem of likelihood-free
+inference with multidimensional data.
+For the sake of the illustration, we will assume 5-dimensional feature $\mathbf{x}$ generated
 from the following process $p_0$:
@@
             $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
-            $z_2 \sim {\text{Mixture}}({\cal N}(\mu=-2, \sigma=1), {\cal N}(\mu=2, \sigma=0.5))$,
-            $z_3 \sim {\text{Exponential}(\lambda=3)}$ and
+            $z_2 \sim {\text{Mixture}}(\sfrac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \sfrac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5))$,
+            $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
             $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
@@
     \centering
-    \begin{subfigure}[t]{0.31\textwidth}
+    \begin{subfigure}[t]{0.45\textwidth}
         \includegraphics[width=\textwidth]{figures/fig5a.pdf}
@@
     ~
-    \begin{subfigure}[t]{0.31\textwidth}
+    \begin{subfigure}[t]{0.45\textwidth}
         \includegraphics[width=\textwidth]{figures/fig5b.pdf}
@@
     ~
-    \begin{subfigure}[t]{0.31\textwidth}
+    \begin{subfigure}[t]{0.45\textwidth}
         \includegraphics[width=\textwidth]{figures/fig5c.pdf}
@@
 
-Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
-our goal is to infer the nominal values $\alpha=1$ and $\beta=-1$ that
-were used for generating this data.
-We use the objective function
+The observations ${\cal D}$ represented in Figure~\ref{fig:4}
+are random samples with $\alpha=1$ and $\beta=-1$. 
+Our goal is to infer the values $\alpha$ and $\beta$ based on ${\cal D}$.
+We construct the log-likelihood ratio 
 \begin{equation}\label{eqn:obj-ex2}
@@
 \end{equation}
-Following recommendations from Section~\ref{sec:param-clf} to reduce
-computational requirements, both terms in the right hand side of Eqn.~\ref{eqn:obj-ex2-expanded} are evaluated using a single
-parameterized approximated  likelihood ratio, in this case based on a parameterized
-2-layer neural network where the alternate is fixed to $\alpha=0, \beta=0$.
+Following the procedure described in Section~\ref{sec:param-clf}, we 
+% to reduce
+%computational requirements, both terms in the right hand side of Eqn.~\ref{eqn:obj-ex2-expanded} are evaluated using 
+a single 2-layer neural network (with 5+2 inputs and one output node) to form the parameterized classifier $s(\mathbf{x}; \theta_0, \theta_1)$ and fix $\theta_1=(\alpha=0, \beta=0)$.
 Since the generative model is not expensive, the classifier output is calibrated
 on-the-fly with histograms for  every candidate parameter pair $(\alpha,\beta)$.
-Minimizing the constructed approximation of the objective with
-Bayesian optimization~\citep{brochu2010tutorial} eventually leads to the
-estimated posterior mean shown in Figure~\ref{fig:5c}, which is close to the
-true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
-estimated minimizer is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$. For
-comparison, Figure~\ref{fig:5b} shows the objective value as directly
-approximated by the proposed method for a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
-roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.
-As discussed in Section~\ref{sec:param-calibration}, parameterized calibration
-is a solution to this issue, but is left for now as an area for future work.
+
+Figure~\ref{fig:5a} shows the exact log-likelihood ratio for this dataset, which has an exact MLE at $(\hat \alpha=1.012, \hat \beta=-0.9221)$.
+Figure~\ref{fig:5b} shows the approximate log-likelihood ratio evaluated on a coarse grid of parameter values.
+Some roughness in the contours is observed, which is due to primarily to variance introduced in the calibration procedure.
+In addition to the statistical fluctuations due to finite calibration samples, there are also fluctuations introduced from changes in the binning of the calibration histograms as $\alpha$ and $\beta$ vary.
+As discussed in Section~\ref{sec:param-calibration}, a parameterized calibration
+procedure should ameliorate this issue, but that is left for now as an area for future work.
+Nevertheless, optimizing the approximate log-likelihood ratio with
+a Bayesian optimization~\citep{brochu2010tutorial} procedure is efficient and effective. 
+After $50$ likelihood evaluations, the maximum likelihood estimate is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$. 
+While the goal of the Bayesian optimization procedure is to find the maximum likelihood, the posterior mean of the 
+internal Gaussian process, shown in Figure~\ref{fig:5c}, is close to the exact log-likelihood ratio illustrated in Figure~\ref{fig:5a}. 
+In each case, the true values $\alpha=1$ and $\beta=-1$ are contained within the $1-\sigma$ likelihood contour.
+%For comparison, Figure~\ref{fig:5b} shows the approximate log likelihood ratio evaluated on a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
+%roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.
+
 Overall, this example further illustrates and confirms the ability
-of the proposed method for inference, even in the case of multi-dimensional data where
-direct approximations $p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$
+of the proposed method for inference with multiple parameters and multi-dimensional data where
+reliable approximations $\hat p(\mathbf{x}|\theta_0)$ and $\hat p(\mathbf{x}|\theta_1)$
 are often difficult to construct.
@@
 
+High energy physics was the original scientific domain that motivated the development of this procedure.
 In high energy physics, we are often searching for some class of events,
@@
 $\mathbf{x}$, with corresponding distributions $p_s(\mathbf{x}|\nu)$ for signal
-and $p_b(\mathbf{x}|\nu)$ for background, and where $\nu$ are nuisance
+and $p_b(\mathbf{x}|\nu)$ for background,  where $\nu$ are nuisance
 parameters describing uncertainties in the underlying physics prediction or
@@
 
-In this setting, a classifier is usually built from a large sample of data
-$\{\mathbf{x}_i, y_i\}$ generated with some nominal values of the parameters
-$\nu=\nu_0$, with $y=0$ for samples drawn from the background component and
-$y=1$ for samples drawn from the signal component. Importantly, the $y=1$ label
-corresponds to the signal only hypothesis, and not to the alternate
-signal-plus-background hypothesis. The resulting classifier approximates the
-regression function
-$\sfrac{p_s(\mathbf{x}|\nu_0)}{p_s(\mathbf{x}|\nu_0)+p_b(\mathbf{x}|\nu_0)}$, which is
-one to one with the likelihood ratio of the null to the alternate
-$\frac{p(\mathbf{x}|\mu=0,\nu_0)}{p(\mathbf{x}|\mu,\nu_0)}$ for all $\mu$ and therefore
-satisfies conditions of Theorem~\ref{thm:ratio-equivalence}. Associating the
-$y=1$ label to the signal component has advantages because it helps the
-classifier focus its capacity on the relevant regions in the feature space,
-particularly when the signal is a very small perturbation to the background
-(i.e. $\mu \ll 1$). Once the classifier is trained, large samples of data are
-generated from $p_s(\mathbf{x} | \nu)$ and $p_b(\mathbf{x} | \nu)$ and we
-estimate the distributions $\hat{p}_s(\hat s(\mathbf{x}) | \nu)$ and
-$\hat{p}_b(\hat s(\mathbf{x}) | \nu)$ continuously parameterized in $\nu$.
+%In this setting, a classifier is usually built from a large sample of data
+%$\{\mathbf{x}_i, y_i\}$ generated with some nominal values of the parameters
+%$\nu=\nu_0$, with $y=0$ for samples drawn from the background component and
+%$y=1$ for samples drawn from the signal component. Importantly, the $y=1$ label
+%corresponds to the signal only hypothesis, and not to the alternate
+%signal-plus-background hypothesis. The resulting classifier approximates the
+%regression function
+%$\sfrac{p_s(\mathbf{x}|\nu_0)}{p_s(\mathbf{x}|\nu_0)+p_b(\mathbf{x}|\nu_0)}$, which is
+%one to one with the likelihood ratio of the null to the alternate
+%$\frac{p(\mathbf{x}|\mu=0,\nu_0)}{p(\mathbf{x}|\mu,\nu_0)}$ for all $\mu$ and therefore
+%satisfies conditions of Theorem~\ref{thm:ratio-equivalence}. Associating the
+%$y=1$ label to the signal component has advantages because it helps the
+%classifier focus its capacity on the relevant regions in the feature space,
+%particularly when the signal is a very small perturbation to the background
+%(i.e. $\mu \ll 1$). Once the classifier is trained, large samples of data are
+%generated from $p_s(\mathbf{x} | \nu)$ and $p_b(\mathbf{x} | \nu)$ and we
+%estimate the distributions $\hat{p}_s(\hat s(\mathbf{x}) | \nu)$ and
+%$\hat{p}_b(\hat s(\mathbf{x}) | \nu)$ continuously parameterized in $\nu$.
 
@@
 Nuisance parameters are an after thought in the typical usage of machine
-learning in high energy physics. In fact, most discussions related to the training and
-optimizing the classifier only consider $p_b(\mathbf{x})$ and $p_s(\mathbf{x})$, with $\nu=\nu_0$
-being implicit. However, as experimentalists we know that we must account for
-various forms of systematic uncertainty, parameterized by nuisance parameters
-$\nu$. In practice, we take the classifier $\hat s$ as fixed and then propagate
-uncertainty by estimating $\hat{p}_s(\hat s(\mathbf{x}) | \nu)$. Building such
-distributions for values of $\nu$ other than the nominal $\nu_0$
-used at training can be thought of as a calibration necessary for
-classical statistical inference. However, this classifier is clearly not optimal
-for $\nu \ne \nu_0$. For the this reason, it is expected that the integrated parameterized approach proposed in this work
+learning in high energy physics. The classifiers are typically trained with 
+data generated using a fixed nominal value of the nuisance parameters $\nu=\nu_0$. 
+However, as experimentalists we know that we must account for
+the systematic uncertainties that correspond to the nuisance parameters
+$\nu$. Thus, typically we take the classifier $\hat s(\mathbf{x})$ as fixed and then propagate
+uncertainty by estimating $\hat{p}_s(\hat s(\mathbf{x}) | \nu)$ with a parametrized calibration procedure. 
+%Building such
+%distributions for values of $\nu$ other than the nominal $\nu_0$
+%used for training is thought of as a calibration necessary for
+%classical statistical inference. 
+However, this classifier is clearly not optimal
+for $\nu \ne \nu_0$. In contrast, a parameterized classifier proposed in this work
 would yield more accurate estimates of the generalized likelihood ratio.
@@
 
-While the original motivation for this work was to improve the treatment of
-systematic uncertainties in new particle searches by parameterizing the
-classifier in terms of the nuisance parameters $\nu$, the same approach can be
-used for parameter inference. In the case of new particle searches, the parameter
-of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
-When measuring particle properties, the distribution of the features also depend
-on parameters such as a particle's mass and quantum numbers. This is easily
-accommodated by extending $p_s(\mathbf{x}|\nu) \to p_s(\mathbf{x}|\theta)$, where $\theta$
-includes both parameters of interest and nuisance parameters.
+In addition to robustness to systematic uncertainties incorporated by the nuisance parameters $\nu$,
+the proposed method can be used to infer parameters of interest. Not only can the mixture coefficient $\mu$
+be inferred using the decomposition procedure, but also physical parameters like particle masses that change the distribution of $\mathbf{x}$.
+%While the original motivation for this work was to improve the treatment of
+%systematic uncertainties in new particle searches by parameterizing the
+%classifier in terms of the nuisance parameters $\nu$, the same approach can be
+%used for parameter inference. In the case of new particle searches, the parameter
+%of interest is the mixture coefficient for the signal component $p_s(x|\nu)$.
+%When measuring particle properties, the distribution of the features also depend
+%on parameters such as a particle's mass and quantum numbers. This is easily
+%accommodated by extending $p_s(\mathbf{x}|\nu) \to p_s(\mathbf{x}|\theta)$, where $\theta$
+%includes both parameters of interest and nuisance parameters.
 This formalism represents a significant step forward in the usage of machine
@@
 quantities we wish to measure.
-The work of \cite{Whiteson:2006ws} is similar,
-as the stochastic optimization was directly trying to minimize the measurement
-uncertainty of a particle's mass; however, the resulting classifier was fixed.
-This approach also offers the advantage that it explicitly reformulates the
-per-experiment optimization to the per-event optimization, which is less
-computationally intensive.
-
-Another approach that is similar in spirit is the so-called matrix element
+
+%The work of \cite{Whiteson:2006ws} used a stochastic optimization technique 
+%to find a binary classifier such that selected events would minimize the measurement
+%uncertainty of a particle's mass; however, the resulting classifier was fixed.
+%This approach also offers the advantage that it explicitly reformulates the
+%per-experiment optimization to the per-event optimization, which is less
+%computationally intensive.
+
+%The work of \cite{Whiteson:2006ws} is similar,
+%as the stochastic optimization was directly trying to minimize the measurement
+%uncertainty of a particle's mass; however, the resulting classifier was fixed.
+%This approach also offers the advantage that it explicitly reformulates the
+%per-experiment optimization to the per-event optimization, which is less
+%computationally intensive.
+
+Another approach for parameter inference with multi-dimensional data specific to
+high energy physics is the so-called matrix element
 method, in which one  directly computes an approximate likelihood ratio by
-performing a computationally intensive integral associated to the detector
+performing a computationally intensive integral associated to a simplified detector
 response~\citep{Volobouev:2011vb}. In the approach considered in this paper, the
-detector response is naturally handled by the Monte Carlo sampling used in the
-simulation of the detector; however, that integral is intractable for the matrix
-element method. Even with drastic simplifications of the detector response, the
+detailed detector response is naturally incorporated by the simulator; however, that integral 
+is intractable for the matrix element method. Even with drastic simplifications of the detector response, the
 matrix element method can take several minutes of CPU time to calculate the
-likelihood ratio for a single event. The work here can be seen as aiming
+likelihood ratio for a single event $\mathbf{x}$. The work here can be seen as aiming
 at the same conceptual target, but relying on machine learning to overcome the
@@
 include the measurement of coefficients to quantum mechanical operators
-describing the decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so
+describing the production and decay of the Higgs boson~\citep{Chen:2014pia} and, if we are so
 lucky, measurement of the mass of supersymmetric particles in cascade
@@
 many events, each with a feature vector $\mathbf{x}$ that has on the order of 10
-components, and a parameter vector $\theta$ with 5-10 parameters of interest and
-possibly many more nuisance parameters. The state of the art for the operator
-coefficients of the Higgs decay uses the so-called matrix element likelihood
-analysis (MELA) in which the equivalent of $s(\mathbf{x}; \theta_0, \theta_1)$ is
-approximated by neglecting detector effects~\citep{Gao:2010qx,Bolognesi:2012mm}.
+components, and a parameter vector $\theta$ with 2-10 parameters of interest and
+possibly many more nuisance parameters. 
+%The state of the art for the operator
+%coefficients of the Higgs decay uses the so-called matrix element likelihood
+%analysis (MELA) in which the equivalent of $s(\mathbf{x}; \theta_0, \theta_1)$ is
+%approximated by neglecting detector effects~\citep{Gao:2010qx,Bolognesi:2012mm}.
 

### 2016-03-14 jasa/paper-jasa.tex
Figure -> Fig.

@@
 on training data with input $(\mathbf{x}, \theta_0, \theta_1)_i$ and target
-$y_i$, as outlined in Algorithm~\ref{alg:training}.
-Somewhat unusually, the
+$y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
 unknown values of the parameters are taken as input to the classifier; their
@@
 limited capacity of the classifier may impact the performance the resulting
-parameterized decision function. These issues are left as an area for future
-work, but preliminary exploration by \cite{Baldi:2016fzo} show that
-Algorithm~\ref{alg:training} is an effective practical approach.
+parameterized decision function.
+Preliminary exploration by \cite{Baldi:2016fzo} shows that
+ a uniform grid scan over parameter space is an effective practical approach;
+ however, we introduce the distributions $\pi_{\Theta_0}$ and $\pi_{\Theta_1}$
+into the Algorithm~\ref{alg:training} to allow for a more
+ sophisticated sampling strategy.
+% this formulation of Algorithm~\ref{alg:training} allows for a more
+% sophisticated sampling strategy.
+% These issues are left as an area for future work.
+%For this reason, we introduce the distributions $\pi_{\Theta_0}$ and $\pi_{\Theta_1}$
+%into the Algorithm~\ref{alg:training} to allow for a more
+% sophisticated sampling strategy.
 
@@
     \State ${\cal T := \{ \}}$;
-    \While{ $\text{size}({\cal L}) < N $ }
-        \State Select or draw $\theta_0$, $\theta_1$ from $\Theta_0$, $\Theta_1$;
+    \While{ $\text{size}({\cal T}) < N $ }
+        %\State Select or draw $\theta_0$, $\theta_1$ from $\Theta_0$, $\Theta_1$;
+        \State Draw $\theta_0 \sim \pi_{\Theta_0}$;
 	    \State Draw $\mathbf{x} \sim p(\mathbf{x}|\theta_0)$;
 		\State ${\cal T} := {\cal T} \cup \{ ((\mathbf{x}, \theta_0, \theta_1), y=0) \}$;
+        \State Draw $\theta_1 \sim \pi_{\Theta_1}$;
 		\State Draw $\mathbf{x} \sim p(\mathbf{x}|\theta_1)$;
@@
 $p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$. Samples drawn for the nominal value $\gamma=0.05$ are shown in
-Figure~\ref{fig:1a} and used later for inference.
-
-Figure~\ref{fig:p_sij} shows the intermediate stages for the decomposition described in Section~\ref{sec:mixture}.
+Fig.~\ref{fig:1a} and used later for inference.
+
+Fig.~\ref{fig:p_sij} shows the intermediate stages for the decomposition described in Section~\ref{sec:mixture}.
 The blue and green curves show $p_{c'}(\hat{s}_{c,c'}(\mathbf{x}))$,  the distributions for the score for the sub-classifiers for the three pair-wise comparisons of the mixture components.
@@
         \includegraphics[width=\textwidth]{figures/fig1c.pdf}
-        \caption{$\log \hat r(\mathbf{x})$ using neural network}
+        \caption{$\log \hat r(\hat{s}(\mathbf{x}))$ using neural network}
         \label{fig:1c}
@@
         \includegraphics[width=\textwidth]{figures/fig1d.pdf}
-        \caption{$\log \hat r(\mathbf{x})$ using random forest}
+        \caption{$\log \hat r(\hat{s}(\mathbf{x}))$ using random forest}
         \label{fig:1d}
@@
         \includegraphics[width=\textwidth]{figures/fig-ratio-dist.pdf}
-        \caption{$p( \log \hat r(\mathbf{x})) \, | \,\gamma = 0.05)$ }
+        \caption{$p( \log \hat r(\hat{s}(\mathbf{x}))) \, | \,\gamma = 0.05)$ }
         \label{fig:decomposed_ratio}
@@
 
-Figure~\ref{fig:decomposed_ratio} shows the distribution of $ \log \hat r(\mathbf{x}))$ for $\gamma=0.05$
+Fig.~\ref{fig:decomposed_ratio} shows the distribution of $ \log \hat r(\mathbf{x}))$ for $\gamma=0.05$
 using the decomposed approximation of $\hat r(\mathbf{x}))$ with neural networks and the distribution
@@
 \begin{equation}
-\log \Lambda( \gamma ) =  \log \frac{ p({\cal D} | \gamma ) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
+\log \Lambda( \gamma ) =  \log \frac{ p({\cal D} | \gamma ) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \;
 \end{equation}
@@
 \begin{equation}
-\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
+\log \frac{ p({\cal D} | \gamma ) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma ) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
 \end{equation}
@@
 as outlined in Section~\ref{sec:generalized-likelihood-ratio}.
-Figure~\ref{fig:likelihood} shows that the exact likelihood curve is very well 
+Fig.~\ref{fig:likelihood} shows that the exact likelihood curve is very well
 approximated by the method, confirming that
@@
 
-\bigskip
 An advantage of this approach compared to ABC is that the
@@
 
-Figure~\ref{fig:2a} shows the empirical distribution of the maximum likelihood estimators (MLEs) from the 
+Fig.~\ref{fig:2a} shows the empirical distribution of the maximum likelihood estimators (MLEs) from the
 approximate likelihood compared to the distribution of the MLEs from the exact likelihood.
@@
 %be evaluated to determine in this case that the null should not be rejected.
-Figure \ref{fig:2b} also confirms this behavior, supporting the applicability of this method for likelihood-based inference techniques in the likelihood-free setting.
+Fig.~\ref{fig:2b} also confirms this behavior, supporting the applicability of this method for likelihood-based inference techniques in the likelihood-free setting.
 % that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
@@
     \end{subfigure}
-    ~
+
     \begin{subfigure}[t]{0.45\textwidth}
@@
 
-The observations ${\cal D}$ represented in Figure~\ref{fig:4}
+The observations ${\cal D}$ represented in Fig.~\ref{fig:4}
 are random samples with $\alpha=1$ and $\beta=-1$.
@@
 \end{equation}
-Following the procedure described in Section~\ref{sec:param-clf}, we 
+Following the procedure described in Section~\ref{sec:param-clf}, we build
 % to reduce
 %computational requirements, both terms in the right hand side of Eqn.~\ref{eqn:obj-ex2-expanded} are evaluated using
-a single 2-layer neural network (with 5+2 inputs and one output node) to form the parameterized classifier $s(\mathbf{x}; \theta_0, \theta_1)$ and fix $\theta_1=(\alpha=0, \beta=0)$.
+a build single 2-layer neural network (with 5+2 inputs and one output node) to form the parameterized classifier $s(\mathbf{x}; \theta_0, \theta_1)$ and fix $\theta_1=(\alpha=0, \beta=0)$.
 Since the generative model is not expensive, the classifier output is calibrated
@@
 While the goal of the Bayesian optimization procedure is to find the maximum likelihood, the posterior mean of the
-internal Gaussian process, shown in Figure~\ref{fig:5c}, is close to the exact log-likelihood ratio illustrated in Figure~\ref{fig:5a}. 
+internal Gaussian process, shown in Fig.~\ref{fig:5c}, is close to the exact log-likelihood ratio illustrated in Fig.~\ref{fig:5a}.
 In each case, the true values $\alpha=1$ and $\beta=-1$ are contained within the $1-\sigma$ likelihood contour.

### 2016-03-14 jasa/paper-jasa.tex
text changes

@@
 
-Fig.~\ref{fig:p_sij} shows the intermediate stages for the decomposition described in Section~\ref{sec:mixture}.
+Figure~\ref{fig:p_sij} shows the intermediate stages for the decomposition described in Section~\ref{sec:mixture}.
 The blue and green curves show $p_{c'}(\hat{s}_{c,c'}(\mathbf{x}))$,  the distributions for the score for the sub-classifiers for the three pair-wise comparisons of the mixture components.
@@
 
-Fig.~\ref{fig:decomposed_ratio} shows the distribution of $ \log \hat r(\mathbf{x}))$ for $\gamma=0.05$
+Figure~\ref{fig:decomposed_ratio} shows the distribution of $ \log \hat r(\mathbf{x}))$ for $\gamma=0.05$
 using the decomposed approximation of $\hat r(\mathbf{x}))$ with neural networks and the distribution
@@
 \begin{equation}
-\log \Lambda( \gamma ) =  \log \frac{ p({\cal D} | \gamma ) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \;
+\log \Lambda( \gamma ) =  \log \frac{ p({\cal D} | \gamma ) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
 \end{equation}
@@
 as outlined in Section~\ref{sec:generalized-likelihood-ratio}.
-Fig.~\ref{fig:likelihood} shows that the exact likelihood curve is very well
+Figure~\ref{fig:likelihood} shows that the exact likelihood curve is very well 
 approximated by the method, confirming that
@@
 
+\bigskip
 An advantage of this approach compared to ABC is that the 
@@
 
-Fig.~\ref{fig:2a} shows the empirical distribution of the maximum likelihood estimators (MLEs) from the
+Figure~\ref{fig:2a} shows the empirical distribution of the maximum likelihood estimators (MLEs) from the 
 approximate likelihood compared to the distribution of the MLEs from the exact likelihood. 
@@
 %be evaluated to determine in this case that the null should not be rejected.  
-Fig.~\ref{fig:2b} also confirms this behavior, supporting the applicability of this method for likelihood-based inference techniques in the likelihood-free setting.
+Figure \ref{fig:2b} also confirms this behavior, supporting the applicability of this method for likelihood-based inference techniques in the likelihood-free setting.
 % that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
@@
 \end{equation}
-Following the procedure described in Section~\ref{sec:param-clf}, we build
+Following the procedure described in Section~\ref{sec:param-clf}, we 
 % to reduce
@@
 Figure~\ref{fig:5a} shows the exact log-likelihood ratio for this dataset, which has an exact MLE at $(\hat \alpha=1.012, \hat \beta=-0.9221)$.
+
 Figure~\ref{fig:5b} shows the approximate log-likelihood ratio evaluated on a coarse grid of parameter values.

### 2016-03-14 jasa/paper-jasa.tex
fix conflicts

@@
 $p_{c_2} := {\cal N}(\mu=1, \sigma^2=0.25)$. Samples drawn for the nominal value $\gamma=0.05$ are shown in
-Fig.~\ref{fig:1a} and used later for inference.
+Figure~\ref{fig:1a} and used later for inference.
 
@@
 %be evaluated to determine in this case that the null should not be rejected.
-Figure \ref{fig:2b} also confirms this behavior, supporting the applicability of this method for likelihood-based inference techniques in the likelihood-free setting.
+Figure~\ref{fig:2b} also confirms this behavior, supporting the applicability of this method for likelihood-based inference techniques in the likelihood-free setting.
 % that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
@@
 \end{equation}
-Following the procedure described in Section~\ref{sec:param-clf}, we 
+Following the procedure described in Section~\ref{sec:param-clf}, we build
 % to reduce
@@
 Figure~\ref{fig:5a} shows the exact log-likelihood ratio for this dataset, which has an exact MLE at $(\hat \alpha=1.012, \hat \beta=-0.9221)$.
-
 Figure~\ref{fig:5b} shows the approximate log-likelihood ratio evaluated on a coarse grid of parameter values.

### 2016-03-15 jasa/paper-jasa.tex
Standalone colorbar
update fig1b, remove old files
adjust alignment of Fig1b
diagnostic plots and text

@@
 $r(\hat{s}(\mathbf{x}))$ directly.
-%There are various strategies.
-For instance, isotonic regression, which is commonly used to transform the classifier score  $\hat{s}(\mathbf{x})$ into $\hat{s}_\textrm{iso}(\mathbf{x})$ that more accurately reflect  the posterior probability $s^*(\mathbf{x})$
+For example, isotonic regression, which is commonly used to transform the classifier score  $\hat{s}(\mathbf{x})$ into $\hat{s}_\textrm{iso}(\mathbf{x})$ that more accurately reflect  the posterior probability $s^*(\mathbf{x})$
 of Eq.~\ref{eqn:best-s-clf}, can be used for calibration.
-This is done by inverting the relationship ${r}(\mathbf{x}) = (1-s^*(\mathbf{x}))/s^*(\mathbf{x})$ to obtain $\hat{r}(\mathbf{x}) = (1-\hat{s}_\textrm{iso}(\mathbf{x}))/\hat{s}_\textrm{iso}(\mathbf{x})$.
-\textbf{Gilles, should we add more about direct density ratio estimation (currently in 5. Related Work) here?}
+This is done by inverting the relationship ${r}(\mathbf{x}) = (1-s^*(\mathbf{x}))/s^*(\mathbf{x})$ to obtain $\hat{r}(\hat{s}(\mathbf{x})) = (1-\hat{s}_\textrm{iso}(\mathbf{x}))/\hat{s}_\textrm{iso}(\mathbf{x})$.
+Additionally, Sec.~\ref{sec:related} describes related work in which the density ratio $\hat{r}(\mathbf{x})$ is estimated directly on the feature space $\mathcal{X}$.
+
+%\textbf{Gilles, should we add more about direct density ratio estimation (currently in 5. Related Work) here?}
 
@@
 p(\hat{s}(\mathbf{x})|\theta)$ or $\hat{r}(\hat{s}(\mathbf{x})) \approx r(\hat{s}(\mathbf{x}))$). Thus, even if the classifier does a poor job at
-learning the optimal decision function~\ref{eqn:best-s-clf} and therefore at
+learning the optimal decision function~\ref{eqn:best-s-clf} and, therefore, at
 reproducing the level sets of the per-sample likelihood ratio, the density of
@@
 use Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
-estimate as
+estimate (MLE) as
 \begin{align}\label{eq:mle_withs}
@@
 
+\subsection{Diagnostics}
+
+While Theorem~\ref{thm:ratio-equivalence} states that the likelihood ratio $r(\mathbf{x}; \theta_0, \theta_1)$ is 
+invariant under the dimensionality reduction $s(\mathbf{x}; \theta_0, \theta_1)$ provided that it is monotonic with 
+$r(\mathbf{x}; \theta_0, \theta_1)$ itself and we know that any universally strongly consistent algorithm can be 
+used to learn such a function, we know that in practice $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$ will not be exact.
+Thus, it is crucial that to have some diagnostic procedures to assess the quality of this approximation.
+This is complicated by the fact that in the likelihood-free setting, we don't have access to the true likelihood ratio. 
+Below we consider two such diagnostic procedures that can be implemented in the likelihood-free setting.
+
+The first diagnostic procedure is related to the procedure for finding the MLE $\hat{\theta}$ in Eqn.~\ref{eq:mle_withs}.
+As pointed out there it is important that one maximizes the likelihood ratio as the surface integral and Jacobian factors related to the dimensionality reduction only cancel in the ratio (see Eqn.~\ref{eqn:factorization}).
+Importantly, they also only cancel if the dimensionality reduction map satisfies the conditions needed for Theorem~\ref{thm:ratio-equivalence} to hold.
+Moreover, the resulting value of $\hat{\theta}$ should be independent of the value of $\theta_1$ used in the denominator of the likelihood ratio.
+Similarly, we have
+%$\log\Lambda(\theta) = \log p(\mathcal{D}|\theta)/p(\mathcal{D}|\hat\theta) = \log p(\mathcal{D}|\theta)/p(\mathcal{D}|\theta_1) -\log p(\mathcal{D}|\hat\theta)/p(\mathcal{D}|\theta_1)$ for all values of $\theta_1$.
+\begin{equation}
+\log\Lambda(\theta) = \log \frac{p(\mathcal{D}|\theta)}{p(\mathcal{D}|\hat\theta)} = \log \frac{p(\mathcal{D}|\theta)}{p(\mathcal{D}|\theta_1)} -\log \frac{p(\mathcal{D}|\hat\theta)}{p(\mathcal{D}|\theta_1)}
+\end{equation}
+for all values of $\theta_1$.
+Thus, by explicitly checking the independence of these quantities on $\theta_1$ we indirectly probe the quality of the approximation $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1)) \approx {r}({s}(\mathbf{x}; \theta_0, \theta_1))$.
+
+The second diagnostic procedure leverages the connection of this technique to direct density ratio estimation and its application to covariate shift and importance sampling.
+The idea is simple: we test the relationship $p(\mathbf{x}|\theta_0) = p(\mathbf{x}|\theta_1) {r}({s}(\mathbf{x}; \theta_0, \theta_1))$ with the approximate ratio $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$ and samples drawn from the generative model.
+More specifically, we can train a classifier to distinguish between unweighted samples from $p(\mathbf{x}|\theta_0)$ and samples from $p(\mathbf{x}|\theta_1)$ weighted by $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$.
+If the classifier can distinguish between the distributions, then $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$ is not a good approximation of ${r}({s}(\mathbf{x}; \theta_0, \theta_1))$.
+In contrast, if the classifier is unable to distinguish between the two distributions, then either $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$ is a good approximation or the discriminator is not effective.
+The two situations can be disentangled to some degree by training another classifier to distinguish between an unweighted distribution of samples from $p(\mathbf{x}|\theta_1)$.
+
+We illustrate these diagnostic procedures in Fig.~\ref{fig:ratio_comp}.
+
 \section{Examples and applications}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig-p_sij.pdf}
+        \hspace{-.2em}\includegraphics[width=1.10\textwidth]{figures/fig1b.pdf}
         \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $1/(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))$}
@@
 
-%    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
-%      %(or a blank line to force the subfigure onto a new line)
-%    \begin{subfigure}[b]{0.3\textwidth}
-%        \includegraphics[width=\textwidth]{figures/fig_ratio_dist.png}
-%        \caption{Likelihood ratio distributions}
-%        \label{fig:decomposed_ratio}
-%    \end{subfigure}
-    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
-      %(or a blank line to force the subfigure onto a new line)
+    \vspace{1em}
+
     \begin{subfigure}[b]{0.4\textwidth}
@@
 
-    \caption{Approximation of the log-likelihood ratio
+    \caption{Example data $\mathcal{D}$ Various stages in Approximation of the log-likelihood ratio
              $\log r(\mathbf{x};\gamma=0.05,\gamma=0)$
@@
     \centering
-    \begin{subfigure}[t]{0.45\textwidth}
+    \begin{subfigure}[t]{0.43\textwidth}
+        \centering
         \includegraphics[width=\textwidth]{figures/fig5a.pdf}
@@
     ~
-    \begin{subfigure}[t]{0.45\textwidth}
+    \begin{subfigure}[t]{0.43\textwidth}
+        \centering
         \includegraphics[width=\textwidth]{figures/fig5b.pdf}
@@
 
-    \begin{subfigure}[t]{0.45\textwidth}
+    \begin{subfigure}[t]{0.43\textwidth}
+        \centering
         \includegraphics[width=\textwidth]{figures/fig5c.pdf}
@@
         \label{fig:5c}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[t]{0.43\textwidth}
+        \centering
+        \hfill\includegraphics[width=0.9\textwidth]{figures/fig5d.pdf}
+        \caption{ }
+        \label{fig:5d}
     \end{subfigure}
@@
 are often difficult to construct.
+
+%\subsubsection{Diagnostics}
+
+\begin{figure}
+    \centering
+    \includegraphics[height=.9\textheight]{figures/diagnostics.pdf}
+    \caption{Diagnostic plots ....}
+    \label{fig:ratio_comp}
+\end{figure}
 

### 2016-03-16 jasa/paper-jasa.tex
adding both types of plots
Add reference to ABC

@@
 %
-% One can think of the ratio $p_1(s)/p_0(s)$ as a way of calibrating the the discriminative classifier and correcting for the monotonic transformation $m$ of the desired likelihood ratio as in Eq.~\ref{eq:montonic}.
+% One can think of the ratio $p_1(s)/p_0(s)$ as a way of calibrating the the discriminative classifier and correcting for the monotonic transformation $m$ of the desired likelihood ratio as in Eqn.~\ref{eq:montonic}.
 
@@
 For example, isotonic regression, which is commonly used to transform the classifier score  $\hat{s}(\mathbf{x})$ into $\hat{s}_\textrm{iso}(\mathbf{x})$ that more accurately reflect  the posterior probability $s^*(\mathbf{x})$
-of Eq.~\ref{eqn:best-s-clf}, can be used for calibration.
+of Eqn.~\ref{eqn:best-s-clf}, can be used for calibration.
 This is done by inverting the relationship ${r}(\mathbf{x}) = (1-s^*(\mathbf{x}))/s^*(\mathbf{x})$ to obtain $\hat{r}(\hat{s}(\mathbf{x})) = (1-\hat{s}_\textrm{iso}(\mathbf{x}))/\hat{s}_\textrm{iso}(\mathbf{x})$.
@@
 %For instance, isotonic regression, which is commonly used to transform the classifier score  $\hat{s}(\mathbf{x})$ into $\hat{s}_\textrm{iso}(\mathbf{x})$ that more accurately reflect  the posterior probability $s^*(\mathbf{x})$
-%of Eq.~\ref{eqn:best-s-clf}, can be used for calibration.
+%of Eqn.~\ref{eqn:best-s-clf}, can be used for calibration.
 This is done by inverting the relationship ${r}(\mathbf{x}; \theta_0, \theta_1) = (1-s^*(\mathbf{x}))/s^*(\mathbf{x})$ to obtain $\hat{r}(\mathbf{x}; \theta_0, \theta_1) = (1-\hat{s}_\textrm{iso}(\mathbf{x}))/\hat{s}_\textrm{iso}(\mathbf{x})$.
@@
 % $\theta_0$ and $\theta_1$. Thus one must implement the embedding of the
-% classifier as in Eq.~\ref{eq:embedding}.
+% classifier as in Eqn.~\ref{eq:embedding}.
 %  A concrete realization of this has
@@
 % classifiers to discriminate between components of a mixture model. First, we
-% generalize Eq.~\ref{eq:hepGen} to a mixture model of several components
+% generalize Eqn.~\ref{eq:hepGen} to a mixture model of several components
 % \begin{equation}
@@
 In the situation where the only free parameters of the  model are the mixture coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio).
-Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parameterized classifier as in Eq.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
+Equation~\ref{eq:decomposedResult} allows one to take advantage of both the parameterized classifier as in Eqn.~\ref{eq:hep_improved} and the capacity focusing technique in the typical HEP usage pattern.
 
@@
 
-An advantage of this approach compared to ABC is that the
+An advantage of this approach compared to Approximate Bayesian Computation~\citep{beaumont2002approximate} is that the
 classifier and calibration -- computationally intensive
@@
     \label{fig:ratio_comp}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \begin{subfigure}[t]{0.4\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/likelihood_comp_0.pdf}
+        \caption{$-2\log \Lambda(\alpha)$: poorly trained, well calibrated. }
+        \label{fig:6a}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[t]{0.4\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/ROC_comp0.pdf}
+        \caption{ROC: poorly trained, well calibrated. }
+        \label{fig:6b}
+    \end{subfigure}
+
+    \begin{subfigure}[t]{0.4\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/likelihood_comp_1.pdf}
+        \caption{$-2\log \Lambda(\alpha)$: poorly calibrated, well trained. }
+        \label{fig:6c}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[t]{0.4\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/ROC_comp1.pdf}
+        \caption{ROC: poorly calibrated, well trained. }
+        \label{fig:6d}
+    \end{subfigure}
+
+    \begin{subfigure}[t]{0.4\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/likelihood_comp_2.pdf}
+        \caption{$-2\log \Lambda(\alpha)$: well trained, well calibrated. }
+        \label{fig:6e}
+    \end{subfigure}
+    ~
+    \begin{subfigure}[t]{0.4\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figures/ROC_comp2.pdf}
+        \caption{ROC: well trained, well calibrated. }
+        \label{fig:6f}
+    \end{subfigure}
+
+    \caption{Alternative presentation of Diagnostic plots.}
+    \label{fig:diagnostics}
 \end{figure}
@@
 % classical statistical setting.
-The proposed method offers an alternative to approximate Bayesian computation
+The proposed method offers an alternative to Approximate Bayesian Computation
 for parameter inference in the likelihood-free setting that can also be used in

### 2016-03-16 jasa/paper-jasa.tex

@@
 
-Figure~\ref{fig:decomposed_ratio} shows the distribution of $ \log \hat r(\mathbf{x}))$ for $\gamma=0.05$
-using the decomposed approximation of $\hat r(\mathbf{x}))$ with neural networks and the distribution
+Figure~\ref{fig:decomposed_ratio} shows the distribution of $ \log \hat r(\mathbf{x})$ for $\gamma=0.05$
+using the decomposed approximation of $\hat r(\mathbf{x})$ with neural networks and the distribution
 of the exact log-likelihood ratio.
@@
 approximate likelihood compared to the distribution of the MLEs from the exact likelihood.
-It clearly demonstrates that in this case the approximate likelihood yields an unbiased estimator with the essentially the same variance as the exact MLE.
+It clearly demonstrates that in this case the approximate likelihood yields an unbiased estimator with essentially the same variance as the exact MLE.
 %($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
@@
 \end{equation}
-Following the procedure described in Section~\ref{sec:param-clf}, we build
+Following the procedure described in Section~\ref{sec:param-clf}, we
 % to reduce
 %computational requirements, both terms in the right hand side of Eqn.~\ref{eqn:obj-ex2-expanded} are evaluated using
-a build single 2-layer neural network (with 5+2 inputs and one output node) to form the parameterized classifier $s(\mathbf{x}; \theta_0, \theta_1)$ and fix $\theta_1=(\alpha=0, \beta=0)$.
+build a single 2-layer neural network (with 5+2 inputs and one output node) to form the parameterized classifier $s(\mathbf{x}; \theta_0, \theta_1)$ and fix $\theta_1=(\alpha=0, \beta=0)$.
 Since the generative model is not expensive, the classifier output is calibrated
@@
 Figure~\ref{fig:5b} shows the approximate log-likelihood ratio evaluated on a coarse grid of parameter values.
-Some roughness in the contours is observed, which is due to primarily to variance introduced in the calibration procedure.
+Some roughness in the contours is observed, which is primarily due to variance introduced in the calibration procedure.
 In addition to the statistical fluctuations due to finite calibration samples, there are also fluctuations introduced from changes in the binning of the calibration histograms as $\alpha$ and $\beta$ vary.

### 2016-03-16 jasa/paper-jasa.tex
updated legends

@@
 
-\subsection{Diagnostics}
+\subsection{Diagnostics}\label{S:diagnostics}
 
@@
         \includegraphics[width=\textwidth]{figures/fig-likelihood.pdf}
-        \caption{$-\log \Lambda(\gamma)$ }
+        \caption{$-2\log \Lambda(\gamma)$ }
         \label{fig:likelihood}
@@
     \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
-    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $(\alpha=1, \beta=-1)$. The size of ${\cal D}$ is set to $500$.}
+    \caption{Scatter plots for the 500 samples in the 5-dimensional data ${\cal D}$ generated for nominal values $(\alpha=1, \beta=-1)$. }
     \label{fig:4}
@@
     to the true values $(\alpha=1, \beta=-1)$ used to generate $\mathcal{D}$, the green dot is the MLE
-    $(\hat \alpha=1.012, \hat \beta=-0.9221)$  from the exact likelihood, while the blue dot is the MLE $(\hat \alpha=1.008, \hat \beta=-1.004)$ from the approximate likelihood.
+     from the exact likelihood, while the blue dot is the MLE  from the approximate likelihood.
     1, 2 and 3-$\sigma$ contours are shown in white.
-             (\ref{fig:5a}) The exact log-likelihood for the observed data ${\cal D}$.
-             (\ref{fig:5b}) The approximate log-likelihood  evaluated on a coarse $15 \times 15$ fixed grid.
-             (\ref{fig:5c}) A Gaussian Process surrogate of the approximate log-likelihood ratio estimated from a Bayesian optimization procedure.
+             (\ref{fig:5a}) The exact $-2\log\Lambda(\alpha,\beta)$ for the observed data ${\cal D}$.
+             (\ref{fig:5b}) The approximate $-2\log\Lambda(\alpha,\beta)$ evaluated on a coarse $15 \times 15$  grid.
+             (\ref{fig:5c}) A Gaussian Process surrogate of $-2\log\Lambda(\alpha,\beta)$ ratio estimated from a Bayesian optimization procedure.
 White dots show the  parameter points sampled during the optimization process.}
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[height=.9\textheight]{figures/diagnostics.pdf}
-    \caption{Diagnostic plots ....}
-    \label{fig:ratio_comp}
-\end{figure}
+%\begin{figure}
+%    \centering
+%    \includegraphics[height=.9\textheight]{figures/diagnostics.pdf}
+%    \caption{Diagnostic plots ....}
+%    \label{fig:ratio_comp}
+%\end{figure}
 
@@
         \includegraphics[width=\textwidth]{figures/likelihood_comp_0.pdf}
-        \caption{$-2\log \Lambda(\alpha)$: poorly trained, well calibrated. }
+        \caption{Poorly trained, well calibrated. }
         \label{fig:6a}
@@
         \includegraphics[width=\textwidth]{figures/ROC_comp0.pdf}
-        \caption{ROC: poorly trained, well calibrated. }
+        \caption{Poorly trained, well calibrated. }
         \label{fig:6b}
@@
         \includegraphics[width=\textwidth]{figures/likelihood_comp_1.pdf}
-        \caption{$-2\log \Lambda(\alpha)$: poorly calibrated, well trained. }
+        \caption{Poorly calibrated, well trained. }
         \label{fig:6c}
@@
         \includegraphics[width=\textwidth]{figures/ROC_comp1.pdf}
-        \caption{ROC: poorly calibrated, well trained. }
+        \caption{Poorly calibrated, well trained. }
         \label{fig:6d}
@@
         \includegraphics[width=\textwidth]{figures/likelihood_comp_2.pdf}
-        \caption{$-2\log \Lambda(\alpha)$: well trained, well calibrated. }
+        \caption{Well trained, well calibrated. }
         \label{fig:6e}
@@
         \includegraphics[width=\textwidth]{figures/ROC_comp2.pdf}
-        \caption{ROC: well trained, well calibrated. }
+        \caption{Well trained, well calibrated. }
         \label{fig:6f}
@@
 
-    \caption{Alternative presentation of Diagnostic plots.}
+    \caption{Results from the diagnostics described in Sec.~\ref{S:diagnostics}.  
+    The rows correspond to the quality of the training and calibration of the classifier.
+The left plots probe the sensitivity to $\theta_1$, while the right plots show the ROC curve for a calibrator trained to discriminate samples from $p(\mathbf{x}|\theta_0)$ and samples from $p(\mathbf{x}|\theta_1)$ weighted as indicated in the legend.
+}
+
+%    \caption{Diagnostic plots for the approximate likelihood ratios.
+The left plots show slices of $-2\log\Lambda(\alpha,\beta=1)$ for the exact likelihood ratio and the approximate likelihood ratio with different choices of $\theta_1$.
+The right plots show the ROC curve for a calibrator trained to discriminate samples from $p(\mathbf{x}|\theta_0)$ and samples from $p(\mathbf{x}|\theta_1)$ weighted as indicated in the legend.
+}
     \label{fig:diagnostics}

### 2016-03-16 jasa/paper-jasa.tex

@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1a.pdf}
+        \includegraphics[height=10.5em]{figures/fig1a.pdf}
         \caption{$p(\mathbf{x}|\gamma)$  for $\gamma=0.05$ and $\gamma=0$}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \hspace{-.2em}\includegraphics[width=1.10\textwidth]{figures/fig1b.pdf}
+        \hspace{-.2em}\includegraphics[height=10.5em]{figures/fig1b.pdf}
         \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $1/(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))$}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1c.pdf}
+        \includegraphics[height=10.5em]{figures/fig1c.pdf}
         \caption{$\log \hat r(\hat{s}(\mathbf{x}))$ using neural network}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1d.pdf}
+        \includegraphics[height=10.5em]{figures/fig1d.pdf}
         \caption{$\log \hat r(\hat{s}(\mathbf{x}))$ using random forest}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig-ratio-dist.pdf}
+        \includegraphics[height=10.5em]{figures/fig-ratio-dist.pdf}
         \caption{$p( \log \hat r(\hat{s}(\mathbf{x}))) \, | \,\gamma = 0.05)$ }
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig-likelihood.pdf}
+        \includegraphics[height=10.5em]{figures/fig-likelihood.pdf}
         \caption{$-2\log \Lambda(\gamma)$ }
@@
 
-    \begin{subfigure}[t]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig2a.pdf}
-        \caption{True MLEs vs. approximated MLEs.}
+    \begin{subfigure}[t]{0.4\textwidth}
+        \includegraphics[height=10.5em]{figures/fig2a.pdf}
+        \caption{Exact vs. approximated MLEs.}
         \label{fig:2a}
@@
       %(or a blank line to force the subfigure onto a new line)
-    \begin{subfigure}[t]{0.45\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig2b.pdf}
+    \begin{subfigure}[t]{0.4\textwidth}
+        \includegraphics[height=10.5em]{figures/fig2b.pdf}
         \caption{$p(-2 \log \Lambda(\gamma=0.05) \, | \, \gamma=0.05)$}
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/likelihood_comp_0.pdf}
+        \includegraphics[height=10.5em]{figures/likelihood_comp_0.pdf}
         \caption{Poorly trained, well calibrated. }
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/ROC_comp0.pdf}
+        \includegraphics[height=10.5em]{figures/ROC_comp0.pdf}
         \caption{Poorly trained, well calibrated. }
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/likelihood_comp_1.pdf}
+        \includegraphics[height=10.5em]{figures/likelihood_comp_1.pdf}
         \caption{Poorly calibrated, well trained. }
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/ROC_comp1.pdf}
+        \includegraphics[height=10.5em]{figures/ROC_comp1.pdf}
         \caption{Poorly calibrated, well trained. }
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/likelihood_comp_2.pdf}
+        \includegraphics[height=10.5em]{figures/likelihood_comp_2.pdf}
         \caption{Well trained, well calibrated. }
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/ROC_comp2.pdf}
+        \includegraphics[height=10.5em]{figures/ROC_comp2.pdf}
         \caption{Well trained, well calibrated. }

### 2016-03-16 jasa/paper-jasa.tex
Should be good now? :s

--- 2016-03-16 jasa/paper-jasa.tex
@@
     \end{subfigure}
-    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc.
-      %(or a blank line to force the subfigure onto a new line)
-%    \begin{subfigure}[b]{0.45\textwidth}
-%        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
-%        \caption{Linear model}
-%        \label{fig:1b}
-%    \end{subfigure}
+    ~
     \begin{subfigure}[b]{0.4\textwidth}
-        \hspace{-.2em}\includegraphics[width=1.10\textwidth]{figures/fig1b.pdf}
+        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
         \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $1/(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))$}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig-ratio-dist.pdf}
+        \includegraphics[width=\textwidth]{figures/fig1e.pdf}
         \caption{$p( \log \hat r(\hat{s}(\mathbf{x}))) \, | \,\gamma = 0.05)$ }
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig-likelihood.pdf}
+        \includegraphics[width=\textwidth]{figures/fig1f.pdf}
         \caption{$-2\log \Lambda(\gamma)$ }
@@
 
-    \begin{subfigure}[t]{0.45\textwidth}
+    \begin{subfigure}[t]{0.4\textwidth}
         \includegraphics[width=\textwidth]{figures/fig2a.pdf}
-        \caption{True MLEs vs. approximated MLEs.}
+        \caption{Exact vs. approximated MLEs.}
         \label{fig:2a}
@@
       %(or a blank line to force the subfigure onto a new line)
-    \begin{subfigure}[t]{0.45\textwidth}
+    \begin{subfigure}[t]{0.4\textwidth}
         \includegraphics[width=\textwidth]{figures/fig2b.pdf}
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/likelihood_comp_0.pdf}
+        \includegraphics[height=10.075em]{figures/likelihood_comp_0.pdf}
         \caption{Poorly trained, well calibrated. }
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/ROC_comp0.pdf}
+        \includegraphics[height=9.5em]{figures/ROC_comp0.pdf}
         \caption{Poorly trained, well calibrated. }
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/likelihood_comp_1.pdf}
+        \includegraphics[height=10.075em]{figures/likelihood_comp_1.pdf}
         \caption{Poorly calibrated, well trained. }
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/ROC_comp1.pdf}
+        \includegraphics[height=9.5em]{figures/ROC_comp1.pdf}
         \caption{Poorly calibrated, well trained. }
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/likelihood_comp_2.pdf}
+        \includegraphics[height=10.075em]{figures/likelihood_comp_2.pdf}
         \caption{Well trained, well calibrated. }
@@
         \centering
-        \includegraphics[width=\textwidth]{figures/ROC_comp2.pdf}
+        \includegraphics[height=9.5em]{figures/ROC_comp2.pdf}
         \caption{Well trained, well calibrated. }

### 2016-03-16 jasa/paper-jasa.tex
text for diagnostic plots

@@
 This is complicated by the fact that in the likelihood-free setting, we don't have access to the true likelihood ratio.
-Below we consider two such diagnostic procedures that can be implemented in the likelihood-free setting.
+Below we consider two such diagnostic procedures that can be implemented in the likelihood-free setting. We illustrate these diagnostic procedures in Fig.~\ref{fig:diagnostics}.
 
@@
 %$\log\Lambda(\theta) = \log p(\mathcal{D}|\theta)/p(\mathcal{D}|\hat\theta) = \log p(\mathcal{D}|\theta)/p(\mathcal{D}|\theta_1) -\log p(\mathcal{D}|\hat\theta)/p(\mathcal{D}|\theta_1)$ for all values of $\theta_1$.
-\begin{equation}
+\begin{equation}\label{eq:theta_1_independence}
 \log\Lambda(\theta) = \log \frac{p(\mathcal{D}|\theta)}{p(\mathcal{D}|\hat\theta)} = \log \frac{p(\mathcal{D}|\theta)}{p(\mathcal{D}|\theta_1)} -\log \frac{p(\mathcal{D}|\hat\theta)}{p(\mathcal{D}|\theta_1)}
@@
 The two situations can be disentangled to some degree by training another classifier to distinguish between an unweighted distribution of samples from $p(\mathbf{x}|\theta_1)$.
-
-We illustrate these diagnostic procedures in Fig.~\ref{fig:ratio_comp}.
 
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1a.pdf}
+        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm, width=\textwidth]{figures/fig1a.pdf}
         \caption{$p(\mathbf{x}|\gamma)$  for $\gamma=0.05$ and $\gamma=0$}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1b.pdf}
+        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1b.pdf}
         \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $1/(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))$}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1c.pdf}
+        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1c.pdf}
         \caption{$\log \hat r(\hat{s}(\mathbf{x}))$ using neural network}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1d.pdf}
+        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1d.pdf}
         \caption{$\log \hat r(\hat{s}(\mathbf{x}))$ using random forest}
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1e.pdf}
+        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1e.pdf}
         \caption{$p( \log \hat r(\hat{s}(\mathbf{x}))) \, | \,\gamma = 0.05)$ }
@@
     \begin{subfigure}[b]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig1f.pdf}
+        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1f.pdf}
         \caption{$-2\log \Lambda(\gamma)$ }
@@
     \begin{subfigure}[t]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig2a.pdf}
+        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig2a.pdf}
         \caption{Exact vs. approximated MLEs.}
@@
     \begin{subfigure}[t]{0.4\textwidth}
-        \includegraphics[width=\textwidth]{figures/fig2b.pdf}
+        \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig2b.pdf}
         \caption{$p(-2 \log \Lambda(\gamma=0.05) \, | \, \gamma=0.05)$}
@@
 After $50$ likelihood evaluations, the maximum likelihood estimate is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$.
-While the goal of the Bayesian optimization procedure is to find the maximum likelihood, the posterior mean of the
+While the objective of the Bayesian optimization procedure is to find the maximum likelihood, the posterior mean of the
 internal Gaussian process, shown in Fig.~\ref{fig:5c}, is close to the exact log-likelihood ratio illustrated in Fig.~\ref{fig:5a}.
@@
 %roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.
+
+Finally, we evaluate the diagnostics described in Sec.~\ref{S:diagnostics} for this example.
+To aid in visualization, we restrict to a 1-dimensional slice of the likelihood along $\alpha$ with $\beta=1$.
+We consider three situations: i) a poorly trained, but well calibrated classifier; ii) a well trained, but poorly calibrated classifier; and iii) a well trained, and well calibrated classifier.
+For each case, we employ two diagnostic tests.
+The first checks for independence of $-2\log\Lambda(\theta)$ with respect to changes in the reference value $\theta_1$ as shown in Eqn.~\ref{eq:theta_1_independence}.
+The second uses a classifier to distinguish between samples from  $p(\mathbf{x}|\theta_0)$ and samples from $p(\mathbf{x}|\theta_1)$ weighted according to $r(\mathbf{x}; \theta_0, \theta_1)$.
+As discussed for Fig.~\ref{fig:5b}, statistical fluctuations in the calibration lead to some noise in the raw approximate likelihood.
+Thus, we show the posterior mean of a Gaussian processes resulting from Bayesian optimization of  the raw approximate likelihood as in Fig.~\ref{fig:5c}.
+In addition, the standard deviation of the Gaussian process is shown for one of the $\theta_1$ reference points to indicate the size of these statistical fluctuations.
+It is clear that in the well calibrated cases that these fluctuations are small, while in the poorly calibrated case these fluctuations are large.
+Moreover, in Fig.~\ref{fig:5a} we see that in the poorly trained, well calibrated case the classifier $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$ has a significant dependence on the $\theta_1$ reference point.
+In contrast, in Fig.~\ref{fig:5c} the likelihood curves vary significantly, but this is comparable to the fluctuations expected from the calibration procedure.
+Finally, Fig.~\ref{fig:5e} shows that in the well trained, well calibrated case that the likelihood curves are all consistent with the exact likelihood within the estimated uncertainty band of the Gaussian process.
+The ROC curves tell a similarly revealing story.
+As expected, the classifier is not able to distinguish between the distributions when $p(\mathbf{x}|\theta_1)$ is weighted by the exact likelihood ratio.
+We can also rule out that this is a deficiency in the classifier because the two distributions are well separated when no weights are applied to $p(\mathbf{x}|\theta_1)$.
+In both Fig.~\ref{fig:5b} and Fig.~\ref{fig:5d} the ROC curve correctly diagnoses deficiencies in the approximate likelihood ratio  $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$.
+Finally, Fig.~\ref{fig:5f} shows that the ROC curve in the well trained, well calibrated case is almost identical with the exact likelihood ratio, confirming the quality of the approximation.
 
@@
         \centering
-        \includegraphics[height=10.075em]{figures/likelihood_comp_0.pdf}
+        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=10.075em]{figures/likelihood_comp_0.pdf}
         \caption{Poorly trained, well calibrated. }
@@
         \centering
-        \includegraphics[height=9.5em]{figures/ROC_comp0.pdf}
+        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=9.5em]{figures/ROC_comp0.pdf}
         \caption{Poorly trained, well calibrated. }
@@
         \centering
-        \includegraphics[height=10.075em]{figures/likelihood_comp_1.pdf}
+        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=10.075em]{figures/likelihood_comp_1.pdf}
         \caption{Poorly calibrated, well trained. }
@@
         \centering
-        \includegraphics[height=9.5em]{figures/ROC_comp1.pdf}
+        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=9.5em]{figures/ROC_comp1.pdf}
         \caption{Poorly calibrated, well trained. }
@@
         \centering
-        \includegraphics[height=10.075em]{figures/likelihood_comp_2.pdf}
+        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=10.075em]{figures/likelihood_comp_2.pdf}
         \caption{Well trained, well calibrated. }
@@
         \centering
-        \includegraphics[height=9.5em]{figures/ROC_comp2.pdf}
+        \includegraphics[clip, trim=0.3cm 0.3cm 0.3cm 0.3cm,height=9.5em]{figures/ROC_comp2.pdf}
         \caption{Well trained, well calibrated. }

### 2016-03-17 jasa/paper-jasa.tex
remove redundant equations

@@
 for the dataset $\mathcal{D}$ shown in  Fig.~\ref{fig:1a}.
-By exploiting the fact that
-\begin{equation}
-\log \frac{ p({\cal D} | \gamma ) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma ) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
-\end{equation}
+By exploiting Eq.~\ref{eq:theta_1_independence},
+% fact that
+%\begin{equation}
+%\log \frac{ p({\cal D} | \gamma ) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma ) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
+%\end{equation}
 the generalized likelihood ratio can be computed by evaluating
@@
 
-    \begin{subfigure}[t]{0.4\textwidth}
+    \begin{subfigure}[t]{0.475\textwidth}
         \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig2a.pdf}
@@
       %(or a blank line to force the subfigure onto a new line)
-    \begin{subfigure}[t]{0.4\textwidth}
+    \begin{subfigure}[t]{0.475\textwidth}
         \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig2b.pdf}
@@
 \end{equation}
-that we calculate by exploiting the fact that
-\begin{equation}\label{eqn:obj-ex2-expanded}
-\log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) } = \log \frac{p({\cal D} | \alpha, \beta)}{ p({\cal D} | \alpha=0, \beta=0) } - \log \frac{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }{ p({\cal D} | \alpha=0, \beta=0)}.
-\end{equation}
+that we calculate via Eqn.~\ref{eq:theta_1_independence}.
+%by exploiting the fact that
+%\begin{equation}\label{eqn:obj-ex2-expanded}
+%\log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) } = \log \frac{p({\cal D} | \alpha, \beta)}{ p({\cal D} | \alpha=0, \beta=0) } - \log \frac{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }{ p({\cal D} | \alpha=0, \beta=0)}.
+%\end{equation}
 Following the procedure described in Section~\ref{sec:param-clf}, we

### 2016-03-17 jasa/paper-jasa.tex
fix figure refs

@@
 As discussed for Fig.~\ref{fig:5b}, statistical fluctuations in the calibration lead to some noise in the raw approximate likelihood.
-Thus, we show the posterior mean of a Gaussian processes resulting from Bayesian optimization of  the raw approximate likelihood as in Fig.~\ref{fig:5c}.
+Thus, we show the posterior mean of a Gaussian processes resulting from Bayesian optimization of  the raw approximate likelihood as in Fig.~\ref{fig:6c}.
 In addition, the standard deviation of the Gaussian process is shown for one of the $\theta_1$ reference points to indicate the size of these statistical fluctuations.
 It is clear that in the well calibrated cases that these fluctuations are small, while in the poorly calibrated case these fluctuations are large.
-Moreover, in Fig.~\ref{fig:5a} we see that in the poorly trained, well calibrated case the classifier $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$ has a significant dependence on the $\theta_1$ reference point.
-In contrast, in Fig.~\ref{fig:5c} the likelihood curves vary significantly, but this is comparable to the fluctuations expected from the calibration procedure.
-Finally, Fig.~\ref{fig:5e} shows that in the well trained, well calibrated case that the likelihood curves are all consistent with the exact likelihood within the estimated uncertainty band of the Gaussian process.
+Moreover, in Fig.~\ref{fig:6a} we see that in the poorly trained, well calibrated case the classifier $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$ has a significant dependence on the $\theta_1$ reference point.
+In contrast, in Fig.~\ref{fig:6c} the likelihood curves vary significantly, but this is comparable to the fluctuations expected from the calibration procedure.
+Finally, Fig.~\ref{fig:6e} shows that in the well trained, well calibrated case that the likelihood curves are all consistent with the exact likelihood within the estimated uncertainty band of the Gaussian process.
 The ROC curves tell a similarly revealing story.
@@
 We can also rule out that this is a deficiency in the classifier because the two distributions are well separated when no weights are applied to $p(\mathbf{x}|\theta_1)$.
-In both Fig.~\ref{fig:5b} and Fig.~\ref{fig:5d} the ROC curve correctly diagnoses deficiencies in the approximate likelihood ratio  $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$.
-Finally, Fig.~\ref{fig:5f} shows that the ROC curve in the well trained, well calibrated case is almost identical with the exact likelihood ratio, confirming the quality of the approximation.
+In both Fig.~\ref{fig:5b} and Fig.~\ref{fig:6d} the ROC curve correctly diagnoses deficiencies in the approximate likelihood ratio  $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$.
+Finally, Fig.~\ref{fig:6f} shows that the ROC curve in the well trained, well calibrated case is almost identical with the exact likelihood ratio, confirming the quality of the approximation.
 
@@
 are often difficult to construct.
-
-%\subsubsection{Diagnostics}
-
-%\begin{figure}
-%    \centering
-%    \includegraphics[height=.9\textheight]{figures/diagnostics.pdf}
-%    \caption{Diagnostic plots ....}
-%    \label{fig:ratio_comp}
-%\end{figure}
 

### 2016-03-17 jasa/paper-jasa.tex
introduce Omega_u notation

@@
 \begin{equation}
-p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) = \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
+p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) = \int_{\mathbf{x}^\prime \in \Omega_u } \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
 \end{equation}
-where $|\nabla s(\mathbf{x}^\prime)| = \sqrt{\sum_{i=1}^p |\frac{\partial}{\partial x_i} s(\mathbf{x}^\prime)|^2}$
-and where $dS_{\mathbf{x}^\prime}$ is the Euclidean surface measure on $\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}$.
+where $\Omega_u = \{x^\prime : s(\mathbf{x}^\prime)=u\}$, $|\nabla s(\mathbf{x}^\prime)| = \sqrt{\sum_{i=1}^p |\frac{\partial}{\partial x_i} s(\mathbf{x}^\prime)|^2}$
+and where $dS_{\mathbf{x}^\prime}$ is the Euclidean surface measure on $\Omega_u$.
 Also, since $s(\mathbf{x})$ is monotonic with
@@
 \begin{align}
-p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
-&= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(u) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
-&= m^{-1}(s(\mathbf{x})) \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
-&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}.
+p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+&= \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(u) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+&= m^{-1}(s(\mathbf{x})) \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
+&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}.
 \label{eqn:factorization}
@@
 \begin{align}
-\frac{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0)}{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
+\frac{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0)}{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
 &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}. \label{eqn:jacob}

### 2016-03-17 jasa/paper-jasa.tex
introduce Omega_u notation

@@
 % used to approximate the density at various parameter points, and an
-% interpolation algorithm is used to approximate the parametrized
+% interpolation algorithm is used to approximate the parameterized
 % model~\citep{Cranmer:2012sba}.
@@
 on a high dimensional feature vector $\mathbf{x}$. The results of this paper outline
-how to extend the use of discriminative classifiers for composite hypotheses (parametrized by $\theta$)
+how to extend the use of discriminative classifiers for composite hypotheses (parameterized by $\theta$)
 in a way
@@
 % separates the quality of the approximation of the target likelihood from the
-% quality of the calibration.  In Section~\ref{S:Related} we discuss the scheme
+% quality of the calibration.  In Sec.~\ref{S:Related} we discuss the scheme
 % sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a
@@
 The rest of the paper is organized as follows. In
-Section~\ref{sec:likelihood-ratio-tests}, we first introduce the likelihood
+Sec.~\ref{sec:likelihood-ratio-tests}, we first introduce the likelihood
 ratio test statistic in the setting of simple hypothesis testing, and then
 outline how it can be computed exactly using calibrated classifiers.
-In Section~\ref{sec:generalized-likelihood-ratio}, we generalize the proposed
+In Sec.~\ref{sec:generalized-likelihood-ratio}, we generalize the proposed
 approach to the case of composite hypothesis testing and discuss directions for
 approximating the statistic efficiently. We then illustrate the proposed
-method in Section~\ref{sec:examples} and outline how it could improve
+method in Sec.~\ref{sec:examples} and outline how it could improve
 statistical analysis within the field of high energy physics. Related work
@@
 % \item $y$: a class label used for training a classifier.
-% \item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
+% \item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parameterized by $\theta_0$ and $\theta_1$
 % %\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
@@
 % from $p(x|\theta_0)$ and $(x,u=1)$ generated from $p(x|\theta_1)$.
-In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the classifier  parametrized in terms of $(\theta_0, \theta_1)$.
+In Sec.~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the classifier  parameterized in terms of $(\theta_0, \theta_1)$.
 %
@@
 $s(\mathbf{x})$ as a discriminative classifier.  In
-Section~\ref{sec:generalized-likelihood-ratio} we extend this result to
+Sec.~\ref{sec:generalized-likelihood-ratio} we extend this result to
 generalized likelihood ratio tests, where it will be useful to have the
@@
 % p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
-p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{du}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq u\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
+p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{du}  \int_{\{\mathbf{x}^\prime : s(\mathbf{x}^\prime) \leq u\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
 % &= \frac{d}{dy} \int_{\mathbb{R}^p} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
@@
 from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies the conditions of
-Theorem~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s} for further details).
+Thm.~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s}).
 In other words, supervised learning yields a sufficient
-procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
+procedure for Thm.~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
 {\it universally strongly consistent} algorithm can be used for learning $s^*$.
 Note however, that it is not a necessary procedure since
-Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
+Thm.~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
 density ratio,  not only for $m(r(\mathbf{x})) = (1 +
@@
 Equivalently,
-Theorem~\ref{thm:ratio-equivalence} shows that in the case that we learn a probabilistic
+Thm.~\ref{thm:ratio-equivalence} shows that in the case that we learn a
 classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
@@
 This is done by inverting the relationship ${r}(\mathbf{x}) = (1-s^*(\mathbf{x}))/s^*(\mathbf{x})$ to obtain $\hat{r}(\hat{s}(\mathbf{x})) = (1-\hat{s}_\textrm{iso}(\mathbf{x}))/\hat{s}_\textrm{iso}(\mathbf{x})$.
-Additionally, Sec.~\ref{sec:related} describes related work in which the density ratio $\hat{r}(\mathbf{x})$ is estimated directly on the feature space $\mathcal{X}$.
+Additionally, Sec.~\ref{sec:related} describes related work in which the ratio $\hat{r}(\mathbf{x})$ is estimated directly on the feature space $\mathcal{X}$.
 
@@
 % continuous model parameters. We postpone the practicalities of training the
-% classifier and estimating the density to Section~\ref{S:classifier} and continue
+% classifier and estimating the density to Sec.~\ref{S:classifier} and continue
 % in the likelihood-free setting with idealized classifiers and their densities.
@@
 % for the observed data and never refer back to the original features $x_e$. In
-% the parametrized setting it is not possible to pre-compute $s(x_e; \theta_0,
+% the parameterized setting it is not possible to pre-compute $s(x_e; \theta_0,
 % \theta_1)$ since $\theta_0$ and $\theta_1$ are unknown.
@@
 This allows us to
-use Theorem~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
+use Thm.~\ref{thm:ratio-equivalence} to reformulate the maximum likelihood
 estimate (MLE) as
@@
 and denominator of Eqn.~\ref{eqn:generalized-lr}, the generalized likelihood
-ratio can be estimated as outlined in Section~\ref{sec:approx}
+ratio can be estimated as outlined in Sec.~\ref{sec:approx}
 for simple hypothesis testing.
@@
 % For completeness, the proposed method is summarized in
-% Algorithm~\ref{alg:training}, while further details for training and calibrating
+% Alg.~\ref{alg:training}, while further details for training and calibrating
 % efficiently a family $s(\mathbf{x};\theta_0,\theta_1)$ of classifiers are
@@
 be done independently for all $\theta_0$ and $\theta_1$, using the procedure
-outlined in Section~\ref{sec:likelihood-ratio-tests}, it is desirable and
+outlined in Sec.~\ref{sec:likelihood-ratio-tests}, it is desirable and
 convenient to have a smooth evolution of the classification score as a function
@@
 on training data with input $(\mathbf{x}, \theta_0, \theta_1)_i$ and target
-$y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
+$y_i$, as outlined in Alg.~\ref{alg:training}. Somewhat unusually, the
 unknown values of the parameters are taken as input to the classifier; their
@@
 learned for the parameter values $\theta_0$ and $\theta_1$ selected in
-Algorithm~\ref{alg:training}, it is not clear whether the optimal decision
+Alg.~\ref{alg:training}, it is not clear whether the optimal decision
 function can be expected for data generated from  $\theta'_0$ and $\theta'_1$
@@
  however, we introduce the distributions $\pi_{\Theta_0}$ and $\pi_{\Theta_1}$
-into the Algorithm~\ref{alg:training} to allow for a more
+into the Alg.~\ref{alg:training} to allow for a more
  sophisticated sampling strategy.
-% this formulation of Algorithm~\ref{alg:training} allows for a more
+% this formulation of Alg.~\ref{alg:training} allows for a more
 % sophisticated sampling strategy.
@@
 %For this reason, we introduce the distributions $\pi_{\Theta_0}$ and $\pi_{\Theta_1}$
-%into the Algorithm~\ref{alg:training} to allow for a more
+%into the Alg.~\ref{alg:training} to allow for a more
 % sophisticated sampling strategy.
@@
 
-Once the parametrized classifier $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$ is trained,
+Once the parameterized classifier $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$ is trained,
 we can use the generative model together with one of the calibration strategies
@@
 $p(s | \theta)$~\citep{read1999linear,Cranmer:2012sba,baak2015interpolation}.
-One can easily imagine a number of approaches to parametrized calibration and the
+One can easily imagine a number of approaches to parameterized calibration and the
 relative merits of those approaches will depend critically on the dimensionality of $\theta$ and the
@@
 % pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features
-% $x_e$. In the parametrized setting it is not possible to pre-compute $\hat
+% $x_e$. In the parameterized setting it is not possible to pre-compute $\hat
 % s(x_e; \theta_0, \theta_1)$ since one does not know the true values of
@@
 %    * compare with other methods (new)
-% - ratio of mixtures and decomposition (special case of generalized likelihood rario tests when parameters only affect the mixture coefficients => no need for parametrized classifiers)
+% - ratio of mixtures and decomposition (special case of generalized likelihood rario tests when parameters only affect the mixture coefficients => no need for parameterized classifiers)
 %   discuss about focusing the capacity of the classifier
@@
 % The second line is a trivial, but useful decomposition into pair-wise classification between $p_{c'}(x|\theta_1)$ and $p_c(x|\theta_0)$.
-The third line uses Theorem~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
+The third line uses Thm.~1 to relate the high-dimensional likelihood ratio into an equivalent calibrated likelihood ratio based on the univariate density of the corresponding classifier, denoted $s_{c,c',\theta_0, \theta_1}$.
 In the situation where the only free parameters of the  model are the mixture coefficients $w_c$, then the distributions $p_{c}(s_{c,c',\theta_0, \theta_1}| \theta)$ are independent of $\theta$ and can be pre-computed (after training the discriminative classifier, but before evaluating the  likelihood ratio).
@@
 $p_c(\mathbf{x}|\theta_0)$.  The third line uses
-Theorem~\ref{thm:ratio-equivalence} to relate the high-dimensional likelihood
+Thm.~\ref{thm:ratio-equivalence} to relate the high-dimensional likelihood
 ratio into an equivalent calibrated likelihood ratio based on the univariate
@@
 
-While Theorem~\ref{thm:ratio-equivalence} states that the likelihood ratio $r(\mathbf{x}; \theta_0, \theta_1)$ is
+While Thm.~\ref{thm:ratio-equivalence} states that the likelihood ratio $r(\mathbf{x}; \theta_0, \theta_1)$ is
 invariant under the dimensionality reduction $s(\mathbf{x}; \theta_0, \theta_1)$ provided that it is monotonic with
@@
 As pointed out there it is important that one maximizes the likelihood ratio as the surface integral and Jacobian factors related to the dimensionality reduction only cancel in the ratio (see Eqn.~\ref{eqn:factorization}).
-Importantly, they also only cancel if the dimensionality reduction map satisfies the conditions needed for Theorem~\ref{thm:ratio-equivalence} to hold.
+Importantly, they also only cancel if the reduction map satisfies the assumptions of Thm.~\ref{thm:ratio-equivalence}.
 Moreover, the resulting value of $\hat{\theta}$ should be independent of the value of $\theta_1$ used in the denominator of the likelihood ratio.
@@
 
-Figure~\ref{fig:p_sij} shows the intermediate stages for the decomposition described in Section~\ref{sec:mixture}.
+Figure~\ref{fig:p_sij} shows the intermediate stages for the decomposition described in Sec.~\ref{sec:mixture}.
 The blue and green curves show $p_{c'}(\hat{s}_{c,c'}(\mathbf{x}))$,  the distributions for the score for the sub-classifiers for the three pair-wise comparisons of the mixture components.
@@
 
-%Using the method outlined in Section~\ref{sec:approx},
+%Using the method outlined in Sec.~\ref{sec:approx},
 
@@
 % When leveraging the fact
-%that densities are mixtures and decomposing their ratio as described in Section~\ref{sec:mixture},
+%that densities are mixtures and decomposing their ratio as described in Sec.~\ref{sec:mixture},
 %the capacity of the underlying classifiers can be more effectively focused on easier supervised learning tasks,
@@
 of the exact log-likelihood ratio.
-While there are some artifacts in the distribution in the low-probability regions and the maximum value of the log-likehood ratio is underestimated, the overall shape of the distribution is well approximated.
+While there are some artifacts in the distribution in the low-probability regions and the maximum value of the log-likelihood ratio is underestimated, the overall shape of the distribution is well approximated.
 
@@
 for the dataset $\mathcal{D}$ shown in  Fig.~\ref{fig:1a}.
-By exploiting Eq.~\ref{eq:theta_1_independence},
+By exploiting Eqn.~\ref{eq:theta_1_independence},
 % fact that
@@
 %can be found
-as outlined in Section~\ref{sec:generalized-likelihood-ratio}.
+as outlined in Sec.~\ref{sec:generalized-likelihood-ratio}.
 Figure~\ref{fig:likelihood} shows that the exact likelihood curve is very well
@@
 %can be found as outlined in
-%Section~\ref{sec:generalized-likelihood-ratio}.
+%Sec.~\ref{sec:generalized-likelihood-ratio}.
 
@@
 %\end{equation}
-Following the procedure described in Section~\ref{sec:param-clf}, we
+Following the procedure described in Sec.~\ref{sec:param-clf}, we
 % to reduce
@@
 In addition to the statistical fluctuations due to finite calibration samples, there are also fluctuations introduced from changes in the binning of the calibration histograms as $\alpha$ and $\beta$ vary.
-As discussed in Section~\ref{sec:param-calibration}, a parameterized calibration
+As discussed in Sec.~\ref{sec:param-calibration}, a parameterized calibration
 procedure should ameliorate this issue, but that is left for now as an area for future work.
@@
 %roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.
-
-Finally, we evaluate the diagnostics described in Sec.~\ref{S:diagnostics} for this example.
-To aid in visualization, we restrict to a 1-dimensional slice of the likelihood along $\alpha$ with $\beta=1$.
-We consider three situations: i) a poorly trained, but well calibrated classifier; ii) a well trained, but poorly calibrated classifier; and iii) a well trained, and well calibrated classifier.
-For each case, we employ two diagnostic tests.
-The first checks for independence of $-2\log\Lambda(\theta)$ with respect to changes in the reference value $\theta_1$ as shown in Eqn.~\ref{eq:theta_1_independence}.
-The second uses a classifier to distinguish between samples from  $p(\mathbf{x}|\theta_0)$ and samples from $p(\mathbf{x}|\theta_1)$ weighted according to $r(\mathbf{x}; \theta_0, \theta_1)$.
-As discussed for Fig.~\ref{fig:5b}, statistical fluctuations in the calibration lead to some noise in the raw approximate likelihood.
-Thus, we show the posterior mean of a Gaussian processes resulting from Bayesian optimization of  the raw approximate likelihood as in Fig.~\ref{fig:6c}.
-In addition, the standard deviation of the Gaussian process is shown for one of the $\theta_1$ reference points to indicate the size of these statistical fluctuations.
-It is clear that in the well calibrated cases that these fluctuations are small, while in the poorly calibrated case these fluctuations are large.
-Moreover, in Fig.~\ref{fig:6a} we see that in the poorly trained, well calibrated case the classifier $\hat{s}(\mathbf{x}; \theta_0, \theta_1)$ has a significant dependence on the $\theta_1$ reference point.
-In contrast, in Fig.~\ref{fig:6c} the likelihood curves vary significantly, but this is comparable to the fluctuations expected from the calibration procedure.
-Finally, Fig.~\ref{fig:6e} shows that in the well trained, well calibrated case that the likelihood curves are all consistent with the exact likelihood within the estimated uncertainty band of the Gaussian process.
-The ROC curves tell a similarly revealing story.
-As expected, the classifier is not able to distinguish between the distributions when $p(\mathbf{x}|\theta_1)$ is weighted by the exact likelihood ratio.
-We can also rule out that this is a deficiency in the classifier because the two distributions are well separated when no weights are applied to $p(\mathbf{x}|\theta_1)$.
-In both Fig.~\ref{fig:5b} and Fig.~\ref{fig:6d} the ROC curve correctly diagnoses deficiencies in the approximate likelihood ratio  $\hat{r}(\hat{s}(\mathbf{x}; \theta_0, \theta_1))$.
-Finally, Fig.~\ref{fig:6f} shows that the ROC curve in the well trained, well calibrated case is almost identical with the exact likelihood ratio, confirming the quality of the approximation.
-
-Overall, this example further illustrates and confirms the ability
-of the proposed method for inference with multiple parameters and multi-dimensional data where
-reliable approximations $\hat p(\mathbf{x}|\theta_0)$ and $\hat p(\mathbf{x}|\theta_1)$
-are often difficult to construct.
 
@@
 
+Finally, we evaluate the diagnostics described in Sec.~\ref{S:diagnostics} for
+this example. To aid in visualization, we restrict to a 1-dimensional slice of
+the likelihood along $\alpha$ with $\beta=1$.   We consider three situations: i)
+a poorly trained, but well calibrated classifier; ii) a well trained, but poorly
+calibrated classifier; and iii) a well trained, and well calibrated classifier.
+For each case, we employ two diagnostic tests. The first checks for independence
+of $-2\log\Lambda(\theta)$ with respect to changes in the reference value
+$\theta_1$ as shown in Eqn.~\ref{eq:theta_1_independence}. The second uses a
+classifier to distinguish between samples from  $p(\mathbf{x}|\theta_0)$ and
+samples from $p(\mathbf{x}|\theta_1)$ weighted according to $r(\mathbf{x};
+\theta_0, \theta_1)$.  As discussed for Fig.~\ref{fig:5b}, statistical
+fluctuations in the calibration lead to some noise in the raw approximate
+likelihood. Thus, we show the posterior mean of a Gaussian processes resulting
+from Bayesian optimization of  the raw approximate likelihood as in
+Fig.~\ref{fig:6c}. In addition, the standard deviation of the Gaussian process
+is shown for one of the $\theta_1$ reference points to indicate the size of
+these statistical fluctuations. It is clear that in the well calibrated cases
+that these fluctuations are small, while in the poorly calibrated case these
+fluctuations are large. Moreover, in Fig.~\ref{fig:6a} we see that in the poorly
+trained, well calibrated case the classifier $\hat{s}(\mathbf{x}; \theta_0,
+\theta_1)$ has a significant dependence on the $\theta_1$ reference point. In
+contrast, in Fig.~\ref{fig:6c} the likelihood curves vary significantly, but
+this is comparable to the fluctuations expected from the calibration procedure.
+Finally, Fig.~\ref{fig:6e} shows that in the well trained, well calibrated case
+that the likelihood curves are all consistent with the exact likelihood within
+the estimated uncertainty band of the Gaussian process.  The ROC curves tell a
+similarly revealing story. As expected, the classifier is not able to
+distinguish between the distributions when $p(\mathbf{x}|\theta_1)$ is weighted
+by the exact likelihood ratio. We can also rule out that this is a deficiency in
+the classifier because the two distributions are well separated when no weights
+are applied to $p(\mathbf{x}|\theta_1)$. In both Fig.~\ref{fig:6b} and
+Fig.~\ref{fig:6d} the ROC curve correctly diagnoses deficiencies in the
+approximate likelihood ratio  $\hat{r}(\hat{s}(\mathbf{x}; \theta_0,
+\theta_1))$. Finally, Fig.~\ref{fig:6f} shows that the ROC curve in the well
+trained, well calibrated case is almost identical with the exact likelihood
+ratio, confirming the quality of the approximation.
+
+Overall, this example further illustrates and confirms the ability
+of the proposed method for inference with multiple parameters and multi-dimensional data where
+reliable approximations $\hat p(\mathbf{x}|\theta_0)$ and $\hat p(\mathbf{x}|\theta_1)$
+are often difficult to construct.
+
 \subsection{High energy physics}
@@
 \end{equation}
-Accordingly, new particle searches at the LHC are typically framed as
+Accordingly, new particle searches are typically framed as
 hypothesis tests where the null corresponds to $\mu=0$, and the generalized
@@
 %$\frac{p(\mathbf{x}|\mu=0,\nu_0)}{p(\mathbf{x}|\mu,\nu_0)}$ for all $\mu$ and therefore
-%satisfies conditions of Theorem~\ref{thm:ratio-equivalence}. Associating the
+%satisfies conditions of Thm.~\ref{thm:ratio-equivalence}. Associating the
 %$y=1$ label to the signal component has advantages because it helps the
@@
 $\nu$. Thus, typically we take the classifier $\hat s(\mathbf{x})$ as fixed and then propagate
-uncertainty by estimating $\hat{p}_s(\hat s(\mathbf{x}) | \nu)$ with a parametrized calibration procedure.
+uncertainty by estimating $\hat{p}_s(\hat s(\mathbf{x}) | \nu)$ with a parameterized calibration procedure.
 %Building such
@@
 % \end{equation}
-% following the approach outlined in Section~\ref{S:GLR}.
+% following the approach outlined in Sec.~\ref{S:GLR}.
 %
@@
 $\hat{\theta}(\mathbf{x})$ proposed by \cite{Neal:2007zz} and leads to
-Theorem~\ref{thm:ratio-equivalence}, which is an exact result in the case of an
+Thm.~\ref{thm:ratio-equivalence}, which is an exact result in the case of an
 ideal classifier. In both cases, the quality of the classifier is factorized
@@
 In this appendix, we show for completeness that the probabilistic classification framework
-yields a reduction $s$ which satisfies conditions of Theorem~\ref{thm:ratio-equivalence}.
+yields a reduction $s$ which satisfies conditions of Thm.~\ref{thm:ratio-equivalence}.
 
@@
 In particular, this regression function satisfies conditions of
-Theorem~\ref{thm:ratio-equivalence} since $s^*(\mathbf{x}) =
+Thm.~\ref{thm:ratio-equivalence} since $s^*(\mathbf{x}) =
 m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$,

### 2016-03-17 jasa/paper-jasa.tex
update acknowledgements

@@
 
-\section*{Acknowledgments} KC and GL are both supported through NSF ACI-1450310.
+\section*{Acknowledgments} KC and GL are both supported through NSF ACI-1450310, 
+additionally KC is supported through PHY-1505463 and PHY-1205376.
+JP was partially supported by the Scientific and Technological Center of Valparaiso (CCTVal) under Fondecyt grant BASAL FB0821.
 KC would like to thank Daniel Whiteson for

### 2016-03-17 jasa/paper-jasa.tex
minor changes

@@
         \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1e.pdf}
-        \caption{$p( \log \hat r(\hat{s}(\mathbf{x}))) \, | \,\gamma = 0.05)$ }
+        \caption{$p( \log \hat r(\hat{s}(\mathbf{x})) \, | \,\gamma = 0.05)$ }
         \label{fig:decomposed_ratio}
@@
 
-    \caption{Example data $\mathcal{D}$ Various stages in Approximation of the log-likelihood ratio
-             $\log r(\mathbf{x};\gamma=0.05,\gamma=0)$
-             with calibrated classifiers. }
+    \caption{Histogram of $\mathcal{D}$ generated from $\gamma=0.05$ and plots illustrating various stages in 
+    the approximation of the log-likelihood ratio  $\log r(\mathbf{x};\gamma=0.05,\gamma=0)$
+             with calibrated classifiers (see text). }
 \end{figure}
@@
     \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator
-            $\hat \gamma$, as empirically estimated 1000 times over re-sampled observed data.}
+            $\hat \gamma$, as empirically estimated from an ensemble of 1000 artificial datasets.}
     \label{fig:2}

### 2016-03-17 jasa/paper-jasa.tex
fixed typo, clarified 1b

@@
 The blue and green curves show $p_{c'}(\hat{s}_{c,c'}(\mathbf{x}))$,  the distributions for the score for the sub-classifiers for the three pair-wise comparisons of the mixture components.
-The red curves in Fig.~\ref{fig:p_sij} show the approximation of the density ratio obtained from those distributions.
+The red curves in Fig.~\ref{fig:p_sij} show the approximation of the density ratio (rescaled as $(1 +
+\hat{r}(\mathbf{x}))^{-1}$) obtained from those distributions.
 
@@
 
-Figures~\ref{fig:1c} and \ref{fig:1d} show the approximate $\log \left( \hat{r}(\mathbf{x}) \right)$ as a function of $\mathbf{x}$ using a
+Figures~\ref{fig:1c} and \ref{fig:1d} show the approximate $\log  \hat{r}(\mathbf{x}) $ as a function of $\mathbf{x}$ using a
 2-layer neural network and a random forest for the classifier $\hat{s}(\mathbf{x})$.
@@
 %forest.
-The blue curves show the exact log-ratio, the green curves show
+The blue curves show the exact $\log {r}(\mathbf{x})$, the green curves show
 $\log \left( (1-\hat s(\mathbf{x}))/\hat s(\mathbf{x}) \right)$
@@
         \includegraphics[clip, trim=0.5cm 0.5cm 0.5cm 0.5cm,width=\textwidth]{figures/fig1b.pdf}
-        \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $1/(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))$}
+        \caption{$\hat{p}_c(\hat{s}_{c,c^\prime}(\mathbf{x}) )$ and $(1+\hat{r}(\hat{s}_{c,c^\prime}(\mathbf{x}))^{-1}$}
         \label{fig:p_sij}
@@
 this example. To aid in visualization, we restrict to a 1-dimensional slice of
-the likelihood along $\alpha$ with $\beta=1$.   We consider three situations: i)
+the likelihood along $\alpha$ with $\beta=-1$.   We consider three situations: i)
 a poorly trained, but well calibrated classifier; ii) a well trained, but poorly

### 2016-03-17 jasa/paper-jasa.tex
fix typos and add tweak conclusion

@@
 never jointly encountered during learning. Similarly, it is not clear how the
-limited capacity of the classifier may impact the performance the resulting
+limited capacity of the classifier may impact the performance of the resulting
 parameterized decision function.
@@
 classifier and calibration -- computationally intensive
-parts of the approximation  -- are independent of the of the dataset $\mathcal{D}$.
+parts of the approximation  -- are independent of the dataset $\mathcal{D}$.
 Thus once trained and calibrated, the approximation can be applied to any dataset $\mathcal{D}$.
@@
 In this work, we have outlined an approach to reformulate generalized likelihood
-ratio testing over a high-dimensional data set in terms of a univariate density
+ratio tests with a high-dimensional data set in terms of univariate densities
 of a classifier score. We have shown that a parameterized family of
@@
 for parameter inference in the likelihood-free setting that can also be used in
-the frequentist formalism without specifying a prior over the parameters. A
-strength of this approach is that it separates the quality of the approximation
+the frequentist formalism without specifying a prior over the parameters. 
+In contrast to approaches that learn the posterior conditional on $\mathcal{D}$, our
+approach can be applied to any observed data $\mathcal{D}$ once trained.
+A strength of this approach is that it separates the quality of the approximation
 of the target likelihood from the quality of the calibration. The former is
-related to the ability of supervised learning approaches to  classification,
-which will continue to improve. The calibration procedure for a particular
+leverages the continuing advances in supervised learning approaches to classification. 
+The calibration procedure for a particular
 parameter point is fairly straightforward since it involves estimating a

### 2016-03-17 jasa/paper-jasa.tex
scikitlearn and gpyopt

@@
 examples where the exact likelihood is known and then discuss its application to high energy physics.
-The code used to produce the results and extended details for these examples is available in Ref.~\citep{carl}.%\footnote{\url{https://github.com/diana-hep/carl}}.
+The code used to produce the results and extended details for these examples is available in Ref.~\citep{carl}, which utilizes the classification and calibration routines in scikit-learn~\citep{scikit-learn}.
+%\footnote{\url{https://github.com/diana-hep/carl}}.
 
@@
 Nevertheless, optimizing the approximate log-likelihood ratio with
-a Bayesian optimization~\citep{brochu2010tutorial} procedure is efficient and effective.
+a Bayesian optimization~\citep{brochu2010tutorial,gpy2014} procedure is efficient and effective.
 After $50$ likelihood evaluations, the maximum likelihood estimate is found at $(\hat \alpha=1.008, \hat \beta=-1.004)$.
@@
 A strength of this approach is that it separates the quality of the approximation
-of the target likelihood from the quality of the calibration. The former is
+of the target likelihood from the quality of the calibration. The former
 leverages the continuing advances in supervised learning approaches to classification. 
@@
 
-For $P(Y=0)=P(Y=1)=\frac{1}{2}$, the best regression function $s^*$ simplifies
+For $P(Y=0)=P(Y=1)=\sfrac{1}{2}$, the best regression function $s^*$ simplifies
 to
@@
 Thm.~\ref{thm:ratio-equivalence} since $s^*(\mathbf{x}) =
-m(\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)})$,
-for $m(r(\mathbf{x})) = \frac{1}{1 + r(\mathbf{x})}$, is monotonic with
-$\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}$.
+m(r(\mathbf{x}; \theta_0, \theta_1))$,
+for $m(r(\mathbf{x})) = ({1 + r(\mathbf{x})})^{-1}$, is monotonic with
+$r(\mathbf{x}; \theta_0, \theta_1)$.
 

### 2016-03-18 jasa/paper-jasa.tex
surrogate model
fix references

@@
 \noindent%
-{\it Keywords:}  likelihood ratio, likelihood-free inference, classification, particle physics
+{\it Keywords:}  likelihood ratio, likelihood-free inference, classification, particle physics, surrogate model
 \vfill
@@
 additionally KC is supported through PHY-1505463 and PHY-1205376.
-JP was partially supported by the Scientific and Technological Center of Valparaiso (CCTVal) under Fondecyt grant BASAL FB0821.
+JP was partially supported by the Scientific and Technological Center of Valpara\'iso (CCTVal) under Fondecyt grant BASAL FB0821.
 KC would like to thank Daniel Whiteson for
